### 2017-03-29 bes-write-up-analysis.Rmd
first commit

---
title: "Whales and Isotopes for BES project write-up"
output: html_notebook
date: "`r Sys.Date()`"
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", 
                      fig.width = 8, fig.height = 6)

```
# Meeting Notes 29-March-2017

* 544 samples
    * 1 blue
    * 6 fin
    * 6 minke
    * 3 sei

The data are...

```{r import-data}
# load required libraries
# library(SIBER)

# read in the data
```

# Data summary & visualisation
Summarise the data here...

```{r summary-stats}
# some summary stats

# some basic visualisation

```

# Models

```{r fit-models}
# fit some models

```

# Results
```{r results}
# summarise model results

```

# Some Notebook shortcuts just to remind me

This is an [R Markdown](http://rmarkdown.rstudio.com) Notebook. When you execute code within the notebook, the results appear beneath the code.

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*.

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

### 2017-03-29 simulated-data/dummy_file.R
added dummy files to make folders appear

@@ -1,0 +1,1 @@
+# dummy file to test github

### 2017-03-30 analysis/bes-write-up-analysis.Rmd
Tidied repo so analysis has own folder

---
title: "Blue Whale - Figures for Paper"
output: html_notebook
date: "`r Sys.Date()`"
author: Andrew L Jackson
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", 
                      fig.width = 8, fig.height = 6)

```

This file, by AJ, is used to generate `ggplot` figures of the data for inclusion in the paper, as per the notes in the `README` file. Currently it creates some simluated data in order to explore how best to render the figures.

```{r import-data}
# load required libraries
# library(SIBER)
library(ggplot2)
library(dplyr)

rm(list=ls())
graphics.off()
set.seed(1)
# read in the data

# simulate some data for now.
t.max <- 60 # number of time steps
increment <- seq(0,t.max)

n.t <- length(increment)

dt <- 1 # the time per increment in days.
time <- increment * dt

# simulate the d13C data
dC <- arima.sim(n = n.t, 
                list(ar = c(0.8897, -0.4858), ma = c(-0.2279, 0.2488)),
                sd = sqrt(0.1796))

# create the dataframe
df <- data.frame(dC = as.numeric(dC), time = time, increment = increment)

# duplicate the dataframe to simulate multiple jittered time series
n.series <- 6
df.list <- vector("list", n.series)
df.list <- lapply(df.list, function(x) {
  x <- df; x$dC <- x$dC + rnorm(nrow(x), 0, 0.2); return(x)
  })

# bind them all into a single data.frame
df.all <- bind_rows(df.list, .id = "id")

```

## Data summary & visualisation
Summarise the data here...

```{r summary-stats}
# some summary stats

# some basic visualisation
plot(dC ~ time, data = df, type = "b")

# same in ggplot
g1 <- ggplot(df.all, aes(time, dC)) + 
  geom_point(aes(color = id)) + 
  geom_line(aes(group = id, color = id)) #+ 
  # geom_bin2d(binwidth = c(1,0.1))
print(g1)

```

## Models

```{r fit-models}
# fit some models

```

## Results
```{r results}
# summarise model results

```

## Some Notebook shortcuts just to remind me

This is an [R Markdown](http://rmarkdown.rstudio.com) Notebook. When you execute code within the notebook, the results appear beneath the code.

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*.

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

### 2017-03-30 manuscript/blue-whale-ms.tex
Initial commit of manuscript draft

% Preamble
\documentclass[a4paper,12pt]{article}

\usepackage[osf]{mathpazo} % palatino
\usepackage{ms}            % load the template
\usepackage[round]{natbib} % author-year citations
\usepackage{graphicx}
\usepackage{parskip}      
\pagenumbering{arabic}    
\linespread{1.66}

% Title page information
\title{Reconstructing the last known movements of one of Nature's giants}
% 90 characters max

\author{
  Natalie Cooper$^{1}$ Jackson2, Trueman3 and Kate3, Coombs1,4, Sabin1 in some order
}
\date{}
\affiliation{\noindent{\footnotesize
  $^1$ Department of Life Sciences, Natural History Museum London, Cromwell Road, London, SW7 5BD, UK. natalie.cooper@nhm.ac.uk. Fax: +44 1 677 8094; Tel: +44 1 896 5083.\\ 
  $^2$ National Oceanographic Centre, University of Southampton, Southampton, UK.\\
  $^3$ School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland. a.jackson@tcd.ie. \\
  $^4$ UCL\\
}}

\vfill

%\runninghead{}
%\keywords{}\\

% End of preamble

\begin{document}
\modulolinenumbers[1]   % Line numbering on every line

\mstitlepage

\parindent = 1.5em
\addtolength{\parskip}{.3em}
% Abstract 200 -300 words max
\section{Abstract}

The blue whale, *Balaenoptera musculus*, is the largest animal to have ever lived, and one of the most heavily influenced by human actions. 
Over 50 years since the International Whaling Commission banned hunting of blue whales, North Atlantic stocks are  still only at \% of pre-whaling levels.
To conserve this endangered species we need to understand its distribution, behaviour and ecology, however, high resolution data on whale ecology are scarce, particularly for the North Atlantic population. 
Here we use baleen samples from the Natural History Museum's iconic blue whale, along with new movement models, to reconstruct the movements of the whale over the last three years of its life.
We show that the whale migrated rather than remaining resident in the northern Atlantic year round. 
This is the first clear quantitative evidence of historical migrations in the species

Museum collections provide a unique opportunity to infer animal movements with non invasive methods.

\section{Main text}
% 1500 words max (excludes refs title, authors, abstract, acknowledgements)

In 1891, a 25m long female blue whale beached off the coast of Wexford in Ireland. In 2017, her skeleton will go on display in the Natural History Museum's (NHM) rennovated Hintze Hall, making her the most famous whale in the UK, if not the world. Ironically, we know very little about this animal, or the species as a whole.

Using stable isotpes (more detail) and a novel model of whale movement that incorporates X, Y and Z (see SI) we were able to trace the most likely movement history of the NHM blue whale. The models do this. We compared these to the carbon traces found for the blue whale (Fig 1C)

This basic lack of understanding is even scarier

Whaling was an intense pressure for whales during this period. The first blue whale fisheries began in XXX, but by 1900 fisheries had moved outside the area because stocks were so depleted. Before whaling, blue whale densities in the North Atlantic are estinated at X-X (depending on the methods used), but by 1900 it is likely that only a few hundred whales were left. The NHM blue whale thus represents an individual from a species at the brink of extinction. Blue whales were the first species that humans legislated to save, and represent a big conservation success story, with numbers of some populations appearing to recover to pre whaling levels.

Even today, high resolution behavioural information for large baleen whales (Balaenopteridae), and cetaceans in general, is rare, particularly for populations in the North Atlantic (by contrast, Pacific populations are more extensively studied, and patches of the North Atlantic are well studied, for example off Iceland)

Our findings have implications beyond this individual whale, and beyond cetaceans. Our models allow us  to ID migrating vs resident pops. Can be used for otehr stuff from whales to seals to sharks. immensely valuable.

\section{Methods (condensed)}
%<3000 words

\subsection{Data Availability}

\section{References}

\section{Supplementary Information}

\section{Acknowledgements}
This work was funded by the British Ecological Society (grant: 5771/6815). 
We thank XXX for comments on drafts.

\section{Author contributions}

\section{Author information}

\section{Tables}

\section{Figure legends}

\bibliographystyle{mee2}
\bibliography{darkside}

\newpage
\section{Figures}

%  \begin{figure}[!htbp]
%    \centering
%      \includegraphics[width=12cm]{XXX.pdf}
%      \caption{}
%      \label{}
%  \end{figure}

\end{document}

### 2017-03-30 manuscript/blue-whale-ms.tex
Edits to abstract

@@
 \linespread{1.66}
+
+%Contact: h.gee#nature.com*
 
@@
 
-\mstitlepage
+%\mstitlepage
 
@@
 \section{Abstract}
+% general topic intro - 1 sentence
+% intro to field - 2 sentences
+% what we find/show - 1 sentence
+% implications - 2-3 sentences. Results in context, how has the paper moved the field forward
 
-The blue whale, *Balaenoptera musculus*, is the largest animal to have ever lived, and one of the most heavily influenced by human actions. 
-Over 50 years since the International Whaling Commission banned hunting of blue whales, North Atlantic stocks are  still only at \% of pre-whaling levels.
-To conserve this endangered species we need to understand its distribution, behaviour and ecology, however, high resolution data on whale ecology are scarce, particularly for the North Atlantic population.
-Here we use baleen samples from the Natural History Museum's iconic blue whale, along with new movement models, to reconstruct the movements of the whale over the last three years of its life.
-We show that the whale migrated rather than remaining resident in the northern Atlantic year round. 
-This is the first clear quantitative evidence of historical migrations in the species
-
-Museum collections provide a unique opportunity to infer animal movements with non invasive methods.
+Over 50 years since the International Whaling Commission banned the hunting of blue whales (\textit{Balaenoptera musculus}), North Atlantic stocks are still at only X\% of pre-whaling levels.
+Conservation efforts are hampered by the scarcity of detailed data on blue whale distribution, behaviour and ecology, for both historical and present-day populations.
+Here we use $\delta^{13}C$ stable isotopes from the baleen of the Natural History Museum's iconic blue whale, along with new movement models, to extract high resolution movement behaviour data for an animal that died over 125 years ago.
+We reconstructed the movements of the blue whale over the final years of its life, and show that it migrated rather than remaining resident in the northern Atlantic year round. 
+This is the first quantitative evidence of historical migrations in blue whales.
+Our new model provides a way to investigate migration across marine vertbrates and highlights the unique opportunities provided by museum collections for inferring past animal movements.
+% About 50 words more
 
@@
 %<3000 words
+% no figures no tables
 
@@
 
+%\bibliographystyle{nature}
+%\bibliography{whale}
+
+Supplementary Information is linked to the online version of the paper at www.nature.com/nature.
+
 \section{Supplementary Information}
@@
 \section{Acknowledgements}
+% adjectives not allowed!
 This work was funded by the British Ecological Society (grant: 5771/6815). 
@@
 \section{Author contributions}
+% rough draft at present feel free to edit
+CT, NC, AJ, RS, and K collected baleen from the NHM collections. 
+K and CT extracted and analysed samples.
+CT created the model.
+EC and NC helped parameterise the model.
+AJ created the figures. 
+NC wrote the paper.
+All authors edited drafts and approved the final version.
 
 \section{Author information}
+Reprints and permissions information is available at www.nature.com/reprints.
+The authors declare no competing financial interests.
+Correspondence and requests for materials should be addressed to natalie.cooper@nhm.ac.uk.
+% One person only allowed
 
@@
 \section{Figure legends}
-
-\bibliographystyle{mee2}
-\bibliography{darkside}
 

### 2017-06-21 code-for-figures-ggplot2.R
creating standalone script files for each figure. starting with figure 1 which is 3 panels

@@ -248,8 +248,14 @@ mp <- ggplot(resTrack, aes(Lon, Lat, group = Rep))
 # add the world map
 mp <- mp + mapWorld
 
+# remove axes and ticks and labels
+mp <- mp + theme(axis.line  = element_blank(),
+                 axis.text  = element_blank(),
+                 axis.ticks = element_blank(),
+                 axis.title = element_blank())
+
 # customise labels and style
-mp <- mp + theme_classic(base_size = 14) + xlab("Longitude") + ylab("Latitude")
+mp <- mp + theme_classic(base_size = 14) #+ xlab("Longitude") + ylab("Latitude")
 
 # zoom into Atlantic region
 mp <- mp +  coord_fixed(xlim = c(-80, 50),  
@@ -257,14 +263,11 @@ mp <- mp +  coord_fixed(xlim = c(-80, 50),
                         ratio = 1.3)
 
 #Now Layer the points on top
-
 mp2 <- mp + geom_point(aes(x=jLon, y=jLat, 
                        color = factor(season)), 
                        size=0.1) + 
   scale_color_viridis(discrete = TRUE, "Quarter") + 
   guides(colour = guide_legend(override.aes = list(size=5)))
-
-  
 
 print(mp2)
 
@@ -315,10 +318,17 @@ ggsave(fn, best_rep_plot, device = png(width = 600, height = 400))
 mp_best <- ggplot(filter(resTrack, Rep == do.this.rep), 
                   aes(jLon, jLat, group = Rep))
 
+# add the world map
 mp_best <- mp_best + mapWorld
 
-mp_best <- mp_best + theme_classic(base_size = 14) + 
-  xlab("Longitude") + ylab("Latitude")
+# apply classic theme
+mp_best <- mp_best + theme_classic(base_size = 14)
+
+# remove axes
+mp_best <- mp_best + theme(axis.line  = element_blank(),
+                           axis.text  = element_blank(),
+                           axis.ticks = element_blank(),
+                           axis.title = element_blank())
 
 # zoom in to Atlantic region
 mp_best <- mp_best + coord_fixed(xlim = c(-80, 50),  

### 2017-06-21 figure1-create-panels.R

@@ -1,0 +1,289 @@
+# Author: Clive Trueman
+# Adapted by: Andrew Jackson 19-Jun-2017
+# About: script to generate figure 1
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+# setwd("/Users/trueman/Desktop/R scripts/migrate.files")
+
+
+library(raster)
+library(gstat)
+library(sp)
+library(maps)
+library(mapdata)
+
+library(tidyr)
+library(dplyr)
+library(ggplot2)
+library(magrittr)
+library(viridis)
+library(lubridate)
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+rm(list = ls())
+graphics.off()
+
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Panel A - raw whale isotope data with two y-axes
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+
+whale_isos <- read.csv("data/KC.NHM.full.csv", header = TRUE, 
+                       stringsAsFactors = FALSE)
+
+KC7 <- filter(whale_isos, Whale == "KC7")
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# convert sample number to days using following parameters
+samp_resolution <- 1 # samples per centimetre
+
+growth_rate <- 13.5  # baleen growth rate in centimetres per year
+
+days <- KC7$Samp.No * samp_resolution * 365.25 / growth_rate
+
+days <- days - days[1]
+
+KC7$days <- days
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# convert to calendar date
+
+# reverse time
+rev_days <- days - days[length(days)]
+
+last_sample <- as.Date("1891-03-01")
+
+sample_date <- last_sample + rev_days
+
+KC7$rev_days <- rev_days
+KC7$date <- sample_date
+KC7$year <- year(sample_date)
+
+# calculate a rolling mean to identify the changes between years
+
+idx_new_year <- c(0, diff(KC7$year))
+
+KC7$day_new_year <- days * idx_new_year - 
+  (idx_new_year * 0.5 * samp_resolution * 365.25 / growth_rate)
+
+KC7$samp_new_year <- KC7$Samp.No * idx_new_year - 
+  (idx_new_year * 0.5)
+
[…]

### 2017-06-21 code-for-figures-ggplot2.R
moved code to remove axes

@@ -257,6 +257,12 @@ mp <- mp + theme(axis.line  = element_blank(),
 # customise labels and style
 mp <- mp + theme_classic(base_size = 14) #+ xlab("Longitude") + ylab("Latitude")
 
+# remove axes
+mp <- mp + theme(axis.line  = element_blank(),
+                 axis.text  = element_blank(),
+                 axis.ticks = element_blank(),
+                 axis.title = element_blank())
+
 # zoom into Atlantic region
 mp <- mp +  coord_fixed(xlim = c(-80, 50),  
                         ylim = c(0, 80), 

### 2017-06-21 Figure-S4-alternative-movement-d13C.R
added S4 script file

@@ -1,0 +1,208 @@
+# Author: Clive Trueman
+# Adapted by: Andrew Jackson 19-Jun-2017
+# About: script to generate figures S4a-d for the blue whale paper
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+# setwd("/Users/trueman/Desktop/R scripts/migrate.files")
+
+
+library(raster)
+library(gstat)
+library(sp)
+library(maps)
+library(mapdata)
+
+library(tidyr)
+library(dplyr)
+library(ggplot2)
+library(magrittr)
+library(viridis)
+library(lubridate)
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+rm(list = ls())
+graphics.off()
+
+# set the constant to add as a correction to the d13C data
+cnst <- 6
+
+
+##########read resident whale model files
+
+resNor    <- read.csv("data/Norway.csv", header = TRUE)
+resIre    <- read.csv("data/west.Ireland.csv", header = TRUE)
+resCan    <- read.csv("data/Canaries.csv", header = TRUE)
+resMidAtl <- read.csv("data/Mid_Atl.csv", header = TRUE)
+
+
+
+
+#plot loess fits to the 30 resident simulated whales for each region
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+#norway resident
+
+# use piping to break up the data.frame by replicate, then 
+# predict the loess
+tmp <- resNor %>% group_by(Rep) %>% 
+  do(., data.frame(Z = predict(loess((.$d13C + cnst) ~ .$Day.No, span = 0.1))))
+
+# append this vector onto the original data.frame
+resNor$lo <- tmp$Z
+
+# plot via ggplot
+nor_plot <- ggplot(resNor, aes(Day.No, lo, group = Rep)) + 
+  geom_line(col = viridis(3)[2], alpha = 0.25) + 
+  xlab("Time") + 
+  ylab(expression(paste(delta^{13}, "C (\u2030)"))) + 
+  theme_classic(base_size = 14) + ylim(-25, -14)
+
+print(nor_plot)
+
+ggsave("figures/Figure-S4a-norway-resident-model-d13C.png", nor_plot, 
+       device = png(width = 600, height = 400))
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+#Ireland resident
+
+# same as for norway example above
[…]

### 2017-06-21 code-for-figures-ggplot2.R

@@ -19,6 +19,7 @@ library(dplyr)
 library(ggplot2)
 library(magrittr)
 library(viridis)
+library(lubridate)
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 rm(list = ls())

### 2017-06-21 manuscript/blue-whale-ms.tex
add Kate's full name
Initial quick transfer of draft to LaTeX

@@
 
-%Contact: h.gee#nature.com*
-
 % Title page information
@@
 \author{
-  Natalie Cooper$^{1}$ Jackson2, Trueman3 and Kate3, Coombs1,4, Sabin1 in some order
+  Natalie Cooper$^{1*}$ Andrew L. Jackson$^{2}$, Katharyn S. Chadwick$^{3}$,\\ Ellen J. Coombs$^{1,4}$,
+  Richard C. Sabin$^{1}$ and Clive N. Trueman$^{3}$ 
 }
@@
 \affiliation{\noindent{\footnotesize
-  $^1$ Department of Life Sciences, Natural History Museum London, Cromwell Road, London, SW7 5BD, UK. natalie.cooper@nhm.ac.uk. Fax: +44 1 677 8094; Tel: +44 1 896 5083.\\ 
-  $^2$ National Oceanographic Centre, University of Southampton, Southampton, UK.\\
-  $^3$ School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland. a.jackson@tcd.ie. \\
-  $^4$ UCL\\
+  $^1$ Department of Life Sciences, Natural History Museum London, Cromwell Road, London, SW7 5BD, UK.\\ 
+  $^2$ School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
+  $^3$ National Oceanographic Centre, University of Southampton, Southampton, UK.\\
+  $^4$ University College London, Gower Street, London, WC1E 6BT, UK.\\
 }}
@@
 %\runninghead{}
-%\keywords{}\\
+%\keywords{}
+%}
 
@@
 
-%\mstitlepage
+\mstitlepage
 
@@
 % Abstract 200 -300 words max
-\section{Abstract}
+% \section{Abstract}
 % general topic intro - 1 sentence
@@
 
-Over 50 years since the International Whaling Commission banned the hunting of blue whales (\textit{Balaenoptera musculus}), North Atlantic stocks are still at only X\% of pre-whaling levels.
+\paragraph{Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change [1,2]. 
+Current methods for monitoring movement behaviours are often expensive, and do not provide information on historical or long-term range movements.
+Animal tissues housed in museum collections present an alternative, providing biochemical records of behaviour that can be used to reconstruct historical individual-level movements relatively inexpensively [3].
+Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct the last years of an iconic blue whale (Balaenoptera musculus) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London (NHM).
+We recover individual-level movement information at high temporal resolution over six years and show distinct periods of foraging and short distance north-south migrations in sub-Arctic waters and calving in subtropical waters.
+We also detect isotopic signatures consistent with pregnancy and lactation in the year before the whale's death. 
+These results supply the first quantitative evidence of historical migrations in blue whales, and suggest that blue whale residency and movement patterns in the North Atlantic have remained similar for hundreds of years, despite changing human pressures in the region.
+Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climatic, ecological conditions and human-induced pressures, and reveal aspects of life history.
+Combining biochemical tracers with simulation modeling unlocks the information recorded in these irreplaceable archives.}
+
+%\section{Main text}
+% 1500 words max (excludes refs title, authors, abstract, acknowledgements)
+\newpage
+
+Migratory species pose a particular challenge for conservation practitioners, because their effective conservation relies on protection at many, often very distant, sites [1]. 
+Such species are also likely to be particularly vulnerable to climate change, as changes may impact various parts of their life cycle [2]. 
+Dealing with these issues is difficult due to the relative scarcity of information on species movements over multiple years, for both historical and present-day populations, particularly for wide-ranging marine species [Ryan et al 2013; Hall-Aspland 2005; Bailey et al 2009].
+
+The blue whale (Balaenoptera musculus), the largest animal to have ever lived on Earth, is one such migratory marine species. 
+Despite blue whale hunting being banned for over 50 years, blue whales are still listed as Endangered by the IUCN Red List (Reilly et al 2008).
 Conservation efforts are hampered by the scarcity of detailed data on blue whale distribution, behaviour and ecology, for both historical and present-day populations. 
-Here we use $\delta^{13}C$ stable isotopes from the baleen of the Natural History Museum's iconic blue whale, along with new movement models, to extract high resolution movement behaviour data for an animal that died over 125 years ago.
-We reconstructed the movements of the blue whale over the final years of its life, and show that it migrated rather than remaining resident in the northern Atlantic year round. 
-This is the first quantitative evidence of historical migrations in blue whales.
-Our new model provides a way to investigate migration across marine vertbrates and highlights the unique opportunities provided by museum collections for inferring past animal movements.
-% About 50 words more
+Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records (Borger 2015; others if space refs MCdonald et al 2006 already used ?silva paper).
+These methods are adding vital behavioural data to our understanding of whale movements, however, they are expensive, often limited to short-term studies on few individuals (Bailey et al 2009; Best et al 2015; Mate et al 2007), and may influence whale behaviour (walker et al 2012).
+An alternative technique is to investigate oscillations in stable isotopes values of whale tissues housed in museum collections (Ryan et al 2013).
+This is a non-invasive method that provides individual-level movement information at high temporal resolution over multiple years. 
+Here we apply this method to baleen from the Natural History Museum's blue whale.
 
-\section{Main text}
-% 1500 words max (excludes refs title, authors, abstract, acknowledgements)
+Baleen is formed of plates of keratin and is used by mysticete whales to filter feed. 
+It is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert (Best and Schell 1996). 
+We sampled keratin from the blue whale’s baleen plate at 1cm intervals and used continuous-flow isotope ratio mass spectrometry to extract $delta^13C$ and $delta^15N$ isotope concentrations. 
+Given the date of stranding (25th March 1891), and estimated baleen growth rates of $13.5 cm y^-1$ (see Supplemental Methods), we reconstructed a timeline for $delta^13C$ and $delta^15N$ fluctuations in the baleen over the last six years of whale’s life (autumn 1885 - spring 1891).
+The $delta^13C$ and $delta^15N$ profiles show two distinct phases in the whale's behaviour.
+In behavioural phase one (month 1885 to May 1889), we find strong cyclical fluctuations in both $delta^13C$ and $delta^15N$ values of an approximately yearly periodicity (Fig.
+1).
+$delta^13C$ and $delta^15N$ exhibit strong negative covariance.
+In behavioural phase two (June 1889 to March 1891) we see a sharp uptick in $delta^13C$, followed by high and relatively constant $delta^13C$ values. 
+This is associated with an overall reduction in $delta^15N$ values and a weakening of the periodic $delta^15N$ fluctuations seen in behavioural phase one.
+There is also no covariance between $delta^13C$ and $delta^15N$ (Fig.
+1).
 
-In 1891, a 25m long female blue whale beached off the coast of Wexford in Ireland. 
-In 2017, her skeleton will go on display in the Natural History Museum's (NHM) rennovated Hintze Hall, making her the most famous whale in the UK, if not the world. 
-Ironically, we know very little about this animal, or the species as a whole.
+%  \begin{figure}[!htbp]
+%    \centering
+%      \includegraphics[width=12cm]{XXX.pdf}
+%      \caption{}
+%      \label{}
+%  \end{figure}
 
-Using stable isotpes (more detail) and a novel model of whale movement that incorporates X, Y and Z (see SI) we were able to trace the most likely movement history of the NHM blue whale.
-The models do this.
-We compared these to the carbon traces found for the blue whale (Fig 1C)
+To interpret these isotopic signatures, we simulated time series of $delta^13C$ and $delta^15N$ values expected under differing movement behaviours, by coupling a novel agent-based model of whale movements (see Supplementary Methods) to models of global phytoplankton $delta^13C$ and $delta^15N$ values (Schmittner and Somes; Magozzi et al 2017).
+We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth and phytoplankton concentration (as a proxy for zooplankton food availability).
+At each time point and location in the simulations, we extracted phytoplankton $delta^13C$ and $delta^15N$ values from the models (Schmittner and Somes, Magozzi et al 2017) resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas.
+We then determined which of these simulated profiles were biogeochemically consistent with the isotopic profile of the blue whale’s baleen (Fig. 1) and use this to interpret the results.
 
-This basic lack of understanding is even scarier
+During behavioural phase one, the whale is foraging at high northern latitudes, likely off the coast of Norway. 
+The pronounced cyclical variations in $delta^13C$ values are characteristic of the isotopic expression of the spring phytoplankton bloom in northern waters (Magozzi et al 2017). 
+This interpretation is also consistent with the relatively negative median $delta^13C$ values (Fig. 1). 
+The periodicity of $delta^13C$ and $delta^15N$ values is explained by short distance seasonal migrations, with the whale overwintering off the coast of Ireland. 
+The relatively positive and invariant $delta^13C$ values seen during behavioural phase two are found in subtropical areas, particularly off the coast of the Azores, suggesting the whale migrated rapidly to southerly latitudes in early 1889, but continued feeding (other mysticete whales fast during migrations, but blue whales are too large to stop feeding for long periods; Busquets-Vass 2017 ?silva paper).
+The whale remained in these warm waters for over a year and was returning to her northern feeding grounds when she became stranded.
 
-Whaling was an intense pressure for whales during this period.
-The first blue whale fisheries began in XXX, but by 1900 fisheries had moved outside the area because stocks were so depleted.
-Before whaling, blue whale densities in the North Atlantic are estinated at X-X (depending on the methods used), but by 1900 it is likely that only a few hundred whales were left.
+Movement to warm waters alone, however, cannot explain the dramatic increase in $delta^13C$ values at the start of behavioural phase two. 
+We believe that this reflects the physiological demands of pregnancy, as similar increases in $delta^13C$ values have been seen in pregnant South American sea lions (Cardona et al 2017). 
+The gradual increase in $delta^15N$ values to the autumn of 1890 is consistent with the isotopic expression associated with lactation (Cardona et al 2017).
+Blue whales have a 10-12 month gestation period, and calves are weaned after 6-7 months (Wilson  Mittermeier 2014).
+The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the known reproductive ecology of blue whales.
+This is the first quantitative evidence of historical movement patterns in blue whales, and reveals a fascinating snapshot into the life of an iconic specimen, where we previously only knew the date and location of its death.
+
+Conventional wisdom suggests that baleen whales migrate each year from feeding grounds near the poles in summer, to overwintering and breeding grounds nearer the equator [Corkeron  Connor 1999; Lockyer and Brown 1981].
+However, a growing body of evidence suggests that not all whales migrate to subtropical waters each year, with some remaining at high latitudes year round (McDonald et al 2006).
+Our results support this conclusion.
+Interestingly, there is still a resident population of blue whales in the region our models place the blue whale within, and females still breed in the subtropical waters near the Azores (Reilly et al 2008).
+This suggests that whale movements have remained similar over hundreds of years, despite changes in anthropogenic pressures.
+Whaling was an intense pressure for blue whales during the period we are analysing.
+Before whaling in the North Atlantic began in 1868 (Reilly et al 2008), there were 10,000-15,000 blue whales in the region (Sigurrjonson 1995). 
+By 1900 fisheries had moved outside the area because stocks were so depleted [Reilly et al 2008]; during this period over 12,000 blue whales were landed (Sigurrjonson 1995). 
 The NHM blue whale thus represents an individual from a species at the brink of extinction. 
-Blue whales were the first species that humans legislated to save, and represent a big conservation success story, with numbers of some populations appearing to recover to pre whaling levels.
+Blue whales were the first species that humans legislated to save, and represent a big conservation success story, although North Atlantic stocks have not recovered to the same extent as other whales in the region (Pike 2009, Sigurrjonson 1995).
+There is also evidence that blue whales are shifting their ranges due to warming waters off Iceland (Vikingsson et al 2015). 
+Thus understanding historical movements of North Atlantic blue whales has never been more vital if we wish to preserve them for future generations.
 
-Even today, high resolution behavioural information for large baleen whales (Balaenopteridae), and cetaceans in general, is rare, particularly for populations in the North Atlantic (by contrast, Pacific populations are more extensively studied, and patches of the North Atlantic are well studied, for example off Iceland)
-
-Our findings have implications beyond this individual whale, and beyond cetaceans.
-Our models allow us  to ID migrating vs resident pops.
-Can be used for otehr stuff from whales to seals to sharks.
-immensely valuable.
+In conclusion, the model we present here allows us to recreate individual-level movement behaviours and life history using tissues from museum collections and other sources. 
+This has applications beyond this individual whale, and beyond cetaceans, as our model can be applied to any marine species with constantly growing tissues, for example seal whiskers, shark skin, and fish scales.
+These techniques can also allow us study extinct species' behaviours and understand how populations were affected by past pressures such as hunting, and present-day pressures such as global change. 
+Finally, our results highlight that museums represent a vast, mostly untapped, treasure trove of information ready for us to explore.
 
@@
 
+\subsection{Stable isotope extractions from baleen}
+Baleen was collected from the Natural History Museum, London (specimen BMNH.1892.3.1.1). 
+The baleen plate was cleaned with ethanol to remove surface contaminants. 
+$~$1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
+78 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section which contains the most recent tissue. 
+Baleen grows at a constant rate, so the samples are equally spaced through time (Best and Schell 1996).
+Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the National Oceanographic Centre (Southampton, UK), using a Vario Isotope select elemental analyser, coupled with Isoprime 100 isotope mass spectrometer.
+Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
+C:N ratios for samples ranged from 3.28$‰$ to 3.72$ppm$, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples (Hobson  Koch 1998).
+
+\subsection{Time calibrating stable isotope profiles}
+Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates  (Hobson and Schell, 1987, Busquets-Vass et al 2017).
+Clear periodicity was evident in $delta^15N$ values across the entire baleen plate, and in $delta^13C$ values in the most distal (oldest) 65cm of the plate. 
+We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis (Cardona et al 2017), revealing a consistent growth rate of 13.5cm $y^-1$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cm y^-1$ estimated for six blue whales from the Californian Pacific (Busquets-Vass et al 2017).
+Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891.
+Cross-correlation analysis (ref?) demonstrated a strong negative covariance between $delta^13C$ and $delta^15N$ values within behavioural phase one (oldest 60cm of the baleen plate), but no relationship between $delta^13C$ and $delta^15N$ values during behavioural phase two.
+
+\subsection{Baseline isotope comparisons}
+Isotope-enabled biogeochemical ocean models (Schmittner and Somes, Magozzi et al 2017) were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds.
+Annual average $delta^15N$ POM (particulate organic matter) values were provided by Somes (pers.comm) based on a 5$^{circ}$ resolution biogeochemical model.
+$delta^13C$ POM values were simulated at 1$^{circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model (Magozzi et al 2017, Yool et al 2013).
+Simulated  $delta^15N$ POM values are relatively positive in the northeast Atlantic north of c.
+60$^{circ}$N, and relatively negative in the central and southern North Atlantic.
+Annual average $delta^13C$ POM values largely vary with latitude, with more negative values in more northerly regions.
+In the central North Atlantic, $delta^13C$ POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig S3).
+The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature (Laws, Magozzi et al 2017).
+Thus temporal variations in $delta^13C$ POM values were superimposed on latitudinal gradients. 
+The scale and nature of temporal variation in $delta^13C$ POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in $delta^13C$ POM values linked to strongly seasonal phytoplankton growth dynamics.
+We therefore used monthly simulated $delta^13C$ POM values to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours.
+
+\subsection{Agent-based whale movement model}
+Movement was coded as set of probabilistic rules. 
+The parameters for these were taken from the literature on blue whale behaviour (e.g. Wilson  Mittermeier 2014). 
+All terms were expressed as probability distributions allowing us to introduce individual variability.
+
+In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and extent (1$^{circ}$ or 2$^{circ}$) of movement are all influenced by the following.
+(i) Behavioural state (migrating north, migrating south, foraging, or nursing). 
+Northerly migrations were possible only in spring, and southerly migrations in autumn.
+Foraging was possible at anytime of year, and was triggered when whales encountered high concentrations of plankton.
+Nursing occurred at temperatures between 18 and 25$^{circ}$C only.
+(ii) Sea surface temperature ($^{circ}$C, Yool et al 2013). 
+When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{circ}$C); whereas whales migrating south sought warmer waters.
+(iii) Water depth (m; what reference is this collected from?). 
+Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
+(iv) Phytoplankton concentration (mmol N m-3, for combined diatom and non-diatom communities, Yool et al 2013). 
+This was included as a proxy for zooplankton food availability. 
+Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state.
+
+We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores; Fig SX; McDonald et al 2006; Reilly et al 2008; sigurjonsson 1995); (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
+Various movement rules were implemented to avoid simulations ending up on land, or becoming “stuck” in the relatively shallow waters of the Bay of Biscay or the Strait of Dover. 
+For a full description of the model and its parameters see Supplementary Methods. We simulated whale movements XX times for each scenario and residency hotspot. 
+For each movement simulation we extracted phytoplankton $delta^13C$ and $delta^15N$ values at each time point and location from the models described above (Magozzi, Somes). 
+We then compared the simulated stable isotope profiles (Fig. S4) to the profile of the blue whale (Fig. 1).
+
 \subsection{Data Availability}
+Data are available on the Natural History Museum, London's data portal (https://data.nhm.ac.uk). 
+R code is available from XXX.
 
@@
 
-%\bibliographystyle{nature}
-%\bibliography{whale}
+\bibliographystyle{nature}
+\bibliography{blue-whale}
 
@@
 \section{Supplementary Information}
+Supplementary Information is linked to the online version of the paper at www.nature.com/nature.
 
@@
 We thank XXX for comments on drafts.
+We thank CJ Somes for providing  δ13C POM data.
 
@@
 % rough draft at present feel free to edit
-CT, NC, AJ, RS, and K collected baleen from the NHM collections. 
-K and CT extracted and analysed samples.
-CT created the model.
+CT, NC, AJ, RS, and KC collected baleen from the NHM collections. 
+KC and CT extracted and analysed samples.
+CT created the movement model.
 EC and NC helped parameterise the model.
@@
 
-\section{Tables}
+%\section{Tables}
 
-\section{Figure legends}
+%\section{Figure legends}
 
-\newpage
-\section{Figures}
+%\newpage
+%\section{Figures}
 

### 2017-06-21 Figure-S4-alternative-movement-d13C.R
added facet-wrap for Fig S4

@@ -200,9 +200,31 @@ all_models_plot <- multiplot(nor_plot2, ire_plot2, can_plot2, MidAtl_plot2,
                              fn = "figures/Figure-S4-all.png",
                              width = 600, height = 400)
 
+# do the same but with facets
 
-# ggsave("figures/Figure-S4-all.png", 
-#        all_models_plot, 
-#        device = png(width = 600, height = 400))
+res_all <- bind_rows(list(Norway = resNor, 
+                          Ireland = resIre, 
+                          Canaries = resCan, 
+                          "Mid-Atlantic" = resMidAtl), .id = "Region")
 
+sp + facet_wrap( ~ day, ncol=2)
+
+# same as for norway example above
+tmp <- res_all %>% group_by(Region) %>% group_by(Rep) %>% 
+  do(., data.frame(Z = predict(loess((.$d13C + cnst) ~ .$Day.No, span = 0.1))))
+
+res_all$lo <- tmp$Z
+
+test_plot <- ggplot(res_all, aes(Day.No, lo, group = Rep)) + 
+  geom_line(col = viridis(3)[2], alpha = 0.25) + 
+  xlab("Time") + 
+  ylab(expression(paste(delta^{13}, "C (\u2030)"))) + 
+  theme_classic(base_size = 14) + ylim(-25, -14)
+
+test_plot <- test_plot + facet_wrap(~Region, ncol = 2)
+
+print(test_plot)
+
+ggsave("figures/Figure-S4-facet-wrap-d13C.png", MidAtl_plot, 
+       device = png(width = 600, height = 400))
 

### 2017-06-21 manuscript/blue-whale-ms.tex
Minor typo in acknowlegments
Fix symbols

@@
 \usepackage{parskip} 
+\usepackage{textcomp} % for parts per mille symbol     
 \pagenumbering{arabic}    
@@
 It is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert (Best and Schell 1996). 
-We sampled keratin from the blue whale’s baleen plate at 1cm intervals and used continuous-flow isotope ratio mass spectrometry to extract $delta^13C$ and $delta^15N$ isotope concentrations. 
-Given the date of stranding (25th March 1891), and estimated baleen growth rates of $13.5 cm y^-1$ (see Supplemental Methods), we reconstructed a timeline for $delta^13C$ and $delta^15N$ fluctuations in the baleen over the last six years of whale’s life (autumn 1885 - spring 1891).
-The $delta^13C$ and $delta^15N$ profiles show two distinct phases in the whale's behaviour.
-In behavioural phase one (month 1885 to May 1889), we find strong cyclical fluctuations in both $delta^13C$ and $delta^15N$ values of an approximately yearly periodicity (Fig.
+We sampled keratin from the blue whale's baleen plate at 1cm intervals and used continuous-flow isotope ratio mass spectrometry to extract $\delta^{13}$C and $\delta^{15}$N isotope concentrations. 
+Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplemental Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the last six years of whale's life (autumn 1885 - spring 1891).
+The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour.
+In behavioural phase one (month 1885 to May 1889), we find strong cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values of an approximately yearly periodicity (Fig.
 1).
-$delta^13C$ and $delta^15N$ exhibit strong negative covariance.
-In behavioural phase two (June 1889 to March 1891) we see a sharp uptick in $delta^13C$, followed by high and relatively constant $delta^13C$ values. 
-This is associated with an overall reduction in $delta^15N$ values and a weakening of the periodic $delta^15N$ fluctuations seen in behavioural phase one.
-There is also no covariance between $delta^13C$ and $delta^15N$ (Fig.
+$\delta^{13}$C and $\delta^{15}$N exhibit strong negative covariance.
+In behavioural phase two (June 1889 to March 1891) we see a sharp uptick in $\delta^{13}$C, followed by high and relatively constant $\delta^{13}$C values. 
+This is associated with an overall reduction in $\delta^{15}$N values and a weakening of the periodic $\delta^{15}$N fluctuations seen in behavioural phase one.
+There is also no covariance between $\delta^{13}$C and $\delta^{15}$N (Fig.
 1).
@@
 
-To interpret these isotopic signatures, we simulated time series of $delta^13C$ and $delta^15N$ values expected under differing movement behaviours, by coupling a novel agent-based model of whale movements (see Supplementary Methods) to models of global phytoplankton $delta^13C$ and $delta^15N$ values (Schmittner and Somes; Magozzi et al 2017).
+To interpret these isotopic signatures, we simulated time series of $\delta^{13}$C and $\delta^{15}$N values expected under differing movement behaviours, by coupling a novel agent-based model of whale movements (see Supplementary Methods) to models of global phytoplankton $\delta^{13}$C and $\delta^{15}$N values (Schmittner and Somes; Magozzi et al 2017).
 We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth and phytoplankton concentration (as a proxy for zooplankton food availability).
-At each time point and location in the simulations, we extracted phytoplankton $delta^13C$ and $delta^15N$ values from the models (Schmittner and Somes, Magozzi et al 2017) resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas.
+At each time point and location in the simulations, we extracted phytoplankton $\delta^{13}$C and $\delta^{15}$N values from the models (Schmittner and Somes, Magozzi et al 2017) resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas.
 We then determined which of these simulated profiles were biogeochemically consistent with the isotopic profile of the blue whale’s baleen (Fig. 1) and use this to interpret the results.
@@
 During behavioural phase one, the whale is foraging at high northern latitudes, likely off the coast of Norway. 
-The pronounced cyclical variations in $delta^13C$ values are characteristic of the isotopic expression of the spring phytoplankton bloom in northern waters (Magozzi et al 2017). 
-This interpretation is also consistent with the relatively negative median $delta^13C$ values (Fig. 1). 
-The periodicity of $delta^13C$ and $delta^15N$ values is explained by short distance seasonal migrations, with the whale overwintering off the coast of Ireland. 
-The relatively positive and invariant $delta^13C$ values seen during behavioural phase two are found in subtropical areas, particularly off the coast of the Azores, suggesting the whale migrated rapidly to southerly latitudes in early 1889, but continued feeding (other mysticete whales fast during migrations, but blue whales are too large to stop feeding for long periods; Busquets-Vass 2017 ?silva paper).
+The pronounced cyclical variations in $\delta^{13}$C values are characteristic of the isotopic expression of the spring phytoplankton bloom in northern waters (Magozzi et al 2017). 
+This interpretation is also consistent with the relatively negative median $\delta^{13}$C values (Fig. 1). 
+The periodicity of $\delta^{13}$C and $\delta^{15}$N values is explained by short distance seasonal migrations, with the whale overwintering off the coast of Ireland. 
+The relatively positive and invariant $\delta^{13}$C values seen during behavioural phase two are found in subtropical areas, particularly off the coast of the Azores, suggesting the whale migrated rapidly to southerly latitudes in early 1889, but continued feeding (other mysticete whales fast during migrations, but blue whales are too large to stop feeding for long periods; Busquets-Vass 2017 ?silva paper).
 The whale remained in these warm waters for over a year and was returning to her northern feeding grounds when she became stranded.
 
-Movement to warm waters alone, however, cannot explain the dramatic increase in $delta^13C$ values at the start of behavioural phase two. 
-We believe that this reflects the physiological demands of pregnancy, as similar increases in $delta^13C$ values have been seen in pregnant South American sea lions (Cardona et al 2017). 
-The gradual increase in $delta^15N$ values to the autumn of 1890 is consistent with the isotopic expression associated with lactation (Cardona et al 2017).
+Movement to warm waters alone, however, cannot explain the dramatic increase in $\delta^{13}$C values at the start of behavioural phase two. 
+We believe that this reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions (Cardona et al 2017). 
+The gradual increase in $\delta^{15}$N values to the autumn of 1890 is consistent with the isotopic expression associated with lactation (Cardona et al 2017).
 Blue whales have a 10-12 month gestation period, and calves are weaned after 6-7 months (Wilson  Mittermeier 2014).
@@
 The baleen plate was cleaned with ethanol to remove surface contaminants. 
-$~$1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
+{\raise.17ex\hbox{$\scriptstyle\sim$}}1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
 78 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section which contains the most recent tissue. 
@@
 Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
-C:N ratios for samples ranged from 3.28$‰$ to 3.72$ppm$, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples (Hobson  Koch 1998).
+C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples (Hobson  Koch 1998).
 
@@
 Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates  (Hobson and Schell, 1987, Busquets-Vass et al 2017).
-Clear periodicity was evident in $delta^15N$ values across the entire baleen plate, and in $delta^13C$ values in the most distal (oldest) 65cm of the plate. 
-We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis (Cardona et al 2017), revealing a consistent growth rate of 13.5cm $y^-1$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cm y^-1$ estimated for six blue whales from the Californian Pacific (Busquets-Vass et al 2017).
+Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the most distal (oldest) 65cm of the plate. 
+We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis (Cardona et al 2017), revealing a consistent growth rate of 13.5cm $y^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cm y{^-1}$ estimated for six blue whales from the Californian Pacific (Busquets-Vass et al 2017).
 Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891.
-Cross-correlation analysis (ref?) demonstrated a strong negative covariance between $delta^13C$ and $delta^15N$ values within behavioural phase one (oldest 60cm of the baleen plate), but no relationship between $delta^13C$ and $delta^15N$ values during behavioural phase two.
+Cross-correlation analysis (ref?) demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase one (oldest 60cm of the baleen plate), but no relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase two.
 
@@
 Isotope-enabled biogeochemical ocean models (Schmittner and Somes, Magozzi et al 2017) were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds.
-Annual average $delta^15N$ POM (particulate organic matter) values were provided by Somes (pers.comm) based on a 5$^{circ}$ resolution biogeochemical model.
-$delta^13C$ POM values were simulated at 1$^{circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model (Magozzi et al 2017, Yool et al 2013).
-Simulated  $delta^15N$ POM values are relatively positive in the northeast Atlantic north of c.
-60$^{circ}$N, and relatively negative in the central and southern North Atlantic.
-Annual average $delta^13C$ POM values largely vary with latitude, with more negative values in more northerly regions.
-In the central North Atlantic, $delta^13C$ POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig S3).
+Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model.
+$\delta^{13}$C POM values were simulated at 1$^{\circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model (Magozzi et al 2017, Yool et al 2013).
+Simulated  $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c.
+60$^{\circ}$N, and relatively negative in the central and southern North Atlantic.
+Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions.
+In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig S3).
 The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature (Laws, Magozzi et al 2017).
-Thus temporal variations in $delta^13C$ POM values were superimposed on latitudinal gradients. 
-The scale and nature of temporal variation in $delta^13C$ POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in $delta^13C$ POM values linked to strongly seasonal phytoplankton growth dynamics.
-We therefore used monthly simulated $delta^13C$ POM values to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours.
+Thus temporal variations in $\delta^{13}$C POM values were superimposed on latitudinal gradients. 
+The scale and nature of temporal variation in $\delta^{13}$C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in $\delta^{13}$C POM values linked to strongly seasonal phytoplankton growth dynamics.
+We therefore used monthly simulated $\delta^{13}$C POM values to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours.
 
@@
 
-In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and extent (1$^{circ}$ or 2$^{circ}$) of movement are all influenced by the following.
+In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and extent (1$^{\circ}$ or 2$^{\circ}$) of movement are all influenced by the following.
 (i) Behavioural state (migrating north, migrating south, foraging, or nursing). 
@@
 Foraging was possible at anytime of year, and was triggered when whales encountered high concentrations of plankton.
-Nursing occurred at temperatures between 18 and 25$^{circ}$C only.
-(ii) Sea surface temperature ($^{circ}$C, Yool et al 2013). 
-When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{circ}$C); whereas whales migrating south sought warmer waters.
+Nursing occurred at temperatures between 18 and 25$^{\circ}$C only.
+(ii) Sea surface temperature ($^{\circ}$C, Yool et al 2013). 
+When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.
 (iii) Water depth (m; what reference is this collected from?). 
 Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
-(iv) Phytoplankton concentration (mmol N m-3, for combined diatom and non-diatom communities, Yool et al 2013). 
+(iv) Phytoplankton concentration (mmol N $m^{-3}$, for combined diatom and non-diatom communities, Yool et al 2013). 
 This was included as a proxy for zooplankton food availability. 
@@
 We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores; Fig SX; McDonald et al 2006; Reilly et al 2008; sigurjonsson 1995); (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
-Various movement rules were implemented to avoid simulations ending up on land, or becoming “stuck” in the relatively shallow waters of the Bay of Biscay or the Strait of Dover. 
+Various movement rules were implemented to avoid simulations ending up on land, or becoming `stuck' in the relatively shallow waters of the Bay of Biscay or the Strait of Dover. 
 For a full description of the model and its parameters see Supplementary Methods. We simulated whale movements XX times for each scenario and residency hotspot. 
-For each movement simulation we extracted phytoplankton $delta^13C$ and $delta^15N$ values at each time point and location from the models described above (Magozzi, Somes). 
+For each movement simulation we extracted phytoplankton $\delta^{13}$C and $\delta^{15}$N values at each time point and location from the models described above (Magozzi, Somes). 
 We then compared the simulated stable isotope profiles (Fig. S4) to the profile of the blue whale (Fig. 1).
@@
 
-\section{Acknowledgements}
+\section{Acknowledgments}
 % adjectives not allowed!
@@
 We thank XXX for comments on drafts.
-We thank CJ Somes for providing  δ13C POM data.
+We thank CJ Somes for providing  $\delta^{15}$N POM data.
 

### 2017-06-21 manuscript/blue-whale-ms.tex
Add references

@@
 \usepackage{ms}            % load the template
-\usepackage[round]{natbib} % author-year citations
+%\usepackage[round]{natbib} % author-year citations
+\usepackage[superscript,biblabel]{cite} % for superscript citations
 \usepackage{graphicx}
@@
 
-\paragraph{Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change [1,2]. 
+\paragraph{Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}. 
 Current methods for monitoring movement behaviours are often expensive, and do not provide information on historical or long-term range movements.
-Animal tissues housed in museum collections present an alternative, providing biochemical records of behaviour that can be used to reconstruct historical individual-level movements relatively inexpensively [3].
+Animal tissues housed in museum collections present an alternative, providing biochemical records of behaviour that can be used to reconstruct historical individual-level movements relatively inexpensively\cite{newsome2010using}.
 Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct the last years of an iconic blue whale (Balaenoptera musculus) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London (NHM).
@@
 
-Migratory species pose a particular challenge for conservation practitioners, because their effective conservation relies on protection at many, often very distant, sites [1]. 
-Such species are also likely to be particularly vulnerable to climate change, as changes may impact various parts of their life cycle [2]. 
-Dealing with these issues is difficult due to the relative scarcity of information on species movements over multiple years, for both historical and present-day populations, particularly for wide-ranging marine species [Ryan et al 2013; Hall-Aspland 2005; Bailey et al 2009].
+Migratory species pose a particular challenge for conservation practitioners, because their effective conservation relies on protection at many, often very distant, sites\cite{runge2014conserving}. 
+Such species are also likely to be particularly vulnerable to climate change, as changes may impact various parts of their life cycle\cite{robinson2009travelling}. 
+Dealing with these issues is difficult due to the relative scarcity of information on species movements over multiple years, for both historical and present-day populations, particularly for wide-ranging marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
 
-The blue whale (Balaenoptera musculus), the largest animal to have ever lived on Earth, is one such migratory marine species. 
-Despite blue whale hunting being banned for over 50 years, blue whales are still listed as Endangered by the IUCN Red List (Reilly et al 2008).
+The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived on Earth, is one such migratory marine species. 
+Despite blue whale hunting being banned for over 50 years, blue whales are still listed as Endangered by the IUCN Red List\cite{reilly2008balaenoptera}.
 Conservation efforts are hampered by the scarcity of detailed data on blue whale distribution, behaviour and ecology, for both historical and present-day populations. 
-Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records (Borger 2015; others if space refs MCdonald et al 2006 already used ?silva paper).
-These methods are adding vital behavioural data to our understanding of whale movements, however, they are expensive, often limited to short-term studies on few individuals (Bailey et al 2009; Best et al 2015; Mate et al 2007), and may influence whale behaviour (walker et al 2012).
-An alternative technique is to investigate oscillations in stable isotopes values of whale tissues housed in museum collections (Ryan et al 2013).
+Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution}).
+These methods are adding vital behavioural data to our understanding of whale movements, however, they are expensive, often limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
+An alternative technique is to investigate oscillations in stable isotopes values of whale tissues housed in museum collections\cite{ryan2013stable}.
 This is a non-invasive method that provides individual-level movement information at high temporal resolution over multiple years. 
@@
 Baleen is formed of plates of keratin and is used by mysticete whales to filter feed. 
-It is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert (Best and Schell 1996). 
+It is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable}. 
 We sampled keratin from the blue whale's baleen plate at 1cm intervals and used continuous-flow isotope ratio mass spectrometry to extract $\delta^{13}$C and $\delta^{15}$N isotope concentrations. 
@@
 $\delta^{13}$C and $\delta^{15}$N exhibit strong negative covariance.
-In behavioural phase two (June 1889 to March 1891) we see a sharp uptick in $\delta^{13}$C, followed by high and relatively constant $\delta^{13}$C values. 
+In behavioural phase two (June 1889 to March 1891) we see a sharp uptick in $\delta^{13}$C, followed by high and relatively constant $\delta^{13}$C values. ,
 This is associated with an overall reduction in $\delta^{15}$N values and a weakening of the periodic $\delta^{15}$N fluctuations seen in behavioural phase one.
@@
 
-To interpret these isotopic signatures, we simulated time series of $\delta^{13}$C and $\delta^{15}$N values expected under differing movement behaviours, by coupling a novel agent-based model of whale movements (see Supplementary Methods) to models of global phytoplankton $\delta^{13}$C and $\delta^{15}$N values (Schmittner and Somes; Magozzi et al 2017).
+To interpret these isotopic signatures, we simulated time series of $\delta^{13}$C and $\delta^{15}$N values expected under differing movement behaviours, by coupling a novel agent-based model of whale movements (see Supplementary Methods) to models of global phytoplankton $\delta^{13}$C and $\delta^{15}$N values\cite{magozzi2017using} (Schmittner and Somes).
 We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth and phytoplankton concentration (as a proxy for zooplankton food availability).
-At each time point and location in the simulations, we extracted phytoplankton $\delta^{13}$C and $\delta^{15}$N values from the models (Schmittner and Somes, Magozzi et al 2017) resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas.
-We then determined which of these simulated profiles were biogeochemically consistent with the isotopic profile of the blue whale’s baleen (Fig. 1) and use this to interpret the results.
+At each time point and location in the simulations, we extracted phytoplankton $\delta^{13}$C and $\delta^{15}$N values from the models\cite{magozzi2017using} (Schmittner and Somes) resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas.
+We then determined which of these simulated profiles were biogeochemically consistent with the isotopic profile of the blue whale's baleen (Fig. 1) and use this to interpret the results.
 
 During behavioural phase one, the whale is foraging at high northern latitudes, likely off the coast of Norway. 
-The pronounced cyclical variations in $\delta^{13}$C values are characteristic of the isotopic expression of the spring phytoplankton bloom in northern waters (Magozzi et al 2017). 
+The pronounced cyclical variations in $\delta^{13}$C values are characteristic of the isotopic expression of the spring phytoplankton bloom in northern waters\cite{magozzi2017using}. 
 This interpretation is also consistent with the relatively negative median $\delta^{13}$C values (Fig. 1). 
 The periodicity of $\delta^{13}$C and $\delta^{15}$N values is explained by short distance seasonal migrations, with the whale overwintering off the coast of Ireland. 
-The relatively positive and invariant $\delta^{13}$C values seen during behavioural phase two are found in subtropical areas, particularly off the coast of the Azores, suggesting the whale migrated rapidly to southerly latitudes in early 1889, but continued feeding (other mysticete whales fast during migrations, but blue whales are too large to stop feeding for long periods; Busquets-Vass 2017 ?silva paper).
+The relatively positive and invariant $\delta^{13}$C values seen during behavioural phase two are found in subtropical areas, particularly off the coast of the Azores, suggesting the whale migrated rapidly to southerly latitudes in early 1889, but continued feeding (other mysticete whales fast during migrations, but blue whales are too large to stop feeding for long periods\cite{busquets2017estimating}?silva paper).
 The whale remained in these warm waters for over a year and was returning to her northern feeding grounds when she became stranded.
@@
 Movement to warm waters alone, however, cannot explain the dramatic increase in $\delta^{13}$C values at the start of behavioural phase two. 
-We believe that this reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions (Cardona et al 2017). 
-The gradual increase in $\delta^{15}$N values to the autumn of 1890 is consistent with the isotopic expression associated with lactation (Cardona et al 2017).
-Blue whales have a 10-12 month gestation period, and calves are weaned after 6-7 months (Wilson  Mittermeier 2014).
+We believe that this reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions\cite{cardona2017temporal}. 
+The gradual increase in $\delta^{15}$N values to the autumn of 1890 is consistent with the isotopic expression associated with lactation\cite{cardona2017temporal}.
+Blue whales have a 10-12 month gestation period, and calves are weaned after 6-7 months\cite{handbook}.
 The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the known reproductive ecology of blue whales.
@@
 
-Conventional wisdom suggests that baleen whales migrate each year from feeding grounds near the poles in summer, to overwintering and breeding grounds nearer the equator [Corkeron  Connor 1999; Lockyer and Brown 1981].
-However, a growing body of evidence suggests that not all whales migrate to subtropical waters each year, with some remaining at high latitudes year round (McDonald et al 2006).
+Conventional wisdom suggests that baleen whales migrate each year from feeding grounds near the poles in summer, to overwintering and breeding grounds nearer the equator\cite{corkeron1999baleen,lockyer1981migration}.
+However, a growing body of evidence suggests that not all whales migrate to subtropical waters each year, with some remaining at high latitudes year round\cite{mcdonald2006biogeographic}. 
 Our results support this conclusion. 
-Interestingly, there is still a resident population of blue whales in the region our models place the blue whale within, and females still breed in the subtropical waters near the Azores (Reilly et al 2008).
+Interestingly, there is still a resident population of blue whales in the region our models place the blue whale within, and females still breed in the subtropical waters near the Azores\cite{reilly2008balaenoptera}.
 This suggests that whale movements have remained similar over hundreds of years, despite changes in anthropogenic pressures. 
 Whaling was an intense pressure for blue whales during the period we are analysing. 
-Before whaling in the North Atlantic began in 1868 (Reilly et al 2008), there were 10,000-15,000 blue whales in the region (Sigurrjonson 1995). 
-By 1900 fisheries had moved outside the area because stocks were so depleted [Reilly et al 2008]; during this period over 12,000 blue whales were landed (Sigurrjonson 1995). 
+Before whaling in the North Atlantic began in 1868\cite{reilly2008balaenoptera}, there were 10,000-15,000 blue whales in the region\cite{sigurjonsson1995life}. 
+By 1900 fisheries had moved outside the area because stocks were so depleted\cite{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed\cite{sigurjonsson1995life}. 
 The NHM blue whale thus represents an individual from a species at the brink of extinction. 
-Blue whales were the first species that humans legislated to save, and represent a big conservation success story, although North Atlantic stocks have not recovered to the same extent as other whales in the region (Pike 2009, Sigurrjonson 1995).
-There is also evidence that blue whales are shifting their ranges due to warming waters off Iceland (Vikingsson et al 2015). 
+Blue whales were the first species that humans legislated to save, and represent a big conservation success story, although North Atlantic stocks have not recovered to the same extent as other whales in the region\cite{sigurjonsson1995life,pike2009note}.
+There is also evidence that blue whales are shifting their ranges due to warming waters off Iceland\cite{10.3389/fevo.2015.00006}. 
 Thus understanding historical movements of North Atlantic blue whales has never been more vital if we wish to preserve them for future generations.
@@
 78 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section which contains the most recent tissue. 
-Baleen grows at a constant rate, so the samples are equally spaced through time (Best and Schell 1996).
+Baleen grows at a constant rate, so the samples are equally spaced through time\cite{best1996stable}.
 Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the National Oceanographic Centre (Southampton, UK), using a Vario Isotope select elemental analyser, coupled with Isoprime 100 isotope mass spectrometer.
@@
 \subsection{Time calibrating stable isotope profiles}
-Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates  (Hobson and Schell, 1987, Busquets-Vass et al 2017).
+Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates\cite{hobson1998stable,busquets2017estimating}.
 Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the most distal (oldest) 65cm of the plate. 
-We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis (Cardona et al 2017), revealing a consistent growth rate of 13.5cm $y^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cm y{^-1}$ estimated for six blue whales from the Californian Pacific (Busquets-Vass et al 2017).
+We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis\cite{cardona2017temporal}, revealing a consistent growth rate of 13.5cm $y^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cm y{^-1}$ estimated for six blue whales from the Californian Pacific\cite{busquets2017estimating}.
 Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891. 
@@
 \subsection{Baseline isotope comparisons}
-Isotope-enabled biogeochemical ocean models (Schmittner and Somes, Magozzi et al 2017) were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds.
+Isotope-enabled biogeochemical ocean models\cite{magozzi2017using} (Schmittner and Somes) were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds.
 Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model. 
-$\delta^{13}$C POM values were simulated at 1$^{\circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model (Magozzi et al 2017, Yool et al 2013).
-Simulated  $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c.
-60$^{\circ}$N, and relatively negative in the central and southern North Atlantic.
+$\delta^{13}$C POM values were simulated at 1$^{\circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model\cite{magozzi2017using,yool2013medusa}. 
+Simulated  $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c. 60$^{\circ}$N, and relatively negative in the central and southern North Atlantic. 
 Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
@@
 Movement was coded as set of probabilistic rules. 
-The parameters for these were taken from the literature on blue whale behaviour (e.g. Wilson  Mittermeier 2014). 
+The parameters for these were taken from the literature on blue whale behaviour (e.g.\cite{handbook}). 
 All terms were expressed as probability distributions allowing us to introduce individual variability.
@@
 Nursing occurred at temperatures between 18 and 25$^{\circ}$C only.
-(ii) Sea surface temperature ($^{\circ}$C, Yool et al 2013). 
+(ii) Sea surface temperature ($^{\circ}$C\cite{yool2013medusa}). 
 When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.
@@
 Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
-(iv) Phytoplankton concentration (mmol N $m^{-3}$, for combined diatom and non-diatom communities, Yool et al 2013). 
+(iv) Phytoplankton concentration (mmol N $m^{-3}$, for combined diatom and non-diatom communities\cite{yool2013medusa}). 
 This was included as a proxy for zooplankton food availability. 
@@
 
-We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores; Fig SX; McDonald et al 2006; Reilly et al 2008; sigurjonsson 1995); (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
+We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Fig SX; (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
 Various movement rules were implemented to avoid simulations ending up on land, or becoming `stuck' in the relatively shallow waters of the Bay of Biscay or the Strait of Dover. 
-For a full description of the model and its parameters see Supplementary Methods. We simulated whale movements XX times for each scenario and residency hotspot. 
-For each movement simulation we extracted phytoplankton $\delta^{13}$C and $\delta^{15}$N values at each time point and location from the models described above (Magozzi, Somes). 
+For a full description of the model and its parameters see Supplementary Methods. 
+We simulated whale movements XX times for each scenario and residency hotspot. 
+For each movement simulation we extracted phytoplankton $\delta^{13}$C and $\delta^{15}$N values at each time point and location from the models described above\cite{magozzi2017using} (Somes). 
 We then compared the simulated stable isotope profiles (Fig. S4) to the profile of the blue whale (Fig. 1).
@@
 
-\section{References}
+% References
 
@@
 \bibliography{blue-whale}
-
-Supplementary Information is linked to the online version of the paper at www.nature.com/nature.
 
@@
 CT created the movement model.
-EC and NC helped parameterise the model.
+EC, NC and RS helped parameterise the model.
 AJ created the figures. 
-NC wrote the paper.
+NC, CT and EC wrote the paper.
 All authors edited drafts and approved the final version.

### 2017-06-21 Figure-S4-alternative-movement-d13C.R
added correlation tests to back up assertion that sims match empirical

@@ -136,6 +136,9 @@ ggsave("figures/Figure-S4d-mid-atlantic-resident-model-d13C.png", MidAtl_plot,
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 
+# I started doing this with a custom function multiplot, but 
+# swapped to ggplot2::facet_wrap below
+
 # create a multiplot using code from 
 # http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
 
@@ -149,57 +152,58 @@ ggsave("figures/Figure-S4d-mid-atlantic-resident-model-d13C.png", MidAtl_plot,
 # then plot 1 will go in the upper left, 2 will go in the upper right, and
 # 3 will go all the way across the bottom.
 #
-multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL, 
-                      fsave = FALSE, fn = NULL, width = NULL, height = NULL) {
-  library(grid)
-  
-  # Make a list from the ... arguments and plotlist
-  plots <- c(list(...), plotlist)
-  
-  numPlots = length(plots)
-  
-  # If layout is NULL, then use 'cols' to determine layout
-  if (is.null(layout)) {
-    # Make the panel
-    # ncol: Number of columns of plots
-    # nrow: Number of rows needed, calculated from # of cols
-    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
-                     ncol = cols, nrow = ceiling(numPlots/cols))
-  }
-  
-  if (numPlots==1) {
-    print(plots[[1]])
-    
-  } else {
-    # Set up the page
-    grid.newpage()
-    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
-    
-    # Make each plot, in the correct location
-    for (i in 1:numPlots) {
-      # Get the i,j matrix positions of the regions that contain this subplot
-      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
-      
-      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
-                                      layout.pos.col = matchidx$col))
-    }
-  }
-  
-  if (fsave) png(filename = fn, width = width, height = height)
-}
+# multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL, 
[…]

### 2017-06-21 figure1-create-panels.R

@@ -277,10 +277,56 @@ ggsave("figures/Figure-1c-migratory-model-d13C.png", track_plot,
        device = png(width = 600, height = 400))
 
 
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# calculate the correlation between the simulated and observed data
+
+zEmpC <- scale(KC7$d13C)
+zDays <- KC7$days + x_shift # align the days between empirical and simulated
+
+
+# predict the loess on the same days
+
+# a function to build the data.frame from the loess fit
+# I need this here as a new "days" vector has to be 
+# created from the length of the predicted loess object
+predict_loess <- function(df, span = 0.1, newdata = NULL){
+  Z <- predict(loess(df$d13C ~ days, 
+                     span = span), newdata = newdata)
+  
+  out <- data.frame(lo = Z, days = seq(1:length(Z)))
+} # end function
+
+
+zSimC <- resTrack %>% group_by(Rep) %>% do(., data.frame(d13C = scale(.$d13C)))
+
+zSimMatched <- zSimC %>% group_by(Rep) %>% 
+  do(., predict_loess(., span = 0.05, newdata = zDays))
 
 
 
+corEmpSim <- zSimMatched %>% group_by(Rep) %>% 
+  summarise(cor = cor(zEmpC, lo), p = cor.test(zEmpC, lo)$p.value)
 
+hist_cor <- ggplot(corEmpSim, aes(cor)) + 
+  geom_histogram(binwidth = 0.1, color = "black", fill = "grey") + 
+  xlim(-0.5, 1) + theme_classic(base_size = 16) + 
+  xlab("Correlation coefficient") + scale_y_continuous(expand = c(0, 0))
+hist_cor
+
+
+hist_cor_p <- ggplot(corEmpSim, aes(p)) + 
+  geom_histogram(binwidth = 0.01, color = "black", fill = "grey") + 
+   theme_classic(base_size = 16) + 
+  xlab("Correlation coefficient") + scale_y_continuous(expand = c(0, 0))
+hist_cor_p
+
+
+# some summary statistics
+corEmpSim %>% summarise(mean = mean(cor), median = median(cor), 
+                        mode = hdrcde::hdr(cor)$mode, sd = sd(cor))
+
+# number of p-values less than 0.05
+p_crit <- sum(corEmpSim$p <= 0.05 / 50)
 
 
 

### 2017-06-21 Figure-S1-S2-code.R
added S1 and S2

@@ -1,0 +1,121 @@
+# Author: Clive Trueman
+# Adapted by: Andrew Jackson 19-Jun-2017
+# About: script to generate figure 1
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+# setwd("/Users/trueman/Desktop/R scripts/migrate.files")
+
+
+library(raster)
+library(gstat)
+library(sp)
+library(maps)
+library(mapdata)
+
+library(tidyr)
+library(dplyr)
+library(ggplot2)
+library(magrittr)
+library(viridis)
+library(lubridate)
+
+library(TSA)
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+rm(list = ls())
+graphics.off()
+
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Panel A - raw whale isotope data with two y-axes
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+
+whale_isos <- read.csv("data/KC.NHM.full.csv", header = TRUE, 
+                       stringsAsFactors = FALSE)
+
+KC7 <- filter(whale_isos, Whale == "KC7")
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# convert sample number to days using following parameters
+samp_resolution <- 1 # samples per centimetre
+
+growth_rate <- 13.5  # baleen growth rate in centimetres per year
+
+days <- KC7$Samp.No * samp_resolution * 365.25 / growth_rate
+
+days <- days - days[1]
+
+KC7$days <- days
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# convert to calendar date
+
+# reverse time
+rev_days <- days - days[length(days)]
+
+last_sample <- as.Date("1891-03-01")
+
+sample_date <- last_sample + rev_days
+
+KC7$rev_days <- rev_days
+KC7$date <- sample_date
+KC7$year <- year(sample_date)
+
+# calculate a rolling mean to identify the changes between years
+
+idx_new_year <- c(0, diff(KC7$year))
+
+KC7$day_new_year <- days * idx_new_year - 
[…]

### 2017-06-21 figure1-create-panels.R
working on getting dates onto simulated tracks

@@ -82,7 +82,7 @@ point_size <- 4
 
 isop <- ggplot(KC7, aes(x = Samp.No)) + 
   theme_classic(base_size = 16) + 
-  xlab("Sample (first sample most recent)") + 
+  xlab("Baleen sample (cm from youngest sample)") + 
   scale_x_continuous(breaks = seq(0,80,20), 
                      labels = seq(80,0,-20))
 
@@ -202,6 +202,12 @@ ggsave("figures/Figure-1b-migratory-model-full-map.png", mp2,
 # ----
 # Panel C
 
+
+
+
+# create a vector of dates corresponding to each day
+resTrack$rev_days <- resTrack$Day.No - max(resTrack$Day.No)
+resTrack$date     <- as.Date(resTrack$rev_days, origin = last_sample)
 
 # set the constant to add as a correction to the d13C data
 cnst <- 6

### 2017-06-21 manuscript/blue-whale-supplementary.tex
Initial conversion of supplemental to LaTeX

% Preamble
\documentclass[a4paper,12pt]{article}

\usepackage[osf]{mathpazo} % palatino
%\usepackage{ms1}            % load the template
%\usepackage[round]{natbib} % author-year citations
\usepackage[superscript,biblabel]{cite} % for superscript citations
\usepackage{graphicx}
\usepackage{parskip} 
\usepackage{amsmath}
\usepackage{textcomp} % for parts per mille symbol  
\usepackage{pdflscape} % landscape   
\pagenumbering{arabic}    
\linespread{1.66}

\renewcommand*{\thefigure}{S\arabic{figure}} % make Fig S1 not Fig 1
\renewcommand*{\thetable}{S\arabic{table}} % make Table S1 not Table 1

% Title page information
\title{Supplementary Information from:\\
\textit{Reconstructing the last known movements of one of Nature's giants}}
% 90 characters max

\author{Natalie Cooper, Andrew L. Jackson, Katharyn S. Chadwick,\\ 
Ellen J. Coombs, Richard C. Sabin and Clive N. Trueman}

\date{}

%\vfill

% End of preamble

\begin{document}
%\title{Supplementary Information - Cooper et al. 2017}
%\modulolinenumbers[1]   % Line numbering on every line

\maketitle

\parindent = 1.5em
\addtolength{\parskip}{.3em}

\section*{Supplementary Methods}

\subsection*{CHARACTERISATION OF THE MEASURED BALEEN}
The baleen plate yielded 78 discrete samples of baleen. 
Across the baleen sample, $\delta^{15}$N values show clear cycles with periods of relatively high $\delta^{15}$N values interspersed by relatively short periods where $\delta^{15}$N values are c.1\text{\textperthousand} lower. 
Fourier analysis (periodogram function in the R package TSA\cite{Chan:2012aa}) revealed a strong periodic repetition with a 13.3cm periodicity. 
$\delta^{13}$C values also show repeated fluctuations in the first (oldest) 60cm of the baleen plate (behavioural phase one), with a similar periodicity of 13.5cm (Fig. \ref{figs1}). 
A distinct change in the isotopic pattern along the baleen plate occurs at around 60cm behavioural phase two), with a transition to relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.

%  \begin{figure}[!htbp]
%    \centering
%      \includegraphics[width=12cm]{XXX.pdf}
%      \caption{Fourier analysis of $delta^{15}$N (A,B) and $delta^{13}$C (C,D) values of the full (A,C) and behavioral phase one only (B,D) components of the baleen plate. Inset values are the two highest frequency spectral densities (in cm). Note the baleen plate is 80cm long. Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent behavioral phase two}
%      \label{figs1}
%  \end{figure}

Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioral phase one (oldest 60cm of the baleen plate; Fig. \ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioral phase two (the final 20cm of the baleen plate).

%  \begin{figure}[!htbp]
%    \centering
%      \includegraphics[width=12cm]{XXX.pdf}
%      \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for behavioural phase one showing strong negative covariance with a periodicity of around 13cm}
%      \label{figs2}
%  \end{figure}

\subsection*{BASELINE ISOTOPE COMPARISONS}
Isotope-enabled biogeochemical ocean models\cite{magozzi2017using} (Schmittner and Somes) were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds. 
Annual average $\delta^{15}$N POM values were provided by C.J. Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model. 
$\delta^{13}$C POM values were simulated using an isotopic extension to the NEMO-Medusa model\cite{magozzi2017using}.

Simulated $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c.60$^{\circ}$N, and relatively negative in the central and southern North Atlantic, associated with high levels of nitrogen fixation. 
Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. \ref{figs3}). 
The temporally coincident periodic, inverse fluctuations in $\delta^{13}$C and $\delta^{15}$N values in the initial phase of the baleen plate are therefore consistent with migration between waters north of 60$^{\circ}$N and more southerly southern waters in the North Atlantic, with northerly occupancy in summer months and southerly occupancy in winter months. 
Incorporation of carbon and nitrogen with isotopic ratios consistent with southerly latitudes during winter months indicates continued feeding during migration and winter residency.

 %  \begin{figure}[!htbp]
%    \centering
%      \includegraphics[width=12cm]{XXX.pdf}
%      \caption{$delta^{13}$C POM values}
%      \label{figs3}
%  \end{figure}

The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature (Laws et al 2017). Thus temporal variations in $\delta^{13}$C POM values are superimposed on latitudinal gradients. The scale and nature of temporal variation in $\delta^{13}$C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in $\delta^{13}$C POM values linked to strongly seasonal phytoplankton growth dynamics. We therefore used monthly simulated $\delta^{13}$C POM values to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours.

\subsection*{ISOTOPE MODEL SIMULATIONS}
We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions of that would be encountered by whales foraging within broad geographic regions (Norwegian Sea, west UK/Irish shelf, Canaries/west Azores, Mid Atlantic ridge; Fig. \ref{figs4}). Strong seasonal dynamics in $\delta^{13}$C values are evident in all northerly regions, characterized by a rapid increase to annual maximum $\delta^{13}$C values associated with the onset of the spring phytoplankton bloom, followed by more gradual decline in $\delta^{13}$C values towards minima in winter conditions (Fig. \ref{figs4}). 
In temperate latitudes around the British Isles, seasonal dynamic cycles are present, but dampened (Fig. \ref{figs4}), whereas in subtropical regions exemplified by the Canaries and western Azores/Mid Atlantic ridge areas, $\delta^{13}$C values are relatively positive and constant throughout the year (Fig. \ref{figs4}).

%ADD THE LOCTION heat maps FOR EACH

  \begin{figure}[!htbp]
    \centering
      \includegraphics[width=14cm]{figures/Figure-S4-all.png}
      \caption{Daily $\delta^{13}$C POM values in regions occupied by 30 simulated blue whales foraging over a two year cycle.}
      \label{figs4}
  \end{figure}

\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase one}
Temporal dynamics in $\delta^{13}$C values simulated in northerly latitudes are similar to the measured $\delta^{13}$C profile during the four years of behavioral phase one. 
Simulated profiles display the same asymmetric isotopic cycles associated with development of the spring bloom, with annual $\delta^{13}$C maxima in June (Fig. \ref{figs4}, Fig. 1). 
Simulated temporal isotopic profiles for northern resident whales, do not display clear isotopic minima preceding the onset of the bloom. 
Adding seasonal north-south migrations within mid-high latitude regions to foraging models, closely matches the measured profile, including $\delta^{13}$C minima preceding the June bloom (Fig. \ref{figs4}, Fig. 1).

\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase two}
Behavioural phase two is characterized by a dramatic shift to maximum $\delta^{13}$C values in June of 1889 followed by approximately 18 months of relatively positive and constant $\delta^{13}$C values. 
$\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity associated with north-south migration is absent in 1890. 
The final sample representing feeding shortly before death in Irish waters implies a return to isotopic conditions more consistent with behavioural phase one. 
The relatively positive and constant carbon isotope composition of behavioural phase two is consistent with residency (and continued feeding) in more southerly latitudes, particularly in the relatively isotopically positive waters southwest of the Azores.

Accordingly we modeled whale movements allowing five years of seasonal north-south migration in the north east northern Atlantic, followed by migration to subtropical waters in the spring of 1890, relative residency for 12 months and return to more northerly waters in the spring of 1891.
The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Fig. 1). 
Consistently low $\delta^{15}$N values in 1890 are also consistent with residency in relatively southern waters.

We note however that this interpretation assumes that the physiological demands of pregnancy and lactation do not create an isotopic expression. 
A captive feeding study in south American sea lions\cite{cardona2017temporal} showed positive $\delta^{13}$C values associated with pregnancy, falling at birth, and a gradual increase in $\delta^{15}$N values during lactation. 
The dramatic positive excursion in $\delta^{13}$C values in the spring/summer of 1888 cannot be replicated purely through movement to warmer waters, thus we tentatively ascribe this signal to the physiological demands of pregnancy.  
$\delta^{15}$N values in the autumn of 1890 show a gradual increase, consistent with the isotopic expression associated with lactation in south American fur seals\cite{cardona2017temporal}.

Blue whales have a gestation period of 10-12 months, and calves suckle for approximately 6-7 months\cite{handbook}. 
The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the assumed reproductive ecology of blue whales.

\subsection*{AGENT BASED WHALE MOVEMENT MODEL}
We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using,yool2013medusa}, where movement behavior is influenced by SST and phytoplankton biomass estimates provided by NEMO-MEDUSA\cite{yool2013medusa}, and bathymetry (Fig. \ref{figs5}).

The agent based model developed employs a simple set of movement rules dictated by behavioural state which is itself fixed according to calendar month by the operator. 
At each daily step, the probability of movement, movement directions and distance traveled are sampled from probability distributions to allow individual variation.

%  \begin{figure}[!htbp]
%    \centering
%      \includegraphics[width=12cm]{XXX.pdf}
%      \caption{Summary diagram of agent-based movement model structure.}
%      \label{figs5}
%  \end{figure}

Initial boundary conditions are defined with a maximum temperature of 25$^{\circ}$C and minimum temperature of 3$^{\circ}$C. The likelihood of movement (i.e. whether to move or not from the current location) is sampled from a binomial distribution with the probability of movement influenced by behavioural state and external conditions. The maximum daily movement distance permitted in each behavioural state is defined as a random sample of a Gaussian distribution with specified mean and standard deviation (see Table \ref{tables1}).

The behavioural states employed in the model simulations are foraging, migrating north, migrating south and nursing. When foraging, the probability of movement and likely direction of movement is primarily controlled by the biomass of phytoplankton in the current and adjacent 1$^{\circ}$ cells. When migrating, a fixed higher probability of movement is given to southerly or northerly movement directions, and the probability of movement is influenced by the ambient sea surface temperature and phytoplankton biomass. When nursing a calf, movement is limited to a daily mean of 25km, as long as sea surface temperatures remain between 18$^{\circ}$C and 25$^{\circ}$C.

Parameters defined for the agent based model best matching the measured profile are summarized in Table \ref{tables1}.

\newpage

\begin{landscape}

\centering
\begin{table}
  \begin{tabular}{|p{5cm}|c|c|c|c|} 
    \hline
    & \multicolumn{4}{|c|}{\textbf{Behavioural state}} \\
    \hline
    \textbf{Parameters} & \textbf{Foraging} & \textbf{Migrating north} & \textbf{Migrating south} & \textbf{Nursing}\\
    \hline
    Month - behavioural phase one (1460 days) & May - Oct & Feb - April & Nov - Jan & NA\\
    \hline
    Month - pregnancy (365 days) & June - Oct & April - May & Nov - March & NA\\
    \hline
    Month - nursing (240 days) & NA & NA & NA & Month???\\
    \hline
    Minimum sea surface temperature ($^{\circ}$C) & 3 & 3 & 3 & 18\\
    \hline
    Maximum sea surface temperature ($^{\circ}$C) & 25 & 25 & 25 & 25\\
    \hline
    Maximum daily distance traveled (km) & $N(\mu=30, \sigma=10)$ & $N(\mu=150, \sigma=10)$ & $N(\mu=150, \sigma=10)$ & $N(\mu=20, \sigma=5)$\\
    \hline
    Plankton biomass threshold for remaining to forage (units) & 50 & NA & NA & NA\\
    \hline
  \end{tabular}
  \caption{Model parameters for agent based model of whale movement.}
  \label{tables1}
\end{table}

\end{landscape}

% References
\newpage
\bibliographystyle{nature}
\bibliography{blue-whale}

\end{document}

### 2017-06-21 figure1-create-panels.R
nailed the placement of year markers

@@ -63,15 +63,19 @@ KC7$rev_days <- rev_days
 KC7$date <- sample_date
 KC7$year <- year(sample_date)
 
-# calculate a rolling mean to identify the changes between years
 
-idx_new_year <- c(0, diff(KC7$year))
+new_years <- with(KC7, seq(min(year), max(year), by = 1))
 
-KC7$day_new_year <- days * idx_new_year - 
-  (idx_new_year * 0.5 * samp_resolution * 365.25 / growth_rate)
+new_years_chr <- paste(new_years, "01", "01", sep = "-")
 
-KC7$samp_new_year <- KC7$Samp.No * idx_new_year - 
-  (idx_new_year * 0.5)
+new_years_date <- as.Date(new_years_chr)
+
+new_years_day <- tail(KC7$days,1) - 
+  (as.numeric(last_sample - new_years_date ))
+
+new_years_samp <- round(new_years_day[2:length(new_years_day)] * growth_rate / 
+                          (samp_resolution * 365.25))
+
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 
@@ -104,8 +108,7 @@ isop <- isop +
     sec.axis = sec_axis(~.+v_shift, 
                         name = expression(paste(delta^{15}, "N (\u2030)"))))
 
-isop <- isop + geom_vline(xintercept = 
-                            with(KC7,samp_new_year[samp_new_year > 0]),
+isop <- isop + geom_vline(xintercept = new_years_samp,
                           color = "grey", lty = 2)
 
 print(isop)
@@ -231,8 +234,8 @@ tmp <- resTrack %>% group_by(Rep) %>%
 
 # data required to add a vertical line indicating the start of migration
 migrate <- as.Date("1889-06-01")
-strand_to_migrate <- tail(KC7$date,1) - migrate
-migrate_day <- tail(KC7$days,1) - strand_to_migrate
+first_sim_day <- as.Date("1885-01-01")
+migrate_day <- as.numeric(migrate - first_sim_day)
 
 # plot the daily tracks
 track_plot <- ggplot(tmp, aes(days, lo, group = Rep)) 
@@ -240,8 +243,17 @@ track_plot <- ggplot(tmp, aes(days, lo, group = Rep))
 # add the vertical line where migration starts in the simulations
 track_plot <- track_plot + geom_vline(xintercept = as.numeric(migrate_day),
                                       lty = 2,
-                                      color = "grey",
+                                      color = "black",
                                       size = 1)
+
[…]

### 2017-06-21 Figure-S1-S2-code.R
point figure writing to manuscript/figures

@@ -77,7 +77,8 @@ KC7$samp_new_year <- KC7$Samp.No * idx_new_year -
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 
-png(filename = "figures/Figure-S1-periodograms.png", width = 1000, height = 600)
+png(filename = "manuscript/figures/Figure-S1-periodograms.png", 
+    width = 1000, height = 600)
 
 par(mfrow = c(2,2))
 
@@ -102,7 +103,8 @@ TSA::acf(cbind(KC7$d13C[1:60], KC7$d15N[1:60]), main = "Oldest 60cm")
 TSA::acf(cbind(KC7$d13C[61:n], KC7$d15N[61:n]), main = "Youngest 20cm")
 
 # save to file
-png(filename = "figures/Figure-S2-cross-cor.png", width = 1000, height = 600)
+png(filename = "manuscript/figures/Figure-S2-cross-cor.png", 
+    width = 1000, height = 600)
 
 par(mfrow = c(1,2))
 TSA::acf(cbind(KC7$d13C[1:60], KC7$d15N[1:60]), main = "Oldest 60cm")

### 2017-06-21 Figure-S4-alternative-movement-d13C.R

@@ -62,8 +62,8 @@ nor_plot <- ggplot(resNor, aes(Day.No, lo, group = Rep)) +
 
 print(nor_plot)
 
-ggsave("figures/Figure-S4a-norway-resident-model-d13C.png", nor_plot, 
-       device = png(width = 600, height = 400))
+# ggsave("figures/Figure-S4a-norway-resident-model-d13C.png", nor_plot, 
+#        device = png(width = 600, height = 400))
 
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
@@ -83,8 +83,8 @@ ire_plot <- ggplot(resIre, aes(Day.No, lo, group = Rep)) +
 
 print(ire_plot)
 
-ggsave("figures/Figure-S4b-ireland-resident-model-d13C.png", ire_plot, 
-       device = png(width = 600, height = 400))
+# ggsave("figures/Figure-S4b-ireland-resident-model-d13C.png", ire_plot,
+#        device = png(width = 600, height = 400))
 
 
 
@@ -109,8 +109,8 @@ can_plot <- ggplot(resCan, aes(Day.No, lo, group = Rep)) +
 
 print(can_plot)
 
-ggsave("figures/Figure-S4c-canada-resident-model-d13C.png", can_plot, 
-       device = png(width = 600, height = 400))
+# ggsave("figures/Figure-S4c-canada-resident-model-d13C.png", can_plot, 
+#        device = png(width = 600, height = 400))
 
 
 
@@ -131,8 +131,8 @@ MidAtl_plot <- ggplot(resMidAtl, aes(Day.No, lo, group = Rep)) +
 
 print(MidAtl_plot)
 
-ggsave("figures/Figure-S4d-mid-atlantic-resident-model-d13C.png", MidAtl_plot, 
-       device = png(width = 600, height = 400))
+# ggsave("figures/Figure-S4d-mid-atlantic-resident-model-d13C.png", MidAtl_plot, 
+#        device = png(width = 600, height = 400))
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 
@@ -230,6 +230,6 @@ sim_facet <- sim_facet + facet_wrap(~Region, ncol = 2)
 
 print(sim_facet)
 
-ggsave("figures/Figure-S4-facet-wrap-d13C.png", sim_facet, 
+ggsave("manuscript/figures/Figure-S4-facet-wrap-d13C.png", sim_facet, 
        device = png(width = 600, height = 400))
 

### 2017-06-21 figure1-create-panels.R

@@ -113,7 +113,7 @@ isop <- isop + geom_vline(xintercept = new_years_samp,
 
 print(isop)
 
-ggsave("figures/Figure-1a-raw-dC-dN-data.png", isop, 
+ggsave("manuscript/figures/Figure-1a-raw-dC-dN-data.png", isop, 
        device = png(width = 600, height = 400))
 
 # with(KC7, day_new_year[day_new_year > 0]))
@@ -194,7 +194,7 @@ mp2 <- mp2 + theme(axis.line  = element_blank(),
 
 print(mp2)
 
-ggsave("figures/Figure-1b-migratory-model-full-map.png", mp2, 
+ggsave("manuscript/figures/Figure-1b-migratory-model-full-map.png", mp2, 
        device = png(width = 600, height = 600, units = "px"))
 
 
@@ -291,7 +291,7 @@ track_plot <- track_plot + geom_point(data = KC7,
 
 print(track_plot)
 
-ggsave("figures/Figure-1c-migratory-model-d13C.png", track_plot, 
+ggsave("manuscript/figures/Figure-1c-migratory-model-d13C.png", track_plot, 
        device = png(width = 600, height = 400))
 
 

### 2017-06-22 manuscript/blue-whale-ms.tex
Some faffing with figure positioning

@@
 \usepackage{parskip} 
+\usepackage{caption}
+\usepackage{subcaption}
 \usepackage{textcomp} % for parts per mille symbol     
@@
 %  \end{figure}
+
+% 2 panel figure of the two traces - real and simulated
+
+\begin{figure}
+  \centering
+    \begin{subfigure}{0.8\textwidth}
+      \centering
+      \includegraphics[width = \linewidth]{figures/Figure-1a-raw-dC-dN-data.png}
+      \label{fig1a}
+    \end{subfigure}
+
+    \begin{subfigure}{0.8\textwidth}
+      \centering
+      \includegraphics[width = \linewidth]{figures/Figure-1c-migratory-model-d13C.png}
+      \label{fig1b}
+    \end{subfigure} 
+  \caption{Two panel figure with two traces}
+\end{figure}
+
+% two panel with the simulations and map
+\begin{figure}
+  \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width = \linewidth]{figures/Figure-1b-migratory-model-full-map.png}
+      \label{fig2a}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width = \linewidth]{figures/Figure-1c-migratory-model-d13C.png}
+      \label{fig2b}
+    \end{subfigure} 
+  \caption{Two panel figure with trace and map}
+\end{figure}
+
+% two panel with the simulations and map
+\begin{figure}
+  \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width = \linewidth]{figures/Figure-1b-migratory-model-full-map.png}
+      \label{figa}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width = \linewidth]{figures/Figure-1c-migratory-model-d13C.png}
+      \label{figb}
+    \end{subfigure} 
+  \caption{Two panel figure with trace and map}
+\end{figure}
+
+% two panel with the simulations and map
+\begin{figure}
+  \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width = \linewidth]{figures/Figure-1c-migratory-model-d13C.png}
+      \label{fig3a}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width = \linewidth]{figures/Figure-1b-migratory-model-full-map.png}
+      \label{fig3b}
+    \end{subfigure} 
+  \caption{Two panel figure with trace and map}
+\end{figure}
+
+% one panel of the real trace
+\begin{figure}
+  \centering
+  \includegraphics[width = \linewidth]{figures/Figure-1a-raw-dC-dN-data.png}
+  \label{fig11a}
+  \caption{One panel figure}
+\end{figure}
+
+% one panel of the real trace
+\begin{figure}
+  \centering
+  \includegraphics[width = \linewidth]{figures/Figure-1b-migratory-model-full-map.png}
+  \label{fig11b}
+  \caption{One panel figure}
+\end{figure}
+
+% one panel of the real trace
+\begin{figure}
+  \centering
+  \includegraphics[width = \linewidth]{figures/Figure-1c-migratory-model-d13C.png}
+  \label{fig11c}
+  \caption{One panel figure}
+\end{figure}
 

### 2017-06-23 manuscript/blue-whale-ms.tex
Add CT edits to abstract

@@
 \paragraph{Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}. 
-Current methods for monitoring movement behaviours are often expensive, and do not provide information on historical or long-term range movements.
-Animal tissues housed in museum collections present an alternative, providing biochemical records of behaviour that can be used to reconstruct historical individual-level movements relatively inexpensively\cite{newsome2010using}.
-Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct the last years of an iconic blue whale (Balaenoptera musculus) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London (NHM).
-We recover individual-level movement information at high temporal resolution over six years and show distinct periods of foraging and short distance north-south migrations in sub-Arctic waters and calving in subtropical waters.
-We also detect isotopic signatures consistent with pregnancy and lactation in the year before the whale's death. 
-These results supply the first quantitative evidence of historical migrations in blue whales, and suggest that blue whale residency and movement patterns in the North Atlantic have remained similar for hundreds of years, despite changing human pressures in the region.
+Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long time scales and under historic climatic and ecological conditions\cite{newsome2010using}.
+Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct movements over the last six years of the life of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London (NHM).
+We identify two distinct phases in the whale's behaviour: repeated annual north-south migrations in sub-Arctic waters, and residency in subtropical waters. 
+Subtropical residency is associated with isotopic signatures consistent with pregnancy and lactation in the year before the whale’s death. 
+These results supply the first quantitative multi-yeat records of historical migrations in individual blue whales, and suggest that blue whale residency and movement patterns in the North Atlantic have remained similar for hundreds of years, despite changing human pressures in the region.
 Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climatic, ecological conditions and human-induced pressures, and reveal aspects of life history.
-Combining biochemical tracers with simulation modeling unlocks the information recorded in these irreplaceable archives.}
+Combining biochemical tracers with simulation modelling unlocks the information recorded in these irreplaceable archives.}
 

### 2017-06-23 manuscript/blue-whale-ms.tex
Add CT edits to paragraph 1 main text

@@
 Such species are also likely to be particularly vulnerable to climate change, as changes may impact various parts of their life cycle\cite{robinson2009travelling}. 
-Dealing with these issues is difficult due to the relative scarcity of information on species movements over multiple years, for both historical and present-day populations, particularly for wide-ranging marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
+Dealing with these issues is difficult due to the relative scarcity of information on individual-level movements over multiple years, for both historical and present-day populations, particularly for wide-ranging marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
 

### 2017-06-23 manuscript/blue-whale-ms.tex
Add CT edits to paragraph 2 main text

@@
 Despite blue whale hunting being banned for over 50 years, blue whales are still listed as Endangered by the IUCN Red List\cite{reilly2008balaenoptera}.
-Conservation efforts are hampered by the scarcity of detailed data on blue whale distribution, behaviour and ecology, for both historical and present-day populations. 
+Conservation efforts are hampered by the scarcity of detailed data on the distribution, behaviour and ecology of blue whales, both for historical and present-day populations. 
 Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution}).
-These methods are adding vital behavioural data to our understanding of whale movements, however, they are expensive, often limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
-An alternative technique is to investigate oscillations in stable isotopes values of whale tissues housed in museum collections\cite{ryan2013stable}.
-This is a non-invasive method that provides individual-level movement information at high temporal resolution over multiple years. 
+These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
+Intrinsic biogeochemical tags provide an alternative geo-location technique, whereby chemical compositions of animal tissues are compared to spatially referenced models of chemical variation across the species’ range (Hobsen review, Bowen BOOK).
+Stable isotope-based geolocation has been used to interpret whale ecology \cite{hobson1998stable,ryan2013stable} Eisenmann et al 2016). 
 Here we apply this method to baleen from the Natural History Museum's blue whale.

### 2017-06-23 manuscript/blue-whale-ms.tex
Add CT edits to paragraph 3 main text

@@
 It is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable}. 
-We sampled keratin from the blue whale's baleen plate at 1cm intervals and used continuous-flow isotope ratio mass spectrometry to extract $\delta^{13}$C and $\delta^{15}$N isotope concentrations. 
+We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth and used continuous-flow isotope ratio mass spectrometry determine the isotopic compositions of carbon and nitrogen (expressed as $\delta^{13}$C and $\delta^{15}$N values).
 Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplemental Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the last six years of whale's life (autumn 1885 - spring 1891).
 The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour.
-In behavioural phase one (month 1885 to May 1889), we find strong cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values of an approximately yearly periodicity (Fig.
-1).
-$\delta^{13}$C and $\delta^{15}$N exhibit strong negative covariance.
-In behavioural phase two (June 1889 to March 1891) we see a sharp uptick in $\delta^{13}$C, followed by high and relatively constant $\delta^{13}$C values. ,
+In behavioural phase one (month 1885 to May 1889), we find strong cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values with strong negative covariance (Fig. \ref{fig1}; Fig. S?). 
+In behavioural phase two (June 1889 to March 1891) we see a sharp increase in $\delta^{13}$C values for approximately eight months.
 This is associated with an overall reduction in $\delta^{15}$N values and a weakening of the periodic $\delta^{15}$N fluctuations seen in behavioural phase one. 
-There is also no covariance between $\delta^{13}$C and $\delta^{15}$N (Fig.
-1).
+There is also no covariance between $\delta^{13}$C and $\delta^{15}$N values in behavioral phase two (Fig. \ref{fig1}).
 

### 2017-06-23 manuscript/blue-whale-ms.tex
Change references to figures now we are splitting into three
Add CT edits to paragraph 4 main text

@@
 
-%  \begin{figure}[!htbp]
-%    \centering
-%      \includegraphics[width=12cm]{XXX.pdf}
-%      \caption{}
-%      \label{}
-%  \end{figure}
-
-% 2 panel figure of the two traces - real and simulated
-
-\begin{figure}
-  \centering
-    \begin{subfigure}{0.8\textwidth}
-      \centering
-      \includegraphics[width = \linewidth]{figures/Figure-1a-raw-dC-dN-data.png}
-      \label{fig1a}
-    \end{subfigure}
-
-    \begin{subfigure}{0.8\textwidth}
-      \centering
-      \includegraphics[width = \linewidth]{figures/Figure-1c-migratory-model-d13C.png}
-      \label{fig1b}
-    \end{subfigure} 
-  \caption{Two panel figure with two traces}
-\end{figure}
-
-% two panel with the simulations and map
-\begin{figure}
-  \centering
-    \begin{subfigure}[b]{0.45\textwidth}
-      \centering
-      \includegraphics[width = \linewidth]{figures/Figure-1b-migratory-model-full-map.png}
-      \label{fig2a}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.45\textwidth}
-      \centering
-      \includegraphics[width = \linewidth]{figures/Figure-1c-migratory-model-d13C.png}
-      \label{fig2b}
-    \end{subfigure} 
-  \caption{Two panel figure with trace and map}
-\end{figure}
-
-% two panel with the simulations and map
-\begin{figure}
-  \centering
-    \begin{subfigure}[b]{0.45\textwidth}
-      \centering
-      \includegraphics[width = \linewidth]{figures/Figure-1b-migratory-model-full-map.png}
-      \label{figa}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.45\textwidth}
-      \centering
-      \includegraphics[width = \linewidth]{figures/Figure-1c-migratory-model-d13C.png}
-      \label{figb}
-    \end{subfigure} 
-  \caption{Two panel figure with trace and map}
-\end{figure}
-
-% two panel with the simulations and map
-\begin{figure}
-  \centering
-    \begin{subfigure}[b]{0.45\textwidth}
-      \centering
-      \includegraphics[width = \linewidth]{figures/Figure-1c-migratory-model-d13C.png}
-      \label{fig3a}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.45\textwidth}
-      \centering
-      \includegraphics[width = \linewidth]{figures/Figure-1b-migratory-model-full-map.png}
-      \label{fig3b}
-    \end{subfigure} 
-  \caption{Two panel figure with trace and map}
-\end{figure}
-
-% one panel of the real trace
+% figure 1
 \begin{figure}
@@
   \includegraphics[width = \linewidth]{figures/Figure-1a-raw-dC-dN-data.png}
-  \label{fig11a}
+  \label{fig1}
   \caption{One panel figure}
@@
 
-% one panel of the real trace
+% figure 2
 \begin{figure}
@@
   \includegraphics[width = \linewidth]{figures/Figure-1b-migratory-model-full-map.png}
-  \label{fig11b}
+  \label{fig2}
   \caption{One panel figure}
@@
 
-% one panel of the real trace
+% figure 3
 \begin{figure}
@@
   \includegraphics[width = \linewidth]{figures/Figure-1c-migratory-model-d13C.png}
-  \label{fig11c}
+  \label{fig3}
   \caption{One panel figure}
@@
 
-To interpret these isotopic signatures, we simulated time series of $\delta^{13}$C and $\delta^{15}$N values expected under differing movement behaviours, by coupling a novel agent-based model of whale movements (see Supplementary Methods) to models of global phytoplankton $\delta^{13}$C and $\delta^{15}$N values\cite{magozzi2017using} (Schmittner and Somes).
-We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth and phytoplankton concentration (as a proxy for zooplankton food availability).
-At each time point and location in the simulations, we extracted phytoplankton $\delta^{13}$C and $\delta^{15}$N values from the models\cite{magozzi2017using} (Schmittner and Somes) resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas.
-We then determined which of these simulated profiles were biogeochemically consistent with the isotopic profile of the blue whale's baleen (Fig. 1) and use this to interpret the results.
+Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values and because the isotopic composition of phytoplankton at the base of marine food webs varies in space and time.
+We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using}, coupled to an agent-based model of whale movements (see Supplementary Methods) to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of blue whales.
+We then compare simulated profiles with the measured isotopic profile of the blue whale's baleen (Fig. \ref{fig1}) to identify the most likely movement history.
 

### 2017-06-23 manuscript/blue-whale-ms.tex
Add CT edits to paragraph 5 main text

@@
 This interpretation is also consistent with the relatively negative median $\delta^{13}$C values (Fig. 1). 
-The periodicity of $\delta^{13}$C and $\delta^{15}$N values is explained by short distance seasonal migrations, with the whale overwintering off the coast of Ireland. 
-The relatively positive and invariant $\delta^{13}$C values seen during behavioural phase two are found in subtropical areas, particularly off the coast of the Azores, suggesting the whale migrated rapidly to southerly latitudes in early 1889, but continued feeding (other mysticete whales fast during migrations, but blue whales are too large to stop feeding for long periods\cite{busquets2017estimating}?silva paper).
-The whale remained in these warm waters for over a year and was returning to her northern feeding grounds when she became stranded.
+The periodicity of $\delta^{13}$C and $\delta^{15}$N values is best explained by short distance seasonal migrations, with the whale overwintering in the central or southern Norwegian Sea. 
+The relatively positive and invariant $\delta^{13}$C values seen during behavioural phase two are found in subtropical areas, particularly to the west of the Azores.
+Continued, but reduced cyclical variation in $\delta^{15}$N values in 1889 and 1890 implies a degree of seasonal north-south migration before rapid migration to southerly latitudes in 1890, but continued feeding (other mysticete whales fast during migrations, but blue whales are too large to stop feeding for long periods\cite{busquets2017estimating}?silva paper).
+The whale remained in these warm waters for around eight months and was conducting a rapid migration returning to her northern feeding grounds when she became stranded.
 

### 2017-06-23 manuscript/blue-whale-ms.tex
Add CT edits to paragraph 6 main text

@@
 
-Movement to warm waters alone, however, cannot explain the dramatic increase in $\delta^{13}$C values at the start of behavioural phase two. 
+The dramatic increase in $\delta^{13}$C values seen between March and July 1889 cannot be simulated purely from movements within the known geographic range.
 We believe that this reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions\cite{cardona2017temporal}. 
@@
 The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the known reproductive ecology of blue whales.
-This is the first quantitative evidence of historical movement patterns in blue whales, and reveals a fascinating snapshot into the life of an iconic specimen, where we previously only knew the date and location of its death.
 
@@
 Interestingly, there is still a resident population of blue whales in the region our models place the blue whale within, and females still breed in the subtropical waters near the Azores\cite{reilly2008balaenoptera}.
-This suggests that whale movements have remained similar over hundreds of years, despite changes in anthropogenic pressures. 
 Whaling was an intense pressure for blue whales during the period we are analysing. 
@@
 The NHM blue whale thus represents an individual from a species at the brink of extinction.
+Our data suggest that the spatial ecology of blue whales in the North Atlantic have remained similar for more than 100 years, despite dramatic changes in anthropogenic pressures.
+
 Blue whales were the first species that humans legislated to save, and represent a big conservation success story, although North Atlantic stocks have not recovered to the same extent as other whales in the region\cite{sigurjonsson1995life,pike2009note}.

### 2017-06-23 manuscript/blue-whale-ms.tex
Add CT edits to paragraph 7/8 main text

@@
 
-Blue whales were the first species that humans legislated to save, and represent a big conservation success story, although North Atlantic stocks have not recovered to the same extent as other whales in the region\cite{sigurjonsson1995life,pike2009note}.
-There is also evidence that blue whales are shifting their ranges due to warming waters off Iceland\cite{10.3389/fevo.2015.00006}. 
-Thus understanding historical movements of North Atlantic blue whales has never been more vital if we wish to preserve them for future generations.
-
-In conclusion, the model we present here allows us to recreate individual-level movement behaviours and life history using tissues from museum collections and other sources. 
-This has applications beyond this individual whale, and beyond cetaceans, as our model can be applied to any marine species with constantly growing tissues, for example seal whiskers, shark skin, and fish scales.
-These techniques can also allow us study extinct species' behaviours and understand how populations were affected by past pressures such as hunting, and present-day pressures such as global change. 
-Finally, our results highlight that museums represent a vast, mostly untapped, treasure trove of information ready for us to explore.
+This is the first quantitative evidence of historical individual-level movement behaviours in blue whales. 
+Our novel simulation modelling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviours over multiple years.
+Such techniques can also be applied to extinct species' behaviours and understanding how populations were affected by past pressures such as hunting, and present-day pressures such as global change. 
+Finally, our results highlight that museums represent a vast, mostly untapped, source of individual-level spatial ecology information.
 

### 2017-06-23 manuscript/blue-whale-ms.tex
Add CT edits to methods first three paragraphs

@@
 
-\section{Methods (condensed)}
+\section{Methods}
 %<3000 words
@@
 Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the most distal (oldest) 65cm of the plate. 
-We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis\cite{cardona2017temporal}, revealing a consistent growth rate of 13.5cm $y^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cm y{^-1}$ estimated for six blue whales from the Californian Pacific\cite{busquets2017estimating}.
+We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis\cite{cardona2017temporal}, revealing a consistent growth rate of 13.5cm $y^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cm y{^-1}$ estimated for six recent blue whales from the Californian Pacific\cite{busquets2017estimating}.
 Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891. 
-Cross-correlation analysis (ref?) demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase one (oldest 60cm of the baleen plate), but no relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase two.
+Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase one (oldest 60cm of the baleen plate), but no relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase two.
 
@@
 Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
-In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig S3). 
-The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature (Laws, Magozzi et al 2017).
+In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S3). 
+The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature\cite{magozzi2017using} Laws et al 1995.
 Thus temporal variations in $\delta^{13}$C POM values were superimposed on latitudinal gradients. 

### 2017-06-23 manuscript/blue-whale-ms.tex
Tiny remaining edits from CT

@@
 \subsection{Agent-based whale movement model}
+We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
+At each time point and location in the simulations, we extracted phytoplankton $\delta^{13}$C values from the spatio-temporal model\cite{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas.
 Movement was coded as set of probabilistic rules. 
@@
 When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.
-(iii) Water depth (m; what reference is this collected from?). 
+(iii) Water depth (m; from the General Bathymetric Chart of the Oceans (GEBCO) global bathymetry dataset). 
 Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
@@
 We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Fig SX; (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
-Various movement rules were implemented to avoid simulations ending up on land, or becoming `stuck' in the relatively shallow waters of the Bay of Biscay or the Strait of Dover. 
 For a full description of the model and its parameters see Supplementary Methods. 
-We simulated whale movements XX times for each scenario and residency hotspot. 
-For each movement simulation we extracted phytoplankton $\delta^{13}$C and $\delta^{15}$N values at each time point and location from the models described above\cite{magozzi2017using} (Somes). 
-We then compared the simulated stable isotope profiles (Fig. S4) to the profile of the blue whale (Fig. 1).
+We simulated whale movements 30 times for each scenario and residency hotspot. 
+For each movement simulation we extracted phytoplankton $\delta^{13}$C values at each time point and location from the models described above\cite{magozzi2017using}. 
+We then compared the simulated stable isotope profiles (Fig. \ref{fig3}) to the profile of the blue whale (Fig. \ref{fig1}).
 
@@
 Data are available on the Natural History Museum, London's data portal (https://data.nhm.ac.uk). 
-R code is available from XXX.
+R code is available from GitHub (github.com/ALJackson/blue-whale-bes).
 
@@
 We thank XXX for comments on drafts.
-We thank CJ Somes for providing  $\delta^{15}$N POM data.
+We thank C.J. Somes for providing  $\delta^{15}$N POM data.
 
@@
 Correspondence and requests for materials should be addressed to natalie.cooper@nhm.ac.uk.
-% One person only allowed
-
-%\section{Tables}
 
@@
 
-%\newpage
 %\section{Figures}
 
-%  \begin{figure}[!htbp]
-%    \centering
-%      \includegraphics[width=12cm]{XXX.pdf}
-%      \caption{}
-%      \label{}
-%  \end{figure}
-
 \end{document}

### 2017-06-23 manuscript/blue-whale-supplementary.tex
Minor edits to supp mat from CT

@@
 Annual average $\delta^{15}$N POM values were provided by C.J. Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model. 
-$\delta^{13}$C POM values were simulated using an isotopic extension to the NEMO-Medusa model\cite{magozzi2017using}.
+$\delta^{13}$C POM values were simulated using an isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using}.
 
@@
 %  \end{figure}
+% should be a two panel fig with the annual average d15N from Somes and d13C from CT
 
-The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature (Laws et al 2017).
+The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature (Laws et al 1995).
 Thus temporal variations in $\delta^{13}$C POM values are superimposed on latitudinal gradients. 
@@
 \ref{figs4}).
-
-%ADD THE LOCTION heat maps FOR EACH
 
@@
 
-Accordingly we modeled whale movements allowing five years of seasonal north-south migration in the north east northern Atlantic, followed by migration to subtropical waters in the spring of 1890, relative residency for 12 months and return to more northerly waters in the spring of 1891.
+Accordingly we modelled whale movements allowing five years of seasonal north-south migration in the north east northern Atlantic, followed by rapid migration to subtropical waters in the spring of 1890, relative residency for eight months and rapid migration returning to more northerly waters in the spring of 1891.
 The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Fig. 1). 
@@
 
-We note however that this interpretation assumes that the physiological demands of pregnancy and lactation do not create an isotopic expression. 
+The dramatic positive excursion in $\delta^{13}$C values seen between March and July 1889 cannot be replicated purely through movement to warmer waters, thus we tentatively ascribe this signal to the physiological demands of pregnancy.
 A captive feeding study in south American sea lions\cite{cardona2017temporal} showed positive $\delta^{13}$C values associated with pregnancy, falling at birth, and a gradual increase in $\delta^{15}$N  values during lactation.
-The dramatic positive excursion in $\delta^{13}$C values in the spring/summer of 1888 cannot be replicated purely through movement to warmer waters, thus we tentatively ascribe this signal to the physiological demands of pregnancy.
-$\delta^{15}$N values in the autumn of 1890 show a gradual increase, consistent with the isotopic expression associated with lactation in south American fur seals\cite{cardona2017temporal}.
+$\delta^{15}$N values in whale baleen during the autumn of 1890 show a gradual increase, consistent with the isotopic expression associated with lactation in south American fur seals\cite{cardona2017temporal}.
 
 Blue whales have a gestation period of 10-12 months, and calves suckle for approximately 6-7 months\cite{handbook}. 
-The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the assumed reproductive ecology of blue whales.
+The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer and autumn 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the assumed reproductive ecology of blue whales.
 
-\subsection*{AGENT BASED WHALE MOVEMENT MODEL}
+\subsection*{AGENT-BASED WHALE MOVEMENT MODEL}
 We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using,yool2013medusa}, where movement behavior is influenced by SST and phytoplankton biomass estimates provided by NEMO-MEDUSA\cite{yool2013medusa}, and bathymetry (Fig.
@@
 
-The agent based model developed employs a simple set of movement rules dictated by behavioural state which is itself fixed according to calendar month by the operator. 
+The agent-based model developed employs a simple set of movement rules dictated by behavioural state which is itself fixed according to calendar month by the operator. 
 At each daily step, the probability of movement, movement directions and distance traveled are sampled from probability distributions to allow individual variation.
@@
 
-Parameters defined for the agent based model best matching the measured profile are summarized in Table \ref{tables1}.
+Parameters defined for the agent-based model best matching the measured profile are summarised in Table \ref{tables1}.
 
@@
   \end{tabular}
-  \caption{Model parameters for agent based model of whale movement.}
+  \caption{Model parameters for agent-based model of whale movement.}
   \label{tables1}

### 2017-06-23 manuscript/blue-whale-ms.tex
Minor edits to figure referencing

@@
 Conservation efforts are hampered by the scarcity of detailed data on the distribution, behaviour and ecology of blue whales, both for historical and present-day populations. 
-Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution}).
+Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}).
 These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
@@
 We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth and used continuous-flow isotope ratio mass spectrometry determine the isotopic compositions of carbon and nitrogen (expressed as $\delta^{13}$C and $\delta^{15}$N values).
-Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplemental Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the last six years of whale's life (autumn 1885 - spring 1891).
+Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplementary Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the last six years of whale's life (autumn 1885 - spring 1891).
 The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
@@
 We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using}, coupled to an agent-based model of whale movements (see Supplementary Methods) to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of blue whales.
-We then compare simulated profiles with the measured isotopic profile of the blue whale's baleen (Fig. \ref{fig1}) to identify the most likely movement history.
+We then compare simulated profiles (Figs. \ref{fig2} and \ref{fig3}) with the measured isotopic profile of the blue whale's baleen (Fig. \ref{fig1}) to identify the most likely movement history.
 
@@
 The pronounced cyclical variations in $\delta^{13}$C values are characteristic of the isotopic expression of the spring phytoplankton bloom in northern waters\cite{magozzi2017using}. 
-This interpretation is also consistent with the relatively negative median $\delta^{13}$C values (Fig. 1). 
-The periodicity of $\delta^{13}$C and $\delta^{15}$N values is best explained by short distance seasonal migrations, with the whale overwintering in the central or southern Norwegian Sea. 
-The relatively positive and invariant $\delta^{13}$C values seen during behavioural phase two are found in subtropical areas, particularly to the west of the Azores.
-Continued, but reduced cyclical variation in $\delta^{15}$N values in 1889 and 1890 implies a degree of seasonal north-south migration before rapid migration to southerly latitudes in 1890, but continued feeding (other mysticete whales fast during migrations, but blue whales are too large to stop feeding for long periods\cite{busquets2017estimating}?silva paper).
+This interpretation is also consistent with the relatively negative median $\delta^{13}$C values (Fig. \ref{fig1}). 
+The periodicity of $\delta^{13}$C and $\delta^{15}$N values is best explained by short distance seasonal migrations, with the whale overwintering in the central or southern Norwegian Sea (Fig.
+\ref{fig2}).
+The relatively positive and invariant $\delta^{13}$C values seen during behavioural phase two are found in subtropical areas, particularly to the west of the Azores (Fig. \ref{fig2}).
+Continued, but reduced cyclical variation in $\delta^{15}$N values in 1889 and 1890 implies a degree of seasonal north-south migration before rapid migration to southerly latitudes in 1890, but continued feeding (other mysticete whales fast during migrations, but blue whales are too large to stop feeding for long periods\cite{busquets2017estimating,silva2013north}).
 The whale remained in these warm waters for around eight months and was conducting a rapid migration returning to her northern feeding grounds when she became stranded.
@@
 Blue whales have a 10-12 month gestation period, and calves are weaned after 6-7 months\cite{handbook}.
-The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the known reproductive ecology of blue whales.
+The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer 1890 and return to northern feeding grounds in early 1891 is therefore consistent with the known reproductive ecology of blue whales.
 
@@
 Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the most distal (oldest) 65cm of the plate. 
-We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis\cite{cardona2017temporal}, revealing a consistent growth rate of 13.5cm $y^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cm y{^-1}$ estimated for six recent blue whales from the Californian Pacific\cite{busquets2017estimating}.
+We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Fig.
+S?), revealing a consistent growth rate of 13.5cm $y^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cm y{^-1}$ estimated for six recent blue whales from the Californian Pacific\cite{busquets2017estimating}.
 Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891. 
-Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase one (oldest 60cm of the baleen plate), but no relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase two.
+Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase one (oldest 60cm of the baleen plate), but no relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase two (Fig.
+S?).
 
 \subsection{Baseline isotope comparisons}
-Isotope-enabled biogeochemical ocean models\cite{magozzi2017using} (Schmittner and Somes) were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds.
-Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model. 
+Isotope-enabled biogeochemical ocean models\cite{magozzi2017using} (Schmittner and Somes) were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Fig.
+S?).
+Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model (Fig. S?). 
 $\delta^{13}$C POM values were simulated at 1$^{\circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model\cite{magozzi2017using,yool2013medusa}. 
@@
 Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
-In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S3). 
+In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S?). 
 The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature\cite{magozzi2017using} Laws et al 1995.
@@
 We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Fig SX; (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
-For a full description of the model and its parameters see Supplementary Methods. 
-We simulated whale movements 30 times for each scenario and residency hotspot. 
+For a full description of the model and its parameters see Supplementary Methods and Table S1. 
+We simulated whale movements 30 times for each scenario and residency hotspot (Fig. S???). 
 For each movement simulation we extracted phytoplankton $\delta^{13}$C values at each time point and location from the models described above\cite{magozzi2017using}. 

### 2017-06-23 manuscript/blue-whale-ms.tex
Minor edits to figure referencing

@@
 Conservation efforts are hampered by the scarcity of detailed data on the distribution, behaviour and ecology of blue whales, both for historical and present-day populations. 
-Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}).
+Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}.
 These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.

### 2017-06-23 manuscript/blue-whale-supplementary.tex
Minor edits to figure referencing

@@
 
-%  \begin{figure}[!htbp]
-%    \centering
-%      \includegraphics[width=12cm]{XXX.pdf}
-%      \caption{Fourier analysis of $delta^{15}$N (A,B) and $delta^{13}$C (C,D) values of the full (A,C) and behavioral phase one only (B,D) components of the baleen plate.
+% figure s1
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width = \linewidth]{figures/Figure-S1-periodograms.png}
+  \caption{Fourier analysis of $delta^{15}$N (A,B) and $delta^{13}$C (C,D) values of the full (A,C) and behavioral phase one only (B,D) components of the baleen plate.
 Inset values are the two highest frequency spectral densities (in cm).
@@
 Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent behavioral phase two}
-%      \label{figs1}
-%  \end{figure}
+  \label{figs1}
+\end{figure}
 
@@
 
-%  \begin{figure}[!htbp]
-%    \centering
-%      \includegraphics[width=12cm]{XXX.pdf}
-%      \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for behavioural phase one showing strong negative covariance with a periodicity of around 13cm}
-%      \label{figs2}
-%  \end{figure}
+% figure s2
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width = \linewidth]{figures/Figure-S2-cross-cor.png}
+  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for behavioural phase one showing strong negative covariance with a periodicity of around 13cm}
+  \label{figs2}
+\end{figure}
 

### 2017-06-24 manuscript/blue-whale-ms.tex
Corrected figure names

@@
   \centering
-  \includegraphics[width = \linewidth]{figures/Figure-1a-raw-dC-dN-data.png}
+  \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
   \label{fig1}
-  \caption{One panel figure}
+  \caption{One panel figure}% Can you add these?
 \end{figure}
+
+Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values and because the isotopic composition of phytoplankton at the base of marine food webs varies in space and time.
+We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using}, coupled to an agent-based model of whale movements (see Supplementary Methods) to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of blue whales.
+We then compare simulated profiles (Figs. \ref{fig2} and \ref{fig3}) with the measured isotopic profile of the blue whale's baleen (Fig. \ref{fig1}) to identify the most likely movement history.
 
@@
   \centering
-  \includegraphics[width = \linewidth]{figures/Figure-1b-migratory-model-full-map.png}
+  \includegraphics[width = \linewidth]{figures/Figure-2-migratory-model-full-map.png}
   \label{fig2}
-  \caption{One panel figure}
+  \caption{One panel figure}% Can you add these?
 \end{figure}
@@
   \centering
-  \includegraphics[width = \linewidth]{figures/Figure-1c-migratory-model-d13C.png}
+  \includegraphics[width = \linewidth]{figures/Figure-3-migratory-model-d13C.png}
   \label{fig3}
-  \caption{One panel figure}
+  \caption{One panel figure}% Can you add these?
 \end{figure}
-
-Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values and because the isotopic composition of phytoplankton at the base of marine food webs varies in space and time.
-We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using}, coupled to an agent-based model of whale movements (see Supplementary Methods) to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of blue whales.
-We then compare simulated profiles (Figs. \ref{fig2} and \ref{fig3}) with the measured isotopic profile of the blue whale's baleen (Fig. \ref{fig1}) to identify the most likely movement history.
 

### 2017-06-24 manuscript/blue-whale-supplementary.tex
Add figures to supp

@@
 Note the baleen plate is 80cm long.
-Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent behavioral phase two}
+Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent behavioral phase two} % check legend
   \label{figs1}
@@
   \includegraphics[width = \linewidth]{figures/Figure-S2-cross-cor.png}
-  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for behavioural phase one showing strong negative covariance with a periodicity of around 13cm}
+  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for behavioural phase one showing strong negative covariance with a periodicity of around 13cm} % check legend
   \label{figs2}
@@
 
- %  \begin{figure}[!htbp]
-%    \centering
-%      \includegraphics[width=12cm]{XXX.pdf}
-%      \caption{$delta^{13}$C POM values}
-%      \label{figs3} 
-%  \end{figure}
-% should be a two panel fig with the annual average d15N from Somes and d13C from CT
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width = \linewidth]{figures/Figure-S3-plankton-d13C-map.png}
+  \caption{$delta^{13}$C POM values} % add legend
+  \label{figs3} 
+\end{figure}
 
@@
     \centering
-      \includegraphics[width=14cm]{figures/Figure-S4-all.png}
-      \caption{Daily $\delta^{13}$C POM values in regions occupied by 30 simulated blue whales foraging over a two year cycle.}
+      \includegraphics[width=\linewidth]{figures/Figure-S4-facet-wrap-d13C.png}
+      \caption{Daily $\delta^{13}$C POM values in regions occupied by 30 simulated blue whales foraging over a two year cycle.} % check legend
       \label{figs4}

### 2017-06-26 manuscript/blue-whale-ms.tex
updated figure legends where i could
saved correct version of figure 2 with arrows.

@@
 \linespread{1.66}
+
+% -------------------------------------------------------------------------------------------------------
+% some custom commands
+
+% Add support for highlighting
+\usepackage{color}
+\newcommand{\hilight}[1]{\colorbox{yellow}{#1}}
+
+% -------------------------------------------------------------------------------------------------------
 
@@
 Subtropical residency is associated with isotopic signatures consistent with pregnancy and lactation in the year before the whale’s death. 
-These results supply the first quantitative multi-yeat records of historical migrations in individual blue whales, and suggest that blue whale residency and movement patterns in the North Atlantic have remained similar for hundreds of years, despite changing human pressures in the region.
+These results supply the first quantitative multi-year records of historical migrations in individual blue whales, and suggest that blue whale residency and movement patterns in the North Atlantic have remained similar for hundreds of years, despite changing human pressures in the region.
 Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climatic, ecological conditions and human-induced pressures, and reveal aspects of life history.
@@
   \label{fig1}
-  \caption{One panel figure}% Can you add these?
+  \caption{Variation in $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right x-axis) longitudinally through the baleen plate (n = 77 samples from a single baleen plate for both isotopes) of the NHM blue whale exhibits strong annual periodicity and cross-correlation (\hilight{Figures S1 and S2}) in both isotopes as well as a change in the mean and magnitude of the cycles during 1890 (approximately at sample 30cm).
+Approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
 \end{figure}
@@
   \label{fig2}
-  \caption{One panel figure}% Can you add these?
+  \caption{Simulated tracks from the migratory movement model (2250 days per each of 50 replicates) with simulated day indicated by colour illustrating the residency period from Ireland to the Barents Sea followed by a migratory period and year spent near the Azores (as indicated by an arrow) and/or Canary Islands before returning north prior to stranding in Wexford, Ireland 1891 (as indicated by an arrow).
+}% Can you add these?
 \end{figure}
@@
   \label{fig3}
-  \caption{One panel figure}% Can you add these?
+  \caption{The corresponding $\delta^{13}$C data from these simulated movements correlate with the empirical data from the Baleen: across the 50 simulations, the correlation coefficient between z-score transformations of empirical and time-matched simulation values had a mean of 0.48 (sd = 0.2), 49 of them were positive,  and 39 of the 50 Pearson correlation tests had a p-value less than the Bonferroni corrected p = 0.05 / 50 = 0.001.
+The initiation of the southward migration around June 1890 is indicated by the vertical broken line.
+The standard deviation of the empirical data have been rescaled by a factor of 3 so as to better visualise the correlation since as a consumer, the whale is expected to integrate and attenuate the fluctuations in its diet, and the end points of the simulations and empirical data have been aligned to coincide.}
 \end{figure}

### 2017-06-26 manuscript/blue-whale-supplementary.tex
updated figure legends where i could
saved correct version of figure 2 with arrows.

@@
 
+% -------------------------------------------------------------------------------------------------------
+% some custom commands
+
+% Add support for highlighting
+\usepackage{color}
+\newcommand{\hilight}[1]{\colorbox{yellow}{#1}}
+
+% figure numbering override
 \renewcommand*{\thefigure}{S\arabic{figure}} % make Fig S1 not Fig 1
 \renewcommand*{\thetable}{S\arabic{table}} % make Table S1 not Table 1
+
+% -------------------------------------------------------------------------------------------------------
 
@@
 Note the baleen plate is 80cm long.
-Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent behavioral phase two} % check legend
+Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent behavioural phase two} % check legend
   \label{figs1}
@@
   \includegraphics[width = \linewidth]{figures/Figure-S3-plankton-d13C-map.png}
-  \caption{$delta^{13}$C POM values} % add legend
+  \caption{Isoscape map of the $delta^{13}$C values from POM used as simulated food in the movement model.}
   \label{figs3} 
@@
       \includegraphics[width=\linewidth]{figures/Figure-S4-facet-wrap-d13C.png}
-      \caption{Daily $\delta^{13}$C POM values in regions occupied by 30 simulated blue whales foraging over a two year cycle.} % check legend
+      \caption{Daily $\delta^{13}$C POM values in regions occupied by 30 simulated blue whales foraging over a two year cycle.
+\hilight{OR} Alternative $\delta^{13}$C series arising from resident-only movement models restricting region to Norway, Ireland, Canaries and the Mid Atlantic ridge as labelled over a simulated 2-year period with daily resolution and 30 replicate simulations per region.
+} % check legend
       \label{figs4}

### 2017-06-26 figures-1-2-3.R

@@ -238,7 +238,7 @@ mp4 <- mp3 + geom_line(data = azores,
 
 print(mp4)
 
-ggsave("manuscript/figures/Figure-2-migratory-model-full-map.png", mp2, 
+ggsave("manuscript/figures/Figure-2-migratory-model-full-map.png", mp4, 
        device = png(width = 600, height = 600, units = "px"))
 
 

### 2017-06-28 manuscript/blue-whale-supplementary.tex
Typo

@@
   \includegraphics[width = \linewidth]{figures/Figure-S1-periodograms.png}
-  \caption{Fourier analysis of $delta^{15}$N (A,B) and $delta^{13}$C (C,D) values of the full (A,C) and behavioral phase one only (B,D) components of the baleen plate.
+  \caption{Fourier analysis of $\delta^{15}$N (A,B) and $\delta^{13}$C (C,D) values of the full (A,C) and behavioral phase one only (B,D) components of the baleen plate.
 Inset values are the two highest frequency spectral densities (in cm).
 Note the baleen plate is 80cm long.
-Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent behavioural phase two} % check legend
+Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent in behavioural phase two}
   \label{figs1}
@@
   \includegraphics[width = \linewidth]{figures/Figure-S3-plankton-d13C-map.png}
-  \caption{Isoscape map of the $delta^{13}$C values from POM used as simulated food in the movement model.}
+  \caption{Isoscape map of the $\delta^{13}$C values from POM used as simulated food in the movement model.}
   \label{figs3} 

### 2017-06-28 manuscript/blue-whale-supplementary.tex
Add Laws ref

@@
 
-The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature (Laws et al 1995).
+The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature\cite{laws1995dependence}.
 Thus temporal variations in $\delta^{13}$C POM values are superimposed on latitudinal gradients. 

### 2017-06-28 manuscript/blue-whale-ms.tex
Moved labels to remove referencing error
Added a few refs

@@
 These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
-Intrinsic biogeochemical tags provide an alternative geo-location technique, whereby chemical compositions of animal tissues are compared to spatially referenced models of chemical variation across the species’ range (Hobsen review, Bowen BOOK).
-Stable isotope-based geolocation has been used to interpret whale ecology \cite{hobson1998stable,ryan2013stable} Eisenmann et al 2016). 
+Intrinsic biogeochemical tags provide an alternative geo-location technique, whereby chemical compositions of animal tissues are compared to spatially referenced models of chemical variation across the species' range\cite{hobson1999tracing} Bowen BOOK).
+Stable isotope-based geolocation has been used to interpret whale ecology \cite{hobson1998stable,ryan2013stable,eisenmann2016isotopic}. 
 Here we apply this method to baleen from the Natural History Museum's blue whale.
@@
 The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
-In behavioural phase one (month 1885 to May 1889), we find strong cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values with strong negative covariance (Fig. \ref{fig1}; Fig. S?). 
+In behavioural phase one (May 1885 to May 1889), we find strong cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values with strong negative covariance (Fig. \ref{fig1}; Fig. S?). 
 In behavioural phase two (June 1889 to March 1891) we see a sharp increase in $\delta^{13}$C values for approximately eight months.
@@
   \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
-  \label{fig1}
   \caption{Variation in $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right x-axis) longitudinally through the baleen plate (n = 77 samples from a single baleen plate for both isotopes) of the NHM blue whale exhibits strong annual periodicity and cross-correlation (\hilight{Figures S1 and S2}) in both isotopes as well as a change in the mean and magnitude of the cycles during 1890 (approximately at sample 30cm).
 Approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
+  \label{fig1}
 \end{figure}
@@
   \includegraphics[width = \linewidth]{figures/Figure-2-migratory-model-full-map.png}
+  \caption{Simulated tracks from the migratory movement model (2250 days per each of 50 replicates) with simulated day indicated by colour illustrating the residency period from Ireland to the Barents Sea followed by a migratory period and year spent near the Azores (as indicated by an arrow) and/or Canary Islands before returning north prior to stranding in Wexford, Ireland 1891 (as indicated by an arrow).
+}
   \label{fig2}
-  \caption{Simulated tracks from the migratory movement model (2250 days per each of 50 replicates) with simulated day indicated by colour illustrating the residency period from Ireland to the Barents Sea followed by a migratory period and year spent near the Azores (as indicated by an arrow) and/or Canary Islands before returning north prior to stranding in Wexford, Ireland 1891 (as indicated by an arrow).
-}% Can you add these?
 \end{figure}
@@
   \includegraphics[width = \linewidth]{figures/Figure-3-migratory-model-d13C.png}
-  \label{fig3}
   \caption{The corresponding $\delta^{13}$C data from these simulated movements correlate with the empirical data from the Baleen: across the 50 simulations, the correlation coefficient between z-score transformations of empirical and time-matched simulation values had a mean of 0.48 (sd = 0.2), 49 of them were positive,  and 39 of the 50 Pearson correlation tests had a p-value less than the Bonferroni corrected p = 0.05 / 50 = 0.001.
@@
 The standard deviation of the empirical data have been rescaled by a factor of 3 so as to better visualise the correlation since as a consumer, the whale is expected to integrate and attenuate the fluctuations in its diet, and the end points of the simulations and empirical data have been aligned to coincide.}
+  \label{fig3}
 \end{figure}
@@
 In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S?). 
-The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature\cite{magozzi2017using} Laws et al 1995.
+The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature\cite{magozzi2017using,laws1995dependence}.
 Thus temporal variations in $\delta^{13}$C POM values were superimposed on latitudinal gradients. 

### 2017-06-29 manuscript/blue-whale-ms.tex
Added RS edits

@@
 It is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable}. 
-We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth and used continuous-flow isotope ratio mass spectrometry determine the isotopic compositions of carbon and nitrogen (expressed as $\delta^{13}$C and $\delta^{15}$N values).
-Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplementary Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the last six years of whale's life (autumn 1885 - spring 1891).
+We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth and used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen (expressed as $\delta^{13}$C and $\delta^{15}$N values).
+Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplementary Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the last six years of the whale's life (autumn 1885 - spring 1891).
 The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
@@
 \subsection{Stable isotope extractions from baleen}
-Baleen was collected from the Natural History Museum, London (specimen BMNH.1892.3.1.1). 
+Baleen was collected from the Natural History Museum, London (specimen NHMUK.1892.3.1.1). 
 The baleen plate was cleaned with ethanol to remove surface contaminants. 
@@
 At each time point and location in the simulations, we extracted phytoplankton $\delta^{13}$C values from the spatio-temporal model\cite{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas.
-Movement was coded as set of probabilistic rules. 
+Movement was coded as a set of probabilistic rules. 
 The parameters for these were taken from the literature on blue whale behaviour (e.g.\cite{handbook}). 
@@
 Northerly migrations were possible only in spring, and southerly migrations in autumn.
-Foraging was possible at anytime of year, and was triggered when whales encountered high concentrations of plankton.
+Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
 Nursing occurred at temperatures between 18 and 25$^{\circ}$C only.

### 2017-06-29 manuscript/blue-whale-ms.tex
Minor edits

@@
 \parindent = 1.5em
-\addtolength{\parskip}{.3em}
+\addtolength{\parskip}{.9em}
 % Abstract 200 -300 words max
@@
 
+\raggedright
+
 \paragraph{Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}. 
-Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long time scales and under historic climatic and ecological conditions\cite{newsome2010using}.
-Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct movements over the last six years of the life of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London (NHM).
+Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}.
+Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct the movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London (NHM).
 We identify two distinct phases in the whale's behaviour: repeated annual north-south migrations in sub-Arctic waters, and residency in subtropical waters. 
-Subtropical residency is associated with isotopic signatures consistent with pregnancy and lactation in the year before the whale’s death. 
+Subtropical residency is associated with isotopic signatures consistent with pregnancy and lactation in the year before the whale's death. 
 These results supply the first quantitative multi-year records of historical migrations in individual blue whales, and suggest that blue whale residency and movement patterns in the North Atlantic have remained similar for hundreds of years, despite changing human pressures in the region.
-Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climatic, ecological conditions and human-induced pressures, and reveal aspects of life history.
+Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climate, ecological conditions and anthropogenic pressures, and reveal aspects of life history.
 Combining biochemical tracers with simulation modelling unlocks the information recorded in these irreplaceable archives.}
@@
 
-Migratory species pose a particular challenge for conservation practitioners, because their effective conservation relies on protection at many, often very distant, sites\cite{runge2014conserving}. 
+Migratory species pose a particular challenge for conservation practitioners, because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}. 
 Such species are also likely to be particularly vulnerable to climate change, as changes may impact various parts of their life cycle\cite{robinson2009travelling}. 
-Dealing with these issues is difficult due to the relative scarcity of information on individual-level movements over multiple years, for both historical and present-day populations, particularly for wide-ranging marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
+Dealing with these issues is difficult due to the relative scarcity of information on individual-level movements over multiple years, for both historical and present-day populations, especially for wide-ranging marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
 
-The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived on Earth, is one such migratory marine species. 
+The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
 Despite blue whale hunting being banned for over 50 years, blue whales are still listed as Endangered by the IUCN Red List\cite{reilly2008balaenoptera}.
@@
 These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
-Intrinsic biogeochemical tags provide an alternative geo-location technique, whereby chemical compositions of animal tissues are compared to spatially referenced models of chemical variation across the species' range\cite{hobson1999tracing} Bowen BOOK).
+Intrinsic biogeochemical tags provide an alternative technique, where the chemical compositions of animal tissues are compared to spatially referenced models of chemical variation across the species' range\cite{hobson1999tracing} Bowen BOOK).
 Stable isotope-based geolocation has been used to interpret whale ecology \cite{hobson1998stable,ryan2013stable,eisenmann2016isotopic}. 
@@
 It is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable}. 
-We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth and used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen (expressed as $\delta^{13}$C and $\delta^{15}$N values).
+We sampled keratin from the blue whale's baleen plate at 1cm intervals, perpendicular to the direction of growth, and used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen (expressed as $\delta^{13}$C and $\delta^{15}$N values).
 Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplementary Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the last six years of the whale's life (autumn 1885 - spring 1891).
 The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
-In behavioural phase one (May 1885 to May 1889), we find strong cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values with strong negative covariance (Fig. \ref{fig1}; Fig. S?). 
+In behavioural phase one (May 1885 to May 1889), we find strong cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values with strong negative covariance (Fig.
+\ref{fig1}; Figs.
+S1 and S2).
 In behavioural phase two (June 1889 to March 1891) we see a sharp increase in $\delta^{13}$C values for approximately eight months.
 This is associated with an overall reduction in $\delta^{15}$N values and a weakening of the periodic $\delta^{15}$N fluctuations seen in behavioural phase one. 
-There is also no covariance between $\delta^{13}$C and $\delta^{15}$N values in behavioral phase two (Fig. \ref{fig1}).
+There is also no covariance between $\delta^{13}$C and $\delta^{15}$N values in behavioral phase two (Fig. \ref{fig1}; Figs. S1 and S2).
 
@@
 
-Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values and because the isotopic composition of phytoplankton at the base of marine food webs varies in space and time.
-We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using}, coupled to an agent-based model of whale movements (see Supplementary Methods) to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of blue whales.
-We then compare simulated profiles (Figs. \ref{fig2} and \ref{fig3}) with the measured isotopic profile of the blue whale's baleen (Fig. \ref{fig1}) to identify the most likely movement history.
+Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values, and because the isotopic composition of phytoplankton at the base of marine food webs varies in space and time.
+We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Fig.
+S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1) to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of blue whales.
+We then compare simulated profiles (Figs.
+\ref{fig2}, \ref{fig3} and Fig.
+S4) with the measured isotopic profile of the blue whale's baleen (Fig.
+\ref{fig1}) to identify the most likely movement history.
 
@@
   \includegraphics[width = \linewidth]{figures/Figure-3-migratory-model-d13C.png}
-  \caption{The corresponding $\delta^{13}$C data from these simulated movements correlate with the empirical data from the Baleen: across the 50 simulations, the correlation coefficient between z-score transformations of empirical and time-matched simulation values had a mean of 0.48 (sd = 0.2), 49 of them were positive,  and 39 of the 50 Pearson correlation tests had a p-value less than the Bonferroni corrected p = 0.05 / 50 = 0.001.
+  \caption{The corresponding $\delta^{13}$C data from these simulated movements correlate with the empirical data from the Baleen: across the 50 simulations, the correlation coefficient between z-score transformations of empirical and time-matched simulation values had a mean of 0.48 ($sd = 0.2$), 49 of them were positive,  and 39 of the 50 Pearson correlation tests had a p-value less than the Bonferroni corrected $p = 0.05/50 = 0.001$.
 The initiation of the southward migration around June 1890 is indicated by the vertical broken line.
-The standard deviation of the empirical data have been rescaled by a factor of 3 so as to better visualise the correlation since as a consumer, the whale is expected to integrate and attenuate the fluctuations in its diet, and the end points of the simulations and empirical data have been aligned to coincide.}
+The standard deviation of the empirical data have been rescaled by a factor of three so as to better visualise the correlation since as a consumer, the whale is expected to integrate and attenuate the fluctuations in its diet, and the end points of the simulations and empirical data have been aligned to coincide.}
   \label{fig3}

### 2017-06-30 Figure-best-simulation.R
plotting of simulations starts at day 101. Added best simulation plots too.

@@ -1,39 +1,11 @@
-# Author: Clive Trueman
-# Adapted by: Andrew Jackson 19-Jun-2017
-# About: script to generate figure 1
-
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-# setwd("/Users/trueman/Desktop/R scripts/migrate.files")
-
-
-library(raster)
-library(gstat)
-library(sp)
-library(maps)
-library(mapdata)
-library(RColorBrewer)
-
-library(tidyr)
-library(dplyr)
-library(ggplot2)
-library(magrittr)
-library(viridis)
-library(lubridate)
-
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 rm(list = ls())
-graphics.off()
-
 
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Panel A - raw whale isotope data with two y-axes
+# Import and Process the empirical data
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 
-
+# Import Whale data
 whale_isos <- read.csv("data/KC.NHM.full.csv", header = TRUE, 
                        stringsAsFactors = FALSE)
 
@@ -83,55 +55,12 @@ new_years_all <- round(new_years_day * growth_rate /
 
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-#----
-# plot the isotope data
-
-point_size <- 4
-
-isop <- ggplot(KC7, aes(x = Samp.No)) + 
-  theme_classic(base_size = 16) + 
-  xlab("Baleen sample (cm from youngest sample)") + 
-  scale_x_continuous(breaks = seq(0,80,20), 
-                     labels = seq(80,0,-20),
-                     sec.axis = sec_axis(~., 
-                                         breaks = new_years_all + 13.5/2,
-                                         labels = new_years,
-                                         name = "Year"))
-
-isop <- isop + geom_line(aes(y = d13C)) + 
-  geom_point(aes(y = d13C), size = point_size, shape = 21, fill = "black") +
-  ylab(expression(paste(delta^{13}, "C (\u2030)")))
-
-# add the d15N data but need to scale it to d13C
-v_shift <- 27.6 #mean(KC7$d15N) - mean(KC7$d13C) 
-# sd <- sd(KC7$d15N) / sd(KC7$d13C) # not currently used in rescaling
-
[…]

### 2017-06-30 manuscript/blue-whale-ms.tex
Add Ellen's comments

@@
 Before whaling in the North Atlantic began in 1868\cite{reilly2008balaenoptera}, there were 10,000-15,000 blue whales in the region\cite{sigurjonsson1995life}. 
-By 1900 fisheries had moved outside the area because stocks were so depleted\cite{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed\cite{sigurjonsson1995life}. 
+In the early 20th century, fisheries moved outside the area because stocks were so depleted\cite{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed\cite{sigurjonsson1995life}.
 The NHM blue whale thus represents an individual from a species at the brink of extinction.
@@
 
-We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Fig SX; (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
+We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Fig SX); (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
 For a full description of the model and its parameters see Supplementary Methods and Table S1. 

### 2017-06-30 Figure-S3-C-N-isoscape.R
added d15N isoscape

@@ -1,0 +1,78 @@
+
+
+library(raster)
+library(RColorBrewer)
+
+
+my_clrs <- brewer.pal(8, "YlOrRd")
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# d13Carbon
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+Atl_carbon   <- raster("data/Atl_Annual_d13C.grd")
+# raster::image(Atl_carbon, xlab = "Long", ylab = "Lat", col = rev(my_clrs))
+
+
+r <- Atl_carbon
+# r <- reclassify(r, c(0,500,1, 500,2000,2))
+val <- getValues(r)
+xy <- as.data.frame(xyFromCell(r,1:ncell(r)))
+xy$val <- val
+
+p <- ggplot(na.omit(xy), aes(x=x, y=y, fill=val)) + 
+  geom_raster() + 
+  borders("world", colour="gray50", fill="gray50")  +
+  coord_fixed(xlim = c(-80, 50),  
+              ylim = c(0, 80), 
+              ratio = 1.3) + 
+  scale_fill_distiller(type = "seq", palette = "YlOrRd", direction = -1,
+                       name = expression(paste(delta^{13}, "C (\u2030)"))) + 
+  theme_classic(base_size = 14) + 
+  theme(axis.line  = element_blank(),
+        axis.text  = element_blank(),
+        axis.ticks = element_blank(),
+        axis.title = element_blank())
+
+print(p)
+
+
+ggsave("manuscript/figures/Figure-S3-plankton-d13C-map.png", p, 
+       device = png(width = 600, height = 400))
+
+
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# d15Nitrogen
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+Atl_nitrogen <- raster("data/Atl_Annual_d15N.grd")
+# raster::image(Atl_nitrogen, xlab = "Long", ylab = "Lat", col = rev(my_clrs))
+
+r2 <- Atl_nitrogen
+# r <- reclassify(r, c(0,500,1, 500,2000,2))
+val2 <- getValues(r2)
+xy2 <- as.data.frame(xyFromCell(r2,1:ncell(r2)))
+xy2$val <- val2
+
+p2 <- ggplot(na.omit(xy2), aes(x=x, y=y, fill=val)) + 
[…]

### 2017-06-30 Figure-S3-carbon-isoscape.R

@@ -1,41 +1,0 @@
-
-
-library(raster)
-library(RColorBrewer)
-
-
-my_clrs <- brewer.pal(12, "YlOrRd")
-
-
-
-Atl_carbon   <- raster("data/Atl_Annual_d13C.grd")
-raster::image(Atl_carbon, xlab = "Long", ylab = "Lat", col = rev(my_clrs))
-
-
-
-
-r <- Atl_carbon
-# r <- reclassify(r, c(0,500,1, 500,2000,2))
-val <- getValues(r)
-xy <- as.data.frame(xyFromCell(r,1:ncell(r)))
-xy <- cbind(xy,val)
-
-p <- ggplot(na.omit(xy), aes(x=x, y=y, fill=val)) + 
-  geom_raster() + 
-  borders("world", colour="gray50", fill="gray50")  +
-  coord_fixed(xlim = c(-80, 50),  
-              ylim = c(0, 80), 
-              ratio = 1.3) + 
-  scale_fill_distiller(type = "seq", palette = "YlOrRd", direction = -1,
-                       name = expression(paste(delta^{13}, "C (\u2030)"))) + 
-  theme_classic(base_size = 14) + 
-  theme(axis.line  = element_blank(),
-        axis.text  = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title = element_blank())
-
-print(p)
-
-
-ggsave("manuscript/figures/Figure-S3-plankton-d13C-map.png", p, 
-       device = png(width = 600, height = 400))

### 2017-06-30 figures-1-2-3.R

@@ -276,9 +276,15 @@ tidy_loess <- function(df, span = 0.1, ...){
 tmp <- resTrack %>% group_by(Rep) %>% 
   do(., tidy_loess(., span = 0.05))
 
+# add the simulated d13C data
+
+# add a vector to enable different alphas for each simulated rep
+tmp$best <- tmp$Rep == -1 # -1 means they are all zero
 
 # plot the daily tracks for simulation days >= 101 only
-track_plot <- ggplot(filter(tmp, days >= 101), aes(days, lo, group = Rep)) 
+track_plot <- ggplot(filter(tmp, days >= 101), aes(days, lo, group = Rep, 
+                                                   alpha = best )) + 
+  scale_alpha_discrete(range = c(0.25, 0.9), guide = FALSE)
 
 # add the vertical line where migration starts in the simulations
 track_plot <- track_plot + geom_vline(xintercept = as.numeric(migrate_day),
@@ -295,8 +301,9 @@ new_years_sim_day <- as.numeric(new_years_date - first_sim_day)
 track_plot <- track_plot + geom_vline(xintercept = new_years_sim_day,
                                       color = "grey", lty = 2)
 
-# add the simulated d13C data
-track_plot <- track_plot  + geom_line(col = viridis(3)[2], alpha = 0.25) + 
+
+
+track_plot <- track_plot  + geom_line(col = viridis(3)[2]) + 
   ylab(expression(paste(delta^{13}, "C (\u2030)"))) + 
   theme_classic(base_size = 14) + 
   scale_x_continuous(name = "Time (years)", breaks = new_years_sim_day, 
@@ -315,14 +322,18 @@ KC7$d13C_scaled <- with(KC7, ((d13C - mean(d13C)) * y_scale) + mean(d13C))
 track_plot <- track_plot + geom_line(data = KC7, 
                                      mapping = aes(x = days + x_shift, 
                                                    y = d13C_scaled, 
-                                                   group = NULL), 
+                                                   group = NULL,
+                                                   alpha = NULL), 
+                                     alpha = 1,
                                      col = viridis::viridis(3)[1],
                                      size = 1)
 
 track_plot <- track_plot + geom_point(data = KC7, 
                                       mapping = aes(x = days + x_shift, 
                                                    y = d13C_scaled, 
-                                                   group = NULL), 
+                                                   group = NULL,
+                                                   alpha = NULL),
+                                      alpha = 1,
                                       col = viridis::viridis(6)[1],
[…]

### 2017-06-30 Figure-S4-alternative-movement-d13C.R
Figure S4 now correctly facet wraps by Region

@@ -212,13 +212,19 @@ res_all <- bind_rows(list(Norway = resNor,
                           "Mid-Atlantic" = resMidAtl), .id = "Region")
 
 
-# same as for norway example above
-tmp <- res_all %>% group_by(Region) %>% group_by(Rep) %>% 
-  do(., data.frame(Z = predict(loess((.$d13C + cnst) ~ .$Day.No, span = 0.1))))
+# same as for norway example above but we are going to plot directly from the 
+# tmp object as the order gets changed due to grouping by Region.
+tmp <- res_all %>% group_by(Region, Rep)  %>% 
+  do(., data.frame(Z = predict(loess((.$d13C + cnst) ~ .$Day.No, span = 0.1)),
+                   Day.No = .$Day.No))
 
-res_all$lo <- tmp$Z
+# re-order the Region Factor
+tmp$Region <- factor(tmp$Region, levels = c("Ireland","Norway",
+                                            "Canaries","Mid-Atlantic"))
 
-sim_facet <- ggplot(res_all, aes(Day.No, lo, group = Rep)) + 
+# res_all$lo <- tmp$Z
+
+sim_facet <- ggplot(tmp, aes(Day.No, Z, group = Rep)) + 
   geom_line(col = viridis(3)[2], alpha = 0.25) + 
   xlab("Time") + 
   ylab(expression(paste(delta^{13}, "C (\u2030)"))) + 

### 2017-06-30 manuscript/blue-whale-ms.tex
Add CT comments

@@
 \author{
-  Natalie Cooper$^{1*}$ Andrew L. Jackson$^{2}$, Katharyn S. Chadwick$^{3}$,\\ Ellen J. Coombs$^{1,4}$,
-  Richard C. Sabin$^{1}$ and Clive N. Trueman$^{3}$ 
+  Clive N. Trueman$^{1}$, Andrew L. Jackson$^{2}$, Katharyn S. Chadwick$^{1}$,\\ Ellen J. Coombs$^{3,4}$,
+  Richard C. Sabin$^{3}$ and Natalie Cooper$^{3*}$
 }
@@
 \affiliation{\noindent{\footnotesize
-  $^1$ Department of Life Sciences, Natural History Museum London, Cromwell Road, London, SW7 5BD, UK.\\ 
+  $^1$ Ocean and Earth Science, University of Southampton Waterfront Campus, Southampton, SO14 3ZH, UK.\\
   $^2$ School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
-  $^3$ National Oceanographic Centre, University of Southampton, Southampton, UK.\\
+  $^3$ Department of Life Sciences, Natural History Museum London, Cromwell Road, London, SW7 5BD, UK.\\ 
   $^4$ University College London, Gower Street, London, WC1E 6BT, UK.\\
@@
 These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
-Intrinsic biogeochemical tags provide an alternative technique, where the chemical compositions of animal tissues are compared to spatially referenced models of chemical variation across the species' range\cite{hobson1999tracing} Bowen BOOK).
+Intrinsic biogeochemical tags provide an alternative technique, where the chemical compositions of animal tissues are compared to spatially referenced models of chemical variation across the species' range\cite{hobson1999tracing,hobson2008tracking}).
 Stable isotope-based geolocation has been used to interpret whale ecology \cite{hobson1998stable,ryan2013stable,eisenmann2016isotopic}. 
@@
 The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
-In behavioural phase one (May 1885 to May 1889), we find strong cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values with strong negative covariance (Fig.
+In behavioural phase one (spring 1885 to spring 1889), we find strong cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values with strong negative covariance (Fig.
 \ref{fig1}; Figs.
 S1 and S2).
-In behavioural phase two (June 1889 to March 1891) we see a sharp increase in $\delta^{13}$C values for approximately eight months.
+In behavioural phase two (summer 1889 to spring 1891) we see a sharp increase in $\delta^{13}$C values for approximately eight months.
 This is associated with an overall reduction in $\delta^{15}$N values and a weakening of the periodic $\delta^{15}$N fluctuations seen in behavioural phase one. 
@@
   \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
-  \caption{Variation in $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right x-axis) longitudinally through the baleen plate (n = 77 samples from a single baleen plate for both isotopes) of the NHM blue whale exhibits strong annual periodicity and cross-correlation (\hilight{Figures S1 and S2}) in both isotopes as well as a change in the mean and magnitude of the cycles during 1890 (approximately at sample 30cm).
+  \caption{Variation in $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right x-axis) longitudinally through the baleen plate (n = 77 samples from a single baleen plate for both isotopes) of the NHM blue whale exhibits strong annual periodicity and cross-correlation (Figs.
+S1 and S2) in both isotopes as well as a change in the mean and magnitude of the cycles during 1890 (approximately at sample 30cm).
 Approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
@@
 
-Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values, and because the isotopic composition of phytoplankton at the base of marine food webs varies in space and time.
+Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values, and because the isotopic composition of phytoplankton at the base of marine food webs varies in both space and time.
 We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Fig.
@@
   \includegraphics[width = \linewidth]{figures/Figure-2-migratory-model-full-map.png}
-  \caption{Simulated tracks from the migratory movement model (2250 days per each of 50 replicates) with simulated day indicated by colour illustrating the residency period from Ireland to the Barents Sea followed by a migratory period and year spent near the Azores (as indicated by an arrow) and/or Canary Islands before returning north prior to stranding in Wexford, Ireland 1891 (as indicated by an arrow).
+  \caption{Simulated tracks from the migratory movement model (2250 days per each of 50 replicates) with simulated day indicated by colour illustrating the residency period from Ireland to the Norwegian/Barents Sea followed by a migratory period and year spent near the Azores (as indicated by an arrow) and/or Canary Islands before returning north prior to stranding in Wexford, Ireland 1891 (as indicated by an arrow).
 }
@@
   \caption{The corresponding $\delta^{13}$C data from these simulated movements correlate with the empirical data from the Baleen: across the 50 simulations, the correlation coefficient between z-score transformations of empirical and time-matched simulation values had a mean of 0.48 ($sd = 0.2$), 49 of them were positive,  and 39 of the 50 Pearson correlation tests had a p-value less than the Bonferroni corrected $p = 0.05/50 = 0.001$.
-The initiation of the southward migration around June 1890 is indicated by the vertical broken line.
+The initiation of the southward migration in the late spring of 1889 is indicated by the vertical broken line.
 The standard deviation of the empirical data have been rescaled by a factor of three so as to better visualise the correlation since as a consumer, the whale is expected to integrate and attenuate the fluctuations in its diet, and the end points of the simulations and empirical data have been aligned to coincide.}
@@
 
-During behavioural phase one, the whale is foraging at high northern latitudes, likely off the coast of Norway. 
+During behavioural phase one, the whale is foraging at high northern latitudes, likely in the Norwegian Sea. 
 The pronounced cyclical variations in $\delta^{13}$C values are characteristic of the isotopic expression of the spring phytoplankton bloom in northern waters\cite{magozzi2017using}. 
@@
 \ref{fig2}).
-The relatively positive and invariant $\delta^{13}$C values seen during behavioural phase two are found in subtropical areas, particularly to the west of the Azores (Fig. \ref{fig2}).
+The relatively positive and invariant $\delta^{13}$C values seen during behavioural phase two are found in subtropical areas.
+Our simulations show high variance during this period (Fig.
+\ref{fig3}), reflecting a range of possible locations, although the most westerly areas, particularly to the west of the Azores (Fig.
+\ref{fig2}), most closely match the measured profile.
 Continued, but reduced cyclical variation in $\delta^{15}$N values in 1889 and 1890 implies a degree of seasonal north-south migration before rapid migration to southerly latitudes in 1890, but continued feeding (other mysticete whales fast during migrations, but blue whales are too large to stop feeding for long periods\cite{busquets2017estimating,silva2013north}).
@@
 
-The dramatic increase in $\delta^{13}$C values seen between March and July 1889 cannot be simulated purely from movements within the known geographic range.
+The dramatic increase in $\delta^{13}$C values seen in the late spring and early summer of 1889 cannot be simulated purely from movements within the known geographic range.
 We believe that this reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions\cite{cardona2017temporal}. 
@@
 Our results support this conclusion. 
-Interestingly, there is still a resident population of blue whales in the region our models place the blue whale within, and females still breed in the subtropical waters near the Azores\cite{reilly2008balaenoptera}.
+Interestingly, there is still a resident population of blue whales in the regions indicated by our model, and females still breed in the subtropical waters near the Azores\cite{reilly2008balaenoptera}.
 Whaling was an intense pressure for blue whales during the period we are analysing. 
@@
 Our novel simulation modelling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviours over multiple years.
-Such techniques can also be applied to extinct species' behaviours and understanding how populations were affected by past pressures such as hunting, and present-day pressures such as global change. 
+Such techniques can also be applied to extinct species' behaviours and used to understand how populations were affected by past pressures such as hunting, and present-day pressures such as global change.
 Finally, our results highlight that museums represent a vast, mostly untapped, source of individual-level spatial ecology information.
@@
 \subsection{Baseline isotope comparisons}
-Isotope-enabled biogeochemical ocean models\cite{magozzi2017using} (Schmittner and Somes) were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Fig.
-S?).
-Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model (Fig. S?). 
+Isotope-enabled biogeochemical ocean models\cite{magozzi2017using,schmittner2016complementary} were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Fig.
+S3).
+Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model (Fig. S3). 
 $\delta^{13}$C POM values were simulated at 1$^{\circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model\cite{magozzi2017using,yool2013medusa}. 
@@
 Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
-In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S?). 
+In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S3). 
 The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature\cite{magozzi2017using,laws1995dependence}.
@@
 
-In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and extent (1$^{\circ}$ or 2$^{\circ}$) of movement are all influenced by the following.
+In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and extent (1$^{\circ}$ or 2$^{\circ}$ in km) of movement are all influenced by the following.
 (i) Behavioural state (migrating north, migrating south, foraging, or nursing). 
 Northerly migrations were possible only in spring, and southerly migrations in autumn. 
-Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
-Nursing occurred at temperatures between 18 and 25$^{\circ}$C only.
+Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton. Nursing occurred at temperatures between 18 and 25$^{\circ}$C only. 
 (ii) Sea surface temperature ($^{\circ}$C\cite{yool2013medusa}). 
@@
 
-We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Fig SX); (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
+We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Fig S4); (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
 For a full description of the model and its parameters see Supplementary Methods and Table S1. 
-We simulated whale movements 30 times for each scenario and residency hotspot (Fig. S???). 
+We simulated whale movements 30 times for each scenario and residency hotspot (Fig. S4). 
 For each movement simulation we extracted phytoplankton $\delta^{13}$C values at each time point and location from the models described above\cite{magozzi2017using}. 

### 2017-06-30 manuscript/blue-whale-ms.tex
minor edit

@@
 Our simulations show high variance during this period (Fig.
-\ref{fig3}), reflecting a range of possible locations, although the most westerly areas, particularly to the west of the Azores (Fig.
+\ref{fig3}), reflecting a range of possible locations for the whale, although the most westerly areas (particularly to the west of the Azores; Fig.
 \ref{fig2}), most closely match the measured profile.
@@
 \subsection{Data Availability}
-Data are available on the Natural History Museum, London's data portal (https://data.nhm.ac.uk). 
+Data are available at XXX. 
 R code is available from GitHub (github.com/ALJackson/blue-whale-bes).
@@
 This work was funded by the British Ecological Society (grant: 5771/6815). 
-We thank XXX for comments on drafts.
 We thank C.J. Somes for providing  $\delta^{15}$N POM data.
@@
 AJ created the figures. 
-NC, CT and EC wrote the paper.
+CT, NC and EC wrote the paper.
 All authors edited drafts and approved the final version.

### 2017-06-30 manuscript/blue-whale-supplementary.tex
Add CT edits to supplemental

@@
 \usepackage{graphicx}
+\usepackage{subcaption}
 \usepackage{parskip} 
@@
 
-\author{Natalie Cooper, Andrew L. Jackson, Katharyn S. Chadwick,\\ 
-Ellen J. Coombs, Richard C. Sabin and Clive N. Trueman}
+\author{Clive N. Trueman, Andrew L. Jackson, Katharyn S. Chadwick,\\ 
+Ellen J. Coombs, Richard C. Sabin and Natalie Cooper}
 
@@
 \subsection*{BASELINE ISOTOPE COMPARISONS}
-Isotope-enabled biogeochemical ocean models\cite{magozzi2017using} (Schmittner and Somes) were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds.
+Isotope-enabled biogeochemical ocean models\cite{magozzi2017using,schmittner2016complementary} were used to characterise the isotopic composition of phytoplankton expected in different potential foraging grounds.
 Annual average $\delta^{15}$N POM values were provided by C.J. Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model. 
@@
 
-Simulated $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c.60$^{\circ}$N, and relatively negative in the central and southern North Atlantic, associated with high levels of nitrogen fixation.
+Simulated $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c.60$^{\circ}$N, and relatively negative in the central and southern North Atlantic, associated with high levels of nitrogen fixation (Fig.
+\ref{figs3}B).
 Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
-In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. \ref{figs3}). 
+In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. \ref{figs3}A). 
 The temporally coincident periodic, inverse fluctuations in $\delta^{13}$C and $\delta^{15}$N values in the initial phase of the baleen plate are therefore consistent with migration between waters north of 60$^{\circ}$N and more southerly southern waters in the North Atlantic, with northerly occupancy in summer months and southerly occupancy in winter months.
@@
     \centering
-  \includegraphics[width = \linewidth]{figures/Figure-S3-plankton-d13C-map.png}
-  \caption{Isoscape map of the $\delta^{13}$C values from POM used as simulated food in the movement model.}
+    \begin{subfigure}[t]{0.45\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{figures/Figure-S3-plankton-d13C-map.png} 
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[t]{0.45\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{figures/Figure-S3-plankton-d15N-map.png} 
+    \end{subfigure}
+    \caption{Spatial model (isoscape) of model-simulated biomass weighted annual average (A) $\delta^{13}$C and (B) $\delta^{15}$N values from POM.
+$\delta^{13}$C values are from Magozzi et al.
+2017\cite{magozzi2017using} and $\delta^{15}$N from Schmittner \& Somes 2016\cite{schmittner2016complementary} and Somes (pers.
+comm.)}
     \label{figs3}
@@
       \includegraphics[width=\linewidth]{figures/Figure-S4-facet-wrap-d13C.png}
-      \caption{Daily $\delta^{13}$C POM values in regions occupied by 30 simulated blue whales foraging over a two year cycle.
-\hilight{OR} Alternative $\delta^{13}$C series arising from resident-only movement models restricting region to Norway, Ireland, Canaries and the Mid Atlantic ridge as labelled over a simulated 2-year period with daily resolution and 30 replicate simulations per region.
-} % check legend
+      \caption{Alternative $\delta^{13}$C series arising from resident-only movement models restricting region to Norway, Ireland, Canaries and the Mid Atlantic ridge as labelled over a simulated two year period with daily resolution and 30 replicate simulations for each region.} % check legend
       \label{figs4}
@@
 \subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase two}
-Behavioural phase two is characterized by a dramatic shift to maximum $\delta^{13}$C values in June of 1889 followed by approximately 18 months of relatively positive and constant $\delta^{13}$C values.
+Behavioural phase two is characterized by a dramatic shift to maximum $\delta^{13}$C values in spring 1889 followed by approximately 18 months of relatively positive and constant $\delta^{13}$C values.
 $\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity associated with north-south migration is absent in 1890. 
@@
 
-Accordingly we modelled whale movements allowing five years of seasonal north-south migration in the north east northern Atlantic, followed by rapid migration to subtropical waters in the spring of 1890, relative residency for eight months and rapid migration returning to more northerly waters in the spring of 1891.
+Accordingly we modelled whale movements allowing five years of seasonal north-south migration in the north east northern Atlantic, followed by rapid migration to subtropical waters in the late spring of 1889, relative residency for eight months and rapid migration returning to more northerly waters in the spring of 1891.
 The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Fig. 1). 
@@
 
-The dramatic positive excursion in $\delta^{13}$C values seen between March and July 1889 cannot be replicated purely through movement to warmer waters, thus we tentatively ascribe this signal to the physiological demands of pregnancy.
+The dramatic positive excursion in $\delta^{13}$C values seen in the late spring and early summer of 1889 cannot be replicated purely through movement to warmer waters, thus we tentatively ascribe this signal to the physiological demands of pregnancy.
 A captive feeding study in south American sea lions\cite{cardona2017temporal} showed positive $\delta^{13}$C values associated with pregnancy, falling at birth, and a gradual increase in $\delta^{15}$N  values during lactation.
@@
 At each daily step, the probability of movement, movement directions and distance traveled are sampled from probability distributions to allow individual variation.
-
-%  \begin{figure}[!htbp]
-%    \centering
-%      \includegraphics[width=12cm]{XXX.pdf}
-%      \caption{Summary diagram of agent-based movement model structure.}
-%      \label{figs5}
-%  \end{figure}
 
@@
 
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[trim = 0cm 14cm 0cm 4cm, clip]{figures/flow-diagram-model.pdf}
+  \caption{Summary diagram of agent-based movement model structure.}
+  \label{figs5}
+\end{figure}
+
+\newpage
+
 \centering
@@
     \hline
-    Month - nursing (240 days) & NA & NA & NA & Month???\\
+    Month - nursing (210 days) & NA & NA & NA & June - Dec\\
     \hline

### 2017-06-30 manuscript/blue-whale-supplementary.tex
Add caption for schematic

@@
 
+  \vspace{-1cm}
 \begin{figure}[!htbp]
   \centering
-  \includegraphics[trim = 0cm 14cm 0cm 4cm, clip]{figures/flow-diagram-model.pdf}
-  \caption{Summary diagram of agent-based movement model structure.}
+  \includegraphics[trim = 0cm 145mm 0cm 4cm, clip]{figures/flow-diagram-model.pdf}
+  \vspace{-1.2cm}
+  \linespread{0.8}
+  \caption{
+  Schematic structure of the agent-based model used to generate hypothetical movement tracks. 
+  At the start of each simulation, max and min sea surface temperature (SST) and food availability thresholds are fixed for all individuals.
+Behavioural states are defined \textit{a priori} based on day of the year.
+Maximum movement distances under each behavioural state are defined for each individual based on a random sample from a Gaussian distribution.
+The model runs with a daily time step.
+Current location and day of simulation is noted, and the behavioural state defined.
+Environmental conditions (SST, food availability and depth) are sensed in the current and eight adjacent 1$^{\circ}$ grid cells.
+The probability of movement is sampled from a binomial distribution defined according to behavioural state and environmental conditions in the current location.
+The direction of movement is sampled based on discrete probabilities associated to each of the eight adjacent 1$^{\circ}$ cells, defined according to behavioural state and environmental conditions.
+Finally the distance moved is sampled from a Poisson distribution conditional on the max distance defined for each individual.
+The new location is then assigned from a linear vector originating at the current location with magnitude defined as the distance moved and direction as 45$^{\circ}$ increments reflecting the relative orientation of each of the eight potential adjacent cells.}
   \label{figs5}
@@
   \end{tabular}
-  \caption{Model parameters for agent-based model of whale movement.}
+  \caption{Model parameters for agent-based model of whale movement. See Fig. \ref{figs5} for details.}
   \label{tables1}

### 2017-06-30 Blue baleen plot.R
updated the acf and ccf plots and annotated where appropriate

@@ -1,0 +1,58 @@
+library(TSA)
+
+# setwd("/Users/trueman/Desktop/R scripts/Whale")
+data<-read.csv("data/KC.NHM.full.csv", header=TRUE)
+par(mfrow=c(1,1))
+colz<-colorRampPalette(c("brown", "cadet blue", "pink"))(12)
+colz<-seq(1:6)
+fin<-data[data$Spp=="fin",]
+minke<-data[data$Spp=="minke",]
+sei<-data[data$Spp=="sei",]
+blue<-data[data$Spp=="blue",]
+
+
+#BLUE
+#setEPS()
+#postscript("blue.eps")
+par(mfrow=c(1,1))
+plot(x=rev(blue$Samp.No[blue$Whale=="KC7"]), y=blue$d13C[blue$Whale=="KC7"], 
+     type="o", xlim=c(80,0), ylim=c(-20, -16.5), xlab="increment no", 
+     ylab=expression(paste(delta^{13},"C")), pch=21, cex=1.5, lwd=0.8, bg=colz[1])
+text(5,-20, "most recent")
+text(80,-20, "oldest")
+text(5, -16.75, "Whale KC1")
+
+par(new=TRUE)
+plot(x=rev(blue$Samp.No[blue$Whale=="KC7"]), y=blue$d15N[blue$Whale=="KC7"], 
+     type="o", xlim=c(80,0), ylim=c(5, 13),  axes=F, xlab="", ylab="", 
+     pch=21, cex=1.5, lwd=0.8, bg="grey")
+axis(side=4)
+mtext(side=4, line=3, expression(paste(delta^{15},"N")))
+#dev.off()
+#par(new=FALSE)
+
+
+
+
+
+## fourier analysis of blue whale - frequency is in cm of baleen
+## fourier for phase 1
+blueP1<-blue[blue$Samp.No<(80-25),]
+
+par(mfrow=c(1,1))
+p=periodogram(blueP1$d15N)
+#plot(p)
+dd<-data.frame(freq=p$freq, spec=p$spec)
+order=dd[order(-dd$spec),]
+top2 = head(order, 2)
+top2
+distance=1/top2$f
+
+text(0.4, 4, paste(distance[1],"cm"))
+#text(0.4, 3.5, paste(distance[2], "cm")
+
+
+## autocorrelation analysis of phase 1 d15N, and cross-correlation of phase 1 d15N and d13C values
+
[…]

### 2017-06-30 Figure-S1-S2-code.R

@@ -77,18 +77,48 @@ KC7$samp_new_year <- KC7$Samp.No * idx_new_year -
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 
+# function to add text for the highest frequency spectral 
+# densities to periodgram plots
+
+periodogramText <- function(p, k){
+  
+  # extract frequencies and spectral densities from the object
+  dd <- data.frame(freq=p$freq, spec=p$spec)
+  
+  # sort them into rank order
+  ord <- dd[order(-dd$spec),]
+  
+  # extract the top two densities
+  top2  <-  head(ord, 2)
+  
+  # convert to distance along our sampling transect
+  distance <- 1/top2$freq
+  
+  # add either the first or second
+  text(0.4, max(p$spec)*0.8, 
+       paste0(sprintf("%.1f", round(distance[k],1))," cm"), 
+       cex = 2)
+  
+  #text(0.4, 3.5, paste(distance[2], "cm")
+}
+
+
 png(filename = "manuscript/figures/Figure-S1-periodograms.png", 
     width = 1000, height = 600)
 
 par(mfrow = c(2,2))
 
-periodogram(KC7$d15N, main = "A - Full d15N")
+p1 <- periodogram(KC7$d15N, main = "A - Full d15N")
+periodogramText(p1, k = 2)
 
-periodogram(KC7$d15N[1:60], main = "B -Oldest 60cm d15N")
+p2 <- periodogram(KC7$d15N[1:60], main = "B -Oldest 60cm d15N")
+periodogramText(p2, k = 1)
 
-periodogram(KC7$d13C, main = "C - Full d13C")
+p3 <- periodogram(KC7$d13C, main = "C - Full d13C")
+periodogramText(p3, k = 2)
 
-periodogram(KC7$d13C[1:60], main = "D - Oldest 60cm d13C")
+p4 <- periodogram(KC7$d13C[1:60], main = "D - Oldest 60cm d13C")
+periodogramText(p4, k = 2)
 
 dev.off()
 
@@ -106,9 +136,18 @@ TSA::acf(cbind(KC7$d13C[61:n], KC7$d15N[61:n]), main = "Youngest 20cm")
 png(filename = "manuscript/figures/Figure-S2-cross-cor.png", 
[…]

### 2017-06-30 manuscript/blue-whale-ms.tex
Change data to zenodo link
Final edits NC

@@
 Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climate, ecological conditions and anthropogenic pressures, and reveal aspects of life history.
-Combining biochemical tracers with simulation modelling unlocks the information recorded in these irreplaceable archives.}
+Combining biochemical tracers with simulation modeling unlocks the information recorded in these irreplaceable archives.}
 
@@
 Stable isotope-based geolocation has been used to interpret whale ecology \cite{hobson1998stable,ryan2013stable,eisenmann2016isotopic}. 
-Here we apply this method to baleen from the Natural History Museum's blue whale.
+Here we apply this method to baleen from the NHM's blue whale.
 
@@
   \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
-  \caption{Variation in $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right x-axis) longitudinally through the baleen plate (n = 77 samples from a single baleen plate for both isotopes) of the NHM blue whale exhibits strong annual periodicity and cross-correlation (Figs.
-S1 and S2) in both isotopes as well as a change in the mean and magnitude of the cycles during 1890 (approximately at sample 30cm).
+  \caption{Variation in $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis) longitudinally through the baleen plate (n = 77 samples from a single baleen plate for both isotopes) of the NHM blue whale exhibits strong annual periodicity and cross-correlation (Figs.
+S1 and S2) in both isotopes as well as a change in the mean and magnitude of the cycles during 1889 (approximately at sample 30cm).
 Approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
@@
 We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Fig.
-S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1) to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of blue whales.
+S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig.
+S5) to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of blue whales.
 We then compare simulated profiles (Figs.
@@
   \includegraphics[width = \linewidth]{figures/Figure-2-migratory-model-full-map.png}
-  \caption{Simulated tracks from the migratory movement model (2250 days per each of 50 replicates) with simulated day indicated by colour illustrating the residency period from Ireland to the Norwegian/Barents Sea followed by a migratory period and year spent near the Azores (as indicated by an arrow) and/or Canary Islands before returning north prior to stranding in Wexford, Ireland 1891 (as indicated by an arrow).
+  \caption{Simulated tracks from the migratory movement model (2250 days for each of 50 replicates) with simulated day indicated by colour illustrating the residency period from Ireland to the Norwegian/Barents Sea followed by a migratory period and year spent near the Azores (as indicated by an arrow) and/or Canary Islands before returning north prior to stranding in Wexford, Ireland 1891 (as indicated by an arrow).
 }
@@
   \includegraphics[width = \linewidth]{figures/Figure-3-migratory-model-d13C.png}
-  \caption{The corresponding $\delta^{13}$C data from these simulated movements correlate with the empirical data from the Baleen: across the 50 simulations, the correlation coefficient between z-score transformations of empirical and time-matched simulation values had a mean of 0.48 ($sd = 0.2$), 49 of them were positive,  and 39 of the 50 Pearson correlation tests had a p-value less than the Bonferroni corrected $p = 0.05/50 = 0.001$.
+  \caption{The corresponding $\delta^{13}$C data from these simulated movements correlate with the empirical data from the baleen: across the 50 simulations, the correlation coefficient between z-score transformations of empirical and time-matched simulation values had a mean of 0.48 ($sd = 0.2$), 49 of them were positive, and 39 of the 50 Pearson correlation tests had a p-value less than the Bonferroni corrected value of 0.001 ($p = 0.05/50 = 0.001$).
 The initiation of the southward migration in the late spring of 1889 is indicated by the vertical broken line.
-The standard deviation of the empirical data have been rescaled by a factor of three so as to better visualise the correlation since as a consumer, the whale is expected to integrate and attenuate the fluctuations in its diet, and the end points of the simulations and empirical data have been aligned to coincide.}
+The standard deviation of the empirical data have been rescaled by a factor of three so as to better visualise the correlation because, as a consumer, the whale is expected to integrate and attenuate the fluctuations in its diet, and the end points of the simulations and empirical data have been aligned to coincide.}
   \label{fig3}
@@
 The NHM blue whale thus represents an individual from a species at the brink of extinction.
-Our data suggest that the spatial ecology of blue whales in the North Atlantic have remained similar for more than 100 years, despite dramatic changes in anthropogenic pressures.
+Our data suggest that the spatial ecology of blue whales in the North Atlantic has remained similar for more than 100 years, despite dramatic changes in anthropogenic pressures.
 
 This is the first quantitative evidence of historical individual-level movement behaviours in blue whales. 
-Our novel simulation modelling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviours over multiple years.
-Such techniques can also be applied to extinct species' behaviours and used to understand how populations were affected by past pressures such as hunting, and present-day pressures such as global change.
+Our novel simulation modeling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviours over multiple years.
+Such techniques can also be applied to extinct species behaviours and used to understand how populations were affected by past pressures such as hunting, and present-day pressures such as global change.
 Finally, our results highlight that museums represent a vast, mostly untapped, source of individual-level spatial ecology information.
@@
 {\raise.17ex\hbox{$\scriptstyle\sim$}}1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
-78 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section which contains the most recent tissue. 
+77 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section which contains the most recent tissue. 
 Baleen grows at a constant rate, so the samples are equally spaced through time\cite{best1996stable}.
@@
 Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
-C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples (Hobson  Koch 1998).
+C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples\cite{hobson1998stable}.
 
@@
 We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Fig.
-S?), revealing a consistent growth rate of 13.5cm $y^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cm y{^-1}$ estimated for six recent blue whales from the Californian Pacific\cite{busquets2017estimating}.
+S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated for six recent blue whales from the Californian Pacific\cite{busquets2017estimating}.
 Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891. 
 Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase one (oldest 60cm of the baleen plate), but no relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase two (Fig.
-S?).
+S2).
 
@@
 We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
-At each time point and location in the simulations, we extracted phytoplankton $\delta^{13}$C values from the spatio-temporal model\cite{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas.
+At each time point and location in the simulations, we extracted phytoplankton $\delta^{13}$C values from the spatio-temporal model\cite{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs 2, 3, and S4).
 Movement was coded as a set of probabilistic rules. 
@@
 Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton. Nursing occurred at temperatures between 18 and 25$^{\circ}$C only. 
-(ii) Sea surface temperature ($^{\circ}$C\cite{yool2013medusa}). 
+(ii) Sea surface temperature\cite{yool2013medusa} ($^{\circ}$C). 
 When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.
@@
 Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
-(iv) Phytoplankton concentration (mmol N $m^{-3}$, for combined diatom and non-diatom communities\cite{yool2013medusa}). 
+(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities\cite{yool2013medusa}). 
 This was included as a proxy for zooplankton food availability. 
@@
 
-We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Fig S4); (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
-For a full description of the model and its parameters see Supplementary Methods and Table S1. 
-We simulated whale movements 30 times for each scenario and residency hotspot (Fig. S4). 
+We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life} ; Fig S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
+For a full description of the model and its parameters see Supplementary Methods: Fig. S5 and Table S1. 
+We simulated whale movements 50 times for each scenario and 30 times for each residency hotspot (Fig. S4). 
 For each movement simulation we extracted phytoplankton $\delta^{13}$C values at each time point and location from the models described above\cite{magozzi2017using}. 
@@
 \subsection{Data Availability}
-Data are available at XXX. 
+Data are available from Zenodo doi: [DOI will be added on acceptance]. 
 R code is available from GitHub (github.com/ALJackson/blue-whale-bes).

### 2017-06-30 manuscript/blue-whale-ms.tex
Edits to Fig 3 legend

@@
   \includegraphics[width = \linewidth]{figures/Figure-3-migratory-model-d13C.png}
-  \caption{The corresponding $\delta^{13}$C data from these simulated movements correlate with the empirical data from the baleen: across the 50 simulations, the correlation coefficient between z-score transformations of empirical and time-matched simulation values had a mean of 0.48 ($sd = 0.2$), 49 of them were positive, and 39 of the 50 Pearson correlation tests had a p-value less than the Bonferroni corrected value of 0.001 ($p = 0.05/50 = 0.001$).
+  \caption{Correlations among $\delta^{13}$C data from 50 movement simulations (green) and empirical data from baleen (blue).
+Across the 50 simulations, the correlation coefficient between z-score transformations of empirical and time-matched simulation values had a mean of 0.48 ($sd = 0.2$), 49 of them were positive, and 39 of the 50 Pearson correlation tests had a p-value less than the Bonferroni corrected value of 0.001 ($p = 0.05/50 = 0.001$).
 The initiation of the southward migration in the late spring of 1889 is indicated by the vertical broken line.

### 2017-06-30 manuscript/CoverLetter.tex
Added crappy cover letter

\documentclass[11pt]{letter}
\usepackage[a4paper,left=2.5cm, right=2.5cm, top=1cm, bottom=1cm]{geometry}
\usepackage[osf]{mathpazo}
\usepackage{url}
\signature{Natalie Cooper on behalf of all coauthors}
\address{Natural History Museum, London \\ Cromwell Road \\ London, SW7 5BD \\ natalie.cooper@nhm.ac.uk}
\longindentation=0pt
\begin{document}

\begin{letter}{}
\opening{Dear Editors,}

Please find attached our letter, `Reconstructing the last known movements of one of Nature's giants', for consideration at Nature.

We get few opportunities to study the behaviour of rare and enigmatic species in the wild, especially wide-ranging marine species. 
Museum collections offer an opportunity to gain information on the movements and life-history of such species. 
In 1891 a 25m long female blue whale stranded in Wexford, Ireland. 
On July 13th 2017, this whale will take pride of place in the Natural History Museum's (NHM) Hintze Hall (we can provide a picture of this as a cover image if desired).
Ironically, we know very little about this animal, or the species as a whole.

Here we use stable isotope analysis of the baleen of the NHM blue whale, along with a custom-built individual-based movement model, to identify the most likely movements of this individual over the last six years of her life. We identify two phases in the behaviour of the whale - short range seasonal migrations at northern latitudes, followed by isotopic signatures consistent with pregnancy, lactation in warmer waters, and then a return to her northern feeding grounds where she stranded. This provides crucial information about historical movements in blue whales at a point where the species was on the brink of extinction in the North Atlantic.

Blue whales were the first species that humans legislated to save, and represent a big conservation success story, although North Atlantic stocks have not recovered to the same extent as other whales in the region. There is also evidence that blue whales are shifting their ranges due to warming waters off Iceland. Thus understanding historical movements of North Atlantic blue whales has never been more vital if we wish to preserve them for future generations.

We look forward to hearing from you,

\closing{Yours sincerely,}

\end{letter}
\end{document}

### 2017-06-30 mark-for-delete/code-for-figures-clive.R
Remove marked for delete files

@@ -1,190 +1,0 @@
-setwd("/Users/trueman/Desktop/R scripts/migrate.files")
-library(raster)
-library(gstat)
-library(sp)
-library(maps)
-library(mapdata)
-
-
-
-##plot model output annual average baseline images
-Atl_carbon   <- raster("data/Atl_Annual_d13C.grd")
-Atl_nitrogen <- raster("data/Atl_Annual_d15N.grd")
-
-image(Atl_carbon, col=colsrev)
-image(Atl_nitrogen, col=colsrev)
-
-
-##########read resident whale model files
-
-resNor    <- read.csv("data/Norway.csv", header = TRUE)
-resIre    <- read.csv("data/west.Ireland.csv", header = TRUE)
-resCan    <- read.csv("data/Canaries.csv", header = TRUE)
-resMidAtl <- read.csv("data/Mid_Atl.csv", header = TRUE)
-
-
-
-
-
-#plot loess fits to the 30 resident simulated whales for each region
-
-
-
-setEPS()
-postscript("resident.eps", width=10, height=5)
-
-par(mfrow=c(2,2))
-
-
-
-
-#norway resident
-inds<-length(unique(resNor$Rep))
-lo<-loess(resNor$d13C[resNor$Rep==1]+6~resNor$Day.No[resNor$Rep==1], span=0.1)
-plot(x=resNor$Day.No[resNor$Rep==1], y=resNor$d13C[resNor$Rep==1], ylim=c(-26,-9), main="Norwegian Sea" , type="p", cex=0, col="red", xlab="Day.No", ylab="d13C")
-lines(resNor$Day.No[resNor$Rep==1],predict(lo), col="red", lwd=0.2)
-par(new=TRUE)
-for(w in 2:inds){
-	lo<-loess(resNor$d13C[resNor$Rep==w]+6~resNor$Day.No[resNor$Rep==w], span=0.1)
-lines(resNor$Day.No[resNor$Rep==w],predict(lo), col="red", lwd=0.2)
-
-par(new=TRUE)
-}
-par(new=FALSE)
-
-
-
-#Ireland resident
-inds<-length(unique(resIre$Rep))
-lo<-loess(resIre$d13C[resNor$Rep==1]+6~resIre$Day.No[resIre$Rep==1], span=0.1)
[…]

### 2017-10-08 manuscript/revision/response.md
Initial commit of revised version

**RE: Nature submission 2017-06-07324A**

Dear Editors,

Our paper was reviewed at Nature and was not accepted for publication. However, it was suggested that we transfer the paper to Nature Ecology and Evolution, which we are glad to do. We are transferring the reviews with the paper, therefore we include a response to reviewers as part of this cover letter. We are grateful to the referees for their helpful and constructive comments, and hope that our edits, along with the move from Nature to Nature Ecology and Evolution, will have allayed their concerns about the manuscript.

We respond to their points below. To improve the clarity of this document we have included the reviewers comments in italics.

**Important note to the reviewers**. During the revisions of this manuscript, which included substantial extra analysis, we discovered an error in our original results. Essentially, the baleen samples had been labelled in the reverse direction to what we expected. The values for all the stable isotope measures remain the same but the timeline is reversed (this occurred due to a misunderstanding with the masters student who ran to original mass spec analyses and subsequently finished her masters project and moved away). We regret this error, but these things happen, and we are relieved that we noticed the error before publication. Interestingly, when we added the extra data to the end of our time series, we find our overall results do not change from our previous manuscript. We communicated with Editor in Chief Patrick Goymer about whether it was still appropriate to cascade our reviews with the paper given this error, and he said this was fine as long as we made this very clear in our response letter. This note is to alert the reveiwers that the results graphs will look different to the original version, though our conclusions have not changed qualitatively.

## Referee 1
### Key results, Originality and Significance

In my opinion, the paper makes two important points:

First: The authors used a mix between a temporal isoscape (measured POM over time and space to create a changing isoscape) and an agent-based model (AI that models the movement of an individual using parameters such as presence of food, time and temperature in adjoining locations to decide if it moves or not). The combination was used to retrace the movement of a blue whale and compare it to baleen isotope records. In my opinion, this is a novel approach that provides more confidence in attributing isotopic signals to place/time, and thus helps reconstruct migration of the individual. However, I feel that it is not a solve-all approach, as just as with isoscapes, you still need to get those initial measurements of time/space isotopes. Nevertheless, it is a very interesting technique.

Second: the baleen isotope records themselves are interesting as there are very few blue whale baleen records published (this is the second one ever to my knowledge, and the first was also published this year). The isotopes also appear to show a mixture of behaviours, such as migration AND permanent residence in a single individual. Such mix of behaviour is rare in the literature, and still debated. Further, the residence appears consistent with pregnancy/lactation, which is even more uncommon in whale baleen literature as far as I know. For these reasons, I think the isotope records will be very interesting to other researchers in the field.

*We thank the referee for their positive comments about our paper.*

### Data and Methodology: Validity

I am satisfied with the SIA methods and results, but I have some slight comments. While I have no experience with modelling, I noted that there was no mention of isotopic fractionation in the model. As far as I know, blue whale feed on zooplankton and small fish, not directly on phytoplankton. If I am correct, the comparison of the baleen records to the modelled isotopes may be incorrect, as it should be affected by trophic shifts. I suppose it depends on whether the model was used for a direct pattern comparison, or if the values modelled and the baleen values were compared to each other. 
Comparable literature in other species (seals, humans) appear to confirm the pregnancy interpretation, although I’m less convinced by the lactation signal as it seems too delayed in time. This could possibly be checked using hormones, although that is also a very novel field. See the following literature on baleen hormones: Hunt 2016 “Longitudinal progesterone profiles in baleen from female North Atlantic right whales (Eubalaena glacialis) match known calving history”; Hunt 2014 “Baleen hormones: a novel tool for retrospective assessment of stress and reproduction in bowhead whales (Balaena mysticetus)”. These are older samples (20 yrs or so) but not as old as this whale so may be loss of hormone signal over time.
I strongly believe this will not prohibit publication, as the results are still very interesting. It is very possible that I am misreading the delay in lactation signal.

### Appropriate use of statistics:

The statistical analyses seem sound and appropriate, although the use of Fourier analysis to calculate periodicity and growth rate in baleen can be difficult with the typically low number of samples per cycle and replication. However, since most of the literature is based on “eyeballing” cycles and visual estimates, the Fourier analysis is a good way to confirm the visual estimates. 
There are no error bars or replication of samples, but this is typical of stable isotopes.

### Suggested improvements/ References:

Some basic stable isotope knowledge, as well as blue whale feeding ecology should probably be mentioned (ie krill/fish feeding, no fasting) or trophic influence on isotopic signal compared to model POM values. Adding sentence on feeding ecology at beginning, and explaining in model why POM is a good proxy for zooplankton should be enough. For example: Line 65: a reference such as West et al, 2006 “Stable isotopes as one of Nature’s ecological records” relevant to SIA basics. Line 82: reference again with basic SIA shifts in marine food webs? Possibly McMahon et al, 2015 “Millennial Scale plankton regime shifts in the subtropical North Pacific Ocean”
The reported baleen growth rates fit the literature, but I believe it may be worth mentioning in Line 61 that baleen gets worn down over time, and that the records are not full life spans. 
According to the model, d15N is relatively higher in the arctic during summer than the Atlantic in winter. In the profile, the winter values in phase one are still comparatively higher than the summer values. Would you be able to explain? I am mostly curious, but it may reflect difficulties with reading the model from a figure.
While the d13C increase fits the sea lion theory, the drop in d13C seen after/at birth happens too quick (late 1889) in the whale to fit the d15N rise consistent with lactation (Almost 6 months late according to timeline, past mid 1890). Human analyses in hair showed a decrease in d15N (1%o) during pregnancy, which can be seen in the baleen if we take the d13C drop as birth time and early 1889 as pregnancy time (see Fuller et al 2004 “Nitrogen Balance and 15N, why you’re not what you eat during pregnancy”). While it’s not a study conducted in marine mammals, it’s a further indication that the birth time may indeed be in late 1889, or at least that the lactation peak is very late. Could indicate a slowing down in baleen growth, or something else? Could be worth checking with hormonal analyses (see above). Another possibility lies in the location of baleen analyses. Since the methods are not clear on that point, can you confirm that the samples were taken past the gum
line? Dried skin/gum could have affected the isotopic signals if samples were taken outside of the erupted zone. Could it be why the isotopic signals are so different in phase two? Busquet Vass paper shows no difference in 13C between skin and baleen so unlikely, but worth confirming.
The final sample interpretation of a return to cyclicity is very open to discussion, as a single point is too weak to be a guarantee. I will concede that the isotopes are consistent with past migratory values, but may be worth softening the comment (line 109-110)?
When describing the different behaviours, such as whales staying in high latitudes instead of migrating, also talk about the ones staying at low lat (isotope blue whale ref. Busquet-Vass) since it is more relevant to your own isotopic interpretation (line 127).

Typos etc:

1. Abstract
Line 20: Should not have abbreviation. Move to line 58 where it is introduced next.
2. Article main text
Line 56: “)” extraneous bracket
3. Supplementary Data
Page 2, line 1: “)” extraneous bracket after “behavioural phase 2”

### Clarity and Context:

The manuscript was clear, but lacked some information on basic isotope literature, and blue whale ecology. While the ecology side is less important, some information on how isotope analyses work could be beneficial to the greater public. I understand that the limited space available prevents a thorough explanation of SIA, but there is no explanation of the general differences between d13C and d15N (geographical versus trophic information). I believe a quick sentence describing the two tracers in line 57 (for example) could help. 
The abstract was clear and direct, providing a good summary of the paper’s approach and conclusions.
The conclusion is clear and justified, although the claim of first quantitative evidence of movement behaviours in blue whale seems unlikely given the slightly older paper from Busquets-Vass also reporting blue whale baleen isotopes.

## Referee 2 :

I appreciated the opportunity to review the submitted manuscript “Reconstructing the last known movements of one of Nature’s giants”.

The authors use stable isotope analysis (SIA) of baleen of a single blue whale, stranded in 1891, to reconstruct a time series of ~6 years. This isotopic reconstruction represents the integrated isotopic ‘signal’ the whale encountered as it moved, allowing reconstruction of past movements. The authors paired these isotopic data with an ‘isoscape’ of values in the North Atlantic, as well as a model that simulated possible whale migration patterns over the isoscape. By combining these two approaches the authors suggest north-south movement for several years, followed by residency in subtropical waters. The authors claim that the similarity of these movements with modern blue whales suggest that movement patterns of blue whales in the North Atlantic have been similar for hundreds of years, despite anthropogenic pressures (e.g., whaling).

The paper is interesting and as the authors state, provides potentially valuable information on (one) historical blue whale movements in the North Atlantic Ocean. However, I have substantial concerns regarding the originality/novelty, assumptions, and conclusions of the study, enough so that I cannot recommend its publication in Nature.

The Abstract and text explicitly or implicitly state the novelty of the following: “first quantitative multi-year records of historical migrations in individual blue whales”, similarity of blue whale movements over hundreds of years, use of archived museum specimens, novel agent based simulation models.

The study does not provide "multi-year records of...individual blue whales", but of one blue whale. The generalization of the results from one individual to the entire population is a central issue to the current paper.

There are also issues with the movement interpretations for this one whale; the most important is that the authors make strong conclusions re: movement when the measured SIA values match those predicted by their model, but then employ a physiological argument for 'phase 2' (pregnancy and lactation) when there is a mismatch between measured data and the model. Of course, this approach makes the assumption that physiology can override the environmental signal, which calls into question any inferred movement patterns throughout the time series.

A main conclusion of the paper, that blue whales then migrated as they do now, simply cannot be made from one whale. I appreciate the difficulty of obtaining specimens like the one here, which makes larger sample sizes difficult to obtain. That limitation does make n = 1 studies reasonable, worthwhile, and interesting; it is my opinion that the results here, both for insights into the movements of this whale and for the methodology employed (combining a movement/isoscape model with measured values), should be published. However, n = 1 studies cannot be used to generalize the movements of an entire wild, historical population.

This paper is also not the first to use historical data to reconstruct past movements of blue whales; see for example Reeves et al (2004) that addresses the same question with whale landings data.

As the authors cite, the reconstruction of whale movements using baleen is well established in the literature; see papers by Best, Schell, Hobson, and others for similar analyses on Bowhead, Right, and other baleen whales.

The penultimate sentence of the abstract and final sentence of the paper suggest that using archived museum specimens is a novel concept in marine ecology. The use of archived museum specimens is not new or overlooked; two immediate examples (of many) include Vo et al (2010) using century-old seabirds and Yeakel et al (2009) on the famous lions of Tsavo. There are papers written that point to the value of using archived collections for data (e.g. Lister 2011), and the approach is common enough that there is a methods paper written on the preparation of archived specimens for SIA (Edwards et al 2002).

The “novel” movement model using SIA is indeed an improvement over many past studies using SIA to reconstruct animal movements. However, there are some studies have employed similar methods (e.g. Flockhart et al 2013 in Monarch butterflies).

In short, the authors applied a common tool in an established way to a single historical specimen of a rare species (though their agent-based movement model is a significant improvement over similar studies). Their model matched their measurements sometimes, but not always, leading to a movement-based interpretation for one period and a physiology-based interpretation (pregnancy and lactation) for the second period.

Perhaps most importantly, if the results and conclusions are considered completely correct, the authors still provide the movements of only a single whale. The conclusion that historical blue whale movements were the same as today’s based on a single individual is simply not appropriate.

By not overextending their results and by addressing a few issues, I do think this could be an interesting and valuable addition to the isotope and marine mammal literature.

See below for additional, specific comments.

Abstract: the inference of pregnancy and lactation seems more questionable than as it is stated here.

As mentioned above, the Abstract overextends the results (e.g., "migrations in blue whales" shouldnt be plural; "hundreds of years" which is not the timeline they present, etc).

Intro: migratory species may be affected by climate change by modifying movements, but aren't necessarily 'vulnerable'. Some would argue that migratory species are the least vulnerable to climate change.

lines 51-52 criticize approaches that are limited to only a few individuals, but this paper provides information on only one individual.

Fig 1 is not very clear. There is no clear indication of the whales movement, and no isotope data on the Figure. I understand the isoscape is in the SM but this figure could be improved to provide better movement and/or isotope data.

The movement simulations in Fig 2 are also confusing. The SM clarifies but this Figure could show more. For example, the data matches the simulations...of what? The reader cannot tell if the differences in simulations are due to isotopic variability or different movement scenarios. Perhaps the 4 movement scenarios shown in SM could be 4 different colors here. In that case, when the measured data match a given slope, peak, or trough, the reader can associate it with a geographic region.

lines 112-122. The conclusion of an isotopic reflection of pregnancy and lactation is stated rather strongly here given the evidence (the authors describe the inference as 'tentative' in the SM).

128: "resident" to where?

137: this point made extensively above....describes the overall "spatial ecology of blue whales in the North Atlantic" based on isotopic inferences over 6 years in a single whale.

139: Has it been demonstrated that whaling can actually modify movement patterns? Should be stated if so.

140-141: OK...but it is movement behavior in a single blue whale, and this has been done in larger sample sets of other baleen whales.

147-149. This sentence makes it seem like hardly anyone is using museum specimens for SIA or other chemical analyses, which simply isn't the case.

*We hope we have responded to all these comments appropriately. Please let us know if you require any further information,*

*Clive Trueman, on behalf of the other authors.*

### 2017-10-08 manuscript/revision/response.md
Minor edits

@@
 However, it was suggested that we transfer the paper to Nature Ecology and Evolution, which we are glad to do.
-We are transferring the reviews with the paper, therefore we include a response to reviewers as part of this cover letter.
+We are transferring the reviews with the paper, therefore we include a response to referees as part of this cover letter.
 We are grateful to the referees for their helpful and constructive comments, and hope that our edits, along with the move from Nature to Nature Ecology and Evolution, will have allayed their concerns about the manuscript.
 
-We respond to their points below. To improve the clarity of this document we have included the reviewers comments in italics.
+We respond to their points below. To improve the clarity of this document we have included our responses to the referees comments in italics.
 
-**Important note to the reviewers**.
+**Important note to the referees**.
 During the revisions of this manuscript, which included substantial extra analysis, we discovered an error in our original results.
@@
 However, since most of the literature is based on “eyeballing” cycles and visual estimates, the Fourier analysis is a good way to confirm the visual estimates.
+
 There are no error bars or replication of samples, but this is typical of stable isotopes.

### 2017-11-07 manuscript/revision/blue-whale-ms-v2.tex
Add edits from CT to draft

--- 2017-06-30 manuscript/blue-whale-ms.tex
@@
 
+%Points here about movement patterns, interactions with human use of oceans, tourism etc
+% Build on lack of information about movements before ,modern use of Atlantic, and lack of information about juvenile behaviour.
+% Introduce historic samples as archive behavoiiral information
+
 Baleen is formed of plates of keratin and is used by mysticete whales to filter feed. 
 It is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable}.
+
+%Refs to support long history of inferring movement from baleen isotopes.
+
 We sampled keratin from the blue whale's baleen plate at 1cm intervals, perpendicular to the direction of growth, and used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen (expressed as $\delta^{13}$C and $\delta^{15}$N values).
-Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplementary Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the last six years of the whale's life (autumn 1885 - spring 1891).
+Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplementary Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the seven full years of the whale's life (early 1884 - spring 1891).
 The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
-In behavioural phase one (spring 1885 to spring 1889), we find strong cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values with strong negative covariance (Fig.
+Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods).  
+In behavioural phase one (from record start in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Fig.
 \ref{fig1}; Figs.
 S1 and S2).
-In behavioural phase two (summer 1889 to spring 1891) we see a sharp increase in $\delta^{13}$C values for approximately eight months.
-This is associated with an overall reduction in $\delta^{15}$N values and a weakening of the periodic $\delta^{15}$N fluctuations seen in behavioural phase one. 
-There is also no covariance between $\delta^{13}$C and $\delta^{15}$N values in behavioral phase two (Fig. \ref{fig1}; Figs. S1 and S2).
+In behavioural phase two (summer 1886 to spring 1891) $\delta^{15}$N values are relatively elevated and we see coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values. 
+In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively positive $\delta^{13}$C values in the second half of 1891.
+The final three months of the record show a continued fall in $\delta^{13}$C values (Fig. \ref{fig1}; Figs. S1 and S2).
+
+% Edit figure legend please
 
@@
 \ref{fig1}) to identify the most likely movement history.
+We assume continued ingestion of new protein throughout the year. 
+Other mysticete whales fast during migrations and during winter, relying on stored lipid reserves to meet metabolic demands, but blue whales are too large to stop feeding for long periods; \cite{busquets2017estimating,silva2013north}).
+%Can we back this up? Are we more saying that we think they need protein to maintain tissue turnover but could meet energetic demands form stored lipids?
+
+%SOME indirect evidence for grey whale continued feeding through migrations -
 
@@
 
-During behavioural phase one, the whale is foraging at high northern latitudes, likely in the Norwegian Sea. 
-The pronounced cyclical variations in $\delta^{13}$C values are characteristic of the isotopic expression of the spring phytoplankton bloom in northern waters\cite{magozzi2017using}. 
-This interpretation is also consistent with the relatively negative median $\delta^{13}$C values (Fig. \ref{fig1}). 
-The periodicity of $\delta^{13}$C and $\delta^{15}$N values is best explained by short distance seasonal migrations, with the whale overwintering in the central or southern Norwegian Sea (Fig.
-\ref{fig2}).
-The relatively positive and invariant $\delta^{13}$C values seen during behavioural phase two are found in subtropical areas.
+The relatively positive and seasonally-invariant $\delta^{13}$C values seen during behavioural phase one are only found in subtropical areas of the North Atlantic.  
 Our simulations show high variance during this period (Fig.
-\ref{fig3}), reflecting a range of possible locations for the whale, although the most westerly areas (particularly to the west of the Azores; Fig.
+\ref{fig3}), reflecting a range of possible locations for the whale, although areas around the Cape Verde Islands, and potentially to the west of the Azores (Fig.
 \ref{fig2}), most closely match the measured profile.
-Continued, but reduced cyclical variation in $\delta^{15}$N values in 1889 and 1890 implies a degree of seasonal north-south migration before rapid migration to southerly latitudes in 1890, but continued feeding (other mysticete whales fast during migrations, but blue whales are too large to stop feeding for long periods\cite{busquets2017estimating,silva2013north}).
-The whale remained in these warm waters for around eight months and was conducting a rapid migration returning to her northern feeding grounds when she became stranded.
+The whale remained in these warm waters for at least two full years. 
+During behavioural phase two, starting at an approximate age of 5-6 years, the observed low $\delta^{13}$C values imply that the whale is foraging in colder, more northerly latitudes. 
+The pronounced cyclical variations in $\delta^{13}$C values are characteristic of the isotopic expression of the spring phytoplankton bloom in northern waters \cite{magozzi2017using}.
 
-The dramatic increase in $\delta^{13}$C values seen in the late spring and early summer of 1889 cannot be simulated purely from movements within the known geographic range.
-We believe that this reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions\cite{cardona2017temporal}. 
-The gradual increase in $\delta^{15}$N values to the autumn of 1890 is consistent with the isotopic expression associated with lactation\cite{cardona2017temporal}. 
-Blue whales have a 10-12 month gestation period, and calves are weaned after 6-7 months\cite{handbook}. 
-The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer 1890 and return to northern feeding grounds in early 1891 is therefore consistent with the known reproductive ecology of blue whales.
+We simulated 1000 individual movement patterns and compared resulting simulated baleen $\delta^{13}$C records to the measured records with simple linear regressions. 
+Simulated baleen $\delta^{13}$C profiles produce a good fit to measured profiles, the median $r^2$ value across 1000 simulated profiles was 0.49, and the maximum was 0.74 (Fig.
+S6, distribution of $r^2$ values).
+The top 10\% best fitting simulated profiles are shown in Fig. \ref{fig2} [line plot top 10\%). 
+Best fitting models predict juvenile residency in the Cape Verde region in behavioural phase 1. 
+Behavioural phase 2 is best simulated by seasonal migrations between summer foraging in northern areas in the Norwegian Sea/Barents Sea/Iceland region, and winter foraging in a broad region between the UK and sub-tropical waters.
+Best fitting models in general predicted a greater latitudinal foraging range, and foraging in more northerly waters (Supplementary materials Fig. S7 [boxplots and kernel density plots]).
+
+The last 500 days of the whale's life are difficult to simulate. 
+Beginning in the winter of 1889/1890, the observed $\delta^{13}$C profile is relatively depleted, and remains constant for approximately six months, before increasing rapidly in the winter of 1890/1891.
+Depleted $\delta^{13}$C values are found in northern waters that show large temporal fluctuations in $\delta^{13}$Cplk values [fig SX residency plot], and the observed values cannot be simulated purely from movements within the known geographic range.
+We infer that the $\delta^{13}$C record in the last period of life reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions\cite{cardona2017temporal}.
+We suggest that the constant, depleted $\delta^{13}$C values reflect use of carbon reserves assimilated in northern latitudes during 6-8 months residency in low latitude waters. 
+After six months the whale began feeding in low latitude waters producing a dramatic increase in $\delta^{13}$C values and began a final migration north. 
+Blue whales have a 10-12 month gestation period, with calving occurring in sub-tropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
+The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, residency in warm waters through the whole of 1890 with feeding resuming in the second half of 1890 and eventual stranding during the return to northern feeding grounds in early 1891, is therefore consistent with the known reproductive ecology of blue whales.
+
+%Possibly more interesting now to talk about juvenile behaviour and potential for prolonged residency in warm waters  - implications for interactions with tourist / shipping etc.
+
+%Kernel density plot gives a visualization of proportional amount of time we infer the whale might have spent in each region
 

### 2017-11-07 manuscript/revision/blue-whale-supplementary-v2.tex
Typo correction for referee 1

--- 2017-06-30 manuscript/blue-whale-supplementary.tex
@@
 $\delta^{13}$C values also show repeated fluctuations in the first (oldest) 60cm of the baleen plate (behavioural phase one), with a similar periodicity of 13.5cm (Fig. \ref{figs1}). 
-A distinct change in the isotopic pattern along the baleen plate occurs at around 60cm behavioural phase two), with a transition to relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.
+A distinct change in the isotopic pattern along the baleen plate occurs at around 60cm behavioural phase two, with a transition to relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.
 
@@
   \includegraphics[width = \linewidth]{figures/Figure-S1-periodograms.png}
-  \caption{Fourier analysis of $\delta^{15}$N (A,B) and $\delta^{13}$C (C,D) values of the full (A,C) and behavioral phase one only (B,D) components of the baleen plate.
-Inset values are the two highest frequency spectral densities (in cm).
-Note the baleen plate is 80cm long.
-Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent in behavioural phase two}
+  \caption{Fourier analysis of $\delta^{15}$N (A,B) and $\delta^{13}$C (C,D) values of the full (A,C) and behavioural phase one only (B,D) components of the baleen plate.
+Inset values are the highest frequency spectral densities (in cm), excluding the values at the edge of the baleen (i.e.
+80cm), and at the sharp increase of $\delta^{13}$C seen at 60cm.
+Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent in behavioural phase two.}
   \label{figs1}
@@
 
-Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioral phase one (oldest 60cm of the baleen plate; Fig.
-\ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioral phase two (the final 20cm of the baleen plate).
+Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase one (oldest 60cm of the baleen plate; Fig.
+\ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioural phase two (the final 20cm of the baleen plate).
 
@@
   \includegraphics[width = \linewidth]{figures/Figure-S2-cross-cor.png}
-  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for behavioural phase one showing strong negative covariance with a periodicity of around 13cm} % check legend
+  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for behavioural phase one showing strong negative covariance with a periodicity of around 13cm.} % check legend
   \label{figs2}
@@
 Simulated $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c.60$^{\circ}$N, and relatively negative in the central and southern North Atlantic, associated with high levels of nitrogen fixation (Fig.
-\ref{figs3}B).
+\ref{figs3}).
 Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
-In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. \ref{figs3}A). 
+In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. \ref{figs3}). 
 The temporally coincident periodic, inverse fluctuations in $\delta^{13}$C and $\delta^{15}$N values in the initial phase of the baleen plate are therefore consistent with migration between waters north of 60$^{\circ}$N and more southerly southern waters in the North Atlantic, with northerly occupancy in summer months and southerly occupancy in winter months.
@@
     \end{subfigure}
-    \caption{Spatial model (isoscape) of model-simulated biomass weighted annual average (A) $\delta^{13}$C and (B) $\delta^{15}$N values from POM.
+    \caption{Spatial model (isoscape) of model-simulated biomass weighted annual average $\delta^{13}$C (left) and $\delta^{15}$N (right) values from POM.
 $\delta^{13}$C values are from Magozzi et al.
-2017\cite{magozzi2017using} and $\delta^{15}$N from Schmittner \& Somes 2016\cite{schmittner2016complementary} and Somes (pers.
-comm.)}
+2017\cite{magozzi2017using} and $\delta^{15}$N from Schmittner \& Somes 2016\cite{schmittner2016complementary} and C.J.
+Somes (pers.
+comm.).}
     \label{figs3}
@@
 \subsection*{ISOTOPE MODEL SIMULATIONS}
-We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions of that would be encountered by whales foraging within broad geographic regions (Norwegian Sea, west UK/Irish shelf, Canaries/west Azores, Mid Atlantic ridge; Fig.
+We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions that would be encountered by whales foraging within broad geographic regions (west UK/Irish shelf, Norwegian/Barents Sea, Canaries/west Azores, Mid Atlantic ridge; Fig.
 \ref{figs4}).
@@
       \includegraphics[width=\linewidth]{figures/Figure-S4-facet-wrap-d13C.png}
-      \caption{Alternative $\delta^{13}$C series arising from resident-only movement models restricting region to Norway, Ireland, Canaries and the Mid Atlantic ridge as labelled over a simulated two year period with daily resolution and 30 replicate simulations for each region.} % check legend
+      \caption{Alternative $\delta^{13}$C series arising from resident-only movement models restricting region to west UK/Irish shelf (Ireland), Norwegian/Barents Sea (Norway), Canaries/west Azores (Canaries), Mid Atlantic ridge (Mid-Atlantic), over a simulated two year period with daily resolution and 30 replicate simulations for each region.} % check legend
       \label{figs4}
@@
 \subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase one}
-Temporal dynamics in $\delta^{13}$C values simulated in northerly latitudes are similar to the measured $\delta^{13}$C profile during the four years of behavioral phase one. 
-Simulated profiles display the same asymmetric isotopic cycles associated with development of the spring bloom, with annual $\delta^{13}$C maxima in June (Fig. \ref{figs4}, Fig. 1). 
-Simulated temporal isotopic profiles for northern resident whales, do not display clear isotopic minima preceding the onset of the bloom. 
+Temporal dynamics in $\delta^{13}$C values simulated in northerly latitudes are similar to the measured $\delta^{13}$C profile during the four years of behavioural phase one. 
+Simulated profiles display the same asymmetric isotopic cycles associated with development of the spring bloom, with annual $\delta^{13}$C maxima in June (Fig. \ref{figs4}, Fig. 3). 
+Simulated temporal isotopic profiles for northern resident whales, do not display clear isotopic minima preceding the onset of the bloom (Fig. \ref{figs4}). 
 Adding seasonal north-south migrations within mid-high latitude regions to foraging models, closely matches the measured profile, including $\delta^{13}$C minima preceding the June bloom (Fig.
@@
 \subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase two}
-Behavioural phase two is characterized by a dramatic shift to maximum $\delta^{13}$C values in spring 1889 followed by approximately 18 months of relatively positive and constant $\delta^{13}$C values.
-$\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity associated with north-south migration is absent in 1890. 
+Behavioural phase two is characterised by a dramatic shift to maximum $\delta^{13}$C values in spring 1889 followed by approximately 18 months of relatively positive and constant $\delta^{13}$C values.
+$\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity associated with north-south migration is absent in 1890 (Fig. 1). 
 The final sample representing feeding shortly before death in Irish waters implies a return to isotopic conditions more consistent with behavioural phase one. 
@@
 
-Accordingly we modelled whale movements allowing five years of seasonal north-south migration in the north east northern Atlantic, followed by rapid migration to subtropical waters in the late spring of 1889, relative residency for eight months and rapid migration returning to more northerly waters in the spring of 1891.
-The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Fig. 1). 
+Accordingly we modeled whale movements allowing five years of seasonal north-south migration in the north east northern Atlantic, followed by rapid migration to subtropical waters in the late spring of 1889, relative residency for eight months and rapid migration returning to more northerly waters in the spring of 1891.
+The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Fig. 3). 
 Consistently low $\delta^{15}$N values in 1890 are also consistent with residency in relatively southern waters.
@@
 \subsection*{AGENT-BASED WHALE MOVEMENT MODEL}
-We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using,yool2013medusa}, where movement behavior is influenced by SST and phytoplankton biomass estimates provided by NEMO-MEDUSA\cite{yool2013medusa}, and bathymetry (Fig.
+We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent-based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using,yool2013medusa}, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA\cite{yool2013medusa}, and bathymetry (Fig.
 \ref{figs5}).
@@
 The agent-based model developed employs a simple set of movement rules dictated by behavioural state which is itself fixed according to calendar month by the operator. 
-At each daily step, the probability of movement, movement directions and distance traveled are sampled from probability distributions to allow individual variation.
+At each daily step, the probability of movement, movement direction and distance traveled are sampled from probability distributions to allow individual variation.
 
@@
   \end{tabular}
-  \caption{Model parameters for agent-based model of whale movement. See Fig. \ref{figs5} for details.}
+  \caption{Model parameters for agent-based model of whale movement. See Fig. \ref{figs5} and text for more details.}
   \label{tables1}

### 2017-11-07 manuscript/revision/response.md
Edits to manuscript and cover letter

@@
-**RE: Nature submission 2017-06-07324A**
+# RE: Nature submission 2017-06-07324A
 
@@
 During the revisions of this manuscript, which included substantial extra analysis, we discovered an error in our original results.
-Essentially, the baleen samples had been labelled in the reverse direction to what we expected.
+Essentially, the baleen samples had been labeled in the reverse direction to what we expected.
 The values for all the stable isotope measures remain the same but the timeline is reversed (this occurred due to a misunderstanding with the masters student who ran to original mass spec analyses and subsequently finished her masters project and moved away).
@@
 We communicated with Editor in Chief Patrick Goymer about whether it was still appropriate to cascade our reviews with the paper given this error, and he said this was fine as long as we made this very clear in our response letter.
-This note is to alert the reveiwers that the results graphs will look different to the original version, though our conclusions have not changed qualitatively.
+This note is to alert the referees that the results graphs will look different to the original version, though our conclusions have not changed qualitatively.
+
+## Original referee comments
 
@@
 
-*We thank the referee for their positive comments about our paper.*
+> We thank the referee for their positive comments about our paper.
 
@@
 I suppose it depends on whether the model was used for a direct pattern comparison, or if the values modelled and the baleen values were compared to each other.
-Comparable literature in other species (seals, humans) appear to confirm the pregnancy interpretation, although I’m less convinced by the lactation signal as it seems too delayed in time.
+
+> The referee is correct
+
+* CLIVE how are we dealing with this???
+
+Comparable literature in other species (seals, humans) appear to confirm the pregnancy interpretation, although I'm less convinced by the lactation signal as it seems too delayed in time.
 This could possibly be checked using hormones, although that is also a very novel field.
@@
 I strongly believe this will not prohibit publication, as the results are still very interesting. It is very possible that I am misreading the delay in lactation signal.
+
+> Measuring hormone levels in the baleen would be a great addition to this work.
+We have colleagues in the US who are planning to extract hormones from this baleen.
+Unfortunately these results have been delayed now for over a year due to US permitting issues (thanks Trump!) and the CITES status of blue whales.
+But we agree that this would be really interesting, if the hormone signals remain over time.
+
+* LACTATION SIGNAL - comments?
 
@@
 
+* Do we need a comment on this?
+
 ### Suggested improvements/ References:
@@
 Line 82: reference again with basic SIA shifts in marine food webs?
-Possibly McMahon et al, 2015 “Millennial Scale plankton regime shifts in the subtropical North Pacific Ocean”
+Possibly McMahon et al, 2015 “Millennial Scale plankton regime shifts in the subtropical North Pacific Ocean”.
+
+> We have added this as suggested
+
+>
+
+* Add it!
+
 The reported baleen growth rates fit the literature, but I believe it may be worth mentioning in Line 61 that baleen gets worn down over time, and that the records are not full life spans.
+
+> We have clarified this in the text at line XXX as follows: "It is ideal for stable isotope studies because keratin grows continuously through an individual's life (although it is worn away at the tips over time, so rarely records an individual's entire lifespan), and once laid down it is metabolically inert."
+
 According to the model, d15N is relatively higher in the arctic during summer than the Atlantic in winter.
@@
 I am mostly curious, but it may reflect difficulties with reading the model from a figure.
+
+> Explain
+
 While the d13C increase fits the sea lion theory, the drop in d13C seen after/at birth happens too quick (late 1889) in the whale to fit the d15N rise consistent with lactation (Almost 6 months late according to timeline, past mid 1890).
-Human analyses in hair showed a decrease in d15N (1%o) during pregnancy, which can be seen in the baleen if we take the d13C drop as birth time and early 1889 as pregnancy time (see Fuller et al 2004 “Nitrogen Balance and 15N, why you’re not what you eat during pregnancy”).
-While it’s not a study conducted in marine mammals, it’s a further indication that the birth time may indeed be in late 1889, or at least that the lactation peak is very late.
+Human analyses in hair showed a decrease in d15N (1%o) during pregnancy, which can be seen in the baleen if we take the d13C drop as birth time and early 1889 as pregnancy time (see Fuller et al 2004 “Nitrogen Balance and 15N, why you're not what you eat during pregnancy”).
+While it's not a study conducted in marine mammals, it's a further indication that the birth time may indeed be in late 1889, or at least that the lactation peak is very late.
 Could indicate a slowing down in baleen growth, or something else?
 Could be worth checking with hormonal analyses (see above).
+
+> Explain
+
 Another possibility lies in the location of baleen analyses.
-Since the methods are not clear on that point, can you confirm that the samples were taken past the gum
-line?
+Since the methods are not clear on that point, can you confirm that the samples were taken past the gum line?
 Dried skin/gum could have affected the isotopic signals if samples were taken outside of the erupted zone.
@@
 Busquet Vass paper shows no difference in 13C between skin and baleen so unlikely, but worth confirming.
+
+> Check
+
 The final sample interpretation of a return to cyclicity is very open to discussion, as a single point is too weak to be a guarantee.
 I will concede that the isotopes are consistent with past migratory values, but may be worth softening the comment (line 109-110)?
+
+> OK
+
 When describing the different behaviours, such as whales staying in high latitudes instead of migrating, also talk about the ones staying at low lat (isotope blue whale ref.
 Busquet-Vass) since it is more relevant to your own isotopic interpretation (line 127).
+
+> Alright
 
@@
 Line 20: Should not have abbreviation. Move to line 58 where it is introduced next.
+
+> Fixed
+
 2. Article main text
 Line 56: “)” extraneous bracket
+
+> Fixed
+
 3. Supplementary Data
 Page 2, line 1: “)” extraneous bracket after “behavioural phase 2”
+
+> Fixed
 
@@
 I believe a quick sentence describing the two tracers in line 57 (for example) could help.
-The abstract was clear and direct, providing a good summary of the paper’s approach and conclusions.
+
+> We have added a brief comment on what blue whales eat on line XXX:.
+
+> We have also added some simpler references to stable isotopes.
+For example.
+We also calrify the units used as this often trips up new users: lin XXX "Note that carbon and nitrogen isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e.
+relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope (West et al 2006)."
+"
+
+The abstract was clear and direct, providing a good summary of the paper's approach and conclusions.
 The conclusion is clear and justified, although the claim of first quantitative evidence of movement behaviours in blue whale seems unlikely given the slightly older paper from Busquets-Vass also reporting blue whale baleen isotopes.
 
-## Referee 2 :
+> Clarify
 
-I appreciated the opportunity to review the submitted manuscript “Reconstructing the last known movements of one of Nature’s giants”.
+## Referee 2:
+
+> If we are interpreting Referee 2's comments correctly (see below), we think they mostly revolve around issues about the novelty of the study, and overstatement of results based on one individual.
+We completely see the points the referee is making.
+Some of this was related to the difficulties of writing in the shorthand Nature style - we have clearly been a little too casual in some of our wording.
+In other cases we haven't been clear that the selling point of this method is that it can easily be applied to hundreds of individuals at relatively low cost.
+Here we test the methods on one individual, but the real value in the method is its broader applicability.
+We have tried to clarify this where possible and have also toned down/removed/clarified a lot of our previous statements about the novelty of the study and interpretations of the results.
+We explain our changes in more detail below.
+
+line XXX: These methods can be applied to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum (NHM) London's blue whale `Hope'.
+
+I appreciated the opportunity to review the submitted manuscript “Reconstructing the last known movements of one of Nature's giants”.
 
@@
 
+> Agreed
+
 There are also issues with the movement interpretations for this one whale; the most important is that the authors make strong conclusions re: movement when the measured SIA values match those predicted by their model, but then employ a physiological argument for 'phase 2' (pregnancy and lactation) when there is a mismatch between measured data and the model.
 Of course, this approach makes the assumption that physiology can override the environmental signal, which calls into question any inferred movement patterns throughout the time series.
+
+> OK
 
@@
 
+> OK keep your hair on
+
 This paper is also not the first to use historical data to reconstruct past movements of blue whales; see for example Reeves et al (2004) that addresses the same question with whale landings data.
 
+> OK
+
 As the authors cite, the reconstruction of whale movements using baleen is well established in the literature; see papers by Best, Schell, Hobson, and others for similar analyses on Bowhead, Right, and other baleen whales.
+
+> OK
 
@@
 
+> OK
+
 The “novel” movement model using SIA is indeed an improvement over many past studies using SIA to reconstruct animal movements.
@@
 
+> OK
+
 In short, the authors applied a common tool in an established way to a single historical specimen of a rare species (though their agent-based movement model is a significant improvement over similar studies).
 Their model matched their measurements sometimes, but not always, leading to a movement-based interpretation for one period and a physiology-based interpretation (pregnancy and lactation) for the second period.
+
+>
 
@@
 
+>
+
 By not overextending their results and by addressing a few issues, I do think this could be an interesting and valuable addition to the isotope and marine mammal literature.
+
+> We hope our edits have addressed these issues.
 
@@
 
+>
+
 As mentioned above, the Abstract overextends the results (e.g., "migrations in blue whales" shouldnt be plural; "hundreds of years" which is not the timeline they present, etc).
+
+>
 
@@
 
+>
+
 lines 51-52 criticize approaches that are limited to only a few individuals, but this paper provides information on only one individual.
+
+>
 
@@
 I understand the isoscape is in the SM but this figure could be improved to provide better movement and/or isotope data.
+
+>
 
@@
 
+>
+
 lines 112-122. The conclusion of an isotopic reflection of pregnancy and lactation is stated rather strongly here given the evidence (the authors describe the inference as 'tentative' in the SM).
+
+>
 
@@
 
+>
+
 137: this point made extensively above....describes the overall "spatial ecology of blue whales in the North Atlantic" based on isotopic inferences over 6 years in a single whale.
+
+>
 
@@
 
+>
+
 140-141: OK...but it is movement behavior in a single blue whale, and this has been done in larger sample sets of other baleen whales.
+
+>
 
@@
 
-*We hope we have responded to all these comments appropriately. Please let us know if you require any further information,*
+>
 
-*Clive Trueman, on behalf of the other authors.*
+> We hope we have responded to all these comments appropriately. Please let us know if you require any further information,
+
+> Clive Trueman, on behalf of the authors.

### 2017-11-07 manuscript/revision/blue-whale-ms-v2.tex
Edits to manuscript and cover letter

@@
 \raggedright
-
+% Need to edit
 \paragraph{Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}. 
 Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}.
-Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct the movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London (NHM).
+Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct the movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
 We identify two distinct phases in the whale's behaviour: repeated annual north-south migrations in sub-Arctic waters, and residency in subtropical waters. 
@@
 These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
-Intrinsic biogeochemical tags provide an alternative technique, where the chemical compositions of animal tissues are compared to spatially referenced models of chemical variation across the species' range\cite{hobson1999tracing,hobson2008tracking}).
-Stable isotope-based geolocation has been used to interpret whale ecology \cite{hobson1998stable,ryan2013stable,eisenmann2016isotopic}. 
-Here we apply this method to baleen from the NHM's blue whale.
+Stable isotope analyses provide an alternative technique\cite{west2006stable}. 
+We can use the isotopic compositions of whale tissues to reconstruct their movement patterns, by comparing them to spatially referenced models of isotopic variation across the species' range\cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
+%line 57,
+These methods can be applied to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum (NHM) London's blue whale `Hope'.
 
@@
 Baleen is formed of plates of keratin and is used by mysticete whales to filter feed. 
-It is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable}.
-
+It is ideal for stable isotope studies because keratin grows continuously through an individual's life (although it is worn away at the tips over time, so rarely records an individual's entire lifespan), and once laid down it is metabolically inert\cite{best1996stable}.
+Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton.
+Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on around two weeks prior to that time point (two weeks is required for the carbon and nitrogen to be assimilated and converted into tissues).
+%REF???
+Baleen stable isotopes have long been used in this way /.... % finish
 %Refs to support long history of inferring movement from baleen isotopes.
 
-We sampled keratin from the blue whale's baleen plate at 1cm intervals, perpendicular to the direction of growth, and used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen (expressed as $\delta^{13}$C and $\delta^{15}$N values).
+We sampled keratin from the blue whale's baleen plate at 1cm intervals for 100cm, perpendicular to the direction of growth, and used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen.
+Note that carbon and nitrogen isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e.
+relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope\cite{west2006stable}.
+
 Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplementary Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the seven full years of the whale's life (early 1884 - spring 1891).
@@
   \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
-  \caption{Variation in $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis) longitudinally through the baleen plate (n = 77 samples from a single baleen plate for both isotopes) of the NHM blue whale exhibits strong annual periodicity and cross-correlation (Figs.
+  \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis).
+Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes) of the NHM blue whale exhibits strong annual periodicity and cross-correlation (Figs.
 S1 and S2) in both isotopes as well as a change in the mean and magnitude of the cycles during 1889 (approximately at sample 30cm).
@@
 
+% Krill has a very similar isotopic composition to phytoplankton
+
 Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values, and because the isotopic composition of phytoplankton at the base of marine food webs varies in both space and time.
+%line 82
 We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Fig.
@@
 
+line 112 - 122 para
 The last 500 days of the whale's life are difficult to simulate. 
@@
 Conventional wisdom suggests that baleen whales migrate each year from feeding grounds near the poles in summer, to overwintering and breeding grounds nearer the equator\cite{corkeron1999baleen,lockyer1981migration}.
-However, a growing body of evidence suggests that not all whales migrate to subtropical waters each year, with some remaining at high latitudes year round\cite{mcdonald2006biogeographic}. 
+However, a growing body of evidence suggests that not all whales migrate to subtropical waters each year, with some remaining at high latitudes year round\cite{mcdonald2006biogeographic}. %line 127
 Our results support this conclusion. 
-Interestingly, there is still a resident population of blue whales in the regions indicated by our model, and females still breed in the subtropical waters near the Azores\cite{reilly2008balaenoptera}.
+Interestingly, there is still a resident%line 128
+ population of blue whales in the regions indicated by our model, and females still breed in the subtropical waters near the Azores\cite{reilly2008balaenoptera}. 
 Whaling was an intense pressure for blue whales during the period we are analysing. 
@@
 The NHM blue whale thus represents an individual from a species at the brink of extinction.
-Our data suggest that the spatial ecology of blue whales in the North Atlantic has remained similar for more than 100 years, despite dramatic changes in anthropogenic pressures.
+Our data suggest that the spatial ecology of blue whales in the North Atlantic has remained similar for more than 100 years, despite dramatic changes in anthropogenic pressures. % line 137 and 139
 
-This is the first quantitative evidence of historical individual-level movement behaviours in blue whales. 
+This is the first quantitative evidence of historical individual-level movement behaviours in blue whales. %line 140-141
 Our novel simulation modeling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviours over multiple years.
 Such techniques can also be applied to extinct species behaviours and used to understand how populations were affected by past pressures such as hunting, and present-day pressures such as global change.
-Finally, our results highlight that museums represent a vast, mostly untapped, source of individual-level spatial ecology information.
+Finally, our results highlight that museums represent a vast, mostly untapped, source of individual-level spatial ecology information. %line 147-149
 

### 2017-11-08 manuscript/revision/response.md
Minor edits wrt referee comments

@@
 
-> We have added this as suggested
+> We have added these as suggested.
 
->
-
-* Add it!
+* In line XX we specifically mention that blue hwale feed on krill ("Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton."). 
+* In lines XX we mention the trophic influence on isotopic signatures [NEED to D/o this]
+* We have added the West et al reference at lines XX and XX
+* We have added an explanation of d13C and d15N at line XX: "Note that carbon and nitrogen isotopes are expressed as d13C and d15N values, i.e.
+relative to a standard and as ratios of the rare (13C, 15N) to common (12C, 14N) isotope"
+* We have added the McMahon reference at line XX [ADD]
 
@@
 
-> Explain
+> Explain. I'm not sure what is going on here - Clive?
 

### 2017-11-08 manuscript/revision/blue-whale-ms-v2.tex
Minor edits wrt referee comments

@@
 
-Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values, and because the isotopic composition of phytoplankton at the base of marine food webs varies in both space and time.
-%line 82
+Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values, and because the isotopic composition of phytoplankton at the base of marine food webs varies in both space and time \cite{west2006stable}.
+%line 82 add ref mentioned by reviewer 1
 We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Fig.
@@
 
-line 112 - 122 para
+%line 112 - 122 para
 The last 500 days of the whale's life are difficult to simulate. 

### 2017-11-14 manuscript/revision/blue-whale-ms-v2.tex
Minor edits

@@
 %\section{Main text}
-% 1500 words max (excludes refs title, authors, abstract, acknowledgements)
 \newpage
@@
 These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
-Stable isotope analyses provide an alternative technique\cite{west2006stable}. 
+Stable isotope analyses provide an alternative technique for investigating blue whale movements\cite{west2006stable}. 
 We can use the isotopic compositions of whale tissues to reconstruct their movement patterns, by comparing them to spatially referenced models of isotopic variation across the species' range\cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
 %line 57,
-These methods can be applied to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum (NHM) London's blue whale `Hope'.
+These methods can be applied to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale `Hope'.
 
@@
 Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton.
-Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on around two weeks prior to that time point (two weeks is required for the carbon and nitrogen to be assimilated and converted into tissues).
+Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on around two weeks prior to that time (two weeks is required for the carbon and nitrogen to be assimilated and converted into tissues).
 %REF???
-Baleen stable isotopes have long been used in this way /.... % finish
+Baleen stable isotopes have a long history of use as biochemical markers to infer movement patterns in whales (REFS). % ADD REFS
 %Refs to support long history of inferring movement from baleen isotopes.
@@
   \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis).
-Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes) of the NHM blue whale exhibits strong annual periodicity and cross-correlation (Figs.
+Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes).
+There is strong annual periodicity and cross-correlation (Figs.
 S1 and S2) in both isotopes as well as a change in the mean and magnitude of the cycles during 1889 (approximately at sample 30cm).

### 2017-11-15 manuscript/revision/response.md
Add responses to referee 1

--- 2017-11-07 manuscript/revision/response.md
@@
 
-> The referee is correct
+> The referee is correct that blue whales feed predominantly on krill (small crustaceans), not directly on phytoplankton.
+However, the trophic shift from phytoplankton to crustacean predators is very small
 
@@
 
-* LACTATION SIGNAL - comments?
+> In terms of the lactation signal, the reviewer makes a good point, and the changes to our results in this version also fail to demonstrate a convincing signal of lactation.
+We have therefore removed the mention of the lactation signal from the manuscript.
+We still see evidence of pregnancy and have retained this interpretation.
 
@@
 
-* Do we need a comment on this?
+> Thank you
 
@@
 
-> We have added this as suggested
+> We have added these as suggested.
 
->
-
-* Add it!
+* In line XX we specifically mention that blue whales feed on krill ("Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton."). 
+* In lines XX we mention the trophic influence on isotopic signatures [NEED to DO this]
+* We have added the West et al reference at lines XX and XX
+* We have added an explanation of d13C and d15N at line XX: "Note that carbon and nitrogen isotopes are expressed as d13C and d15N values, i.e.
+relative to a standard and as ratios of the rare (13C, 15N) to common (12C, 14N) isotope"
+* We have added the McMahon reference at line XX [ADD]
 
@@
 
-> Explain
+> As mentioned above, we are unable to complete hormonal analyses at this time (and are unsure if the signal will remain for over 100 years).
+However, we agree with the reviewers concerns and have removed mention of lactation signal from the manuscript, particularly after our reanalysis removed the strength of any signal.
 
@@
 
-> Check
+> We sampled starting at 0.5cm from the proximal end of the baleen, which is likely to cross the gum line.
+However, the baleen plate was cleaned thoroughly before analysis, and even prior to this no trace of skin or gum could be seen on the plates (this material was carefully cleaned back in the 1800s when the baleen was first stored).
+We therefore do not believe this reflects skin or gum influencing the isotopic signals.
+We had mentioned cleaning the plate in the methods previously, but have now explicitly mentioned that this cleaning removes skin/gum (and that this can influence stable isotope readings) at line XX: "The baleen plate was cleaned with ethanol to remove surface contaminants such as skin/gum or other lipids that can influence isotopic signals."
 
@@
 
-> OK
+> This is a fair point. We have now edited this whole section because of the reverse of the timescale of the analysis and so have removed this comment.
 
@@
 
-> We have added a brief comment on what blue whales eat on line XXX:.
+> We have added a brief comment on what blue whales eat on line XXX: "Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton".
 
-> We have also added some simpler references to stable isotopes.
-For example.
-We also calrify the units used as this often trips up new users: lin XXX "Note that carbon and nitrogen isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e.
-relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope (West et al 2006)."
-"
+> We have also added some basic information on stable isotopes, but have tried to avoid increasing the word count too much.
+We clarify the units used as this often trips up new users.
+line XXX "We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin.
+These isotopes are expressed as d13C and d15N values, i.e.
+relative to a standard and as ratios of the rare (13C and 15N) to common (12C and 14N) isotope (West et al 2006).
+Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; isotopes provide information on the geographical locations of an individual, whereas nitrogen isotopes provide trophic and physiological information (West et al 2006)."
 
@@
 
-> Clarify
+> We have removed this comment from the manuscript, and discuss the novelty of our work more in the response to referee 2 below.
 

### 2017-11-15 manuscript/revision/blue-whale-ms-v2.tex
Add responses to referee 1

@@
 We identify two distinct phases in the whale's behaviour: repeated annual north-south migrations in sub-Arctic waters, and residency in subtropical waters. 
-Subtropical residency is associated with isotopic signatures consistent with pregnancy and lactation in the year before the whale's death. 
+Subtropical residency is associated with isotopic signatures consistent with pregnancy in the year before the whale's death. 
 These results supply the first quantitative multi-year records of historical migrations in individual blue whales, and suggest that blue whale residency and movement patterns in the North Atlantic have remained similar for hundreds of years, despite changing human pressures in the region.
@@
 It is ideal for stable isotope studies because keratin grows continuously through an individual's life (although it is worn away at the tips over time, so rarely records an individual's entire lifespan), and once laid down it is metabolically inert\cite{best1996stable}.
-Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton.
+Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton \cite{handbook}.
 Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on around two weeks prior to that time (two weeks is required for the carbon and nitrogen to be assimilated and converted into tissues).
@@
 
-We sampled keratin from the blue whale's baleen plate at 1cm intervals for 100cm, perpendicular to the direction of growth, and used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen.
-Note that carbon and nitrogen isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e.
+We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth. 
+This resulted in XX discrete samples of baleen. % add number
+We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin. 
+These isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e.
 relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope\cite{west2006stable}.
+Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; isotopes provide information on the geographical locations of an individual, whereas nitrogen isotopes provide trophic and physiological information\cite{west2006stable}.
 
@@
 
-Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values, and because the isotopic composition of phytoplankton at the base of marine food webs varies in both space and time \cite{west2006stable}.
-%line 82 add ref mentioned by reviewer 1
+Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values, and because the isotopic composition of phytoplankton at the base of marine food webs varies in both space and time \cite{west2006stable,mcmahon2015millennial}.
 We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Fig.
@@
 Baleen was collected from the Natural History Museum, London (specimen NHMUK.1892.3.1.1). 
-The baleen plate was cleaned with ethanol to remove surface contaminants. 
+The baleen plate was cleaned with ethanol to remove surface contaminants such as skin/gum or other lipids that can influence isotopic signals. 
 {\raise.17ex\hbox{$\scriptstyle\sim$}}1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
-77 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section which contains the most recent tissue. 
+77% edit number
+ samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section which contains the most recent tissue. 
 Baleen grows at a constant rate, so the samples are equally spaced through time\cite{best1996stable}.

### 2017-11-15 manuscript/revision/response.md
Edits to main text in response to referee 2

@@
 We have tried to clarify this where possible and have also toned down/removed/clarified a lot of our previous statements about the novelty of the study and interpretations of the results.
-We explain our changes in more detail below.
-
-line XXX: These methods can be applied to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum (NHM) London's blue whale `Hope'.
+We explain our changes in more detail below, and apologise for any offense caused by our apparent disregard for previous excellent work in this area.
 
@@
 
-> Agreed
+> This is a fair point, and partially the result of some lax wording on our part. We have toned down our claims of novelty throughout and removed this sentence from the abstract.
 
@@
 
-> OK
+> The referee is correct in that
 
@@
 
-> OK keep your hair on
+> This is a fair point again.
+As stated above this is partially an issue of our wording and attempts to write this in a Nature style.
+We have toned down these comments throughout, and focus instead on promoting the methodological approach we have used, which we do believe is novel.
 
@@
 
-> OK
-
 As the authors cite, the reconstruction of whale movements using baleen is well established in the literature; see papers by Best, Schell, Hobson, and others for similar analyses on Bowhead, Right, and other baleen whales.
 
-> OK
+> Yes we are aware of these other papers, and cite many of them already.
+We agree that our wording in the original manuscript was not very clear and overly stated the novelty of our findings.
+We have tried to tone this down throughout, and provide more balanced coverage of previous work.
 
@@
 
-> OK
+> Again we realise that our wording was lax in places and overstated the novelty of this approach.
+This was not our intention - our intention was to encourage more ecologists to consider these archives of data in their work.
+We are often surprised by how under-utilised our collections are for this kind of work (even in the NHM) so wanted to highlight this.
+We have toned down our claims of novelty and have added more references to clarify that this is an established field of research.
 
@@
 
-> OK
+> OK ???? HAVE THEY???
 
@@
 
->
+Perhaps most importantly, if the results and conclusions are considered completely correct, the authors still provide the movements of only a single whale.
+The conclusion that historical blue whale movements were the same as today's based on a single individual is simply not appropriate.
 
-Perhaps most importantly, if the results and conclusions are considered completely correct, the authors still provide the movements of only a single whale.
-The conclusion that historical blue whale movements were the same as today’s based on a single individual is simply not appropriate.
-
->
+> We have removed this from the text and have toned down interpretations that relate to just a single individual.
 
@@
 
->
+> We have removed the reference to lactation, and reworded to make it clear that the pregnancy conclusion is more tenuous than it appeared in the original abstract.
 
@@
 
->
+> We have reworded the abstract and removed this inaccurate comment.
 
@@
 
->
+> We have reworded this to "Such species may also be particularly affected by climate change".
 
@@
 
->
+> Fair point.
+We were trying to clarify that our approach is more scalable, but agree that this was poorly worded.
+Instead we now state on line XXX: "These methods can be applied to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum (NHM) London's blue whale `Hope'."
 
@@
 
->
+> We are a little confused by this comment. I assume the referee meant Figure 2? Figure 1 is a trace of d13C and d15N, and quite clearly contains isotope data. 
+FIX!!!
 
@@
 
->
+> Again, we assume the referee is referring to Fig 3 not Fig 2.
+This figure has changed now we have changed our movement models and added more data.
+We have clarified what the figure is showing in the legend as follows: :
+FIX!!!
 
@@
 
->
+> We have rephrased this paragraph in response to the new data and analyses we have added. We have also toned down the claims of pregnancy and removed those about lactation.
 
@@
 
->
+> We have removed this statement and toned down claims based on this one whale throughout.
 
@@
 
->
+> We have rephrased this.
 

### 2017-11-15 manuscript/revision/blue-whale-ms-v2.tex
Edits to main text in response to referee 2

@@
 % Need to edit
-\paragraph{Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}. 
+
+\section{Abstract}
+Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}. 
 Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}.
 Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct the movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
-We identify two distinct phases in the whale's behaviour: repeated annual north-south migrations in sub-Arctic waters, and residency in subtropical waters. 
-Subtropical residency is associated with isotopic signatures consistent with pregnancy in the year before the whale's death. 
-These results supply the first quantitative multi-year records of historical migrations in individual blue whales, and suggest that blue whale residency and movement patterns in the North Atlantic have remained similar for hundreds of years, despite changing human pressures in the region.
+%We identify two distinct phases in the whale's behaviour: repeated annual north-south migrations in sub-Arctic waters, and residency in subtropical waters. 
+%Subtropical residency is associated with isotopic signatures consistent with pregnancy in the year before the whale's death. % make clear this is tenuous.
+% Juvenile behaviour?
+
 Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climate, ecological conditions and anthropogenic pressures, and reveal aspects of life history.
-Combining biochemical tracers with simulation modeling unlocks the information recorded in these irreplaceable archives.}
+Combining biochemical tracers with simulation modeling unlocks the information recorded in these irreplaceable archives.
 
-%\section{Main text}
 \newpage
-
+\section{Main text}
 Migratory species pose a particular challenge for conservation practitioners, because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}. 
-Such species are also likely to be particularly vulnerable to climate change, as changes may impact various parts of their life cycle\cite{robinson2009travelling}. 
+Such species may also be particularly affected by climate change, as changes can impact various parts of their life cycle\cite{robinson2009travelling}. 
 Dealing with these issues is difficult due to the relative scarcity of information on individual-level movements over multiple years, for both historical and present-day populations, especially for wide-ranging marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
@@
 Despite blue whale hunting being banned for over 50 years, blue whales are still listed as Endangered by the IUCN Red List\cite{reilly2008balaenoptera}.
-Conservation efforts are hampered by the scarcity of detailed data on the distribution, behaviour and ecology of blue whales, both for historical and present-day populations. 
+Conservation efforts are hampered by the scarcity of detailed data on the distribution, behaviour and ecology of blue whales, particularly for juveniles. 
 Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}.
 These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
-Stable isotope analyses provide an alternative technique for investigating blue whale movements\cite{west2006stable}. 
+They also cannot provide information on historical movement patterns of blue whales prior to overexploitation and modern use of the oceans, which could be important for understanding the drivers of their declines.
+An alternative technique for investigating blue whale movements in both present-day and historical populations is to use stable isotopes taken from individuals in the field or from museum specimens\cite{west2006stable,busquets2017estimating,hobson2008tracking}.
+Historic samples from museum collections form a rich archive of behavioural information \cite{lister2011natural} that is often under-utilised.
 We can use the isotopic compositions of whale tissues to reconstruct their movement patterns, by comparing them to spatially referenced models of isotopic variation across the species' range\cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
-%line 57,
 These methods can be applied to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale `Hope'.
 
-%Points here about movement patterns, interactions with human use of oceans, tourism etc
-% Build on lack of information about movements before ,modern use of Atlantic, and lack of information about juvenile behaviour.
-% Introduce historic samples as archive behavoiiral information
-
+Baleen stable isotopes have a long history of use as biochemical markers to infer movement patterns in whales\cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}.
 Baleen is formed of plates of keratin and is used by mysticete whales to filter feed. 
@@
 %REF???
-Baleen stable isotopes have a long history of use as biochemical markers to infer movement patterns in whales (REFS). % ADD REFS
-%Refs to support long history of inferring movement from baleen isotopes.
 
@@
 This resulted in XX discrete samples of baleen. % add number
-We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin. 
+We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin. ,
 These isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e.
 relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope\cite{west2006stable}.
-Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; isotopes provide information on the geographical locations of an individual, whereas nitrogen isotopes provide trophic and physiological information\cite{west2006stable}.
+Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; carbon isotopes provide information on the geographical locations of an individual, whereas nitrogen isotopes provide trophic and physiological information\cite{west2006stable}.
 
@@
 Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods).  
-In behavioural phase one (from record start in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Fig.
+In behavioural phase one (from the start of the record in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Fig.
 \ref{fig1}; Figs.
@@
 % figure 1
-\begin{figure}
-  \centering
-  \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
-  \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis).
+%\begin{figure}
+ % \centering
+  %\includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
+  %\caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis).
 Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes).
@@
 Approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
-  \label{fig1}
-\end{figure}
+  %\label{fig1}
+%\end{figure}
 
@@
 We assume continued ingestion of new protein throughout the year. 
-Other mysticete whales fast during migrations and during winter, relying on stored lipid reserves to meet metabolic demands, but blue whales are too large to stop feeding for long periods; \cite{busquets2017estimating,silva2013north}).
+Other mysticete whales fast during migrations and during winter, relying on stored lipid reserves to meet metabolic demands, but blue whales are too large to stop feeding for long periods; \cite{busquets2017estimating,silva2013north}.
 %Can we back this up? Are we more saying that we think they need protein to maintain tissue turnover but could meet energetic demands form stored lipids?
@@
 % figure 2
-\begin{figure}
-  \centering
-  \includegraphics[width = \linewidth]{figures/Figure-2-migratory-model-full-map.png}
-  \caption{Simulated tracks from the migratory movement model (2250 days for each of 50 replicates) with simulated day indicated by colour illustrating the residency period from Ireland to the Norwegian/Barents Sea followed by a migratory period and year spent near the Azores (as indicated by an arrow) and/or Canary Islands before returning north prior to stranding in Wexford, Ireland 1891 (as indicated by an arrow).
+%\begin{figure}
+ % \centering
+ % \includegraphics[width = \linewidth]{figures/Figure-2-migratory-model-full-map.png}
+  %\caption{Simulated tracks from the migratory movement model (2250 days for each of 50 replicates) with simulated day indicated by colour illustrating the residency period from Ireland to the Norwegian/Barents Sea followed by a migratory period and year spent near the Azores (as indicated by an arrow) and/or Canary Islands before returning north prior to stranding in Wexford, Ireland 1891 (as indicated by an arrow).
 }
-  \label{fig2}
-\end{figure}
+  %\label{fig2}
+%\end{figure}
 
 % figure 3
-\begin{figure}
-  \centering
-  \includegraphics[width = \linewidth]{figures/Figure-3-migratory-model-d13C.png}
-  \caption{Correlations among $\delta^{13}$C data from 50 movement simulations (green) and empirical data from baleen (blue).
+%\begin{figure}
+ % \centering
+  %\includegraphics[width = \linewidth]{figures/Figure-3-migratory-model-d13C.png}
+  %\caption{Correlations among $\delta^{13}$C data from 50 movement simulations (green) and empirical data from baleen (blue).
 Across the 50 simulations, the correlation coefficient between z-score transformations of empirical and time-matched simulation values had a mean of 0.48 ($sd = 0.2$), 49 of them were positive, and 39 of the 50 Pearson correlation tests had a p-value less than the Bonferroni corrected value of 0.001 ($p = 0.05/50 = 0.001$).
@@
 The standard deviation of the empirical data have been rescaled by a factor of three so as to better visualise the correlation because, as a consumer, the whale is expected to integrate and attenuate the fluctuations in its diet, and the end points of the simulations and empirical data have been aligned to coincide.}
-  \label{fig3}
-\end{figure}
+  %\label{fig3}
+%\end{figure}
 
@@
 Best fitting models predict juvenile residency in the Cape Verde region in behavioural phase 1. 
-Behavioural phase 2 is best simulated by seasonal migrations between summer foraging in northern areas in the Norwegian Sea/Barents Sea/Iceland region, and winter foraging in a broad region between the UK and sub-tropical waters.
+Behavioural phase 2 is best simulated by seasonal migrations between summer foraging in northern areas in the Norwegian Sea/Barents Sea/Iceland region, and winter foraging in a broad region between the UK and subtropical waters.
 Best fitting models in general predicted a greater latitudinal foraging range, and foraging in more northerly waters (Supplementary materials Fig. S7 [boxplots and kernel density plots]).
@@
 After six months the whale began feeding in low latitude waters producing a dramatic increase in $\delta^{13}$C values and began a final migration north. 
-Blue whales have a 10-12 month gestation period, with calving occurring in sub-tropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
+Blue whales have a 10-12 month gestation period, with calving occurring in subtropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
 The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, residency in warm waters through the whole of 1890 with feeding resuming in the second half of 1890 and eventual stranding during the return to northern feeding grounds in early 1891, is therefore consistent with the known reproductive ecology of blue whales.
+
+%%% STILL TO FIX BELOW %%%%
 

### 2017-11-15 manuscript/revision/blue-whale-ms-v2.tex
Minor tweaks

@@
 Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton \cite{handbook}.
-Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on around two weeks prior to that time (two weeks is required for the carbon and nitrogen to be assimilated and converted into tissues).
+Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on around three weeks prior to that time (this time is required for the carbon and nitrogen to be assimilated and converted into tissues).
 %REF???
@@
 {\raise.17ex\hbox{$\scriptstyle\sim$}}1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
-77% edit number
- samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section which contains the most recent tissue. 
+XX samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section which contains the most recent tissue. 
 Baleen grows at a constant rate, so the samples are equally spaced through time\cite{best1996stable}. 
@@
 Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates\cite{hobson1998stable,busquets2017estimating}.
-Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the most distal (oldest) 65cm of the plate. 
+Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70cm of the plate. 
 We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Fig.
@@
 Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891. 
-Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase one (oldest 60cm of the baleen plate), but no relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase two (Fig.
+Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase two (70cm to 18cm on the baleen plate), but no relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase one (the oldest 30cm of the baleen plate, i.e.
+100cm - 70cm; Fig.
 S2).
+% CHECK!!!
 
@@
 In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and extent (1$^{\circ}$ or 2$^{\circ}$ in km) of movement are all influenced by the following.
-(i) Behavioural state (migrating north, migrating south, foraging, or nursing). 
+(i) Behavioural state (migrating north, migrating south, foraging, or nursing). % this has changed
 Northerly migrations were possible only in spring, and southerly migrations in autumn. 
@@
 
-We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life} ; Fig S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
+We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life} ; Fig.
+S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
 For a full description of the model and its parameters see Supplementary Methods: Fig. S5 and Table S1. 

### 2017-11-15 quarantine/Figure-best-simulation.R
all old code still being modified

@@ -1,0 +1,273 @@
+rm(list = ls())
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Import and Process the empirical data
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+# Import Whale data
+whale_isos <- read.csv("data/KC.NHM.full.csv", header = TRUE, 
+                       stringsAsFactors = FALSE)
+
+KC7 <- filter(whale_isos, Whale == "KC7")
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# convert sample number to days using following parameters
+samp_resolution <- 1 # samples per centimetre
+
+growth_rate <- 13.5  # baleen growth rate in centimetres per year
+
+days <- KC7$Samp.No * samp_resolution * 365.25 / growth_rate
+
+days <- days - days[1]
+
+KC7$days <- days
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# convert to calendar date
+
+# reverse time
+rev_days <- days - days[length(days)]
+
+last_sample <- as.Date("1891-03-01")
+
+sample_date <- last_sample + rev_days
+
+KC7$rev_days <- rev_days
+KC7$date <- sample_date
+KC7$year <- year(sample_date)
+
+
+new_years <- with(KC7, seq(min(year), max(year), by = 1))
+
+new_years_chr <- paste(new_years, "01", "01", sep = "-")
+
+new_years_date <- as.Date(new_years_chr)
+
+new_years_day <- tail(KC7$days,1) - 
+  (as.numeric(last_sample - new_years_date ))
+
+new_years_samp <- round(new_years_day[2:length(new_years_day)] * growth_rate / 
+                          (samp_resolution * 365.25))
+
+new_years_all <- round(new_years_day * growth_rate / 
+                         (samp_resolution * 365.25))
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Import and process the simulation data
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# ----
+
+# which Rep will we plot
+do_rep = 43
+
+# Map plot of all simulated tracks for the full migratory model
+
+resTrack <- read.csv("data/record_migrate5.csv", header=TRUE)
+
+# remove NAs from the dataset
+# resTrack <- filter(resTrack, !is.na(d13C))
+
[…]

### 2017-11-15 quarantine/Figure2.R

@@ -1,0 +1,346 @@
+# Author: Clive Trueman
+# Adapted by: Andrew Jackson June 2017 & Natalie Cooper Oct 2017.
+# About: Script to generate figure 1
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Load libraries
+
+#library(raster)
+#library(gstat)
+#library(sp)
+##library(maps)
+#library(mapdata)
+#library(RColorBrewer)
+
+#library(tidyr)
+#library(dplyr)
+#library(ggplot2)
+#library(magrittr)
+#library(viridis)
+
+library(tidyverse)
+library(lubridate)
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Figure 2 - Map plot of all simulated tracks for the full migratory model
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+resTrack <- read_csv("data/Model.sims.csv")
+res <- resTrack
+
+# cut out last rep if it didn't work
+resTrack <- slice(res, -c(3213051:3213211))
+
+
+
+res <- filter(resTrack, Rep == 1)
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+x <- 
+  res %>%
+  group_by(Rep) %>%
+  summarise(max = max(count2)) %>%
+  #mutate(days = 1:max)
+  map(mutate, days = 1:max)
+
+remove <- x$Rep[x$max < 3019]
+
+resTrack <- slice(resTrack, -remove)
+
+
+
+# remove NAs from the dataset
+# resTrack <- filter(resTrack, !is.na(d13C))
+
+# create a new sequence of days on which to predict our
+# loess smoothed function and add it to the resTrack data.frame for 
+# each Rep
+days <- rep(seq(from = 1, to = 3019), length(resTrack$Rep))
+# resTrack$days 
+resTrack3 <- 
+  resTrack %>% 
+  group_by(Rep) %>% 
+  data.frame(days)
+
+
+# jitter the points
+resTrack$jLat <- jitter(resTrack$Lat, factor = 2)
+resTrack$jLon <- jitter(resTrack$Lon, factor = 1)
+
+# data required to add a vertical line indicating the start of migration
+migrate <- as.Date("1889-06-01")
+first_sim_day <- as.Date("1885-01-01")
+migrate_day <- as.numeric(migrate - first_sim_day)
+
[…]

### 2017-11-15 quarantine/Figure3.R

@@ -1,0 +1,226 @@
+# Author: Clive Trueman
+# Adapted by: Andrew Jackson June 2017 & Natalie Cooper Oct 2017.
+# About: Script to generate figure 1
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Load libraries
+
+#library(raster)
+#library(gstat)
+#library(sp)
+##library(maps)
+#library(mapdata)
+#library(RColorBrewer)
+
+#library(tidyr)
+#library(dplyr)
+#library(ggplot2)
+#library(magrittr)
+
+library(viridis)
+library(tidyverse)
+library(lubridate)
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Add raw whale isotope data
+# See figure 1 code for more detail
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# read in data and extract just the blue whale
+whale_isos <- read_csv("data/raw-whale-isotope-data.csv")
+KC7 <- filter(whale_isos, whale_ID == "KC7")
+growth_rate <- 13.5  # baleen growth rate in centimetres per year
+last_sample <- ymd("1891-03-01") # final date of last keratin sample
+
+KC7 <- 
+  KC7 %>%
+  mutate(days = sample_cm * 365.25 / growth_rate) %>%
+  mutate(days = days - days[1]) %>%
+  mutate(rev_days = days - days[length(days)]) %>%
+  mutate(sample_date = last_sample + rev_days) %>%
+  mutate(year = year(sample_date))
+
+# Create a variable to label the years
+year_labels <- with(KC7, seq(min(year), max(year), by = 1))
+year_breaks <- which(duplicated(KC7$year) == FALSE)
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# plot the isotope data
+point_size <- 4
+
+isop <- ggplot(KC7, aes(x = rev(sample_cm))) + 
+  theme_classic(base_size = 16) + 
+  xlab("Baleen sample (cm from youngest sample)") + 
+  scale_x_continuous(breaks = seq(0, 100, 20), 
+                     labels = seq(100, 0, -20),
+                     sec.axis = sec_axis(~., 
+                                         breaks = rev(year_breaks),
+                                         labels = rev(year_labels),
+                                         name = "Year"))
+
+isop <- isop + geom_line(aes(y = d13C)) + 
[…]

### 2017-11-15 quarantine/code-for-figures-ggplot2.R

@@ -1,0 +1,398 @@
+# Author: Clive Trueman
+# Adapted by: Andrew Jackson 19-Jun-2017
+# About: script to generate figures for the blue whale paper
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+# setwd("/Users/trueman/Desktop/R scripts/migrate.files")
+
+
+library(raster)
+library(gstat)
+library(sp)
+library(maps)
+library(mapdata)
+
+library(tidyr)
+library(dplyr)
+library(ggplot2)
+library(magrittr)
+library(viridis)
+library(lubridate)
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+rm(list = ls())
+graphics.off()
+
+
+##plot model output annual average baseline images
+Atl_carbon   <- raster("data/Atl_Annual_d13C.grd")
+Atl_nitrogen <- raster("data/Atl_Annual_d15N.grd")
+
+image(Atl_carbon)
+image(Atl_nitrogen)
+
+# set the constant to add as a correction to the d13C data
+cnst <- 6
+
+
+##########read resident whale model files
+
+resNor    <- read.csv("data/Norway.csv", header = TRUE)
+resIre    <- read.csv("data/west.Ireland.csv", header = TRUE)
+resCan    <- read.csv("data/Canaries.csv", header = TRUE)
+resMidAtl <- read.csv("data/Mid_Atl.csv", header = TRUE)
+
+
+
+
+#plot loess fits to the 30 resident simulated whales for each region
+
+
+
+# setEPS()
+# postscript("resident.eps", width=10, height=5)
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+#norway resident
+
+# use piping to break up the data.frame by replicate, then 
+# predict the loess
+tmp <- resNor %>% group_by(Rep) %>% 
+  do(., data.frame(Z = predict(loess((.$d13C + cnst) ~ .$Day.No, span = 0.1))))
+
+# append this vector onto the original data.frame
+resNor$lo <- tmp$Z
+  
+# plot via ggplot
+nor_plot <- ggplot(resNor, aes(Day.No, lo, group = Rep)) + 
+  geom_line(col = viridis(3)[2], alpha = 0.25) + 
+  xlab("Time") + 
+  ylab(expression(paste(delta^{13}, "C (\u2030)"))) + 
+  theme_classic(base_size = 14)
+print(nor_plot)
+
[…]

### 2017-11-15 quarantine/figures-1-2-3.R

@@ -1,0 +1,415 @@
+# Author: Clive Trueman
+# Adapted by: Andrew Jackson 19-Jun-2017
+# About: script to generate figure 1
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+# setwd("/Users/trueman/Desktop/R scripts/migrate.files")
+
+
+library(raster)
+library(gstat)
+library(sp)
+library(maps)
+library(mapdata)
+library(RColorBrewer)
+
+library(tidyr)
+library(dplyr)
+library(ggplot2)
+library(magrittr)
+library(viridis)
+library(lubridate)
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+rm(list = ls())
+graphics.off()
+
+
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Panel A - raw whale isotope data with two y-axes
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+
+whale_isos <- read.csv("data/KC.NHM.full.csv", header = TRUE, 
+                       stringsAsFactors = FALSE)
+
+KC7 <- filter(whale_isos, Whale == "KC7")
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# convert sample number to days using following parameters
+samp_resolution <- 1 # samples per centimetre
+
+growth_rate <- 13.5  # baleen growth rate in centimetres per year
+
+days <- KC7$Samp.No * samp_resolution * 365.25 / growth_rate
+
+days <- days - days[1]
+
+KC7$days <- days
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# convert to calendar date
+
+# reverse time
+rev_days <- days - days[length(days)]
+
+last_sample <- as.Date("1891-03-01")
+
+sample_date <- last_sample + rev_days
+
+KC7$rev_days <- rev_days
+KC7$date <- sample_date
+KC7$year <- year(sample_date)
+
+
+new_years <- with(KC7, seq(min(year), max(year), by = 1))
+
+new_years_chr <- paste(new_years, "01", "01", sep = "-")
+
+new_years_date <- as.Date(new_years_chr)
+
+new_years_day <- tail(KC7$days,1) - 
+  (as.numeric(last_sample - new_years_date ))
+
+new_years_samp <- round(new_years_day[2:length(new_years_day)] * growth_rate / 
[…]

### 2017-11-15 quarantine/figures-1-2-3_new.R

@@ -1,0 +1,362 @@
+# Author: Clive Trueman
+# Adapted by: Andrew Jackson 19-Jun-2017
+# And Natalie Cooper Oct 2017.
+# About: script to generate figure 1, 2 and 3
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Load libraries
+
+library(raster)
+library(gstat)
+library(sp)
+library(maps)
+library(mapdata)
+library(RColorBrewer)
+
+library(tidyr)
+library(dplyr)
+library(ggplot2)
+library(magrittr)
+library(viridis)
+library(lubridate)
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Figure 1 - raw whale isotope data with two y-axes
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+# read in data and extract just the blue whale
+whale_isos <- read.csv("data/raw-whale-isotope-data.csv", header = TRUE, 
+                       stringsAsFactors = FALSE)
+
+KC7 <- filter(whale_isos, whale_ID == "KC7")
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# convert year and day number to date 
+KC7$date <- ymd(paste0(KC7$year, "-01-01")) + KC7$day
+
+# Create a variable to label the years
+year_labels <- with(KC7, seq(min(year), max(year), by = 1))
+year_breaks <- which(duplicated(KC7$year) == FALSE)
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+# plot the isotope data
+point_size <- 4
+
+isop <- ggplot(KC7, aes(x = rev(sample_cm))) + 
+  theme_classic(base_size = 16) + 
+  xlab("Baleen sample (cm from youngest sample)") + 
+  scale_x_continuous(breaks = seq(0, 100, 20), 
+                     labels = seq(100, 0, -20),
+                     sec.axis = sec_axis(~., 
+                                         breaks = rev(year_breaks),
+                                         labels = rev(year_labels),
+                                         name = "Year"))
+
+isop <- isop + geom_line(aes(y = d13C)) + 
+  geom_point(aes(y = d13C), size = point_size, shape = 21, fill = "black") +
+  ylab(expression(paste(delta^{13}, "C (\u2030)")))
+
+# add the d15N data but need to scale it to d13C
[…]

### 2017-11-15 code/Figure-1-code.R
Updated versions of code for revision

@@ -1,0 +1,71 @@
+# Author: Clive Trueman
+# Adapted by: Andrew Jackson June 2017 & Natalie Cooper Oct 2017.
+# About: Script to generate figure 1
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Load libraries
+library(tidyverse)
+library(lubridate)
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Figure 1 - raw whale isotope data with two y-axes
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# read in data and extract just the blue whale
+whale_isos <- read_csv("data/raw-whale-isotope-data.csv")
+
+KC7 <- filter(whale_isos, whale_ID == "KC7")
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# convert sample number to days based on growth rate of baleen
+growth_rate <- 13.5  # baleen growth rate in centimetres per year
+last_sample <- ymd("1891-03-01") # final date of last keratin sample
+
+KC7 <- 
+  KC7 %>%
+  mutate(days = sample_cm * 365.25 / growth_rate) %>%
+  mutate(days = days - days[1]) %>%
+  mutate(rev_days = days - days[length(days)]) %>%
+  mutate(sample_date = last_sample + rev_days) %>%
+  mutate(year = year(sample_date))
+
+# Create a variable to label the years
+year_labels <- with(KC7, seq(min(year), max(year), by = 1))
+year_breaks <- which(duplicated(KC7$year) == FALSE)
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# plot the isotope data
+point_size <- 4
+
+isop <- ggplot(KC7, aes(x = rev(sample_cm))) + 
+  theme_classic(base_size = 16) + 
+  xlab("Baleen sample (cm from youngest sample)") + 
+  scale_x_continuous(breaks = seq(0, 100, 20), 
+                     labels = seq(100, 0, -20),
+                     sec.axis = sec_axis(~., 
+                                         breaks = rev(year_breaks),
+                                         labels = rev(year_labels),
+                                         name = "Year"))
+
+isop <- isop + geom_line(aes(y = d13C)) + 
+  geom_point(aes(y = d13C), size = point_size, shape = 21, fill = "black") +
[…]

### 2017-11-15 code/Figure-S1-S2-code.R

@@ -1,0 +1,94 @@
+# Author: Clive Trueman
+# Adapted by: Andrew Jackson 19-Jun-2017
+# About: script to generate figures S1 & S2
+# Modified by Natalie Cooper Oct 2017
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+library(tidyverse)
+library(TSA)
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Read in raw whale isotope data
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+whale_isos <- read_csv("data/raw-whale-isotope-data.csv")
+
+KC7 <- filter(whale_isos, whale_ID == "KC7")
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# function to add text for the highest frequency spectral 
+# densities to periodogram plots
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+periodogramText <- function(p, k){
+  
+  # extract frequencies and spectral densities from the object
+  dd <- data.frame(freq = p$freq, spec = p$spec)
+  
+  # sort them into rank order
+  ord <- dd[order(-dd$spec), ]
+  
+  # extract the top two densities
+  top2  <-  head(ord, 2)
+  
+  # convert to distance along our sampling transect
+  distance <- 1/top2$freq
+  
+  # add either the first or second
+  text(0.4, max(p$spec) * 0.8, 
+       paste0(sprintf("%.1f", round(distance[k], 1))," cm"), 
+       cex = 2)
+}
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Figure S1 - periodograms
+# phase 1 = 100 - 70cm
+# phase 2 = 70 - 18cm
+# phase 3 = 18 - 1cm
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+png(filename = "manuscript/revision/figures/Figure-S1-periodograms.png", 
+    width = 1000, height = 600)
+
+par(mfrow = c(2, 2))
+
+p1 <- periodogram(KC7$d15N, main = "A - Full d15N")
+periodogramText(p1, k = 2)
+
[…]

### 2017-11-15 code/Figure-S3-code.R

@@ -1,0 +1,71 @@
+# Author: Clive Trueman
+# Adapted by: Andrew Jackson 19-Jun-2017
+# About: script to generate figure S3
+# Modified by Natalie Cooper Oct 2017
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Load libraries
+library(tidyverse)
+library(raster)
+library(RColorBrewer)
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# d13Carbon
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+# Read in data
+carbon <- raster("data/Atl_Annual_d13C.grd")
+
+# Extract values
+val <- getValues(carbon)
+
+# Create dataframe of values per cell
+xy <- as.data.frame(xyFromCell(carbon, 1:ncell(carbon)))
+xy$val <- val
+
+# Plot map of values
+p <- 
+  ggplot(na.omit(xy), aes(x = x, y = y, fill = val)) + 
+    geom_raster() + 
+    borders("world", colour = "gray50", fill = "gray50")  +
+    coord_fixed(xlim = c(-80, 50),  
+                ylim = c(0, 80), 
+                ratio = 1.3) + 
+    scale_fill_distiller(type = "seq", palette = "YlOrRd", direction = -1,
+                         name = expression(paste(delta^{13}, "C (\u2030)"))) + 
+    theme_classic(base_size = 14) + 
+    theme(axis.line  = element_blank(),
+          axis.text  = element_blank(),
+          axis.ticks = element_blank(),
+          axis.title = element_blank())
+
+ggsave("manuscript/revision/figures/Figure-S3-plankton-d13C-map.png", p, 
+       device = png(width = 600, height = 400))
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# d15Nitrogen
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+nitrogen <- raster("data/Atl_Annual_d15N.grd")
+
+val2 <- getValues(nitrogen)
+xy2 <- as.data.frame(xyFromCell(nitrogen, 1:ncell(nitrogen)))
+xy2$val2 <- val2
+
+p2 <- 
+  ggplot(na.omit(xy2), aes(x = x, y = y, fill = val2)) + 
+    geom_raster() + 
+    borders("world", colour = "gray50", fill = "gray50")  +
+    coord_fixed(xlim = c(-80, 50),  
[…]

### 2017-11-15 code/Figure-S4-code.R

@@ -1,0 +1,58 @@
+# Author: Clive Trueman
+# Adapted by: Andrew Jackson 19-Jun-2017
+# About: script to generate figures S4a-d for the blue whale paper
+# Tidied by Natalie Cooper Nov 2017
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+library(tidyverse)
+library(viridis)
+library(lubridate)
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# set the constant to add as a correction to the d13C data
+cnst <- 6
+
+# Read in resident whale model data
+resNor <- read.csv("data/Norway.csv", header = TRUE)
+resIre <- read.csv("data/west.Ireland.csv", header = TRUE)
+resCan <- read.csv("data/Canaries.csv", header = TRUE)
+resMidAtl <- read.csv("data/Mid_Atl.csv", header = TRUE)
+
+# Combine datasets
+res_all <- bind_rows(list(Norway = resNor, 
+                          Ireland = resIre, 
+                          Canaries = resCan, 
+                          "Mid-Atlantic" = resMidAtl), .id = "Region")
+
+# Remove lines with no d13C values
+res_all <- res_all[which(!is.na(res_all$d13C)), ]
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Plot loess fits to the 30 resident simulated whales for 
+# each region
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+# Group data by replicate and region
+# And add Day.No variable
+tmp2 <- res_all %>% group_by(Region, Rep)  %>% 
+  do(., data.frame(Z = predict(loess((.$d13C + cnst) ~ .$Day.No, span = 0.1)),
+                   Day.No = .$Day.No))
+
+# re-order the Region Factor
+tmp2$Region <- factor(tmp2$Region, levels = c("Ireland","Norway",
+                                              "Canaries","Mid-Atlantic"))
+
+# Plot and facet wrap by region
+sim_facet <- 
+  ggplot(tmp2, aes(Day.No, Z, group = Rep)) + 
+    geom_line(col = viridis(3)[2], alpha = 0.25) + 
+    xlab("Time") + 
+    ylab(expression(paste(delta^{13}, "C (\u2030)"))) + 
+    theme_classic(base_size = 14) + ylim(-25, -14) + 
+    theme(strip.background = element_blank(),
[…]

### 2017-11-15 manuscript/revision/blue-whale-ms-v2.tex
small edits related to Fig S2

@@
 Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates\cite{hobson1998stable,busquets2017estimating}.
-Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70cm of the plate. 
+Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
 We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Fig.
@@
 Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891. 
-Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase two (70cm to 18cm on the baleen plate), but no relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase one (the oldest 30cm of the baleen plate, i.e.
+Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase two (70cm to 18cm on the baleen plate), but no clear relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase one (the oldest 30cm of the baleen plate, i.e.
 100cm - 70cm; Fig.
+S2) or the final 500 days of the whale's life (youngest 18cm of the baleen plate; Fig.
 S2).
-% CHECK!!!
 

### 2017-11-15 Blue baleen plot.R
Tidy repo

@@ -1,58 +1,0 @@
-library(TSA)
-
-# setwd("/Users/trueman/Desktop/R scripts/Whale")
-data<-read.csv("data/KC.NHM.full.csv", header=TRUE)
-par(mfrow=c(1,1))
-colz<-colorRampPalette(c("brown", "cadet blue", "pink"))(12)
-colz<-seq(1:6)
-fin<-data[data$Spp=="fin",]
-minke<-data[data$Spp=="minke",]
-sei<-data[data$Spp=="sei",]
-blue<-data[data$Spp=="blue",]
-
-
-#BLUE
-#setEPS()
-#postscript("blue.eps")
-par(mfrow=c(1,1))
-plot(x=rev(blue$Samp.No[blue$Whale=="KC7"]), y=blue$d13C[blue$Whale=="KC7"], 
-     type="o", xlim=c(80,0), ylim=c(-20, -16.5), xlab="increment no", 
-     ylab=expression(paste(delta^{13},"C")), pch=21, cex=1.5, lwd=0.8, bg=colz[1])
-text(5,-20, "most recent")
-text(80,-20, "oldest")
-text(5, -16.75, "Whale KC1")
-
-par(new=TRUE)
-plot(x=rev(blue$Samp.No[blue$Whale=="KC7"]), y=blue$d15N[blue$Whale=="KC7"], 
-     type="o", xlim=c(80,0), ylim=c(5, 13),  axes=F, xlab="", ylab="", 
-     pch=21, cex=1.5, lwd=0.8, bg="grey")
-axis(side=4)
-mtext(side=4, line=3, expression(paste(delta^{15},"N")))
-#dev.off()
-#par(new=FALSE)
-
-
-
-
-
-## fourier analysis of blue whale - frequency is in cm of baleen
-## fourier for phase 1
-blueP1<-blue[blue$Samp.No<(80-25),]
-
-par(mfrow=c(1,1))
-p=periodogram(blueP1$d15N)
-#plot(p)
-dd<-data.frame(freq=p$freq, spec=p$spec)
-order=dd[order(-dd$spec),]
-top2 = head(order, 2)
-top2
-distance=1/top2$f
-
-text(0.4, 4, paste(distance[1],"cm"))
-#text(0.4, 3.5, paste(distance[2], "cm")
-
-
-## autocorrelation analysis of phase 1 d15N, and cross-correlation of phase 1 d15N and d13C values
-
[…]

### 2017-11-15 Figure-S1-S2-code.R

@@ -1,105 +1,0 @@
-# Author: Clive Trueman
-# Adapted by: Andrew Jackson 19-Jun-2017
-# About: script to generate figures S1 & S2
-# Modified by Natalie Cooper Oct 2017
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-library(raster)
-library(gstat)
-library(sp)
-library(maps)
-library(mapdata)
-
-library(tidyr)
-library(dplyr)
-library(ggplot2)
-library(magrittr)
-library(viridis)
-library(lubridate)
-
-library(TSA)
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Read in raw whale isotope data
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-whale_isos <- read.csv("data/raw-whale-isotope-data.csv", header = TRUE, 
-                       stringsAsFactors = FALSE)
-
-KC7 <- filter(whale_isos, whale_ID == "KC7")
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# function to add text for the highest frequency spectral 
-# densities to periodogram plots
-
-periodogramText <- function(p, k){
-  
-  # extract frequencies and spectral densities from the object
-  dd <- data.frame(freq = p$freq, spec = p$spec)
-  
-  # sort them into rank order
-  ord <- dd[order(-dd$spec), ]
-  
-  # extract the top two densities
-  top2  <-  head(ord, 2)
-  
-  # convert to distance along our sampling transect
-  distance <- 1/top2$freq
-  
-  # add either the first or second
-  text(0.4, max(p$spec) * 0.8, 
-       paste0(sprintf("%.1f", round(distance[k], 1))," cm"), 
-       cex = 2)
-}
-
-png(filename = "manuscript/figures/Figure-S1-periodograms.png", 
-    width = 1000, height = 600)
-
-par(mfrow = c(2, 2))
-
-p1 <- periodogram(KC7$d15N, main = "A - Full d15N")
-periodogramText(p1, k = 2)
-
-p2 <- periodogram(KC7$d15N[1:60], main = "B - Oldest 60cm d15N")
-periodogramText(p2, k = 1)
-
[…]

### 2017-11-15 Figure-S3-C-N-isoscape.R

@@ -1,78 +1,0 @@
-
-
-library(raster)
-library(RColorBrewer)
-
-
-my_clrs <- brewer.pal(8, "YlOrRd")
-
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# d13Carbon
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-Atl_carbon   <- raster("data/Atl_Annual_d13C.grd")
-# raster::image(Atl_carbon, xlab = "Long", ylab = "Lat", col = rev(my_clrs))
-
-
-r <- Atl_carbon
-# r <- reclassify(r, c(0,500,1, 500,2000,2))
-val <- getValues(r)
-xy <- as.data.frame(xyFromCell(r,1:ncell(r)))
-xy$val <- val
-
-p <- ggplot(na.omit(xy), aes(x=x, y=y, fill=val)) + 
-  geom_raster() + 
-  borders("world", colour="gray50", fill="gray50")  +
-  coord_fixed(xlim = c(-80, 50),  
-              ylim = c(0, 80), 
-              ratio = 1.3) + 
-  scale_fill_distiller(type = "seq", palette = "YlOrRd", direction = -1,
-                       name = expression(paste(delta^{13}, "C (\u2030)"))) + 
-  theme_classic(base_size = 14) + 
-  theme(axis.line  = element_blank(),
-        axis.text  = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title = element_blank())
-
-print(p)
-
-
-ggsave("manuscript/figures/Figure-S3-plankton-d13C-map.png", p, 
-       device = png(width = 600, height = 400))
-
-
-
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# d15Nitrogen
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-Atl_nitrogen <- raster("data/Atl_Annual_d15N.grd")
-# raster::image(Atl_nitrogen, xlab = "Long", ylab = "Lat", col = rev(my_clrs))
-
-r2 <- Atl_nitrogen
-# r <- reclassify(r, c(0,500,1, 500,2000,2))
-val2 <- getValues(r2)
-xy2 <- as.data.frame(xyFromCell(r2,1:ncell(r2)))
-xy2$val <- val2
-
-p2 <- ggplot(na.omit(xy2), aes(x=x, y=y, fill=val)) + 
[…]

### 2017-11-15 Figure-S4-alternative-movement-d13C.R

@@ -1,241 +1,0 @@
-# Author: Clive Trueman
-# Adapted by: Andrew Jackson 19-Jun-2017
-# About: script to generate figures S4a-d for the blue whale paper
-
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-# setwd("/Users/trueman/Desktop/R scripts/migrate.files")
-
-
-library(raster)
-library(gstat)
-library(sp)
-library(maps)
-library(mapdata)
-
-library(tidyr)
-library(dplyr)
-library(ggplot2)
-library(magrittr)
-library(viridis)
-library(lubridate)
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-rm(list = ls())
-graphics.off()
-
-# set the constant to add as a correction to the d13C data
-cnst <- 6
-
-
-##########read resident whale model files
-
-resNor    <- read.csv("data/Norway.csv", header = TRUE)
-resIre    <- read.csv("data/west.Ireland.csv", header = TRUE)
-resCan    <- read.csv("data/Canaries.csv", header = TRUE)
-resMidAtl <- read.csv("data/Mid_Atl.csv", header = TRUE)
-
-
-
-
-#plot loess fits to the 30 resident simulated whales for each region
-
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-#norway resident
-
-# use piping to break up the data.frame by replicate, then 
-# predict the loess
-tmp <- resNor %>% group_by(Rep) %>% 
-  do(., data.frame(Z = predict(loess((.$d13C + cnst) ~ .$Day.No, span = 0.1))))
-
-# append this vector onto the original data.frame
-resNor$lo <- tmp$Z
-
-# plot via ggplot
-nor_plot <- ggplot(resNor, aes(Day.No, lo, group = Rep)) + 
-  geom_line(col = viridis(3)[2], alpha = 0.25) + 
-  xlab("Time") + 
-  ylab(expression(paste(delta^{13}, "C (\u2030)"))) + 
-  theme_classic(base_size = 14) + ylim(-25, -14)
-
-print(nor_plot)
-
-# ggsave("figures/Figure-S4a-norway-resident-model-d13C.png", nor_plot, 
-#        device = png(width = 600, height = 400))
-
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-#Ireland resident
-
-# same as for norway example above
[…]

### 2017-11-15 figures-1-2-3_new.R

@@ -1,362 +1,0 @@
-# Author: Clive Trueman
-# Adapted by: Andrew Jackson 19-Jun-2017
-# And Natalie Cooper Oct 2017.
-# About: script to generate figure 1, 2 and 3
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Load libraries
-
-library(raster)
-library(gstat)
-library(sp)
-library(maps)
-library(mapdata)
-library(RColorBrewer)
-
-library(tidyr)
-library(dplyr)
-library(ggplot2)
-library(magrittr)
-library(viridis)
-library(lubridate)
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Figure 1 - raw whale isotope data with two y-axes
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-# read in data and extract just the blue whale
-whale_isos <- read.csv("data/raw-whale-isotope-data.csv", header = TRUE, 
-                       stringsAsFactors = FALSE)
-
-KC7 <- filter(whale_isos, whale_ID == "KC7")
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# convert year and day number to date 
-KC7$date <- ymd(paste0(KC7$year, "-01-01")) + KC7$day
-
-# Create a variable to label the years
-year_labels <- with(KC7, seq(min(year), max(year), by = 1))
-year_breaks <- which(duplicated(KC7$year) == FALSE)
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-# plot the isotope data
-point_size <- 4
-
-isop <- ggplot(KC7, aes(x = rev(sample_cm))) + 
-  theme_classic(base_size = 16) + 
-  xlab("Baleen sample (cm from youngest sample)") + 
-  scale_x_continuous(breaks = seq(0, 100, 20), 
-                     labels = seq(100, 0, -20),
-                     sec.axis = sec_axis(~., 
-                                         breaks = rev(year_breaks),
-                                         labels = rev(year_labels),
-                                         name = "Year"))
-
-isop <- isop + geom_line(aes(y = d13C)) + 
-  geom_point(aes(y = d13C), size = point_size, shape = 21, fill = "black") +
-  ylab(expression(paste(delta^{13}, "C (\u2030)")))
-
-# add the d15N data but need to scale it to d13C
[…]

### 2017-11-15 manuscript/revision/blue-whale-ms-v2.tex
Few edits to figure legend

@@
 
-% Edit figure legend please
-
 % figure 1
-%\begin{figure}
- % \centering
-  %\includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
-  %\caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis).
+\begin{figure}
+  \centering
+  \includegraphics[width = \linewidth]{revision/figures/Figure-1-raw-dC-dN-data.png}
+  \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis).
 Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes).
 There is strong annual periodicity and cross-correlation (Figs.
-S1 and S2) in both isotopes as well as a change in the mean and magnitude of the cycles during 1889 (approximately at sample 30cm).
-Approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
-  %\label{fig1}
-%\end{figure}
-
-% Krill has a very similar isotopic composition to phytoplankton
+S1 and S2) in both isotopes.
+The approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
+  \label{fig1}
+\end{figure}
 
@@
 
-%line 112 - 122 para
 The last 500 days of the whale's life are difficult to simulate. 
@@
 Conventional wisdom suggests that baleen whales migrate each year from feeding grounds near the poles in summer, to overwintering and breeding grounds nearer the equator\cite{corkeron1999baleen,lockyer1981migration}.
-However, a growing body of evidence suggests that not all whales migrate to subtropical waters each year, with some remaining at high latitudes year round\cite{mcdonald2006biogeographic}. %line 127
+However, a growing body of evidence suggests that not all whales migrate each year, with some remaining at high or low latitudes year round\cite{mcdonald2006biogeographic,busquets2017estimating}. 
 Our results support this conclusion. 

### 2017-11-16 manuscript/revision/blue-whale-ms-v2.tex
Edit intro paragraphs

@@
 These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
-They also cannot provide information on historical movement patterns of blue whales prior to overexploitation and modern use of the oceans, which could be important for understanding the drivers of their declines.
-An alternative technique for investigating blue whale movements in both present-day and historical populations is to use stable isotopes taken from individuals in the field or from museum specimens\cite{west2006stable,busquets2017estimating,hobson2008tracking}.
-Historic samples from museum collections form a rich archive of behavioural information \cite{lister2011natural} that is often under-utilised.
+Additionally, they do not provide information on historical movement patterns that could be important for understanding the drivers of whale declines.
+An alternative technique for investigating blue whale movements is to use stable isotopes\cite{west2006stable,busquets2017estimating,hobson2008tracking}. 
+Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioural information that are often under-utilised \cite{lister2011natural}.
 We can use the isotopic compositions of whale tissues to reconstruct their movement patterns, by comparing them to spatially referenced models of isotopic variation across the species' range\cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
-These methods can be applied to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale `Hope'.
+These methods can be applied fairly inexpensively to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale `Hope'.
 
@@
 %REF???
+% add line about it being ok to use phytoplankton?
 
@@
 This resulted in XX discrete samples of baleen. % add number
-We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin. ,
+We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin.
 These isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e.
@@
   \centering
-  \includegraphics[width = \linewidth]{revision/figures/Figure-1-raw-dC-dN-data.png}
+  \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
   \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis).

### 2017-11-16 manuscript/revision/blue-whale-ms-v2.tex
Minor edits up to Discussion

@@
 We assume continued ingestion of new protein throughout the year. 
-Other mysticete whales fast during migrations and during winter, relying on stored lipid reserves to meet metabolic demands, but blue whales are too large to stop feeding for long periods; \cite{busquets2017estimating,silva2013north}.
+Other mysticete whales fast during migrations and during winter, relying on stored lipid reserves to meet metabolic demands, but blue whales are too large to stop feeding for long periods\cite{busquets2017estimating,silva2013north}.
 %Can we back this up? Are we more saying that we think they need protein to maintain tissue turnover but could meet energetic demands form stored lipids?
@@
 The whale remained in these warm waters for at least two full years. 
+The whale was estimated to be 15 years old when it died (based on vertebral epiphyseal fusion; R.Sabin \textit{pers.comm}), so was approximately 3-5 years old during this period.
 During behavioural phase two, starting at an approximate age of 5-6 years, the observed low $\delta^{13}$C values imply that the whale is foraging in colder, more northerly latitudes. 

### 2017-11-16 manuscript/revision/response.md
Complete first draft response and edits

@@
 
-**Important note to the referees**.
+## Important note to the referees 
 During the revisions of this manuscript, which included substantial extra analysis, we discovered an error in our original results.
@@
 This note is to alert the referees that the results graphs will look different to the original version, though our conclusions have not changed qualitatively.
+
+## New analyses
+As well as responding to the referee comments, and substantially rewriting sections of the paper, this new version contains several new analyses. In brief:
+
+1. We took further keratin samples from the fronds of the baleen, this has increased our number of samples from 78 to XX, and the time series from six to seven full years of records.
+2.
+We have increased the number of simulations on our models from 100 to 1000, to better explore parameter space.
+We provide X new figures in the supplementary information exploring these models more thoroughly.
 
@@
 I suppose it depends on whether the model was used for a direct pattern comparison, or if the values modelled and the baleen values were compared to each other.
-
 > The referee is correct that blue whales feed predominantly on krill (small crustaceans), not directly on phytoplankton.
@@
 relative to a standard and as ratios of the rare (13C, 15N) to common (12C, 14N) isotope"
-* We have added the McMahon reference at line XX [ADD]
+* We have added the McMahon reference at line XX
 
@@
 
-> Alright
+> We have added this reference.
 
@@
 
-> The referee is correct in that
+> The referee is correct in that ???
 
@@
 
->
+> We have removed this comment from the text.
 
@@
 
->
+> Not per se, though we see changes in pre and post whaling distributions, as seen in Reeves et al 2004. We have modified this statement.
 
@@
 
-> We have rephrased this.
+> We have rephrased this to make it clear we are highlighting something others have also highlighted in the past.
 

### 2017-11-16 manuscript/revision/blue-whale-ms-v2.tex
Complete first draft response and edits

@@
 
-Conventional wisdom suggests that baleen whales migrate each year from feeding grounds near the poles in summer, to overwintering and breeding grounds nearer the equator\cite{corkeron1999baleen,lockyer1981migration}.
+Data on the movement patterns of blue whales, particularly of juveniles, are scarce. 
+Although based on a single individual, our results highlight some interesting patterns worthy of further study.
+Firstly, next to nothing is known about juvenile behaviour in blue whales - calves become independent of their mothers at around seven months, and become sexually mature around 8-10 years old \cite{handbook}.
+Our analyses suggest that between these points, juvenile North Atlantic blue whales may spend extended periods in warm waters before migrating to higher latitude feeding grounds.
+This has implications for blue whale conservation, as these warm waters around the Azores and Cape Verde are also tourist and shipping hotspots, increasing the likelihood of ship strikes and anthropogenic effects on the juvenile whales.
+Secondly, conventional wisdom suggests that baleen whales migrate each year from feeding grounds near the poles in summer, to overwintering and breeding grounds nearer the equator\cite{corkeron1999baleen,lockyer1981migration}.
 However, a growing body of evidence suggests that not all whales migrate each year, with some remaining at high or low latitudes year round\cite{mcdonald2006biogeographic,busquets2017estimating}. 
-Our results support this conclusion. 
-Interestingly, there is still a resident%line 128
- population of blue whales in the regions indicated by our model, and females still breed in the subtropical waters near the Azores\cite{reilly2008balaenoptera}. 
+Our results support this conclusion, with the whale remaining at relatively high latitudes for several years, before migrating south, presumably to calve.
+
 Whaling was an intense pressure for blue whales during the period we are analysing. 
@@
 The NHM blue whale thus represents an individual from a species at the brink of extinction.
-Our data suggest that the spatial ecology of blue whales in the North Atlantic has remained similar for more than 100 years, despite dramatic changes in anthropogenic pressures. % line 137 and 139
+Despite dramatic changes in anthropogenic pressures since 1891, the NHM whale and present-day blue whales in the North Atlantic have similar spatial ecologies; there are still populations of blue whales in the regions indicated by our model, and females still breed in the subtropical waters near the Azores\cite{reilly2008balaenoptera}.
+This may reflect behavioural responses to whaling, as blue whales were more widely distributed in the past\cite{reeves2004historical}, or the selective extinction of populations outside of these areas.
 
-This is the first quantitative evidence of historical individual-level movement behaviours in blue whales. %line 140-141
-Our novel simulation modeling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviours over multiple years.
+Our movement simulation modeling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviours over multiple years.
 Such techniques can also be applied to extinct species behaviours and used to understand how populations were affected by past pressures such as hunting, and present-day pressures such as global change.
-Finally, our results highlight that museums represent a vast, mostly untapped, source of individual-level spatial ecology information. %line 147-149
+Our results also add to previous research\cite{lister2011natural,ryan2013stable} highlighting museums as vast sources of individual-level spatial ecology information.
 

### 2018-01-03 code/Extract-models.R
Add code to extract only midpoint of tracks

@@ -98,12 +98,13 @@ resTrack$phase[resTrack$count2 < 1000] <- 1
 resTrack$phase[resTrack$count2 >= 1000 & resTrack$count2 < 2500] <- 2
 resTrack$phase[resTrack$count2 > 2500] <- 3
 
+# extract top 10% and bottom 10% best fit models
 topX <- ddply(resTrack, "Rep", function(x) {
-  newSeries <- x[x$Rep%in%unique(resTrack$Rep)[order(r2)[limit+1:length(r2)]], ]
+  newSeries <- x[x$Rep%in%unique(resTrack$Rep)[order(r2)[limit + 1:length(r2)]], ]
 })
 
 bottomY <- ddply(resTrack, "Rep", function(x) {
-  newSeries <- x[x$Rep%in%unique(resTrack$Rep)[order(r2)[1:limitB]],]
+  newSeries <- x[x$Rep%in%unique(resTrack$Rep)[order(r2)[1:limitB]], ]
 })
 
 # Write to file
@@ -149,3 +150,67 @@ write.csv(file = "data/max.lat.csv", max.lat.ds, quote = FALSE)
 write.csv(file = "data/sd.lat.csv", sd.lat.ds, quote = FALSE)
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# extract top 10% and bottom 10% best fit models for 
+# phase 2 only (day 1000 - day 2499)
+# write out for figures 
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+# Extract only phase 2 days from simulations 
+mid <- ddply(resTrack, "Rep", function(x) {
+  newSeries <- x[x$count2%in%seq(from = 1000, to = 2499, by = 1), ]
+})
+
+# Get sample days
+mo_no.mid <- 1500/30 - 1
+sampleD.mid <- rnorm(mo_no.mid, 30, 1)
+sampledays.mid <- as.integer(cumsum(sampleD.mid)) + 1000 - 30
+
+# Extract only values for days of sample
+mid.test <- ddply(mid, "Rep", function(x) {
+  newSeries <- x[x$count2%in%sampledays.mid, ]
+})
+
+# Run loess through series and predict for same days as measured
+mid.test3 <- ddply(mid.test, "Rep", function(x) {
+  lo <- predict(loess(x$d13C ~ x$count2, span = 0.2))[1:48]
+})
+
+# Transpose
[…]

### 2018-01-03 code/Figure-2-code.R
tidy code

@@ -13,7 +13,7 @@ library(viridis)
 top10 <- read.csv("data/top100.csv")
 
 # Pick colours
-mycols <- c(viridis_pal()(10)[1],viridis_pal()(10)[5],viridis_pal()(10)[9])
+mycols <- c(viridis_pal()(10)[1], viridis_pal()(10)[5], viridis_pal()(10)[9])
 
 # Jitter points of lat and long to get smooth density plot
 jLat <- jitter(top10$Lat, factor = 2)
@@ -23,7 +23,7 @@ jLon <- jitter(top10$Lon, factor = 1)
 # Points plot with phases coloured
 png("manuscript/revision/figures/Figure-2-points.png", width = 800, height = 600)
 
-# Top 10 %
+# Create empty backgrounds
 plot(x = NA, y = NA, xlim = c(-80, 50), ylim = c(0, 80), 
      xlab = "", ylab = "", axes = FALSE)
 rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], 

### 2018-01-03 code/Extract-models.R
Minor edits

@@ -184,15 +184,17 @@ mid.test3 <- as.data.frame(mid.test3[-1, ])
 
 # Link to the measured data to allow regression comparison
 mid.test3$Day <- mid.test$count2[mid.test$Rep == 1][1:48]
-mid.test3$Blue<-rev(blue$d13C[blue$Day.sim < 2500 & blue$Day.sim >= 1000])
+mid.test3$Blue <-rev(blue$d13C[blue$Day.sim < 2500 & blue$Day.sim >= 1000])
 
 # Remove runs with missing data
-mid.test4<-mid.test3[, -which(colMeans(is.na(mid.test3)) > 0.5)]
+# In my 1000 models there are no missing data
+#mid.test4 <- mid.test3[, -which(colMeans(is.na(mid.test3)) > 0.5)]
+mid.test4 <- mid.test3
 
 # Run regressions for each model simulation and save r square values
 mid.r2 <- vector()
 
-for(run in 3:dim(mid.test4)[2] - 2){
+for(run in 3:length(unique(mid$Rep))){
   mid.r2[run]<-summary(lm(mid.test4[, run] ~ mid.test4$Blue))$adj.r.squared
 }
 

### 2018-01-03 code/Figure-4-code.R
Code for Figure 4 monthly plots of phase 2

@@ -1,0 +1,60 @@
+# Author: Clive Trueman
+# About: script to plot track for top 10% of models by month
+# for behavioural phase 2.
+# Tidied by Natalie Cooper Jan 2018
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Load libraries
+library(maps)
+library(mapdata)
+library(spatstat)
+library(viridis)
+
+# Read in data
+mid.top10 <- read.csv("data/mid.top100.csv")
+
+# Pick colours
+mycols <- c(viridis_pal()(10)[1], viridis_pal()(10)[5], viridis_pal()(10)[9])
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Points plot with phases coloured
+png("manuscript/revision/figures/Figure-4-monthly.png", width = 800, height = 600)
+
+# Make 12 plots
+par(mfrow = c(2, 6))
+par(mar = c(1, 1, 2, 1))
+
+# Loop through months  
+for (Mn in 1:12) {
+  
+  # Make background of plots
+  plot(x = NA, y = NA, xlim = c(-80, 50), ylim = c(0, 80), 
+       xlab = "", ylab = "", axes = FALSE, main = month.abb[Mn])
+  rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], 
+       col = "grey90")
+  
+  # Select month
+  Month <- mid.top10[mid.top10$Month == Mn, ]
+  
+  # Jitter points of lat and long to get smooth density plot
+  jLat <- jitter(Month$Lat, factor = 2)
+  jLon <- jitter(Month$Lon, factor = 1)
+  
+  # Add points coloured by phase
+  points(jLon, jLat, pch = 16, col = mycols[2], cex = 0.05, lwd = 0.01)
+
+  # Add world map
+  map('world', col = "black", fill = TRUE, add = TRUE, lwd = 0.25)
+  # If you get this error, 
+  # Error in as_mapper(.f, ...) : argument ".f" is missing, with no default
+  # You need to detach the package purrr (detach(package:purrr)) and reload
+  # the maps library (library(map))
+  
+  
+  # Add mont hheader somewhere
+  # remove massive margins
+  par(new=FALSE)
+}
+
+dev.off()
+

### 2018-01-03 code/Figure-4-code.R
Edits to title size and point sizes

@@ -13,7 +13,7 @@ library(viridis)
 # Read in data
 mid.top10 <- read.csv("data/mid.top100.csv")
 
-# Pick colours
+# Pick colours to match phases in Figure 2
 mycols <- c(viridis_pal()(10)[1], viridis_pal()(10)[5], viridis_pal()(10)[9])
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
@@ -29,7 +29,7 @@ for (Mn in 1:12) {
   
   # Make background of plots
   plot(x = NA, y = NA, xlim = c(-80, 50), ylim = c(0, 80), 
-       xlab = "", ylab = "", axes = FALSE, main = month.abb[Mn])
+       xlab = "", ylab = "", axes = FALSE, main = month.abb[Mn], cex.main = 2)
   rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], 
        col = "grey90")
   
@@ -41,7 +41,7 @@ for (Mn in 1:12) {
   jLon <- jitter(Month$Lon, factor = 1)
   
   # Add points coloured by phase
-  points(jLon, jLat, pch = 16, col = mycols[2], cex = 0.05, lwd = 0.01)
+  points(jLon, jLat, pch = 16, col = mycols[2], cex = 0.2)
 
   # Add world map
   map('world', col = "black", fill = TRUE, add = TRUE, lwd = 0.25)
@@ -50,9 +50,6 @@ for (Mn in 1:12) {
   # You need to detach the package purrr (detach(package:purrr)) and reload
   # the maps library (library(map))
   
-  
-  # Add mont hheader somewhere
-  # remove massive margins
   par(new=FALSE)
 }
 

### 2018-01-03 manuscript/revision/blue-whale-ms-v2.tex
Edits to abstract from CT

@@
 Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}. 
-Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}.
-Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct the movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
-%We identify two distinct phases in the whale's behaviour: repeated annual north-south migrations in sub-Arctic waters, and residency in subtropical waters. 
-%Subtropical residency is associated with isotopic signatures consistent with pregnancy in the year before the whale's death. % make clear this is tenuous.
-% Juvenile behaviour?
-
+Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}, but interpreting these records can be challenging.
+Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct seven years of movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
+We provide the longest continuous record of movement for an individual blue whale in the north Atlantic. 
+We show that the whale spent at least two years resident in sub-tropical waters as a juvenile, before commencing annual latitudinal migrations consistent with patterns of movements of blue whales in the northeast Atlantic inferred from observations, tagging and historic whaling data.
+The isotopic record is also suggestive of pregnancy and calf-rearing in the year before death. 
 Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climate, ecological conditions and anthropogenic pressures, and reveal aspects of life history.

### 2018-01-03 manuscript/revision/blue-whale-ms-v2.tex
Edits from CT as far as Fig 1

@@
 \section{Main text}
-Migratory species pose a particular challenge for conservation practitioners, because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}. 
-Such species may also be particularly affected by climate change, as changes can impact various parts of their life cycle\cite{robinson2009travelling}. 
-Dealing with these issues is difficult due to the relative scarcity of information on individual-level movements over multiple years, for both historical and present-day populations, especially for wide-ranging marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
+Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}. 
+Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle\cite{robinson2009travelling}.
+Identifying threats to migratory species and developing effective conservation measures requires information on individual-level movements spanning multiple years, for both historical and present-day populations.
+Such data are scarce, especially for rare, wide-ranging and long-lived marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
 
 The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
-Despite blue whale hunting being banned for over 50 years, blue whales are still listed as Endangered by the IUCN Red List\cite{reilly2008balaenoptera}.
-Conservation efforts are hampered by the scarcity of detailed data on the distribution, behaviour and ecology of blue whales, particularly for juveniles. 
-Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}.
-These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
-Additionally, they do not provide information on historical movement patterns that could be important for understanding the drivers of whale declines.
-An alternative technique for investigating blue whale movements is to use stable isotopes\cite{west2006stable,busquets2017estimating,hobson2008tracking}. 
-Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioural information that are often under-utilised \cite{lister2011natural}.
-We can use the isotopic compositions of whale tissues to reconstruct their movement patterns, by comparing them to spatially referenced models of isotopic variation across the species' range\cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
-These methods can be applied fairly inexpensively to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale `Hope'.
+Despite hunting for blue whales being banned for over 50 years, blue whales are still listed as Endangered on the IUCN Red List\cite{reilly2008balaenoptera}.
+Blue whales in the northeast Atlantic appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations (REF).  
+The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland\cite{pike2009note}.
+Their small population size and oceanic habit makes blue whale movements extremely hard to study. 
+Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
+A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands\cite{baines2014upwellings}.
+Northward migration in the spring may occur along mid-Atlantic corridors with peak sightings in the Azores around April-May.
+Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas\cite{pike2009note}.
+Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals\cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
+While direct tagging methods are adding vital behavioural data to our understanding of whale movements\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
+Currently the longest period of continuous monitoring via satellite tagging of an individual blue whale in the northeast Atlantic is X days, while photographic identification has matched a single blue whale with sightings in Iceland and Mauritania.
+Conservation efforts are consequently hampered by the lack of detailed data on the distribution, behaviour and ecology of blue whales, particularly during the winter and for the juvenile phase of life.
+An alternative technique for investigating whale movements retrospectively is to use intrinsic biochemical information such as stable isotopes\cite{west2006stable,busquets2017estimating,hobson2008tracking}.
+The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
+Stable isotope data therefore provide information on diet and location, but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values.
+Here we reconstruct movements of a historic whale by comparing isotopic compositions of samples from museum collections, rich archives of behavioural information that are often under-utilised \cite{lister2011natural}, to spatially referenced models of isotopic variation in diet across the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
 
-Baleen stable isotopes have a long history of use as biochemical markers to infer movement patterns in whales\cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}.
-Baleen is formed of plates of keratin and is used by mysticete whales to filter feed. 
-It is ideal for stable isotope studies because keratin grows continuously through an individual's life (although it is worn away at the tips over time, so rarely records an individual's entire lifespan), and once laid down it is metabolically inert\cite{best1996stable}.
+Mysticete (baleen) whales are characterized by the development of baleen plates, keratinous structures in the upper jaw used to filter food items from seawater. 
+Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable}. 
+Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although baleen is worn away at the tips over time, so rarely records an individual's entire lifespan.
+Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales\cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}, but have yet to be applied to blue whales from the north Atlantic population.
+Here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale ``Hope''. 
+Hope is estimated to be a young adult c.15 years at death, suggesting that her baleen may yield information about juvenile behaviours in north Atlantic blue whales around the peak of industrial whaling.
+
 Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton \cite{handbook}.
@@
 
-We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth. 
-This resulted in XX discrete samples of baleen. % add number
+We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth, resulting in 97 discrete samples of baleen.
 We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin.
@@
 relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope\cite{west2006stable}.
-Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; carbon isotopes provide information on the geographical locations of an individual, whereas nitrogen isotopes provide trophic and physiological information\cite{west2006stable}.
+Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; both isotopes provide information on the geographical locations of an individual, but nitrogen isotopes are more strongly fractionated during protein metabolism and thus contain additional trophic and physiological information\cite{west2006stable}.
 
-Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplementary Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the seven full years of the whale's life (early 1884 - spring 1891).
+Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods). 
+Therefore, given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$, we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over seven full years of the whale's life (early 1884 - spring 1891).
 The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
-Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods).  
 In behavioural phase one (from the start of the record in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Fig.
@@
 S1 and S2).
-In behavioural phase two (summer 1886 to spring 1891) $\delta^{15}$N values are relatively elevated and we see coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values. 
+In behavioural phase two (summer 1886 to spring 1891) $\delta^{15}$N values are relatively elevated and $\delta^{13}$C values are relatively depleted with coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values.
 In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively positive $\delta^{13}$C values in the second half of 1891.
-The final three months of the record show a continued fall in $\delta^{13}$C values (Fig. \ref{fig1}; Figs. S1 and S2).
+The final three months of the record show a progressive fall in $\delta^{13}$C values (Fig. \ref{fig1}; Figs. S1 and S2).
 
@@
 S1 and S2) in both isotopes.
-The approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
+The approximate relationship to years assuming a growth rate of 13.5cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
   \label{fig1}

### 2018-01-03 manuscript/revision/blue-whale-ms-v2.tex
Edits from CT as far as Fig 2/3

@@
 The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland\cite{pike2009note}.
-Their small population size and oceanic habit makes blue whale movements extremely hard to study. 
+Their small population size and oceanic habit make blue whale movements extremely hard to study. 
 Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
-A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands\cite{baines2014upwellings}.
+A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggest that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands\cite{baines2014upwellings}.
 Northward migration in the spring may occur along mid-Atlantic corridors with peak sightings in the Azores around April-May.
@@
 While direct tagging methods are adding vital behavioural data to our understanding of whale movements\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
-Currently the longest period of continuous monitoring via satellite tagging of an individual blue whale in the northeast Atlantic is X days, while photographic identification has matched a single blue whale with sightings in Iceland and Mauritania.
+Currently the longest period of continuous monitoring via satellite tagging of an individual blue whale in the northeast Atlantic is X days, while photographic identification has matched a single blue whale with sightings in Iceland and Mauritania.\cite{baines2014upwellings}  % How many days?
 Conservation efforts are consequently hampered by the lack of detailed data on the distribution, behaviour and ecology of blue whales, particularly during the winter and for the juvenile phase of life.
@@
 Stable isotope data therefore provide information on diet and location, but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values.
-Here we reconstruct movements of a historic whale by comparing isotopic compositions of samples from museum collections, rich archives of behavioural information that are often under-utilised \cite{lister2011natural}, to spatially referenced models of isotopic variation in diet across the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
+Here we reconstruct movements of an historic whale by comparing isotopic compositions of samples from museum collections, rich archives of behavioural information\cite{lister2011natural}, to spatially referenced models of isotopic variation in diet across the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
 
@@
 Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton \cite{handbook}.
-Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on around three weeks prior to that time (this time is required for the carbon and nitrogen to be assimilated and converted into tissues).
-%REF???
-% add line about it being ok to use phytoplankton?
-
+Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on.
 We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth, resulting in 97 discrete samples of baleen.
@@
 
-Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values, and because the isotopic composition of phytoplankton at the base of marine food webs varies in both space and time \cite{west2006stable,mcmahon2015millennial}.
-We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Fig.
+Relating marine animal tissue isotopic compositions to likely locations is complicated by a lack of knowledge of spatio-temporal variations in the isotopic composition of diet\cite{west2006stable,mcmahon2015millennial}.
+We develop a new approach to address this problem using temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Fig.
 S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig.
-S5) to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of blue whales.
+S5).
+We use our coupled model to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of north Atlantic blue whales.
 We then compare simulated profiles (Figs.
 \ref{fig2}, \ref{fig3} and Fig.
-S4) with the measured isotopic profile of the blue whale's baleen (Fig.
+S4) to the measured isotopic profile of the blue whale's baleen (Fig.
 \ref{fig1}) to identify the most likely movement history.
-We assume continued ingestion of new protein throughout the year. 
-Other mysticete whales fast during migrations and during winter, relying on stored lipid reserves to meet metabolic demands, but blue whales are too large to stop feeding for long periods\cite{busquets2017estimating,silva2013north}.
-%Can we back this up? Are we more saying that we think they need protein to maintain tissue turnover but could meet energetic demands form stored lipids?
-
-%SOME indirect evidence for grey whale continued feeding through migrations -
+Blue whales are commonly assumed to feed mainly in summer in productive, high latitude regions and rely largely on stored lipid reserves to meet metabolic demands in lower latitudes in winter. 
+However, in winter blue whales are associated with upwellings\cite{baines2017autumn}, while northward migrations in spring coincide with the onset and development of phytoplankton blooms\cite{silva2013north,visser2011timing,busquets2017estimating}, strongly implying feeding throughout the year.
+We therefore assume opportunistic ingestion of new protein throughout the year.
 
@@
  % \centering
- % \includegraphics[width = \linewidth]{figures/Figure-2-migratory-model-full-map.png}
+ % \includegraphics[width = \linewidth]{figures/Figure-2-points.png}
   %\caption{Simulated tracks from the migratory movement model (2250 days for each of 50 replicates) with simulated day indicated by colour illustrating the residency period from Ireland to the Norwegian/Barents Sea followed by a migratory period and year spent near the Azores (as indicated by an arrow) and/or Canary Islands before returning north prior to stranding in Wexford, Ireland 1891 (as indicated by an arrow).
@@
  % \centering
-  %\includegraphics[width = \linewidth]{figures/Figure-3-migratory-model-d13C.png}
+  %\includegraphics[width = \linewidth]{figures/Figure-3-blue-sims.png}
   %\caption{Correlations among $\delta^{13}$C data from 50 movement simulations (green) and empirical data from baleen (blue).

### 2018-01-03 code/Extract-models.R
Extract r2 for histogram

@@ -80,6 +80,8 @@ for(run in 1:length(unique(resTrack$Rep))){
 	r2[run] <- summary(lm(test3[, run] ~ test3$Blue))$adj.r.squared
 }
 
+write.csv(file = "data/all.r2.csv", r2, quote = FALSE, row.names = FALSE)
+
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = 
 # extract top 10% and bottom 10% best fit models for whole series
 # write out for figures 

### 2018-01-03 code/Figure-S6-hist-code.R
Code to make r2 historgram

@@ -1,0 +1,25 @@
+# Author: Natalie Cooper
+# About: script to plot histogram of r2 values
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Load libraries
+library(tidyverse)
+
+# Load data
+r2 <- read.csv("data/all.r2.csv")
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Plot
+
+p1 <- 
+  ggplot(r2, aes(x = x)) +
+  geom_histogram(fill = "grey", binwidth = 0.05) +  
+  theme_classic(base_size = 16) + 
+  xlab(expression(paste(r^2))) +
+  #xlab(expression(paste(r^2, " from models comparing real to simulated ", 
+  #                delta^{13}, "C (\u2030)", " values"))) +
+  expand_limits(x = c(0, 1),  y = c(0, 200))
+
+
+ggsave("manuscript/revision/figures/Figure-S6-r2.png", p1, 
+       device = png(width = 600, height = 400))

### 2018-01-04 code/00-fix-simulations-file.R
Edit to remove reps with < 3000 days

@@ -1,9 +1,7 @@
 # Fixing up movement model simulations dataset
-# For some reason the data has 1078 replicates not 1000
-# Therefore this code removes the excess 78
-# I've arbitrarily chosen to remove the 78 with the fewest
-# with fewest days in the run < 2957.
-# Natalie Cooper Nov 2017
+# Some of the runs have < 3000 days so these need to be cut
+# before we start (where the simulation got stuck and "stranded")
+# Natalie Cooper Jan 2018
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Load libraries
 library(tidyverse)
@@ -12,17 +10,12 @@ library(tidyverse)
 # Takes a bit of time
 resTrack <- read_csv("data/model.sims.csv")
 
-# Remove rep that didn't work
-resTrack <- resTrack[-c(3213051:3213211), ]
-
-# Remove excess 78 simulations
-# Remove those with fewest days - < 2957
-
+# Remove reps with < 3000 days
 remove <- 
   resTrack %>%
   group_by(Rep) %>%
   summarise(ln = length(count2)) %>%
-  filter(ln < 2957)
+  filter(ln < 3000)
 
 toremove <- match(resTrack$Rep, remove$Rep, nomatch = 0)
 
@@ -30,4 +23,4 @@ resTrack2 <- resTrack[toremove == 0, ]
 
 # Write to data folder
 # Use this for later analyses
-write.csv(resTrack2, "data/model.sims1000.csv", quote = FALSE, row.names = FALSE)
+write.csv(resTrack2, "data/model.sims.3000.csv", quote = FALSE, row.names = FALSE)

### 2018-01-04 code/Extract-models.R
Identified some potentially unecessary code

@@ -12,10 +12,12 @@ library(mapdata)
 library(tidyverse)
 library(raster)
 
+### NOT SURE THIS IS NEEDED HERE
 # Read in the measured whale data and select blue whale
 whale_isos <- read_csv("data/raw-whale-isotope-data.csv")
 blue <- filter(whale_isos, whale_ID == "KC7")
 
+### NOT SURE THIS IS NEEDED HERE
 # Fix day numbers reflecting monthly samples 
 # (for loess sampling - fixed to have same no.s)
 predict.days <- seq(from = 30, to = (8 * 365) + 100, by = 31)
@@ -43,6 +45,8 @@ test <- ddply(resTrack, "Rep", function(x) {
 	newSeries <- x[x$count2%in%sampledays, ]
 })
 
+# Remove any simulated values where the day > 3020
+# which is the last day of the real data
 test2 <- ddply(test, "Rep", function(x) {
 	newSeries <- x[x$count2 < 3021, ]
 })
@@ -76,7 +80,6 @@ test3$Blue <- rev(blue$d13C)
 r2 <- vector()
 
 for(run in 1:length(unique(resTrack$Rep))){
-	testX <- lm(test3[, run] ~ test3$Blue)
 	r2[run] <- summary(lm(test3[, run] ~ test3$Blue))$adj.r.squared
 }
 

### 2018-01-04 code/Figure-S5-code.R
Change names of supp figures

@@ -1,52 +1,25 @@
-# Author: Clive Trueman
-# About: script to plot max and sd of latitude 
-# for the top 10% and bottom 10% of models
-# Modified by Natalie Cooper Nov 2017
+# Author: Natalie Cooper
+# About: script to plot histogram of r2 values
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Load libraries
 library(tidyverse)
-library(gridExtra)
 
 # Load data
-max <- read.csv("data/max.lat.csv")
-sdev <- read.csv("data/sd.lat.csv")
+r2 <- read.csv("data/all.r2.csv")
 
-# Relevel so top 10% is first
-max$group <-relevel(max$group, ref = "top")
-sdev$group <-relevel(sdev$group, ref = "top")
-
-# Get means
-mean.max <- summarise(group_by(max, group), mn = mean(V1))
-mean.sd <- summarise(group_by(sdev, group), mn = mean(V1))
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Plot
 
 p1 <- 
-  ggplot(max, aes(x = group, y = V1, fill = group)) +
-  geom_boxplot() +  
+  ggplot(r2, aes(x = x)) +
+  geom_histogram(fill = "grey", binwidth = 0.05) +  
   theme_classic(base_size = 16) + 
-  xlab("") +
-  ylab("Maximum latitude") +
-  scale_x_discrete(labels = c("Top 10%", "Bottom 10%")) +
-  scale_y_continuous(breaks = c(20, 40, 60, 80, 100),
-                     labels = c(20, 40, 60, 80, 100)) +
-  scale_fill_manual(values = c("white", "gray")) +
-  theme(legend.position = "none")
+  xlab(expression(paste(r^2))) +
+  #xlab(expression(paste(r^2, " from models comparing real to simulated ", 
+  #                delta^{13}, "C (\u2030)", " values"))) +
+  expand_limits(x = c(0, 1),  y = c(0, 200))
 
-p2 <- 
-  ggplot(sdev, aes(x = group, y = V1, fill = group)) +
-  geom_boxplot() +  
-  theme_classic(base_size = 16) + 
-  xlab("") +
-  ylab("Standard deviation of latitude") +
-  scale_x_discrete(labels = c("Top 10%", "Bottom 10%")) +
-  scale_y_continuous(breaks = c(10, 20, 30, 40, 50),
[…]

### 2018-01-04 code/Figure-S6-code.R

@@ -1,46 +1,52 @@
 # Author: Clive Trueman
-# About: script to plot kernal densities for top 10% and bottom 10% of models
-# Tidied by Natalie Cooper Nov 2017
+# About: script to plot max and sd of latitude 
+# for the top 10% and bottom 10% of models
+# Modified by Natalie Cooper Nov 2017
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Load libraries
-library(maps)
-library(mapdata)
-library(spatstat)
+library(tidyverse)
+library(gridExtra)
 
-# Read in data
-top10 <- read.csv("data/top10percent.csv")
-bottom10 <- read.csv("data/bottom10percent.csv")
+# Load data
+max <- read.csv("data/max.lat.csv")
+sdev <- read.csv("data/sd.lat.csv")
 
+# Relevel so top 10% is first
+max$group <-relevel(max$group, ref = "top")
+sdev$group <-relevel(sdev$group, ref = "top")
+
+# Get means
+mean.max <- summarise(group_by(max, group), mn = mean(V1))
+mean.sd <- summarise(group_by(sdev, group), mn = mean(V1))
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Kernal density plots
+# Plot
 
-# Jitter points of lat and long to get smooth density plot
-jLat <- jitter(top10$Lat, factor = 2)
-jLon <- jitter(top10$Lon, factor = 1)
-jLatB <- jitter(bottom10$Lat, factor = 2)
-jLonB <- jitter(bottom10$Lon, factor = 1)
+p1 <- 
+  ggplot(max, aes(x = group, y = V1, fill = group)) +
+  geom_boxplot() +  
+  theme_classic(base_size = 16) + 
+  xlab("") +
+  ylab("Maximum latitude") +
+  scale_x_discrete(labels = c("Top 10%", "Bottom 10%")) +
+  scale_y_continuous(breaks = c(20, 40, 60, 80, 100),
+                     labels = c(20, 40, 60, 80, 100)) +
+  scale_fill_manual(values = c("white", "gray")) +
+  theme(legend.position = "none")
 
-# Create point patterns to plot
-pts <- ppp(jLon, jLat, window = owin(c(-80, 55), c(-10, 85)))
[…]

### 2018-01-04 code/Figure-S6-hist-code.R

@@ -1,25 +1,0 @@
-# Author: Natalie Cooper
-# About: script to plot histogram of r2 values
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Load libraries
-library(tidyverse)
-
-# Load data
-r2 <- read.csv("data/all.r2.csv")
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Plot
-
-p1 <- 
-  ggplot(r2, aes(x = x)) +
-  geom_histogram(fill = "grey", binwidth = 0.05) +  
-  theme_classic(base_size = 16) + 
-  xlab(expression(paste(r^2))) +
-  #xlab(expression(paste(r^2, " from models comparing real to simulated ", 
-  #                delta^{13}, "C (\u2030)", " values"))) +
-  expand_limits(x = c(0, 1),  y = c(0, 200))
-
-
-ggsave("manuscript/revision/figures/Figure-S6-r2.png", p1, 
-       device = png(width = 600, height = 400))

### 2018-01-04 code/Figure-S7-code.R

@@ -1,0 +1,46 @@
+# Author: Clive Trueman
+# About: script to plot kernal densities for top 10% and bottom 10% of models
+# Tidied by Natalie Cooper Nov 2017
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Load libraries
+library(maps)
+library(mapdata)
+library(spatstat)
+
+# Read in data
+top10 <- read.csv("data/top10percent.csv")
+bottom10 <- read.csv("data/bottom10percent.csv")
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Kernal density plots
+
+# Jitter points of lat and long to get smooth density plot
+jLat <- jitter(top10$Lat, factor = 2)
+jLon <- jitter(top10$Lon, factor = 1)
+jLatB <- jitter(bottom10$Lat, factor = 2)
+jLonB <- jitter(bottom10$Lon, factor = 1)
+
+# Create point patterns to plot
+pts <- ppp(jLon, jLat, window = owin(c(-80, 55), c(-10, 85)))
+ptsB <- ppp (jLonB, jLatB, window = owin(c(-80, 55), c(-10, 85)))
+
+
+png("manuscript/revision/figures/Figure-S6-kernals.png", width = 1200, height = 800)
+
+par(mfrow = c(1, 2))
+par(mar = c(2, 1, 3, 1))
+
+# Top 10 %
+plot(x = NA, y = NA, xlim = c(-80,55), ylim = c(-10, 85), 
+     xlab = "", ylab = "", axes = FALSE, main = "Top 10%")
+plot(density(pts), add = TRUE)
+map('world', col = "grey", fill = TRUE, add = TRUE, lwd = 0.25)
+
+# Bottom 10 %
+plot(x = NA, y = NA, xlim = c(-80,55), ylim = c(-10, 85), 
+     xlab = "", ylab = "", axes = FALSE, main = "Bottom 10%")
+plot(density(ptsB), add = TRUE)
+map('world', col = "grey", fill = TRUE, add = TRUE, lwd = 0.25)
+
+dev.off()

### 2018-01-04 code/Figure-S5-code.R
Correct figure numbers

@@ -21,5 +21,5 @@ p1 <-
   expand_limits(x = c(0, 1),  y = c(0, 200))
 
 
-ggsave("manuscript/revision/figures/Figure-S6-r2.png", p1, 
+ggsave("manuscript/revision/figures/Figure-S5-r2.png", p1, 
        device = png(width = 600, height = 400))

### 2018-01-04 code/Figure-S6-code.R

@@ -48,5 +48,5 @@ p2 <-
 
 boxes <- grid.arrange(p1, p2, ncol = 2)
 
-ggsave("manuscript/revision/figures/Figure-S5-boxplots.png", boxes, 
+ggsave("manuscript/revision/figures/Figure-S6-boxplots.png", boxes, 
        device = png(width = 600, height = 400))

### 2018-01-04 code/Figure-S7-code.R

@@ -26,7 +26,7 @@ pts <- ppp(jLon, jLat, window = owin(c(-80, 55), c(-10, 85)))
 ptsB <- ppp (jLonB, jLatB, window = owin(c(-80, 55), c(-10, 85)))
 
 
-png("manuscript/revision/figures/Figure-S6-kernals.png", width = 1200, height = 800)
+png("manuscript/revision/figures/Figure-S7-kernals.png", width = 1200, height = 800)
 
 par(mfrow = c(1, 2))
 par(mar = c(2, 1, 3, 1))

### 2018-01-05 quarantine/Whale model code NEE.R
Initial commit of model code

@@ -1,0 +1,951 @@
+# Whale individual based movement model
+# Author: Clive Trueman (trueman@noc.soton.ac.uk)
+# This version Sept. 2017
+
+
+
+# run the Lat.Lon code to get the km conversion first.
+# Degree to km function
+
+# Calculate distance in kilometers between two points
+earth.dist <- function (long1, lat1, long2, lat2)
+{
+	
+rad <- pi/180
+a1 <- lat1 * rad
+a2 <- long1 * rad
+b1 <- lat2 * rad
+b2 <- long2 * rad
+dlon <- b2 - a2
+dlat <- b1 - a1
+a <- (sin(dlat/2))^2 + cos(a1) * cos(b1) * (sin(dlon/2))^2
+c <- 2 * atan2(sqrt(a), sqrt(1 - a))
+R <- 6378.145
+d <- R * c
+return(d)
+}
+
+# build array for distance vectors
+Lat.Lon<-array(data=0, dim=c(180, 360,2))
+
+	#longitude distance varies with lat
+	Lt=vector(length=179)
+	count=1
+	
+	for(Lat in 89:-90){
+		x<-earth.dist(-180,Lat, -179,Lat)
+		Lt[count]<-x
+		count=count+1
+		}
+
+Lat.Lon[1:180,1,2]<-0
+	Lat.Lon[1:180,2,2]<-Lt
+	for(Lon in 3:359){
+	Lat.Lon[1:180,Lon,2]<-Lat.Lon[,Lon-1,2]+Lat.Lon[,2,2]
+	}
+	
+		#latitude distance constant at 111km
+
+	Lt2=rep(0,179)
+	count=2
+		for(Lat in -89:89){
+
+		Lt2[count]<-Lt2[count-1]+111
+		count=count+1
+		}
+
+Lat.Lon[,,1]<-Lt2
+
+
+Lon2<-Lat.Lon[,1:359,2]/Lat.Lon[,359,2]*360
+
+
+#########################################################
+
+library(R.matlab)
+library(maps)
+library(mapdata)
+library(scales)
+library(TTR)
+require(maptools)
+require(raster) 
+
+
+
+# link to monthly isoscape layers 
[…]

### 2018-01-05 manuscript/Nature_submission/blue-whale-supplementary.tex
Move all Nature draft to new folder

--- 2017-11-07 manuscript/revision/blue-whale-supplementary-v2.tex
@@
 $\delta^{13}$C values also show repeated fluctuations in the first (oldest) 60cm of the baleen plate (behavioural phase one), with a similar periodicity of 13.5cm (Fig. \ref{figs1}). 
-A distinct change in the isotopic pattern along the baleen plate occurs at around 60cm behavioural phase two, with a transition to relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.
+A distinct change in the isotopic pattern along the baleen plate occurs at around 60cm behavioural phase two), with a transition to relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.
 

### 2018-01-23 manuscript/revision/blue-whale-ms-v2.tex
Minor edit to abstract

--- 2018-01-03 manuscript/revision/blue-whale-ms-v2.tex
@@
 Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}.
-Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}, but interpreting these records can be challenging.
-Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct seven years of movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
-We provide the longest continuous record of movement for an individual blue whale in the north Atlantic. 
-We show that the whale spent at least two years resident in sub-tropical waters as a juvenile, before commencing annual latitudinal migrations consistent with patterns of movements of blue whales in the northeast Atlantic inferred from observations, tagging and historic whaling data.
-The isotopic record is also suggestive of pregnancy and calf-rearing in the year before death. 
+Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}.
+Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct the movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
+%We identify two distinct phases in the whale's behaviour: repeated annual north-south migrations in sub-Arctic waters, and residency in subtropical waters. 
+%Subtropical residency is associated with isotopic signatures consistent with pregnancy in the year before the whale's death. % make clear this is tenuous.
+% Juvenile behaviour?
+
 Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climate, ecological conditions and anthropogenic pressures, and reveal aspects of life history.

### 2018-01-23 manuscript/revision/blue-whale-ms-v2.tex

--- 2018-01-03 manuscript/revision/blue-whale-ms-v2.tex
@@
 The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
-In behavioural phase one (from the start of the record in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Fig.
-\ref{fig1}; Figs.
-S1 and S2).
-In behavioural phase two (summer 1886 to spring 1891) $\delta^{15}$N values are relatively elevated and $\delta^{13}$C values are relatively depleted with coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values.
+In behavioural phase one (from the start of the record in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Figure \ref{fig1}; Supplementary Figures S1 and S2).
+In behavioural phase two (summer 1886 to spring 1890) $\delta^{15}$N values are relatively elevated and $\delta^{13}$C values are relatively depleted with coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values.
 In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively positive $\delta^{13}$C values in the second half of 1891.
-The final three months of the record show a progressive fall in $\delta^{13}$C values (Fig. \ref{fig1}; Figs. S1 and S2).
+The final three months of the record show a progressive fall in $\delta^{13}$C values (Figure \ref{fig1}; Supplementary Figures S1 and S2).
 
@@
 Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes).
-There is strong annual periodicity and cross-correlation (Figs.
-S1 and S2) in both isotopes.
+There is strong annual periodicity and cross-correlation (Supplementary Figures S1 and S2) in both isotopes.
 The approximate relationship to years assuming a growth rate of 13.5cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
@@
 Relating marine animal tissue isotopic compositions to likely locations is complicated by a lack of knowledge of spatio-temporal variations in the isotopic composition of diet\cite{west2006stable,mcmahon2015millennial}.
-We develop a new approach to address this problem using temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Fig.
-S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig.
-S5).
+We develop a new approach to address this problem using temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Supplementary Figure S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Supplementary Table S1; Supplementary Figure S5).
 We use our coupled model to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of north Atlantic blue whales.
-We then compare simulated profiles (Figs.
-\ref{fig2}, \ref{fig3} and Fig.
-S4) to the measured isotopic profile of the blue whale's baleen (Fig.
-\ref{fig1}) to identify the most likely movement history.
+We then compare simulated profiles (Figures \ref{fig2}, \ref{fig3} and Supplementary Figure S4) to the measured isotopic profile of the blue whale's baleen (Figure \ref{fig1}) to identify the most likely movement history.
 Blue whales are commonly assumed to feed mainly in summer in productive, high latitude regions and rely largely on stored lipid reserves to meet metabolic demands in lower latitudes in winter. 
 However, in winter blue whales are associated with upwellings\cite{baines2017autumn}, while northward migrations in spring coincide with the onset and development of phytoplankton blooms\cite{silva2013north,visser2011timing,busquets2017estimating}, strongly implying feeding throughout the year.
-We therefore assume opportunistic ingestion of new protein throughout the year.
+We therefore assume opportunistic ingestion of protein throughout the year.
 
-% figure 2
-%\begin{figure}
- % \centering
- % \includegraphics[width = \linewidth]{figures/Figure-2-points.png}
-  %\caption{Simulated tracks from the migratory movement model (2250 days for each of 50 replicates) with simulated day indicated by colour illustrating the residency period from Ireland to the Norwegian/Barents Sea followed by a migratory period and year spent near the Azores (as indicated by an arrow) and/or Canary Islands before returning north prior to stranding in Wexford, Ireland 1891 (as indicated by an arrow).
-}
-  %\label{fig2}
-%\end{figure}
+% figure 2{}
+\begin{figure}
+ \centering
+ \includegraphics[width = \linewidth]{figures/Figure-2-points.png}
+  \caption{Simulated locations of the whale taken from the top 10\% best fitting migratory movement models. 
+  Colours reflect the behavioural phase. 
+  Phase one is early 1884 to spring 1886, phase two is summer 1886 to spring 1890, and phase three is spring 1890 to spring 1891.}
+  \label{fig2}
+\end{figure}
 
 % figure 3
-%\begin{figure}
- % \centering
-  %\includegraphics[width = \linewidth]{figures/Figure-3-blue-sims.png}
-  %\caption{Correlations among $\delta^{13}$C data from 50 movement simulations (green) and empirical data from baleen (blue).
-Across the 50 simulations, the correlation coefficient between z-score transformations of empirical and time-matched simulation values had a mean of 0.48 ($sd = 0.2$), 49 of them were positive, and 39 of the 50 Pearson correlation tests had a p-value less than the Bonferroni corrected value of 0.001 ($p = 0.05/50 = 0.001$).
-The initiation of the southward migration in the late spring of 1889 is indicated by the vertical broken line.
-The standard deviation of the empirical data have been rescaled by a factor of three so as to better visualise the correlation because, as a consumer, the whale is expected to integrate and attenuate the fluctuations in its diet, and the end points of the simulations and empirical data have been aligned to coincide.}
-  %\label{fig3}
-%\end{figure}
+\begin{figure}
+ \centering
+  \includegraphics[width = \linewidth]{figures/Figure-3-blue-sims.png}
+  \caption{Correlations among simulated $\delta^{13}$C from the top 10\% best fitting migratory movement models (grey lines, right hand y-axis) and $\delta^{13}$C from baleen (black line, left hand y-axis; see Figure \ref{fig1}).
+  The standard deviation of the empirical data have been rescaled by a factor of six so as to better visualise the correlation because, as a consumer, the whale is expected to integrate and attenuate the fluctuations in its diet, and the end points of the simulations and empirical data have been aligned to coincide.}
+  \label{fig3}
+\end{figure}
 
 The relatively positive and seasonally-invariant $\delta^{13}$C values seen during behavioural phase one are only found in subtropical areas of the North Atlantic.  
-Our simulations show high variance during this period (Fig.
-\ref{fig3}), reflecting a range of possible locations for the whale, although areas around the Cape Verde Islands, and potentially to the west of the Azores (Fig.
-\ref{fig2}), most closely match the measured profile.
+Our simulations show high variance during this period (Figure \ref{fig3}), reflecting a range of possible locations for the whale, although areas around the Mauritanian coast and Cape Verde Islands, a known current and historic winter feeding area for blue whales \cite{baines2014upwellings,reeves2004historical}, and potentially to the west of the Azores (Figure \ref{fig2}), most closely match the measured profile.
 The whale remained in these warm waters for at least two full years. 
-The whale was estimated to be 15 years old when it died (based on vertebral epiphyseal fusion; R.Sabin \textit{pers.comm}), so was approximately 3-5 years old during this period.
-During behavioural phase two, starting at an approximate age of 5-6 years, the observed low $\delta^{13}$C values imply that the whale is foraging in colder, more northerly latitudes. 
-The pronounced cyclical variations in $\delta^{13}$C values are characteristic of the isotopic expression of the spring phytoplankton bloom in northern waters \cite{magozzi2017using}.
+Hope was estimated to be 15 years old when she died (based on vertebral epiphyseal fusion; R.Sabin \textit{pers.comm}), so was approximately 3-5 years old during this period.
+During behavioural phase two, starting at an approximate age of 5-6 years, the observed low $\delta^{13}$C values imply foraging in colder, more northerly latitudes. 
+The pronounced cyclical variations in $\delta^{13}$C values observed during behavioural phase two could reflect either the isotopic expression of the spring phytoplankton bloom in northern waters \cite{magozzi2017using}, latitudinal migrations, or a combination of both.
 
-We simulated 1000 individual movement patterns and compared resulting simulated baleen $\delta^{13}$C records to the measured records with simple linear regressions. 
-Simulated baleen $\delta^{13}$C profiles produce a good fit to measured profiles, the median $r^2$ value across 1000 simulated profiles was 0.49, and the maximum was 0.74 (Fig.
-S6, distribution of $r^2$ values).
-The top 10\% best fitting simulated profiles are shown in Fig. \ref{fig2} [line plot top 10\%). 
-Best fitting models predict juvenile residency in the Cape Verde region in behavioural phase 1. 
-Behavioural phase 2 is best simulated by seasonal migrations between summer foraging in northern areas in the Norwegian Sea/Barents Sea/Iceland region, and winter foraging in a broad region between the UK and subtropical waters.
-Best fitting models in general predicted a greater latitudinal foraging range, and foraging in more northerly waters (Supplementary materials Fig. S7 [boxplots and kernel density plots]).
+We simulated 1000 individual movement patterns and compared the resulting simulated baleen $\delta^{13}$C records to the measured records with simple linear regressions. 
+Simulated baleen $\delta^{13}$C profiles produce a good fit to measured profiles, the median $r^2$ value across 1000 simulated profiles was 0.49, and the maximum was 0.74 (Supplementary Figure S6).
+% update these values
+The top 10\% best fitting simulated profiles are shown in Figure \ref{fig2}. 
+Best fitting models predict juvenile residency in the Cape Verde region in behavioural phase one. 
+Behavioural phase two is best simulated by seasonal migrations between summer foraging in northern areas in the Norwegian Sea/Barents Sea/Iceland region, and winter foraging in a broad region between the UK and more southerly, subtropical waters.
+Better fitting models in general were those predicting a greater latitudinal foraging range, and foraging in more northerly waters (Supplementary Figure S7 [boxplots and kernel density plots]).
+% what is fig S7?
+Best-fit model distributions are largely consistent with our current understanding of blue whale distributions in the northeast Atlantic\cite{reeves2004historical,baines2014upwellings,baines2017autumn,reeves2004historical}, with perhaps greater importance of winter foraging in temperate regions (Figure \ref{fig4}).
+
+% figure 4
+\begin{figure}
+ \centering
+  \includegraphics[width = \linewidth]{figures/Figure-4-monthly.png}
+  \caption{Simulated locations by month taken from the top 10\% best fitting migratory movement models for behavioral phase two (summer 1886 to spring 1890) only.}
+  \label{fig4}
+\end{figure}
 
@@
 Beginning in the winter of 1889/1890, the observed $\delta^{13}$C profile is relatively depleted, and remains constant for approximately six months, before increasing rapidly in the winter of 1890/1891.
-Depleted $\delta^{13}$C values are found in northern waters that show large temporal fluctuations in $\delta^{13}$Cplk values [fig SX residency plot], and the observed values cannot be simulated purely from movements within the known geographic range.
-We infer that the $\delta^{13}$C record in the last period of life reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions\cite{cardona2017temporal}.
-We suggest that the constant, depleted $\delta^{13}$C values reflect use of carbon reserves assimilated in northern latitudes during 6-8 months residency in low latitude waters. 
-After six months the whale began feeding in low latitude waters producing a dramatic increase in $\delta^{13}$C values and began a final migration north. 
+Depleted $\delta^{13}$C values are found in northern waters, but these regions also show large temporal fluctuations in $\delta^{13}$Cplk values (Supplementary Figure S4), and the observed values cannot be simulated purely from movements within the known geographic range.
+We infer that the $\delta^{13}$C record in the last period of life reflects pregnancy and calf-rearing. 
+We suggest that the constant, depleted $\delta^{13}$C values reflect exclusive use of carbon reserves assimilated in northern latitudes (i.e. a period of time with no opportunistic foraging). 
+After six months the whale began ingesting protein with high $\delta^{13}$C  values, indicating feeding in low latitude waters shortly followed by a final northward migration. 
 Blue whales have a 10-12 month gestation period, with calving occurring in subtropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
-The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, residency in warm waters through the whole of 1890 with feeding resuming in the second half of 1890 and eventual stranding during the return to northern feeding grounds in early 1891, is therefore consistent with the known reproductive ecology of blue whales.
-
-%%% STILL TO FIX BELOW %%%%
-
-%Possibly more interesting now to talk about juvenile behaviour and potential for prolonged residency in warm waters  - implications for interactions with tourist / shipping etc.
-
-%Kernel density plot gives a visualization of proportional amount of time we infer the whale might have spent in each region
+We therefore interpret the final year of the isotopic record as indicating: (a) birth of a calf in the winter of 1889/1890, presumably in sub-tropical waters, (b) 6 months of weaning during which time the maternal whale did not feed, followed by (c) a short period of feeding in sub-tropical waters in the second half of 1890, potentially during a final northward migration, and (d) eventual stranding during the return to northern feeding grounds in early 1891.
 
@@
 Although based on a single individual, our results highlight some interesting patterns worthy of further study.
-Firstly, next to nothing is known about juvenile behaviour in blue whales - calves become independent of their mothers at around seven months, and become sexually mature around 8-10 years old \cite{handbook}.
-Our analyses suggest that between these points, juvenile North Atlantic blue whales may spend extended periods in warm waters before migrating to higher latitude feeding grounds.
-This has implications for blue whale conservation, as these warm waters around the Azores and Cape Verde are also tourist and shipping hotspots, increasing the likelihood of ship strikes and anthropogenic effects on the juvenile whales.
-Secondly, conventional wisdom suggests that baleen whales migrate each year from feeding grounds near the poles in summer, to overwintering and breeding grounds nearer the equator\cite{corkeron1999baleen,lockyer1981migration}.
-However, a growing body of evidence suggests that not all whales migrate each year, with some remaining at high or low latitudes year round\cite{mcdonald2006biogeographic,busquets2017estimating}. 
-Our results support this conclusion, with the whale remaining at relatively high latitudes for several years, before migrating south, presumably to calve.
+Firstly, next to nothing is known about juvenile behaviour in blue whales - calves become independent of their mothers at around seven months, and become sexually mature around 8-10 years old\cite{handbook}.
+Our analyses suggest that between these points, juvenile North Atlantic blue whales may spend extended periods in warm, highly productive waters associated with upwelling systems before migrating to higher latitude feeding grounds.
+This has implications for blue whale conservation, as the upwelling regions around Mauritania and the Cape Verde islands are areas with emerging offshore oil and gas exploration and are fishing, tourist and shipping hotspots, increasing threats to juvenile whales.
 
@@
 The NHM blue whale thus represents an individual from a species at the brink of extinction.
-Despite dramatic changes in anthropogenic pressures since 1891, the NHM whale and present-day blue whales in the North Atlantic have similar spatial ecologies; there are still populations of blue whales in the regions indicated by our model, and females still breed in the subtropical waters near the Azores\cite{reilly2008balaenoptera}.
+Despite dramatic changes in anthropogenic pressures since 1891, the NHM whale and present-day blue whales in the North Atlantic have similar spatial ecologies; there are still populations of blue whales in the regions indicated by our model.
 This may reflect behavioural responses to whaling, as blue whales were more widely distributed in the past\cite{reeves2004historical}, or the selective extinction of populations outside of these areas.
 
-Our movement simulation modeling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviours over multiple years.
+Our use of movement simulation modeling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviours over multiple years.
 Such techniques can also be applied to extinct species behaviours and used to understand how populations were affected by past pressures such as hunting, and present-day pressures such as global change.
-Our results also add to previous research\cite{lister2011natural,ryan2013stable} highlighting museums as vast sources of individual-level spatial ecology information.
+Our results also add to previous research\cite{lister2011natural,ryan2013stable} highlighting museums as rich archives of individual-level behavioural information.
 
@@
 {\raise.17ex\hbox{$\scriptstyle\sim$}}1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
-XX samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section which contains the most recent tissue. 
+97 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section that contains the most recent tissue. 
 Baleen grows at a constant rate, so the samples are equally spaced through time\cite{best1996stable}. 
@@
 Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
-We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Fig.
-S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated for six recent blue whales from the Californian Pacific\cite{busquets2017estimating}.
+We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Supplementary Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated for six recent blue whales from the Californian Pacific\cite{busquets2017estimating}.
 Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891. 
 Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase two (70cm to 18cm on the baleen plate), but no clear relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase one (the oldest 30cm of the baleen plate, i.e.
-100cm - 70cm; Fig.
-S2) or the final 500 days of the whale's life (youngest 18cm of the baleen plate; Fig.
-S2).
+100cm - 70cm; Supplementary Figure S2) or the final 500 days of the whale's life (youngest 18cm of the baleen plate; Supplementary Figure S2).
 
 \subsection{Baseline isotope comparisons}
-Isotope-enabled biogeochemical ocean models\cite{magozzi2017using,schmittner2016complementary} were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Fig.
-S3).
-Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model (Fig. S3). 
+Isotope-enabled biogeochemical ocean models\cite{magozzi2017using,schmittner2016complementary} were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Supplementary Figure S3).
+Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model (Supplementary Figure S3). 
 $\delta^{13}$C POM values were simulated at 1$^{\circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model\cite{magozzi2017using,yool2013medusa}. 
@@
 Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
-In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S3). 
+In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Supplementary Figure S3). 
 The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature\cite{magozzi2017using,laws1995dependence}.
@@
 We therefore used monthly simulated $\delta^{13}$C POM values to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours.
+Whales feed on krill, not POM. 
+Assimilation of carbon into krill tissues will dampen the temporal variability seen in POM, effectively producing a temporal average over the timescale of isotopic turnover within krill. 
+We estimate turnover to be complete between four and six months [REF] and therefore we resampled the $\delta^{13}$C POM values in each one degree cell to reflect a six month moving average. 
+Carbon isotope values are also likely to be fractionated during transfer from plankton to krill as $^{12}$C  is preferentially lost through respiration. 
+The degree of such trophic fractionation is unclear, however and we do not draw interpretations based on absolute $\delta^{13}$C values, rather on the relative $\delta^{13}$C  values across the length of the baleen plate.
 
@@
 We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
-At each time point and location in the simulations, we extracted phytoplankton $\delta^{13}$C values from the spatio-temporal model\cite{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs 2, 3, and S4).
+At each time point and location in the simulations, we extracted time averaged $\delta^{13}$C values from the spatio-temporal model\cite{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figures \ref{fig2}, \ref{fig3}, and Supplementary Figure S4).
 Movement was coded as a set of probabilistic rules. 
 The parameters for these were taken from the literature on blue whale behaviour (e.g.\cite{handbook}). 
-All terms were expressed as probability distributions allowing us to introduce individual variability.
+All terms were expressed as probability distributions, yielding multiple potential movement tracks.
 
-In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and extent (1$^{\circ}$ or 2$^{\circ}$ in km) of movement are all influenced by the following.
-(i) Behavioural state (migrating north, migrating south, foraging, or nursing). % this has changed
+In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and linear distance of movement are all influenced by the following. 
+(i) Behavioural state (migrating north, migrating south, or foraging).
 Northerly migrations were possible only in spring, and southerly migrations in autumn. 
-Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton. Nursing occurred at temperatures between 18 and 25$^{\circ}$C only. 
+Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
 (ii) Sea surface temperature\cite{yool2013medusa} ($^{\circ}$C). 
@@
 
-We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life} ; Fig.
-S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
-For a full description of the model and its parameters see Supplementary Methods: Fig. S5 and Table S1. 
-We simulated whale movements 50 times for each scenario and 30 times for each residency hotspot (Fig. S4). 
+We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplementary Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
+Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four years of simulations.
+For a full description of the model and its parameters see Supplementary Methods: Supplementary Figure S5 and Supplementary Table S1. 
+We simulated whale movements 1000 times and 30 times for each residency hotspot (Supplementary Figure S4). 
 For each movement simulation we extracted phytoplankton $\delta^{13}$C values at each time point and location from the models described above\cite{magozzi2017using}. 
-We then compared the simulated stable isotope profiles (Fig. \ref{fig3}) to the profile of the blue whale (Fig. \ref{fig1}).
+We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).
 
 \subsection{Data Availability}
-Data are available from Zenodo doi: [DOI will be added on acceptance]. 
-R code is available from GitHub (github.com/ALJackson/blue-whale-bes).
+Data are available from the NHM Data Portal [DOI will be added on acceptance]. 
+R code is available from GitHub (github.com/nhcooper123/blue-whale-bes).
 
@@
 
-\section{Supplementary Information}
-Supplementary Information is linked to the online version of the paper at www.nature.com/nature.
+%\section{Supplementary Information}
 
@@
 
-\section{Author information}
-Reprints and permissions information is available at www.nature.com/reprints.
-The authors declare no competing financial interests.
-Correspondence and requests for materials should be addressed to natalie.cooper@nhm.ac.uk.
+%\section{Author information}
+%Reprints and permissions information is available at www.nature.com/reprints.
+%The authors declare no competing financial interests.
+%Correspondence and requests for materials should be addressed to natalie.cooper@nhm.ac.uk.
 

### 2018-01-23 movement-model/whale-movement-model.R
Add CT's movement model code

@@ -1,0 +1,951 @@
+# Whale individual based movement model
+# Author: Clive Trueman (trueman@noc.soton.ac.uk)
+# This version Sept. 2017
+# If you want to use this and get confused please email Clive :)
+
+
+# run the Lat.Lon code to get the km conversion first.
+# Degree to km function
+
+# Calculate distance in kilometers between two points
+earth.dist <- function (long1, lat1, long2, lat2)
+{
+	
+rad <- pi/180
+a1 <- lat1 * rad
+a2 <- long1 * rad
+b1 <- lat2 * rad
+b2 <- long2 * rad
+dlon <- b2 - a2
+dlat <- b1 - a1
+a <- (sin(dlat/2))^2 + cos(a1) * cos(b1) * (sin(dlon/2))^2
+c <- 2 * atan2(sqrt(a), sqrt(1 - a))
+R <- 6378.145
+d <- R * c
+return(d)
+}
+
+# build array for distance vectors
+Lat.Lon<-array(data=0, dim=c(180, 360,2))
+
+	#longitude distance varies with lat
+	Lt=vector(length=179)
+	count=1
+	
+	for(Lat in 89:-90){
+		x<-earth.dist(-180,Lat, -179,Lat)
+		Lt[count]<-x
+		count=count+1
+		}
+
+Lat.Lon[1:180,1,2]<-0
+	Lat.Lon[1:180,2,2]<-Lt
+	for(Lon in 3:359){
+	Lat.Lon[1:180,Lon,2]<-Lat.Lon[,Lon-1,2]+Lat.Lon[,2,2]
+	}
+	
+		#latitude distance constant at 111km
+
+	Lt2=rep(0,179)
+	count=2
+		for(Lat in -89:89){
+
+		Lt2[count]<-Lt2[count-1]+111
+		count=count+1
+		}
+
+Lat.Lon[,,1]<-Lt2
+
+
+Lon2<-Lat.Lon[,1:359,2]/Lat.Lon[,359,2]*360
+
+
+#########################################################
+
+library(R.matlab)
+library(maps)
+library(mapdata)
+library(scales)
+library(TTR)
+require(maptools)
+require(raster) 
+
+
+
+# link to monthly isoscape layers 
[…]

### 2018-01-23 manuscript/revision/blue-whale-supplementary-v2.tex
CT edits to paragraph 1

@@
 Fourier analysis (periodogram function in the R package TSA\cite{Chan:2012aa}) revealed a strong periodic repetition with a 13.3cm periodicity. 
-$\delta^{13}$C values also show repeated fluctuations in the first (oldest) 60cm of the baleen plate (behavioural phase one), with a similar periodicity of 13.5cm (Fig. \ref{figs1}). 
-A distinct change in the isotopic pattern along the baleen plate occurs at around 60cm behavioural phase two, with a transition to relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.
+The oldest 35cm of the baleen plate  record relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.
+$\delta^{13}$C values also show repeated fluctuations in the middle 50cm of the baleen plate (behavioural phase two), with a similar periodicity of 13.5cm (Fig. \ref{figs1}). 
+A distinct change in the isotopic pattern along the baleen plate occurs at around 84cm in behavioural phase two, with a transition to uniform, relatively low $\delta^{13}$C values followed by an abrupt transition to positive $\delta^{13}$C values and a decline in $\delta^{13}$C values in the last 3cm of the record.
+
+Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N  values within behavioural phase two (Fig.
+\ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioural phase one.
 

### 2018-01-23 manuscript/revision/blue-whale-supplementary-v2.tex
Edits to table s1

@@
 
-Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N  values within behavioural phase two (Fig.
-\ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioural phase one.
-
 % figure s1
@@
 
-Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase one (oldest 60cm of the baleen plate; Fig.
-\ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioural phase two (the final 20cm of the baleen plate).
+Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N  values within behavioural phase two (Fig.
+\ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioural phase one.
 
@@
 \subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase one}
-Temporal dynamics in $\delta^{13}$C values simulated in northerly latitudes are similar to the measured $\delta^{13}$C profile during the four years of behavioural phase one. 
-Simulated profiles display the same asymmetric isotopic cycles associated with development of the spring bloom, with annual $\delta^{13}$C maxima in June (Fig. \ref{figs4}, Fig. 3). 
-Simulated temporal isotopic profiles for northern resident whales, do not display clear isotopic minima preceding the onset of the bloom (Fig. \ref{figs4}). 
+Behavioural phase one is characterised by approximately 24 months of relatively positive and constant $\delta^{13}$C values. 
+$\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity is absent (Fig. \ref{figs1}). 
+Positive and temporally constant $\delta^{13}$C values are characteristic of subtropical regions particularly upwelling areas around Mauritania, Cape Verde and the Azores region.
+
+\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase two}
+Temporal dynamics in $\delta^{13}$C values simulated in northerly latitudes cannot be reproduced from simulated temporal trends in $\delta^{13}$C values in krill within static areas (Fig.
+\ref{figs4}, Fig.
+3).
 Adding seasonal north-south migrations within mid-high latitude regions to foraging models, closely matches the measured profile, including $\delta^{13}$C minima preceding the June bloom (Fig.
@@
 
-\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase two}
-Behavioural phase two is characterised by a dramatic shift to maximum $\delta^{13}$C values in spring 1889 followed by approximately 18 months of relatively positive and constant $\delta^{13}$C values.
-$\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity associated with north-south migration is absent in 1890 (Fig. 1). 
-The final sample representing feeding shortly before death in Irish waters implies a return to isotopic conditions more consistent with behavioural phase one. 
-The relatively positive and constant carbon isotope composition of behavioural phase two is consistent with residency (and continued feeding) in more southerly latitudes, particularly in the relatively isotopically positive waters southwest of the Azores.
-
-Accordingly we modeled whale movements allowing five years of seasonal north-south migration in the north east northern Atlantic, followed by rapid migration to subtropical waters in the late spring of 1889, relative residency for eight months and rapid migration returning to more northerly waters in the spring of 1891.
+Accordingly we modeled whale movements allowing three years of residency in warm waters of the Azores or Cape Verde/Mauritania region followed by a further 3.5 years of seasonal north-south migration in the northeast northern Atlantic, followed by residency in subtropical waters in 1890 and rapid migration returning to more northerly waters in the spring of 1891.
 The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Fig. 3). 
-Consistently low $\delta^{15}$N values in 1890 are also consistent with residency in relatively southern waters.
-
-The dramatic positive excursion in $\delta^{13}$C values seen in the late spring and early summer of 1889 cannot be replicated purely through movement to warmer waters, thus we tentatively ascribe this signal to the physiological demands of pregnancy.
-A captive feeding study in south American sea lions\cite{cardona2017temporal} showed positive $\delta^{13}$C values associated with pregnancy, falling at birth, and a gradual increase in $\delta^{15}$N  values during lactation.
-$\delta^{15}$N values in whale baleen during the autumn of 1890 show a gradual increase, consistent with the isotopic expression associated with lactation in south American fur seals\cite{cardona2017temporal}.
-
-Blue whales have a gestation period of 10-12 months, and calves suckle for approximately 6-7 months\cite{handbook}. 
+Blue whales have a gestation period of 10-12 months, and calves suckle for approximately 6-7 months \cite{handbook}. 
 The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer and autumn 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the assumed reproductive ecology of blue whales.
@@
 
-The behavioural states employed in the model simulations are foraging, migrating north, migrating south and nursing.
+The behavioural states employed in the model simulations are foraging, migrating north and migrating south.
 When foraging, the probability of movement and likely direction of movement is primarily controlled by the biomass of phytoplankton in the current and adjacent 1$^{\circ}$ cells.
 When migrating, a fixed higher probability of movement is given to southerly or northerly movement directions, and the probability of movement is influenced by the ambient sea surface temperature and phytoplankton biomass.
-When nursing a calf, movement is limited to a daily mean of 25km, as long as sea surface temperatures remain between 18$^{\circ}$C and 25$^{\circ}$C.
 

### 2018-01-23 manuscript/revision/blue-whale-supplementary-v2.tex
Add CT edits for table s1

@@
 \begin{table}
-  \begin{tabular}{|p{5cm}|c|c|c|c|} 
+  \begin{tabular}{|p{5cm}|c|c|c|} 
     \hline
-    & \multicolumn{4}{|c|}{\textbf{Behavioural state}} \\
+    & \multicolumn{3}{|c|}{\textbf{Behavioural state}} \\
     \hline
-    \textbf{Parameters} & \textbf{Foraging} & \textbf{Migrating north} & \textbf{Migrating south} & \textbf{Nursing}\\
+    \textbf{Parameters} & \textbf{Foraging} & \textbf{Migrating north} & \textbf{Migrating south}\\
     \hline
-    Month - behavioural phase one (1460 days) & May - Oct & Feb - April & Nov - Jan & NA\\
+    Behavioural phase one (1095 days) & Jan - Dec & NA & NA\\
     \hline
-    Month - pregnancy (365 days) & June - Oct & April - May & Nov - March & NA\\
+    Behavioural phase two (1278 days) & June - Oct & April - May & Nov - March\\
     \hline
-    Month - nursing (210 days) & NA & NA & NA & June - Dec\\
+    Pregnancy, nursing, return (447 days) & Jan - Dec & NA & NA\\
     \hline
-    Minimum sea surface temperature ($^{\circ}$C) & 3 & 3 & 3 & 18\\
+    Post nursing migration (100 days) & NA & Jan - June & NA\\
     \hline
-    Maximum sea surface temperature ($^{\circ}$C) & 25 & 25 & 25 & 25\\
+    Minimum sea surface temperature$^*$ ($^{\circ}$C) & 3 & 3 & 3\
     \hline
-    Maximum daily distance traveled (km) & $N(\mu=30, \sigma=10)$ & $N(\mu=150, \sigma=10)$ & $N(\mu=150, \sigma=10)$ & $N(\mu=20, \sigma=5)$\\
+    Maximum sea surface temperature ($^{\circ}$C) & 25 & 25 & 25\\
     \hline
-    Plankton biomass threshold for remaining to forage (units) & 50 & NA & NA & NA\\
+    Maximum daily distance traveled (km)$^**$ & $N(\mu=50, \sigma=25)$ & $N(\mu=150, \sigma=75)$ & $N(\mu=150, \sigma=75)$\\
+    \hline
+    Plankton biomass threshold for remaining to forage (units) & 50 & NA & NA\\
     \hline
   \end{tabular}
-  \caption{Model parameters for agent-based model of whale movement. See Fig. \ref{figs5} and text for more details.}
+  \caption{Model parameters for agent-based model of whale movement. 
+  See Fig. \ref{figs5} and text for more details. 
+  $^*$During nursing minimum sea surface temperature is set to 18$^{\circ}$C. 
+  $^**$During behavioural phase one, mean maximum daily distance is set to 20km.}
   \label{tables1}

### 2018-01-23 manuscript/revision/blue-whale-ms-v2.tex
Add data portal location ref and doi

@@
   \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis).
-Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes).
+Samples were taken longitudinally through the baleen plate (n = 97 samples from a single baleen plate for both isotopes).
 There is strong annual periodicity and cross-correlation (Supplementary Figures S1 and S2) in both isotopes.
@@
 C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples\cite{hobson1998stable}.
+All data are available from the NHM Data Portal\cite{data-set}.
 
@@
 For a full description of the model and its parameters see Supplementary Methods: Supplementary Figure S5 and Supplementary Table S1. 
-We simulated whale movements 1000 times and 30 times for each residency hotspot (Supplementary Figure S4). 
+We simulated whale movements 1000 times and 30 times for each residency hotspot (Supplementary Figure S4). % check this
 For each movement simulation we extracted phytoplankton $\delta^{13}$C values at each time point and location from the models described above\cite{magozzi2017using}. 
@@
 \subsection{Data Availability}
-Data are available from the NHM Data Portal [DOI will be added on acceptance]. 
+Data are available from the NHM Data Portal [doi.org/10.5519/0093278](https://doi.org/10.5519/0093278). 
 R code is available from GitHub (github.com/nhcooper123/blue-whale-bes).
-
+{}
 % References

### 2018-01-23 manuscript/revision/blue-whale-supplementary-v2.tex
Typo on table
Add data portal location ref and doi

@@
 The direction of movement is sampled based on discrete probabilities associated to each of the eight adjacent 1$^{\circ}$ cells, defined according to behavioural state and environmental conditions.
-Finally the distance moved is sampled from a Poisson distribution conditional on the max distance defined for each individual.
+Finally the distance moved is sampled from a Poisson distribution conditional on the maximum distance defined for each individual.
 The new location is then assigned from a linear vector originating at the current location with magnitude defined as the distance moved and direction as 45$^{\circ}$ increments reflecting the relative orientation of each of the eight potential adjacent cells.}
@@
 \begin{table}
-  \begin{tabular}{|p{5cm}|c|c|c|} 
+  \begin{tabular}{|p{8cm}|c|c|c|} 
     \hline
@@
     \hline
-    Minimum sea surface temperature$^*$ ($^{\circ}$C) & 3 & 3 & 3\
+    Minimum sea surface temperature$^*$ ($^{\circ}$C) & 3 & 3 & 3\\
     \hline
@@
     \hline
-    Maximum daily distance traveled (km)$^**$ & $N(\mu=50, \sigma=25)$ & $N(\mu=150, \sigma=75)$ & $N(\mu=150, \sigma=75)$\\
+    Maximum daily distance traveled (km)$^{**}$ & $N(\mu=50, \sigma=25)$ & $N(\mu=150, \sigma=75)$ & $N(\mu=150, \sigma=75)$\\
     \hline
@@
   $^*$During nursing minimum sea surface temperature is set to 18$^{\circ}$C. 
-  $^**$During behavioural phase one, mean maximum daily distance is set to 20km.}
+  $^{**}$During behavioural phase one, mean maximum daily distance is set to 20km.}
   \label{tables1}
@@
 \end{landscape}
+
+\newpage
+
+\subsection*{ADDITIONAL RESULTS FIGURES}
+
+  \begin{figure}[!htbp]
+    \centering
+      \includegraphics[width=\linewidth]{figures/Figure-S5-r2.png}
+      \caption{Histogram of $r^2$ values from linear regressions comparing model simulation $\delta^{13}$C values to those of the real data (Fig. 1).} % check legend
+      \label{figs6}
+  \end{figure}
+
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[width=\linewidth]{figures/Figure-S6-boxplots.png}
+      \caption{Boxplots showing the maximum and standard devations of latitudinal positions of simulated whales taken from the top 10\%, and bottom 10\%, of movement model simulations.} % check legend
+      \label{figs7}
+  \end{figure}
+
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[width=\linewidth]{figures/Figure-S7-kernals.png}
+      \caption{Kernal density plots showing the positions of simulated whales taken from the top 10\%, and bottom 10\%, of movement model simulations. 
+      High densities are in yellow, low densities are in blue.} % check legend
+      \label{figs8}
+  \end{figure}
 

### 2018-01-24 manuscript/revision/blue-whale-supplementary-v2.tex
CT edits to supp mat

@@
 \subsection*{CHARACTERISATION OF THE MEASURED BALEEN}
-The baleen plate yielded 78 discrete samples of baleen. 
+The baleen plate yielded 97 discrete samples of baleen. 
 Across the baleen sample, $\delta^{15}$N values show clear cycles with periods of relatively high $\delta^{15}$N values interspersed by relatively short periods where $\delta^{15}$N values are c.1\text{\textperthousand} lower.
 Fourier analysis (periodogram function in the R package TSA\cite{Chan:2012aa}) revealed a strong periodic repetition with a 13.3cm periodicity. 
-The oldest 35cm of the baleen plate  record relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.
+The oldest (most distal) 35cm of the baleen plate  record relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.
 $\delta^{13}$C values also show repeated fluctuations in the middle 50cm of the baleen plate (behavioural phase two), with a similar periodicity of 13.5cm (Fig. \ref{figs1}). 
-A distinct change in the isotopic pattern along the baleen plate occurs at around 84cm in behavioural phase two, with a transition to uniform, relatively low $\delta^{13}$C values followed by an abrupt transition to positive $\delta^{13}$C values and a decline in $\delta^{13}$C values in the last 3cm of the record.
+A distinct change in the isotopic pattern along the baleen plate occurs at around 84cm from the distal end in behavioural phase two, with a transition to uniform, relatively low $\delta^{13}$C values followed by an abrupt transition to positive $\delta^{13}$C values and a decline in $\delta^{13}$C values in the most recent (most gingival) 3cm of the record.
 
@@
 In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. \ref{figs3}). 
-The temporally coincident periodic, inverse fluctuations in $\delta^{13}$C and $\delta^{15}$N values in the initial phase of the baleen plate are therefore consistent with migration between waters north of 60$^{\circ}$N and more southerly southern waters in the North Atlantic, with northerly occupancy in summer months and southerly occupancy in winter months.
+The temporally coincident periodic, inverse fluctuations in $\delta^{13}$C and $\delta^{15}$N values in the second phase of the baleen isotopic record are therefore consistent with migration between waters north of 60$^{\circ}$N and more southerly southern waters in the North Atlantic, with northerly occupancy in summer months and southerly occupancy in winter months.
 Incorporation of carbon and nitrogen with isotopic ratios consistent with southerly latitudes during winter months indicates continued feeding during migration and winter residency.
@@
 \subsection*{ISOTOPE MODEL SIMULATIONS}
-We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions that would be encountered by whales foraging within broad geographic regions (west UK/Irish shelf, Norwegian/Barents Sea, Canaries/west Azores, Mid Atlantic ridge; Fig.
+We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions that would be encountered by whales foraging within broad geographic regions (west UK/Irish shelf, Norwegian/Barents Sea, Canaries/west Azores, Mid Atlantic ridge, Cape Verde/Mauritanean upwelling area; Fig.
 \ref{figs4}).
@@
 In temperate latitudes around the British Isles, seasonal dynamic cycles are present, but dampened (Fig.
-\ref{figs4}), whereas in subtropical regions exemplified by the Canaries and western Azores/Mid Atlantic ridge areas, $\delta^{13}$C values are relatively positive and constant throughout the year (Fig.
+\ref{figs4}), whereas in subtropical regions exemplified by the Canaries, Azores and particularly Mauritanean upwelling areas, $\delta^{13}$C values are relatively positive and constant throughout the year (Fig.
 \ref{figs4}).

### 2018-01-24 manuscript/revision/blue-whale-ms-v2.tex
Replace 1000 sims with 1049

@@
 
-We simulated 1000 individual movement patterns and compared the resulting simulated baleen $\delta^{13}$C records to the measured records with simple linear regressions. 
-Simulated baleen $\delta^{13}$C profiles produce a good fit to measured profiles, the median $r^2$ value across 1000 simulated profiles was 0.49, and the maximum was 0.74 (Supplementary Figure S6).
-% update these values
+We simulated 1049 individual movement patterns and compared the resulting simulated baleen $\delta^{13}$C records to the measured records with simple linear regressions. 
+There are 1049 simulations because we ran 1500 simulations, then excluded simulations where the virtual whale stranded before reaching the 3019 days represented in the baleen record.
+Simulated baleen $\delta^{13}$C profiles produce a good fit to measured profiles, the median $r^2$ value across 1049 simulated profiles was 0.49, and the maximum was 0.71 (Supplementary Figure S6).
 The top 10\% best fitting simulated profiles are shown in Figure \ref{fig2}. 
@@
 Behavioural phase two is best simulated by seasonal migrations between summer foraging in northern areas in the Norwegian Sea/Barents Sea/Iceland region, and winter foraging in a broad region between the UK and more southerly, subtropical waters.
-Better fitting models in general were those predicting a greater latitudinal foraging range, and foraging in more northerly waters (Supplementary Figure S7 [boxplots and kernel density plots]).
-% what is fig S7?
+Better fitting models in general were those predicting a greater latitudinal foraging range, and foraging in more northerly waters (Supplementary Figure S7).
 Best-fit model distributions are largely consistent with our current understanding of blue whale distributions in the northeast Atlantic\cite{reeves2004historical,baines2014upwellings,baines2017autumn,reeves2004historical}, with perhaps greater importance of winter foraging in temperate regions (Figure \ref{fig4}).
@@
 
-We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplementary Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
+We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanean upwelling area\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplementary Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
 Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four years of simulations.
 For a full description of the model and its parameters see Supplementary Methods: Supplementary Figure S5 and Supplementary Table S1. 
-We simulated whale movements 1000 times and 30 times for each residency hotspot (Supplementary Figure S4). % check this
+We simulated whale movements 1500 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.
 For each movement simulation we extracted phytoplankton $\delta^{13}$C values at each time point and location from the models described above\cite{magozzi2017using}. 

### 2018-01-24 manuscript/revision/blue-whale-ms-v2.tex
finessing of 1000 simulations issue

@@
 
-We simulated 1049 individual movement patterns and compared the resulting simulated baleen $\delta^{13}$C records to the measured records with simple linear regressions. 
-There are 1049 simulations because we ran 1500 simulations, then excluded simulations where the virtual whale stranded before reaching the 3019 days represented in the baleen record.
+We simulated 1200 individual movement patterns, then excluded simulations where the virtual whale stranded before reaching the 3019 days represented in the baleen.
+With then compared the remaining 1049 simulations and the resulting simulated baleen $\delta^{13}$C records to the measured records with simple linear regressions. 
 Simulated baleen $\delta^{13}$C profiles produce a good fit to measured profiles, the median $r^2$ value across 1049 simulated profiles was 0.49, and the maximum was 0.71 (Supplementary Figure S6).
@@
 For a full description of the model and its parameters see Supplementary Methods: Supplementary Figure S5 and Supplementary Table S1. 
-We simulated whale movements 1500 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.
+We simulated whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.
 For each movement simulation we extracted phytoplankton $\delta^{13}$C values at each time point and location from the models described above\cite{magozzi2017using}. 

### 2018-01-24 code/Figure-S4-code.R
Edits to FigS4 code with new data added

@@ -1,7 +1,7 @@
 # Author: Clive Trueman
 # Adapted by: Andrew Jackson 19-Jun-2017
-# About: script to generate figures S4a-d for the blue whale paper
-# Tidied by Natalie Cooper Nov 2017
+# About: script to generate figures S4 for the blue whale paper
+# Tidied by Natalie Cooper Jan 2017
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 
 library(tidyverse)
@@ -13,16 +13,54 @@ library(lubridate)
 cnst <- 6
 
 # Read in resident whale model data
-resNor <- read.csv("data/Norway.csv", header = TRUE)
-resIre <- read.csv("data/west.Ireland.csv", header = TRUE)
-resCan <- read.csv("data/Canaries.csv", header = TRUE)
-resMidAtl <- read.csv("data/Mid_Atl.csv", header = TRUE)
+resNor <- read.csv("data/Norway.Res.TL.csv", header = TRUE)
+resIre <- read.csv("data/Ireland.Res.TL.csv", header = TRUE)
+resCan <- read.csv("data/Canaries.Res.TL.csv", header = TRUE)
+resMidAtl <- read.csv("data/Atl.Res.TL.csv", header = TRUE)
+resCV <- read.csv("data/CV.TL.csv", header = TRUE)
+#resMaur <- read.csv("data/Maur.TL.csv", header = TRUE)
+
+# Remove B.state variable as it messes up bind.rows
+resNor <- 
+  resNor %>%
+  select(-B.state)
+
+resNor <- 
+  resNor %>%
+  select(-B.state)
+
+resIre <- 
+  resIre %>%
+  select(-B.state)
+
+resCan <- 
+  resCan %>%
+  select(-B.state)
+
+resMidAtl <- 
+  resMidAtl %>%
+  select(-B.state)
+
+#resMaur <- 
+#  resMaur %>%
+#  select(-B.state)
+
+# Here count is correct, but Day.No is not
+# so switch them to get the next steps to work
+# Also shift up start to 120 days to match others
+resCV <- 
+  resCV %>%
+  select(-B.state, -Day.No) %>%
+  mutate(Day.No = count + 119)
+  
 
 # Combine datasets
 res_all <- bind_rows(list(Norway = resNor, 
                           Ireland = resIre, 
                           Canaries = resCan, 
-                          "Mid-Atlantic" = resMidAtl), .id = "Region")
[…]

### 2018-01-24 quarantine/fig.s4 code.R
CT code for Fig S4

@@ -1,0 +1,119 @@
+setwd("/Users/trueman/Desktop/R scripts/migrate.files")
+library(raster)
+
+
+##########read resident whale model files with the trophic attenuation d13C values
+setwd("/Users/trueman/Desktop/R scripts/migrate.files/Model.files/blue.res")
+
+resNor<-read.csv("Norway.Res.TL.csv", header=TRUE)
+resIre<-read.csv("Ireland.Res.TL.csv", header=TRUE)
+resCan<-read.csv("Canaries.Res.TL.csv", header=TRUE)
+resMidAtl<-read.csv("Atl.Res.TL.csv", header=TRUE)
+resCV<-read.csv("CV.TL.csv", header=TRUE)
+
+
+#remove samples with NAs
+resNor<-na.omit(resNor)
+resIre<-na.omit(resIre)
+resCan<-na.omit(resCan)
+resMidAtl<-na.omit(resMidAtl)
+resCV<-na.omit(resCV)
+
+
+
+#plot loess fits to the 30 resident simulated whales for each region
+
+
+
+setEPS()
+postscript("resident.eps", width=10, height=5)
+par(new=FALSE)
+par(mfrow=c(2,3))
+
+
+
+#norway resident
+
+#get series length
+inds<-length(unique(resNor$Rep))
+
+#run loess on day number
+lo<-loess(resNor$TL4[resNor$Rep==1]+6~resNor$Day.No[resNor$Rep==1], span=0.1)
+
+# plot blank
+plot(x=resNor$Day.No[resNor$Rep==1], y=resNor$TL4[resNor$Rep==1], ylim=c(-23,-12), main="Norwegian Sea" , type="p", cex=0, col="red", xlab="Day.No", ylab="d13C")
+#add loess line
+lines(resNor$Day.No[resNor$Rep==1],predict(lo), col="red", lwd=0.2)
+
+par(new=TRUE)
+# run loess and plot for each of the 30 simulations 
+
+for(w in 2:inds){
+	lo<-loess(resNor$TL4[resNor$Rep==w]+6~resNor$Day.No[resNor$Rep==w], span=0.1)
+lines(resNor$Day.No[resNor$Rep==w],predict(lo), col="red", lwd=0.2)
+
+par(new=TRUE)
+}
+par(new=FALSE)
+
+
[…]

### 2018-01-24 code/Figure-S4-code.R
Edits to names of regions to match CT code

@@ -55,8 +55,8 @@ resCV <-
   
 
 # Combine datasets
-res_all <- bind_rows(list(Norway = resNor, 
-                          Ireland = resIre, 
+res_all <- bind_rows(list("Norwegian Sea" = resNor, 
+                          "West Ireland" = resIre, 
                           Canaries = resCan, 
                           "Mid-Atlantic" = resMidAtl,
                           #"Mauritania" = resMaur,
@@ -77,7 +77,7 @@ tmp2 <- res_all %>% group_by(Region, Rep)  %>%
                    Day.No = .$Day.No))
 
 # re-order the Region Factor
-tmp2$Region <- factor(tmp2$Region, levels = c("west Ireland","Norwegian Sea",
+tmp2$Region <- factor(tmp2$Region, levels = c("West Ireland","Norwegian Sea",
                                               "Canaries","Mid-Atlantic",
                                               #"Mauritania",
                                               "Cape Verde"))
@@ -86,7 +86,7 @@ tmp2$Region <- factor(tmp2$Region, levels = c("west Ireland","Norwegian Sea",
 sim_facet <- 
   ggplot(tmp2, aes(Day.No, Z, group = Rep)) + 
     geom_line(col = viridis(3)[2], alpha = 0.25) + 
-    xlab("Time") + 
+    xlab("Time (days)") + 
     ylab(expression(paste(delta^{13}, "C (\u2030)"))) + 
     theme_classic(base_size = 14) + ylim(-25, -14) + 
     theme(strip.background = element_blank(),

### 2018-01-24 manuscript/revision/blue-whale-ms-v2.tex
add CT edits to main text

@@
 Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}, but interpreting these records can be challenging.
-Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct seven years of movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
+Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct seven years of movements of an iconic female blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
 We provide the longest continuous record of movement for an individual blue whale in the north Atlantic. 
 We show that the whale spent at least two years resident in sub-tropical waters as a juvenile, before commencing annual latitudinal migrations consistent with patterns of movements of blue whales in the northeast Atlantic inferred from observations, tagging and historic whaling data.
-The isotopic record is also suggestive of pregnancy and calf-rearing in the year before death. 
+The isotopic record is consistent with observations of opportunistic feeding throughout the year in blue whales, and suggests pregnancy and calf-rearing in the year before death.
 Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climate, ecological conditions and anthropogenic pressures, and reveal aspects of life history.
@@
 Despite hunting for blue whales being banned for over 50 years, blue whales are still listed as Endangered on the IUCN Red List\cite{reilly2008balaenoptera}.
-Blue whales in the northeast Atlantic appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations (REF).  
+In the northeast Atlantic, blue whales appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations (REF).  
 The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland\cite{pike2009note}.
-Their small population size and oceanic habit make blue whale movements extremely hard to study. 
+Following movements of individual blue whales is extremely challenging due to their scarcity and oceanic habit.
 Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
@@
 Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas\cite{pike2009note}.
-Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals\cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
+Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southward-migrating animals\cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
 While direct tagging methods are adding vital behavioural data to our understanding of whale movements\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
@@
 The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
-Stable isotope data therefore provide information on diet and location, but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values.
+Stable isotope data therefore provide information on diet and location (i.e.
+`trophic geography'\cite{bird2018global}), but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values.
 Here we reconstruct movements of an historic whale by comparing isotopic compositions of samples from museum collections, rich archives of behavioural information\cite{lister2011natural}, to spatially referenced models of isotopic variation in diet across the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
@@
 Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable}. 
-Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although baleen is worn away at the tips over time, so rarely records an individual's entire lifespan.
+Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an iindividual whale, although as baleen is worn away at the tips over time, an individual's entire lifespan is rarely recorded.
 Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales\cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}, but have yet to be applied to blue whales from the north Atlantic population.
 Here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale ``Hope''. 
-Hope is estimated to be a young adult c.15 years at death, suggesting that her baleen may yield information about juvenile behaviours in north Atlantic blue whales around the peak of industrial whaling.
+Hope is female, and estimated to be a young adult c.15 years at death, suggesting that her baleen may yield information about juvenile behaviours in north Atlantic blue whales around the peak of industrial whaling.
 
@@
 In behavioural phase one (from the start of the record in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Figure \ref{fig1}; Supplementary Figures S1 and S2).
-In behavioural phase two (summer 1886 to spring 1890) $\delta^{15}$N values are relatively elevated and $\delta^{13}$C values are relatively depleted with coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values.
-In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively positive $\delta^{13}$C values in the second half of 1891.
+In behavioural phase two (summer 1886 to spring 1890) $\delta^{15}$N values are relatively high and $\delta^{13}$C values are relatively low with coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values.
+In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively high $\delta^{13}$C values in the second half of 1891.
 The final three months of the record show a progressive fall in $\delta^{13}$C values (Figure \ref{fig1}; Supplementary Figures S1 and S2).
@@
 
-Relating marine animal tissue isotopic compositions to likely locations is complicated by a lack of knowledge of spatio-temporal variations in the isotopic composition of diet\cite{west2006stable,mcmahon2015millennial}.
+Relating the isotopic compositions of animal tissues to likely locations is complicated by a lack of knowledge of spatio-temporal variations in the isotopic composition of diet\cite{west2006stable,mcmahon2015millennial}.
 We develop a new approach to address this problem using temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Supplementary Figure S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Supplementary Table S1; Supplementary Figure S5).
@@
 Blue whales are commonly assumed to feed mainly in summer in productive, high latitude regions and rely largely on stored lipid reserves to meet metabolic demands in lower latitudes in winter. 
-However, in winter blue whales are associated with upwellings\cite{baines2017autumn}, while northward migrations in spring coincide with the onset and development of phytoplankton blooms\cite{silva2013north,visser2011timing,busquets2017estimating}, strongly implying feeding throughout the year.
+However, in winter blue whales in the northeast Atlantic are associated with upwellings\cite{baines2017autumn}, while northward migrations in spring coincide with the onset and development of phytoplankton blooms\cite{silva2013north,visser2011timing,busquets2017estimating}, strongly implying feeding throughout the year.
 We therefore assume opportunistic ingestion of protein throughout the year.
@@
 
-The relatively positive and seasonally-invariant $\delta^{13}$C values seen during behavioural phase one are only found in subtropical areas of the North Atlantic.  
+The relatively positive and seasonally-invariant $\delta^{13}$C values seen during behavioural phase one are only found in sub-tropical areas of the North Atlantic.  
 Our simulations show high variance during this period (Figure \ref{fig3}), reflecting a range of possible locations for the whale, although areas around the Mauritanian coast and Cape Verde Islands, a known current and historic winter feeding area for blue whales \cite{baines2014upwellings,reeves2004historical}, and potentially to the west of the Azores (Figure \ref{fig2}), most closely match the measured profile.
-The whale remained in these warm waters for at least two full years. 
+The whale remained in relatively warm waters for at least two full years. 
 Hope was estimated to be 15 years old when she died (based on vertebral epiphyseal fusion; R.Sabin \textit{pers.comm}), so was approximately 3-5 years old during this period.
@@
 Best fitting models predict juvenile residency in the Cape Verde region in behavioural phase one. 
-Behavioural phase two is best simulated by seasonal migrations between summer foraging in northern areas in the Norwegian Sea/Barents Sea/Iceland region, and winter foraging in a broad region between the UK and more southerly, subtropical waters.
+Behavioural phase two is best simulated by seasonal migrations between summer foraging in northern areas in the Norwegian Sea/Barents Sea/Iceland region, and winter foraging in a broad region between the UK and more southerly, sub-tropical waters.
 Better fitting models in general were those predicting a greater latitudinal foraging range, and foraging in more northerly waters (Supplementary Figure S7).
@@
 
-The last 500 days of the whale's life are difficult to simulate. 
-Beginning in the winter of 1889/1890, the observed $\delta^{13}$C profile is relatively depleted, and remains constant for approximately six months, before increasing rapidly in the winter of 1890/1891.
-Depleted $\delta^{13}$C values are found in northern waters, but these regions also show large temporal fluctuations in $\delta^{13}$Cplk values (Supplementary Figure S4), and the observed values cannot be simulated purely from movements within the known geographic range.
+The isotopic record of the last 500 days of the whale's life are difficult to simulate. 
+Beginning in the winter of 1889/1890, the observed $\delta^{13}$C values are relatively low, and remain constant for approximately six months, before increasing rapidly in the winter of 1890/1891. 
+Low $\delta^{13}$C values are found in northern waters, but these regions also show large temporal fluctuations in $\delta^{13}$Cplk values (Supplementary Figure S4), and the observed values cannot be simulated purely from movements within the known geographic range.
 We infer that the $\delta^{13}$C record in the last period of life reflects pregnancy and calf-rearing. 
@@
 After six months the whale began ingesting protein with high $\delta^{13}$C  values, indicating feeding in low latitude waters shortly followed by a final northward migration. 
-Blue whales have a 10-12 month gestation period, with calving occurring in subtropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
+Blue whales have a 10-12 month gestation period, with calving occurring in sub-tropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
 We therefore interpret the final year of the isotopic record as indicating: (a) birth of a calf in the winter of 1889/1890, presumably in sub-tropical waters, (b) 6 months of weaning during which time the maternal whale did not feed, followed by (c) a short period of feeding in sub-tropical waters in the second half of 1890, potentially during a final northward migration, and (d) eventual stranding during the return to northern feeding grounds in early 1891.
@@
 Our analyses suggest that between these points, juvenile North Atlantic blue whales may spend extended periods in warm, highly productive waters associated with upwelling systems before migrating to higher latitude feeding grounds.
-This has implications for blue whale conservation, as the upwelling regions around Mauritania and the Cape Verde islands are areas with emerging offshore oil and gas exploration and are fishing, tourist and shipping hotspots, increasing threats to juvenile whales.
+Extended periods of juvenile residency in sub-tropical waters could have implications for blue whale conservation in the northeast Atlantic, as the upwelling regions around Mauritania and the Cape Verde islands are areas with emerging offshore oil and gas exploration and are fishing, tourist and shipping hotspots, potentially increasing threats to juvenile whales.
 
 Whaling was an intense pressure for blue whales during the period we are analysing. 
-Before whaling in the North Atlantic began in 1868\cite{reilly2008balaenoptera}, there were 10,000-15,000 blue whales in the region\cite{sigurjonsson1995life}. 
+Before whaling in the North Atlantic began in 1868\cite{reilly2008balaenoptera}, there were an estimated 10,000-15,000 blue whales in the region\cite{sigurjonsson1995life}. 
 In the early 20th century, fisheries moved outside the area because stocks were so depleted\cite{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed\cite{sigurjonsson1995life}.
-The NHM blue whale thus represents an individual from a species at the brink of extinction.
+The NHM blue whale thus represents an individual from a species at the brink of extirpation.
 Despite dramatic changes in anthropogenic pressures since 1891, the NHM whale and present-day blue whales in the North Atlantic have similar spatial ecologies; there are still populations of blue whales in the regions indicated by our model.
@@
 {\raise.17ex\hbox{$\scriptstyle\sim$}}1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
-97 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section that contains the most recent tissue. 
+97 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal (gingival) section that contains the most recent tissue. 
 Baleen grows at a constant rate, so the samples are equally spaced through time\cite{best1996stable}. 
-Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the National Oceanographic Centre (Southampton, UK), using a Vario Isotope select elemental analyser, coupled with Isoprime 100 isotope mass spectrometer.
+Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the University of Southampton SEAPORT Stable Isotope Ratio Mass Spectrometry Laboratory (Southampton, UK), using a Vario Isotope select elemental analyser, coupled to an Isoprime 100 isotope mass spectrometer.
 Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
@@
 When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.
-(iii) Water depth (m; from the General Bathymetric Chart of the Oceans (GEBCO) global bathymetry dataset). 
+(iii) Water depth\cite{bathy} (m; extracted via the marmap package in R\cite{marmap}). 
 Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
@@
 
-We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanean upwelling area\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplementary Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
+We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplementary Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and sub-tropical latitudes.
 Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four years of simulations.
@@
 This work was funded by the British Ecological Society (grant: 5771/6815). 
-We thank C.J. Somes for providing  $\delta^{15}$N POM data.
+We thank C.J.
+Somes for providing  $\delta^{15}$N POM data, and Bastian Hambach and Megan Spencer at the University of Southampton SEAPORT isotope laboratory for assistance with stable isotope analyses.
 

### 2018-01-24 manuscript/revision/blue-whale-supplementary-v2.tex
Add some line breaks
Move link to flowchart fig

@@
     \centering
-      \includegraphics[width=\linewidth]{figures/Figure-S4-facet-wrap-d13C.png}
-      \caption{Alternative $\delta^{13}$C series arising from resident-only movement models restricting region to west UK/Irish shelf (Ireland), Norwegian/Barents Sea (Norway), Canaries/west Azores (Canaries), Mid Atlantic ridge (Mid-Atlantic), over a simulated two year period with daily resolution and 30 replicate simulations for each region.} % check legend
+      \includegraphics[width=\linewidth]{figures/Figure-S4-regions-d13C.png}
+      \caption{Alternative $\delta^{13}$C series arising from resident-only movement models restricting region to west UK/Irish shelf (Ireland), Norwegian/Barents Sea (Norway), Canaries/west Azores (Canaries), Mid Atlantic ridge (Mid-Atlantic), Cape Verde, and the Cape Verde/Mauritanian upwelling area over a simulated two year period with daily resolution and 30 replicate simulations for each region.} % check legend
       \label{figs4}
@@
   \centering
-  \includegraphics[trim = 0cm 145mm 0cm 4cm, clip]{figures/flow-diagram-model.pdf}
+  \includegraphics[trim = 0cm 145mm 0cm 4cm, clip]{revision/figures/flow-diagram-model.pdf}
   \vspace{-1.2cm}

### 2018-01-25 code/Figure-S4-code.R
Edits to remove incorrect region

@@ -18,7 +18,6 @@ resIre <- read.csv("data/Ireland.Res.TL.csv", header = TRUE)
 resCan <- read.csv("data/Canaries.Res.TL.csv", header = TRUE)
 resMidAtl <- read.csv("data/Atl.Res.TL.csv", header = TRUE)
 resCV <- read.csv("data/CV.TL.csv", header = TRUE)
-#resMaur <- read.csv("data/Maur.TL.csv", header = TRUE)
 
 # Remove B.state variable as it messes up bind.rows
 resNor <- 
@@ -41,26 +40,19 @@ resMidAtl <-
   resMidAtl %>%
   select(-B.state)
 
-#resMaur <- 
-#  resMaur %>%
-#  select(-B.state)
-
 # Here count is correct, but Day.No is not
 # so switch them to get the next steps to work
-# Also shift up start to 120 days to match others
 resCV <- 
   resCV %>%
   select(-B.state, -Day.No) %>%
-  mutate(Day.No = count + 119)
+  rename(Day.No = count)
   
-
 # Combine datasets
 res_all <- bind_rows(list("Norwegian Sea" = resNor, 
                           "West Ireland" = resIre, 
                           Canaries = resCan, 
                           "Mid-Atlantic" = resMidAtl,
-                          #"Mauritania" = resMaur,
-                          "Cape Verde" = resCV), .id = "Region")
+                          "Cape Verde/Mauritania" = resCV), .id = "Region")
 
 # Remove lines with no d13C values
 res_all <- res_all[which(!is.na(res_all$d13C)), ]
@@ -79,8 +71,7 @@ tmp2 <- res_all %>% group_by(Region, Rep)  %>%
 # re-order the Region Factor
 tmp2$Region <- factor(tmp2$Region, levels = c("West Ireland","Norwegian Sea",
                                               "Canaries","Mid-Atlantic",
-                                              #"Mauritania",
-                                              "Cape Verde"))
+                                              "Cape Verde/Mauritania"))
 
 # Plot and facet wrap by region
 sim_facet <- 

### 2018-01-25 code/Figure-S4-code.R
Remove excess code

@@ -17,13 +17,9 @@ resNor <- read.csv("data/Norway.Res.TL.csv", header = TRUE)
 resIre <- read.csv("data/Ireland.Res.TL.csv", header = TRUE)
 resCan <- read.csv("data/Canaries.Res.TL.csv", header = TRUE)
 resMidAtl <- read.csv("data/Atl.Res.TL.csv", header = TRUE)
-resCV <- read.csv("data/CV.TL.csv", header = TRUE)
+resCV <- read.csv("data/CV.Res.TL.csv", header = TRUE)
 
 # Remove B.state variable as it messes up bind.rows
-resNor <- 
-  resNor %>%
-  select(-B.state)
-
 resNor <- 
   resNor %>%
   select(-B.state)

### 2018-01-26 manuscript/revision/blue-whale-ms-v2.tex
Add SM as author and change author contributions

@@
 \author{
-  Clive N. Trueman$^{1}$, Andrew L. Jackson$^{2}$, Katharyn S. Chadwick$^{1}$,\\ Ellen J. Coombs$^{3,4}$,
-  Richard C. Sabin$^{3}$ and Natalie Cooper$^{3*}$
+  Clive N. Trueman$^{1}$, Andrew L. Jackson$^{2}$, Katharyn S. Chadwick$^{1}$,\\ 
+  Ellen J. Coombs$^{3,4}$, Sarah Magozzi$^{1,5}$, Richard C. Sabin$^{3}$ \\
+  and Natalie Cooper$^{3*}$
 }
@@
   $^3$ Department of Life Sciences, Natural History Museum London, Cromwell Road, London, SW7 5BD, UK.\\ 
-  $^4$ University College London, Gower Street, London, WC1E 6BT, UK.\\
+  $^4$ Department of Earth Sciences, University College London, Gower Street, London, WC1E 6BT, UK.\\
+  $^5$ Department of Geology and Geophysics, University of Utah, Salt Lake City, UT 84112-0102, USA.\\
 }}
@@
 We thank C.J.
-Somes for providing  $\delta^{15}$N POM data, and Bastian Hambach and Megan Spencer at the University of Southampton SEAPORT isotope laboratory for assistance with stable isotope analyses.
+Somes for providing  $\delta^{15}$N POM data, Bastian Hambach and Megan Spencer at the University of Southampton SEAPORT isotope laboratory for assistance with stable isotope analyses, and Andrew Yool for allowing us to use and share NEMO-MEDUSA outputs.
 
@@
 CT created the movement model.
-EC, NC and RS helped parameterise the model.
-AJ created the figures. 
-CT, NC and EC wrote the paper.
+EC, NC, SM and RS helped parameterise the model.
+AJ and NC created the figures. 
+CT and NC wrote the paper.
 All authors edited drafts and approved the final version.

### 2018-01-31 manuscript/revision/blue-whale-ms-v2.tex
Edit Fig 3 legend

@@
   \caption{Correlations among simulated $\delta^{13}$C from the top 10\% best fitting migratory movement models (grey lines, right hand y-axis) and $\delta^{13}$C from baleen (black line, left hand y-axis; see Figure \ref{fig1}).
-  The standard deviation of the empirical data have been rescaled by a factor of six so as to better visualise the correlation because, as a consumer, the whale is expected to integrate and attenuate the fluctuations in its diet, and the end points of the simulations and empirical data have been aligned to coincide.}
+  Simulated $\delta^{13}$C values six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location, reflecting temporal integration of phytoplankton $\delta^{13}$C values within the food chain before ingestion by the whale as krill.
+  The end points of the simulations and empirical data have been aligned to coincide.
+}
   \label{fig3}

### 2018-01-31 manuscript/revision/blue-whale-supplementary-v2.tex
Fix link to flowchart fig
Fix trimming of flowchart fig
Edits to Fig S2

@@
   \includegraphics[width = \linewidth]{figures/Figure-S2-cross-cor.png}
-  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for behavioural phase one showing strong negative covariance with a periodicity of around 13cm.} % check legend
+  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for the three behavioural phases, showing limited covariation in behavioural phases one and three, but strong negative co-variance in behavioural phase two with a periodicity of around 13cm.
+  }
   \label{figs2}
@@
   \centering
-  \includegraphics[trim = 0cm 145mm 0cm 4cm, clip]{revision/figures/flow-diagram-model.pdf}
-  \vspace{-1.2cm}
+  \includegraphics[trim = {3cm 0 0 0}, clip]{figures/flow-diagram-model.pdf}
+  %\vspace{-1.2cm}
   \linespread{0.8}

### 2018-01-31 manuscript/revision/blue-whale-supplementary-v2.tex
Fig S4 legend edits

@@
       \includegraphics[width=\linewidth]{figures/Figure-S4-regions-d13C.png}
-      \caption{Alternative $\delta^{13}$C series arising from resident-only movement models restricting region to west UK/Irish shelf (Ireland), Norwegian/Barents Sea (Norway), Canaries/west Azores (Canaries), Mid Atlantic ridge (Mid-Atlantic), Cape Verde, and the Cape Verde/Mauritanian upwelling area over a simulated two year period with daily resolution and 30 replicate simulations for each region.} % check legend
+      \caption{$\delta^{13}$C values expected in baleen plates for whales simulated with resident foraging models (i.e.
+restricting movements to broad regions) (i)  Norwegian/Barents Sea (Norway), (ii) UK/Irish shelf (Ireland), (iii) Mid-Atlantic ridge (Mid-Atlantic), (D) Cape Verde/Mauritanean upwelling (Cape Verde), over a simulated two year period with daily resolution and 30 replicate simulations for each region.
+      Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location.}
       \label{figs4}

### 2018-01-31 manuscript/revision/blue-whale-supplementary-v2.tex
change Fig to Figure

@@
 The oldest (most distal) 35cm of the baleen plate  record relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.
-$\delta^{13}$C values also show repeated fluctuations in the middle 50cm of the baleen plate (behavioural phase two), with a similar periodicity of 13.5cm (Fig. \ref{figs1}). 
+$\delta^{13}$C values also show repeated fluctuations in the middle 50cm of the baleen plate (behavioural phase two), with a similar periodicity of 13.5cm (Figure \ref{figs1}). 
 A distinct change in the isotopic pattern along the baleen plate occurs at around 84cm from the distal end in behavioural phase two, with a transition to uniform, relatively low $\delta^{13}$C values followed by an abrupt transition to positive $\delta^{13}$C values and a decline in $\delta^{13}$C values in the most recent (most gingival) 3cm of the record.
@@
 
-Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N  values within behavioural phase two (Fig.
-\ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioural phase one.
+Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N  values within behavioural phase two (Figure \ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioural phase one.
 
@@
 
-Simulated $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c.60$^{\circ}$N, and relatively negative in the central and southern North Atlantic, associated with high levels of nitrogen fixation (Fig.
-\ref{figs3}).
+Simulated $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c.60$^{\circ}$N, and relatively negative in the central and southern North Atlantic, associated with high levels of nitrogen fixation (Figure \ref{figs3}).
 Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
-In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. \ref{figs3}). 
+In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Figure \ref{figs3}). 
 The temporally coincident periodic, inverse fluctuations in $\delta^{13}$C and $\delta^{15}$N values in the second phase of the baleen isotopic record are therefore consistent with migration between waters north of 60$^{\circ}$N and more southerly southern waters in the North Atlantic, with northerly occupancy in summer months and southerly occupancy in winter months.
@@
 \subsection*{ISOTOPE MODEL SIMULATIONS}
-We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions that would be encountered by whales foraging within broad geographic regions (west UK/Irish shelf, Norwegian/Barents Sea, Canaries/west Azores, Mid Atlantic ridge, Cape Verde/Mauritanean upwelling area; Fig.
-\ref{figs4}).
-Strong seasonal dynamics in $\delta^{13}$C values are evident in all northerly regions, characterized by a rapid increase to annual maximum $\delta^{13}$C values associated with the onset of the spring phytoplankton bloom, followed by more gradual decline in $\delta^{13}$C values towards minima in winter conditions (Fig.
-\ref{figs4}).
-In temperate latitudes around the British Isles, seasonal dynamic cycles are present, but dampened (Fig.
-\ref{figs4}), whereas in subtropical regions exemplified by the Canaries, Azores and particularly Mauritanean upwelling areas, $\delta^{13}$C values are relatively positive and constant throughout the year (Fig.
-\ref{figs4}).
+We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions that would be encountered by whales foraging within broad geographic regions (west UK/Irish shelf, Norwegian/Barents Sea, Canaries/west Azores, Mid Atlantic ridge, Cape Verde/Mauritanean upwelling area; Figure \ref{figs4}).
+Strong seasonal dynamics in $\delta^{13}$C values are evident in all northerly regions, characterized by a rapid increase to annual maximum $\delta^{13}$C values associated with the onset of the spring phytoplankton bloom, followed by more gradual decline in $\delta^{13}$C values towards minima in winter conditions (Figure \ref{figs4}).
+In temperate latitudes around the British Isles, seasonal dynamic cycles are present, but dampened (Figure \ref{figs4}), whereas in subtropical regions exemplified by the Canaries, Azores and particularly Mauritanean upwelling areas, $\delta^{13}$C values are relatively positive and constant throughout the year (Figure \ref{figs4}).
 
@@
 Behavioural phase one is characterised by approximately 24 months of relatively positive and constant $\delta^{13}$C values. 
-$\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity is absent (Fig. \ref{figs1}). 
+$\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity is absent (Figure \ref{figs1}). 
 Positive and temporally constant $\delta^{13}$C values are characteristic of subtropical regions particularly upwelling areas around Mauritania, Cape Verde and the Azores region.
@@
 \subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase two}
-Temporal dynamics in $\delta^{13}$C values simulated in northerly latitudes cannot be reproduced from simulated temporal trends in $\delta^{13}$C values in krill within static areas (Fig.
-\ref{figs4}, Fig.
-3).
-Adding seasonal north-south migrations within mid-high latitude regions to foraging models, closely matches the measured profile, including $\delta^{13}$C minima preceding the June bloom (Fig.
-\ref{figs4}, Fig.
-1).
+Temporal dynamics in $\delta^{13}$C values simulated in northerly latitudes cannot be reproduced from simulated temporal trends in $\delta^{13}$C values in krill within static areas (Figure \ref{figs4}, Figure 3).
+Adding seasonal north-south migrations within mid-high latitude regions to foraging models, closely matches the measured profile, including $\delta^{13}$C minima preceding the June bloom (Figure \ref{figs4}, Figure 1).
 
 Accordingly we modeled whale movements allowing three years of residency in warm waters of the Azores or Cape Verde/Mauritania region followed by a further 3.5 years of seasonal north-south migration in the northeast northern Atlantic, followed by residency in subtropical waters in 1890 and rapid migration returning to more northerly waters in the spring of 1891.
-The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Fig. 3). 
+The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Figure 3). 
 Blue whales have a gestation period of 10-12 months, and calves suckle for approximately 6-7 months \cite{handbook}. 
@@
 \subsection*{AGENT-BASED WHALE MOVEMENT MODEL}
-We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent-based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using,yool2013medusa}, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA\cite{yool2013medusa}, and bathymetry (Fig.
-\ref{figs5}).
+We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent-based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using,yool2013medusa}, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA\cite{yool2013medusa}, and bathymetry (Figure \ref{figs5}).
 
@@
   \caption{Model parameters for agent-based model of whale movement. 
-  See Fig. \ref{figs5} and text for more details. 
+  See Figure \ref{figs5} and text for more details. 
   $^*$During nursing minimum sea surface temperature is set to 18$^{\circ}$C. 
@@
       \includegraphics[width=\linewidth]{figures/Figure-S5-r2.png}
-      \caption{Histogram of $r^2$ values from linear regressions comparing model simulation $\delta^{13}$C values to those of the real data (Fig. 1).} % check legend
+      \caption{Histogram of $r^2$ values from linear regressions comparing model simulation $\delta^{13}$C values to those of the real data (Figure 1).} % check legend
       \label{figs6}

### 2018-01-31 manuscript/revision/blue-whale-supplementary-v2.tex
Edits to supp figures

@@
 \author{Clive N. Trueman, Andrew L. Jackson, Katharyn S. Chadwick,\\ 
-Ellen J. Coombs, Richard C. Sabin and Natalie Cooper}
+Ellen J. Coombs, Sarah Magozzi, Richard C. Sabin\\
+and Natalie Cooper}
 
@@
   \includegraphics[width = \linewidth]{figures/Figure-S1-periodograms.png}
-  \caption{Fourier analysis of $\delta^{15}$N (A,B) and $\delta^{13}$C (C,D) values of the full (A,C) and behavioural phase one only (B,D) components of the baleen plate.
+  \caption{Fourier analysis of $\delta^{15}$N (A, B) and $\delta^{13}$C (C, D) values of the full (A, C) and behavioural phase one only (B, D) components of the baleen plate.
 Inset values are the highest frequency spectral densities (in cm), excluding the values at the edge of the baleen (i.e.
@@
   \includegraphics[width = \linewidth]{figures/Figure-S2-cross-cor.png}
-  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for the three behavioural phases, showing limited covariation in behavioural phases one and three, but strong negative co-variance in behavioural phase two with a periodicity of around 13cm.
+  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for the three behavioural phases, showing limited covariation in behavioural phases one and three, but strong negative covariance in behavioural phase two with a periodicity of around 13cm.
   }
@@
 \subsection*{ISOTOPE MODEL SIMULATIONS}
-We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions that would be encountered by whales foraging within broad geographic regions (west UK/Irish shelf, Norwegian/Barents Sea, Canaries/west Azores, Mid Atlantic ridge, Cape Verde/Mauritanean upwelling area; Figure \ref{figs4}).
+We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions that would be encountered by whales foraging within broad geographic regions (west UK/Irish shelf, Norwegian/Barents Sea, Canaries/west Azores, Mid-Atlantic ridge, Cape Verde/Mauritanian upwelling area; Figure \ref{figs4}).
 Strong seasonal dynamics in $\delta^{13}$C values are evident in all northerly regions, characterized by a rapid increase to annual maximum $\delta^{13}$C values associated with the onset of the spring phytoplankton bloom, followed by more gradual decline in $\delta^{13}$C values towards minima in winter conditions (Figure \ref{figs4}).
@@
       \caption{$\delta^{13}$C values expected in baleen plates for whales simulated with resident foraging models (i.e.
-restricting movements to broad regions) (i)  Norwegian/Barents Sea (Norway), (ii) UK/Irish shelf (Ireland), (iii) Mid-Atlantic ridge (Mid-Atlantic), (D) Cape Verde/Mauritanean upwelling (Cape Verde), over a simulated two year period with daily resolution and 30 replicate simulations for each region.
+restricting movements to broad regions) for west UK/Irish shelf (West Ireland), Norwegian/Barents Sea (Norwegian Sea), Canaries/west Azores (Canaries), Mid-Atlantic ridge (Mid-Atlantic), and the Cape Verde/Mauritanian upwelling area (Cape Verde/Mauritania), over a simulated two year period with daily resolution and 30 replicate simulations for each region.
       Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location.}
@@
     \centering
-      \includegraphics[width=\linewidth]{figures/Figure-S5-r2.png}
+      \includegraphics[width=\linewidth]{figures/Figure-S6-r2.png}
       \caption{Histogram of $r^2$ values from linear regressions comparing model simulation $\delta^{13}$C values to those of the real data (Figure 1).} % check legend
@@
     \centering
-      \includegraphics[width=\linewidth]{figures/Figure-S6-boxplots.png}
+      \includegraphics[width=\linewidth]{figures/Figure-S7-boxplots.png}
       \caption{Boxplots showing the maximum and standard devations of latitudinal positions of simulated whales taken from the top 10\%, and bottom 10\%, of movement model simulations.} % check legend
@@
     \centering
-      \includegraphics[width=\linewidth]{figures/Figure-S7-kernals.png}
+      \includegraphics[width=\linewidth]{figures/Figure-S8-kernals.png}
       \caption{Kernal density plots showing the positions of simulated whales taken from the top 10\%, and bottom 10\%, of movement model simulations. 

### 2018-01-31 code/Figure-S5-code.R
Fix code names to match supp figs

@@ -1,25 +1,0 @@
-# Author: Natalie Cooper
-# About: script to plot histogram of r2 values
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Load libraries
-library(tidyverse)
-
-# Load data
-r2 <- read.csv("data/all.r2.csv")
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Plot
-
-p1 <- 
-  ggplot(r2, aes(x = x)) +
-  geom_histogram(fill = "grey", binwidth = 0.05) +  
-  theme_classic(base_size = 16) + 
-  xlab(expression(paste(r^2))) +
-  #xlab(expression(paste(r^2, " from models comparing real to simulated ", 
-  #                delta^{13}, "C (\u2030)", " values"))) +
-  expand_limits(x = c(0, 1),  y = c(0, 200))
-
-
-ggsave("manuscript/revision/figures/Figure-S5-r2.png", p1, 
-       device = png(width = 600, height = 400))

### 2018-01-31 code/Figure-S6-code.R

@@ -1,52 +1,25 @@
-# Author: Clive Trueman
-# About: script to plot max and sd of latitude 
-# for the top 10% and bottom 10% of models
-# Modified by Natalie Cooper Nov 2017
+# Author: Natalie Cooper
+# About: script to plot histogram of r2 values
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Load libraries
 library(tidyverse)
-library(gridExtra)
 
 # Load data
-max <- read.csv("data/max.lat.csv")
-sdev <- read.csv("data/sd.lat.csv")
+r2 <- read.csv("data/all.r2.csv")
 
-# Relevel so top 10% is first
-max$group <-relevel(max$group, ref = "top")
-sdev$group <-relevel(sdev$group, ref = "top")
-
-# Get means
-mean.max <- summarise(group_by(max, group), mn = mean(V1))
-mean.sd <- summarise(group_by(sdev, group), mn = mean(V1))
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Plot
 
 p1 <- 
-  ggplot(max, aes(x = group, y = V1, fill = group)) +
-  geom_boxplot() +  
+  ggplot(r2, aes(x = x)) +
+  geom_histogram(fill = "grey", binwidth = 0.05) +  
   theme_classic(base_size = 16) + 
-  xlab("") +
-  ylab("Maximum latitude") +
-  scale_x_discrete(labels = c("Top 10%", "Bottom 10%")) +
-  scale_y_continuous(breaks = c(20, 40, 60, 80, 100),
-                     labels = c(20, 40, 60, 80, 100)) +
-  scale_fill_manual(values = c("white", "gray")) +
-  theme(legend.position = "none")
+  xlab(expression(paste(r^2))) +
+  #xlab(expression(paste(r^2, " from models comparing real to simulated ", 
+  #                delta^{13}, "C (\u2030)", " values"))) +
+  expand_limits(x = c(0, 1),  y = c(0, 200))
 
-p2 <- 
-  ggplot(sdev, aes(x = group, y = V1, fill = group)) +
-  geom_boxplot() +  
-  theme_classic(base_size = 16) + 
-  xlab("") +
-  ylab("Standard deviation of latitude") +
-  scale_x_discrete(labels = c("Top 10%", "Bottom 10%")) +
-  scale_y_continuous(breaks = c(10, 20, 30, 40, 50),
[…]

### 2018-01-31 code/Figure-S7-code.R

@@ -1,46 +1,52 @@
 # Author: Clive Trueman
-# About: script to plot kernal densities for top 10% and bottom 10% of models
-# Tidied by Natalie Cooper Nov 2017
+# About: script to plot max and sd of latitude 
+# for the top 10% and bottom 10% of models
+# Modified by Natalie Cooper Nov 2017
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Load libraries
-library(maps)
-library(mapdata)
-library(spatstat)
+library(tidyverse)
+library(gridExtra)
 
-# Read in data
-top10 <- read.csv("data/top10percent.csv")
-bottom10 <- read.csv("data/bottom10percent.csv")
+# Load data
+max <- read.csv("data/max.lat.csv")
+sdev <- read.csv("data/sd.lat.csv")
 
+# Relevel so top 10% is first
+max$group <-relevel(max$group, ref = "top")
+sdev$group <-relevel(sdev$group, ref = "top")
+
+# Get means
+mean.max <- summarise(group_by(max, group), mn = mean(V1))
+mean.sd <- summarise(group_by(sdev, group), mn = mean(V1))
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Kernal density plots
+# Plot
 
-# Jitter points of lat and long to get smooth density plot
-jLat <- jitter(top10$Lat, factor = 2)
-jLon <- jitter(top10$Lon, factor = 1)
-jLatB <- jitter(bottom10$Lat, factor = 2)
-jLonB <- jitter(bottom10$Lon, factor = 1)
+p1 <- 
+  ggplot(max, aes(x = group, y = V1, fill = group)) +
+  geom_boxplot() +  
+  theme_classic(base_size = 16) + 
+  xlab("") +
+  ylab("Maximum latitude") +
+  scale_x_discrete(labels = c("Top 10%", "Bottom 10%")) +
+  scale_y_continuous(breaks = c(20, 40, 60, 80, 100),
+                     labels = c(20, 40, 60, 80, 100)) +
+  scale_fill_manual(values = c("white", "gray")) +
+  theme(legend.position = "none")
 
-# Create point patterns to plot
-pts <- ppp(jLon, jLat, window = owin(c(-80, 55), c(-10, 85)))
[…]

### 2018-01-31 code/Figure-S6-code.R
change supp figure numbers in code

@@ -21,5 +21,5 @@ p1 <-
   expand_limits(x = c(0, 1),  y = c(0, 200))
 
 
-ggsave("manuscript/revision/figures/Figure-S5-r2.png", p1, 
+ggsave("manuscript/revision/figures/Figure-S6-r2.png", p1, 
        device = png(width = 600, height = 400))

### 2018-01-31 code/Figure-S7-code.R

@@ -48,5 +48,5 @@ p2 <-
 
 boxes <- grid.arrange(p1, p2, ncol = 2)
 
-ggsave("manuscript/revision/figures/Figure-S6-boxplots.png", boxes, 
+ggsave("manuscript/revision/figures/Figure-S7-boxplots.png", boxes, 
        device = png(width = 600, height = 400))

### 2018-01-31 code/Figure-S8-code.R

@@ -26,7 +26,7 @@ pts <- ppp(jLon, jLat, window = owin(c(-80, 55), c(-10, 85)))
 ptsB <- ppp (jLonB, jLatB, window = owin(c(-80, 55), c(-10, 85)))
 
 
-png("manuscript/revision/figures/Figure-S7-kernals.png", width = 1200, height = 800)
+png("manuscript/revision/figures/Figure-S8-kernals.png", width = 1200, height = 800)
 
 par(mfrow = c(1, 2))
 par(mar = c(2, 1, 3, 1))

### 2018-01-31 manuscript/revision/blue-whale-ms-v2.tex
Correct supp figure numbers

@@
 
-Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods). 
+Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods; Supplementary Figures S1 and S2).
 Therefore, given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$, we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over seven full years of the whale's life (early 1884 - spring 1891).
@@
 The relatively positive and seasonally-invariant $\delta^{13}$C values seen during behavioural phase one are only found in sub-tropical areas of the North Atlantic.  
-Our simulations show high variance during this period (Figure \ref{fig3}), reflecting a range of possible locations for the whale, although areas around the Mauritanian coast and Cape Verde Islands, a known current and historic winter feeding area for blue whales \cite{baines2014upwellings,reeves2004historical}, and potentially to the west of the Azores (Figure \ref{fig2}), most closely match the measured profile.
+Our simulations show high variance during this period (Figure \ref{fig3}), reflecting a range of possible locations for the whale, although areas around the Mauritanian coast and Cape Verde islands, a known current and historic winter feeding area for blue whales \cite{baines2014upwellings,reeves2004historical}, and potentially to the west of the Azores (Figure \ref{fig2}), most closely match the measured profile.
 The whale remained in relatively warm waters for at least two full years. 
@@
 Behavioural phase two is best simulated by seasonal migrations between summer foraging in northern areas in the Norwegian Sea/Barents Sea/Iceland region, and winter foraging in a broad region between the UK and more southerly, sub-tropical waters.
-Better fitting models in general were those predicting a greater latitudinal foraging range, and foraging in more northerly waters (Supplementary Figure S7).
+Better fitting models in general were those predicting a greater latitudinal foraging range, and foraging in more northerly waters (Supplementary Figures S7 and S8).
 Best-fit model distributions are largely consistent with our current understanding of blue whale distributions in the northeast Atlantic\cite{reeves2004historical,baines2014upwellings,baines2017autumn,reeves2004historical}, with perhaps greater importance of winter foraging in temperate regions (Figure \ref{fig4}).
@@
 Blue whales have a 10-12 month gestation period, with calving occurring in sub-tropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
-We therefore interpret the final year of the isotopic record as indicating: (a) birth of a calf in the winter of 1889/1890, presumably in sub-tropical waters, (b) 6 months of weaning during which time the maternal whale did not feed, followed by (c) a short period of feeding in sub-tropical waters in the second half of 1890, potentially during a final northward migration, and (d) eventual stranding during the return to northern feeding grounds in early 1891.
+We therefore interpret the final year of the isotopic record as indicating: (a) birth of a calf in the winter of 1889/1890, presumably in sub-tropical waters, (b) six months of weaning during which time the maternal whale did not feed, followed by (c) a short period of feeding in sub-tropical waters in the second half of 1890, potentially during a final northward migration, and (d) eventual stranding during the return to northern feeding grounds in early 1891.
 

### 2018-02-01 manuscript/revision/blue-whale-ms-v2.tex
Deal with minor remaining changes for ms

@@
 \raggedright
-% Need to edit
 
@@
 Despite hunting for blue whales being banned for over 50 years, blue whales are still listed as Endangered on the IUCN Red List\cite{reilly2008balaenoptera}.
-In the northeast Atlantic, blue whales appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations (REF).  
+In the northeast Atlantic, blue whales appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations\cite{pike2009note}.  
 The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland\cite{pike2009note}.
@@
 While direct tagging methods are adding vital behavioural data to our understanding of whale movements\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
-Currently the longest period of continuous monitoring via satellite tagging of an individual blue whale in the northeast Atlantic is X days, while photographic identification has matched a single blue whale with sightings in Iceland and Mauritania.\cite{baines2014upwellings}  % How many days?
 Conservation efforts are consequently hampered by the lack of detailed data on the distribution, behaviour and ecology of blue whales, particularly during the winter and for the juvenile phase of life.
@@
 Assimilation of carbon into krill tissues will dampen the temporal variability seen in POM, effectively producing a temporal average over the timescale of isotopic turnover within krill. 
-We estimate turnover to be complete between four and six months [REF] and therefore we resampled the $\delta^{13}$C POM values in each one degree cell to reflect a six month moving average. 
+We estimate turnover to be complete between four and six months and therefore we resampled the $\delta^{13}$C POM values in each one degree cell to reflect a six month moving average. 
 Carbon isotope values are also likely to be fractionated during transfer from plankton to krill as $^{12}$C  is preferentially lost through respiration. 
@@
 Data are available from the NHM Data Portal [doi.org/10.5519/0093278](https://doi.org/10.5519/0093278). 
-R code is available from GitHub (github.com/nhcooper123/blue-whale-bes).
-{}
+R code is available from GitHub [github.com/nhcooper123/blue-whale-bes](https://github.com/nhcooper123/blue-whale-bes).
 % References

### 2018-02-01 manuscript/revision/blue-whale-ms-v2.tex
Typo

@@
 Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable}. 
-Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an iindividual whale, although as baleen is worn away at the tips over time, an individual's entire lifespan is rarely recorded.
+Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although as baleen is worn away at the tips over time, an individual's entire lifespan is rarely recorded.
 Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales\cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}, but have yet to be applied to blue whales from the north Atlantic population.

### 2018-02-01 manuscript/revision/blue-whale-ms-v2.tex
Typo

@@
 Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
-A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggest that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands\cite{baines2014upwellings}.
+A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggest that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde islands\cite{baines2014upwellings}.
 Northward migration in the spring may occur along mid-Atlantic corridors with peak sightings in the Azores around April-May.

### 2018-02-01 manuscript/revision/blue-whale-ms-v2.tex
Typos

@@
 Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
-A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggest that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde islands\cite{baines2014upwellings}.
+A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggest that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde slands\cite{baines2014upwellings}.
 Northward migration in the spring may occur along mid-Atlantic corridors with peak sightings in the Azores around April-May.
@@
 Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although as baleen is worn away at the tips over time, an individual's entire lifespan is rarely recorded.
-Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales\cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}, but have yet to be applied to blue whales from the north Atlantic population.
+Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales\cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}, but have yet to be applied to blue whales from the North Atlantic population.
 Here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale ``Hope''. 
-Hope is female, and estimated to be a young adult c.15 years at death, suggesting that her baleen may yield information about juvenile behaviours in north Atlantic blue whales around the peak of industrial whaling.
+Hope is female, and estimated to be a young adult c.15 years at death, suggesting that her baleen may yield information about juvenile behaviours in North Atlantic blue whales around the peak of industrial whaling.
 
@@
 We develop a new approach to address this problem using temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Supplementary Figure S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Supplementary Table S1; Supplementary Figure S5).
-We use our coupled model to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of north Atlantic blue whales.
+We use our coupled model to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of North Atlantic blue whales.
 We then compare simulated profiles (Figures \ref{fig2}, \ref{fig3} and Supplementary Figure S4) to the measured isotopic profile of the blue whale's baleen (Figure \ref{fig1}) to identify the most likely movement history.
@@
   \caption{Correlations among simulated $\delta^{13}$C from the top 10\% best fitting migratory movement models (grey lines, right hand y-axis) and $\delta^{13}$C from baleen (black line, left hand y-axis; see Figure \ref{fig1}).
-  Simulated $\delta^{13}$C values six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location, reflecting temporal integration of phytoplankton $\delta^{13}$C values within the food chain before ingestion by the whale as krill.
+  Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location, reflecting temporal integration of phytoplankton $\delta^{13}$C values within the food chain before ingestion by the whale as krill.
   The end points of the simulations and empirical data have been aligned to coincide.
@@
 The whale remained in relatively warm waters for at least two full years. 
-Hope was estimated to be 15 years old when she died (based on vertebral epiphyseal fusion; R.Sabin \textit{pers.comm}), so was approximately 3-5 years old during this period.
+Hope was estimated to be 15 years old when she died (based on vertebral epiphyseal fusion; R.C. Sabin \textit{pers.comm}), so was approximately 3-5 years old during this period.
 During behavioural phase two, starting at an approximate age of 5-6 years, the observed low $\delta^{13}$C values imply foraging in colder, more northerly latitudes. 
@@
 We simulated 1200 individual movement patterns, then excluded simulations where the virtual whale stranded before reaching the 3019 days represented in the baleen.
-With then compared the remaining 1049 simulations and the resulting simulated baleen $\delta^{13}$C records to the measured records with simple linear regressions. 
+We then compared the remaining 1049 simulations and the resulting simulated baleen $\delta^{13}$C records, to the measured records with simple linear regressions. 
 Simulated baleen $\delta^{13}$C profiles produce a good fit to measured profiles, the median $r^2$ value across 1049 simulated profiles was 0.49, and the maximum was 0.71 (Supplementary Figure S6).
@@
 Firstly, next to nothing is known about juvenile behaviour in blue whales - calves become independent of their mothers at around seven months, and become sexually mature around 8-10 years old\cite{handbook}.
-Our analyses suggest that between these points, juvenile North Atlantic blue whales may spend extended periods in warm, highly productive waters associated with upwelling systems before migrating to higher latitude feeding grounds.
-Extended periods of juvenile residency in sub-tropical waters could have implications for blue whale conservation in the northeast Atlantic, as the upwelling regions around Mauritania and the Cape Verde islands are areas with emerging offshore oil and gas exploration and are fishing, tourist and shipping hotspots, potentially increasing threats to juvenile whales.
+Our analyses suggest that between these points, juvenile North Atlantic blue whales may spend extended periods in warm, highly productive waters associated with upwelling systems, before migrating to higher latitude feeding grounds.
+Extended periods of juvenile residency in sub-tropical waters could have implications for blue whale conservation in the northeast Atlantic, as the upwelling regions around Mauritania and the Cape Verde islands are areas with emerging offshore oil and gas exploration, and are fishing, tourist and shipping hotspots, potentially increasing threats to juvenile whales.
 
@@
 Isotope-enabled biogeochemical ocean models\cite{magozzi2017using,schmittner2016complementary} were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Supplementary Figure S3).
-Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model (Supplementary Figure S3). 
+Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by C.J.
+Somes (\textit{pers.comm}) based on a 5$^{\circ}$ resolution biogeochemical model (Supplementary Figure S3).
 $\delta^{13}$C POM values were simulated at 1$^{\circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model\cite{magozzi2017using,yool2013medusa}. 
@@
 
-We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplementary Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and sub-tropical latitudes.
+We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid-Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplementary Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and sub-tropical latitudes.
 Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four years of simulations.
@@
 \subsection{Data Availability}
-Data are available from the NHM Data Portal [doi.org/10.5519/0093278](https://doi.org/10.5519/0093278). 
-R code is available from GitHub [github.com/nhcooper123/blue-whale-bes](https://github.com/nhcooper123/blue-whale-bes).
+Data are available from the NHM Data Portal (https://doi.org/10.5519/0093278). 
+R code is available from GitHub (https://github.com/nhcooper123/blue-whale-bes).
 % References

### 2018-02-02 manuscript/revision/blue-whale-supplementary-v2.tex
Typo

@@
 Fourier analysis (periodogram function in the R package TSA\cite{Chan:2012aa}) revealed a strong periodic repetition with a 13.3cm periodicity. 
-The oldest (most distal) 35cm of the baleen plate  record relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.
+The oldest (most distal) 35cm of the baleen plate records relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.
 $\delta^{13}$C values also show repeated fluctuations in the middle 50cm of the baleen plate (behavioural phase two), with a similar periodicity of 13.5cm (Figure \ref{figs1}). 
@@
 Isotope-enabled biogeochemical ocean models\cite{magozzi2017using,schmittner2016complementary} were used to characterise the isotopic composition of phytoplankton expected in different potential foraging grounds.
-Annual average $\delta^{15}$N POM values were provided by C.J. Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model. 
+Annual average $\delta^{15}$N POM values were provided by C.J. Somes (\textit{pers.comm}) based on a 5$^{\circ}$ resolution biogeochemical model. 
 $\delta^{13}$C POM values were simulated using an isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using}.

### 2018-02-02 code/Figure-4-code.R
Make 3 * 4 not 6 * 2 plot arrangement in Fig 4

@@ -21,7 +21,7 @@ mycols <- c(viridis_pal()(10)[1], viridis_pal()(10)[5], viridis_pal()(10)[9])
 png("manuscript/revision/figures/Figure-4-monthly.png", width = 800, height = 600)
 
 # Make 12 plots
-par(mfrow = c(2, 6))
+par(mfrow = c(3, 4))
 par(mar = c(1, 1, 2, 1))
 
 # Loop through months  

### 2018-02-14 manuscript/NatureEE/blue-whale-supplementary-v2.tex
Change name of revision folder

--- 2018-02-02 manuscript/revision/blue-whale-supplementary-v2.tex
@@
   \includegraphics[width = \linewidth]{figures/Figure-S1-periodograms.png}
-  \caption{Fourier analysis of $\delta^{15}$N (A, B) and $\delta^{13}$C (C, D) values of the full (A, C) and behavioural phase one only (B, D) components of the baleen plate.
+  \caption{Fourier analysis of $\delta^{15}$N (A, B) and $\delta^{13}$C (C, D) values of the full (A, C) and behavioural phase two only (B, D) components of the baleen plate.
 Inset values are the highest frequency spectral densities (in cm), excluding the values at the edge of the baleen (i.e.
@@
 In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Figure \ref{figs3}). 
-The temporally coincident periodic, inverse fluctuations in $\delta^{13}$C and $\delta^{15}$N values in the second phase of the baleen isotopic record are therefore consistent with migration between waters north of 60$^{\circ}$N and more southerly southern waters in the North Atlantic, with northerly occupancy in summer months and southerly occupancy in winter months.
+The temporally coincident periodic, inverse fluctuations in $\delta^{13}$C and $\delta^{15}$N values in the second phase of the baleen isotopic record are therefore consistent with migration between waters north of 60$^{\circ}$N and more southerly waters in the North Atlantic, with northerly occupancy in summer months and southerly occupancy in winter months.
 Incorporation of carbon and nitrogen with isotopic ratios consistent with southerly latitudes during winter months indicates continued feeding during migration and winter residency.
@@
     \end{subfigure}
-    \caption{Spatial model (isoscape) of model-simulated biomass weighted annual average $\delta^{13}$C (left) and $\delta^{15}$N (right) values from POM.
+    \caption{Spatial model (isoscape) of model-simulated biomass, using weighted annual average $\delta^{13}$C (left) and $\delta^{15}$N (right) values from particulate organic matter (POM).
 $\delta^{13}$C values are from Magozzi et al.
 2017\cite{magozzi2017using} and $\delta^{15}$N from Schmittner \& Somes 2016\cite{schmittner2016complementary} and C.J.
-Somes (pers.
-comm.).}
+Somes (\textit{pers.
+comm.}).}
     \label{figs3}
@@
 We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions that would be encountered by whales foraging within broad geographic regions (west UK/Irish shelf, Norwegian/Barents Sea, Canaries/west Azores, Mid-Atlantic ridge, Cape Verde/Mauritanian upwelling area; Figure \ref{figs4}).
-Strong seasonal dynamics in $\delta^{13}$C values are evident in all northerly regions, characterized by a rapid increase to annual maximum $\delta^{13}$C values associated with the onset of the spring phytoplankton bloom, followed by more gradual decline in $\delta^{13}$C values towards minima in winter conditions (Figure \ref{figs4}).
-In temperate latitudes around the British Isles, seasonal dynamic cycles are present, but dampened (Figure \ref{figs4}), whereas in subtropical regions exemplified by the Canaries, Azores and particularly Mauritanean upwelling areas, $\delta^{13}$C values are relatively positive and constant throughout the year (Figure \ref{figs4}).
+Strong seasonal dynamics in $\delta^{13}$C values are evident in all northerly regions, characterized by a rapid increase to annual maximum $\delta^{13}$C values associated with the onset of the spring phytoplankton bloom, followed by a more gradual decline in $\delta^{13}$C values towards minima in winter conditions (Figure \ref{figs4}).
+In temperate latitudes around the British Isles, seasonal dynamic cycles are present, but dampened (Figure \ref{figs4}), whereas in sub-tropical regions exemplified by the Canaries, Azores and particularly Mauritanian upwelling areas, $\delta^{13}$C values are relatively positive and constant throughout the year (Figure \ref{figs4}).
 
@@
 $\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity is absent (Figure \ref{figs1}). 
-Positive and temporally constant $\delta^{13}$C values are characteristic of subtropical regions particularly upwelling areas around Mauritania, Cape Verde and the Azores region.
+Positive and temporally constant $\delta^{13}$C values are characteristic of sub-tropical regions particularly upwelling areas around Mauritania, Cape Verde and the Azores region.
 
@@
 
-Accordingly we modeled whale movements allowing three years of residency in warm waters of the Azores or Cape Verde/Mauritania region followed by a further 3.5 years of seasonal north-south migration in the northeast northern Atlantic, followed by residency in subtropical waters in 1890 and rapid migration returning to more northerly waters in the spring of 1891.
+Accordingly we modeled whale movements allowing three years of residency in warm waters of the Azores or Cape Verde/Mauritania region followed by a further 3.5 years of seasonal north-south migration in the northeast northern Atlantic, followed by residency in sub-tropical waters in 1890 and rapid migration returning to more northerly waters in the spring of 1891.
 The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Figure 3). 
@@
 \subsection*{AGENT-BASED WHALE MOVEMENT MODEL}
-We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent-based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using,yool2013medusa}, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA\cite{yool2013medusa}, and bathymetry (Figure \ref{figs5}).
+We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent-based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using,yool2013medusa}, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA\cite{yool2013medusa}, and bathymetry from\cite{bathy} and extracted using the marmap R pakckage\cite{marmap} (Figure \ref{figs5}).
 
@@
     \hline
-    Minimum sea surface temperature$^*$ ($^{\circ}$C) & 3 & 3 & 3\\
+    Minimum sea surface temperature ($^{\circ}$C) & 3 & 3 & 3\\
     \hline
@@
     \hline
-    Maximum daily distance traveled (km)$^{**}$ & $N(\mu=50, \sigma=25)$ & $N(\mu=150, \sigma=75)$ & $N(\mu=150, \sigma=75)$\\
+    Maximum daily distance traveled (km)$^{*}$ & $N(\mu=50, \sigma=25)$ & $N(\mu=150, \sigma=75)$ & $N(\mu=150, \sigma=75)$\\
     \hline
@@
   See Figure \ref{figs5} and text for more details. 
-  $^*$During nursing minimum sea surface temperature is set to 18$^{\circ}$C. 
-  $^{**}$During behavioural phase one, mean maximum daily distance is set to 20km.}
+  %$^*$During nursing minimum sea surface temperature is set to 18$^{\circ}$C. 
+  $^{*}$During behavioural phase one, mean maximum daily distance traveled is set to 20km.}
   \label{tables1}
@@
 
-\subsection*{ADDITIONAL RESULTS FIGURES}
+\section*{Supplementary Results}
 
@@
       \includegraphics[width=\linewidth]{figures/Figure-S7-boxplots.png}
-      \caption{Boxplots showing the maximum and standard devations of latitudinal positions of simulated whales taken from the top 10\%, and bottom 10\%, of movement model simulations.} % check legend
+      \caption{Boxplots showing the maximum and standard devation of latitudinal positions of simulated whales taken from the top 10\%, and bottom 10\%, of movement model simulations.} 
       \label{figs7}
@@
 
+\begin{landscape}
     \begin{figure}[!htbp]
@@
       \caption{Kernal density plots showing the positions of simulated whales taken from the top 10\%, and bottom 10\%, of movement model simulations. 
-      High densities are in yellow, low densities are in blue.} % check legend
+      High densities are in yellow, low densities are in blue.}
       \label{figs8}
   \end{figure}
+\end{landscape}
 

### 2018-02-14 manuscript/CurrentBiology/blue-whale-ms.tex
Edits for Current Biology formatting

--- 2018-02-01 manuscript/revision/blue-whale-ms-v2.tex
@@
 %\usepackage[round]{natbib} % author-year citations
-\usepackage[superscript,biblabel]{cite} % for superscript citations
+%\usepackage[superscript,biblabel]{cite} % for superscript citations
 \usepackage{graphicx}
@@
 \linespread{1.66}
+
+% remove brackets from bibliography list
+\makeatletter 
+\renewcommand\@biblabel[1]{#1} 
+\makeatother
 
@@
 \title{Reconstructing the last known movements of one of Nature's giants}
-% 90 characters max
+% 150 characters max
 
@@
   $^5$ Department of Geology and Geophysics, University of Utah, Salt Lake City, UT 84112-0102, USA.\\
+  $^*$Correspondence: natalie.cooper@nhm.ac.uk; @nhcooper123. 
 }}
@@
 %\runninghead{}
-%\keywords{}
-%}
+\keywords{stable isotope analysis, movement models, museum collections, blue whale, baleen}
 
@@
 \addtolength{\parskip}{.9em}
-% Abstract 200 -300 words max
-% \section{Abstract}
-% general topic intro - 1 sentence
-% intro to field - 2 sentences
-% what we find/show - 1 sentence
-% implications - 2-3 sentences. Results in context, how has the paper moved the field forward
+% Summary 250 words max
 
@@
 
-\section{Abstract}
-Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}. 
-Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}, but interpreting these records can be challenging.
-Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct seven years of movements of an iconic female blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
+\section{Summary}
+Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change. 
+Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions, but interpreting these records can be challenging.
+Here we combine stable isotope analyses of baleen with agent-based simulation models, to reconstruct seven years of movements of an iconic female blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
 We provide the longest continuous record of movement for an individual blue whale in the north Atlantic. 
@@
 \newpage
-\section{Main text}
-Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}. 
-Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle\cite{robinson2009travelling}.
+\section{Introduction}
+Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites \cite{runge2014conserving}.
+Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle \cite{robinson2009travelling}.
 Identifying threats to migratory species and developing effective conservation measures requires information on individual-level movements spanning multiple years, for both historical and present-day populations.
-Such data are scarce, especially for rare, wide-ranging and long-lived marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
+Such data are scarce, especially for rare, wide-ranging and long-lived marine species \cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
 
 The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
-Despite hunting for blue whales being banned for over 50 years, blue whales are still listed as Endangered on the IUCN Red List\cite{reilly2008balaenoptera}.
-In the northeast Atlantic, blue whales appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations\cite{pike2009note}.  
-The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland\cite{pike2009note}.
+Despite hunting for blue whales being banned for over 50 years, blue whales are still listed as Endangered on the IUCN Red List \cite{reilly2008balaenoptera}.
+In the northeast Atlantic, blue whales appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations \cite{pike2009note}.  
+The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland \cite{pike2009note}.
 Following movements of individual blue whales is extremely challenging due to their scarcity and oceanic habit.
 Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
-A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggest that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde slands\cite{baines2014upwellings}.
+A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggest that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde islands \cite{baines2014upwellings}.
 Northward migration in the spring may occur along mid-Atlantic corridors with peak sightings in the Azores around April-May.
-Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas\cite{pike2009note}.
-Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southward-migrating animals\cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
-While direct tagging methods are adding vital behavioural data to our understanding of whale movements\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
+Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas \cite{pike2009note}.
+Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southward-migrating animals \cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
+While direct tagging methods are adding vital behavioural data to our understanding of whale movements \cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies \cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour \cite{walker2012review}.
 Conservation efforts are consequently hampered by the lack of detailed data on the distribution, behaviour and ecology of blue whales, particularly during the winter and for the juvenile phase of life.
-An alternative technique for investigating whale movements retrospectively is to use intrinsic biochemical information such as stable isotopes\cite{west2006stable,busquets2017estimating,hobson2008tracking}.
+An alternative technique for investigating whale movements retrospectively is to use intrinsic biochemical information such as stable isotopes \cite{west2006stable,busquets2017estimating,hobson2008tracking}.
 The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
 Stable isotope data therefore provide information on diet and location (i.e.
-`trophic geography'\cite{bird2018global}), but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values.
-Here we reconstruct movements of an historic whale by comparing isotopic compositions of samples from museum collections, rich archives of behavioural information\cite{lister2011natural}, to spatially referenced models of isotopic variation in diet across the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
+`trophic geography' \cite{bird2018global}), but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values.
+Here we reconstruct movements of an historic whale by comparing isotopic compositions of samples from museum collections, rich archives of behavioural information \cite{lister2011natural}, to spatially referenced models of isotopic variation in diet across the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
+Mysticete (baleen) whales are characterized by the development of baleen plates, keratinous structures in the upper jaw used to filter food items from seawater. 
+Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert \cite{best1996stable}. 
+Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although as baleen is worn away at the tips over time, an individual's entire lifespan is rarely recorded.
+Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales \cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}, but have yet to be applied to blue whales from the North Atlantic population.
+Here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale ``Hope'', a female whale estimated to be a young adult c.15 years at her death in 1891.
 
-Mysticete (baleen) whales are characterized by the development of baleen plates, keratinous structures in the upper jaw used to filter food items from seawater. 
-Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable}. 
-Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although as baleen is worn away at the tips over time, an individual's entire lifespan is rarely recorded.
-Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales\cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}, but have yet to be applied to blue whales from the North Atlantic population.
-Here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale ``Hope''. 
-Hope is female, and estimated to be a young adult c.15 years at death, suggesting that her baleen may yield information about juvenile behaviours in North Atlantic blue whales around the peak of industrial whaling.
+\section{Results and Discussion}
 
@@
 These isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e.
-relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope\cite{west2006stable}.
-Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; both isotopes provide information on the geographical locations of an individual, but nitrogen isotopes are more strongly fractionated during protein metabolism and thus contain additional trophic and physiological information\cite{west2006stable}.
+relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope \cite{west2006stable}.
+Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; both isotopes provide information on the geographical locations of an individual, but nitrogen isotopes are more strongly fractionated during protein metabolism and thus contain additional trophic and physiological information \cite{west2006stable}.
 
-Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods; Supplementary Figures S1 and S2).
+Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplemental Methods; Supplemental Figures S1 and S2).
 Therefore, given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$, we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over seven full years of the whale's life (early 1884 - spring 1891).
 The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
-In behavioural phase one (from the start of the record in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Figure \ref{fig1}; Supplementary Figures S1 and S2).
+In behavioural phase one (from the start of the record in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Figure \ref{fig1}; Supplemental Figures S1 and S2).
 In behavioural phase two (summer 1886 to spring 1890) $\delta^{15}$N values are relatively high and $\delta^{13}$C values are relatively low with coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values.
 In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively high $\delta^{13}$C values in the second half of 1891.
-The final three months of the record show a progressive fall in $\delta^{13}$C values (Figure \ref{fig1}; Supplementary Figures S1 and S2).
+The final three months of the record show a progressive fall in $\delta^{13}$C values (Figure \ref{fig1}; Supplemental Figures S1 and S2).
 
@@
 Samples were taken longitudinally through the baleen plate (n = 97 samples from a single baleen plate for both isotopes).
-There is strong annual periodicity and cross-correlation (Supplementary Figures S1 and S2) in both isotopes.
+There is strong annual periodicity and cross-correlation (Supplemental Figures S1 and S2) in both isotopes.
 The approximate relationship to years assuming a growth rate of 13.5cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
@@
 
-Relating the isotopic compositions of animal tissues to likely locations is complicated by a lack of knowledge of spatio-temporal variations in the isotopic composition of diet\cite{west2006stable,mcmahon2015millennial}.
-We develop a new approach to address this problem using temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Supplementary Figure S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Supplementary Table S1; Supplementary Figure S5).
+Relating the isotopic compositions of animal tissues to likely locations is complicated by a lack of knowledge of spatio-temporal variations in the isotopic composition of diet \cite{west2006stable,mcmahon2015millennial}.
+We develop a new approach to address this problem using temporally dynamic models of $\delta^{13}$C values in the global ocean \cite{magozzi2017using} (Supplemental Figure S3), coupled to an agent-based model of whale movements (see Supplemental Methods; Supplemental Table S1; Supplemental Figure S5).
 We use our coupled model to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of North Atlantic blue whales.
-We then compare simulated profiles (Figures \ref{fig2}, \ref{fig3} and Supplementary Figure S4) to the measured isotopic profile of the blue whale's baleen (Figure \ref{fig1}) to identify the most likely movement history.
+We then compare simulated profiles (Figures \ref{fig2}, \ref{fig3} and Supplemental Figure S4) to the measured isotopic profile of the blue whale's baleen (Figure \ref{fig1}) to identify the most likely movement history.
 Blue whales are commonly assumed to feed mainly in summer in productive, high latitude regions and rely largely on stored lipid reserves to meet metabolic demands in lower latitudes in winter. 
-However, in winter blue whales in the northeast Atlantic are associated with upwellings\cite{baines2017autumn}, while northward migrations in spring coincide with the onset and development of phytoplankton blooms\cite{silva2013north,visser2011timing,busquets2017estimating}, strongly implying feeding throughout the year.
+However, in winter blue whales in the northeast Atlantic are associated with upwellings \cite{baines2017autumn}, while northward migrations in spring coincide with the onset and development of phytoplankton blooms \cite{silva2013north,visser2011timing,busquets2017estimating}, strongly implying feeding throughout the year.
 We therefore assume opportunistic ingestion of protein throughout the year.
 
-% figure 2{}
+% figure 2
 \begin{figure}
@@
 We then compared the remaining 1049 simulations and the resulting simulated baleen $\delta^{13}$C records, to the measured records with simple linear regressions. 
-Simulated baleen $\delta^{13}$C profiles produce a good fit to measured profiles, the median $r^2$ value across 1049 simulated profiles was 0.49, and the maximum was 0.71 (Supplementary Figure S6).
+Simulated baleen $\delta^{13}$C profiles produce a good fit to measured profiles, the median $r^2$ value across 1049 simulated profiles was 0.49, and the maximum was 0.71 (Supplemental Figure S6).
 The top 10\% best fitting simulated profiles are shown in Figure \ref{fig2}. 
@@
 Behavioural phase two is best simulated by seasonal migrations between summer foraging in northern areas in the Norwegian Sea/Barents Sea/Iceland region, and winter foraging in a broad region between the UK and more southerly, sub-tropical waters.
-Better fitting models in general were those predicting a greater latitudinal foraging range, and foraging in more northerly waters (Supplementary Figures S7 and S8).
-Best-fit model distributions are largely consistent with our current understanding of blue whale distributions in the northeast Atlantic\cite{reeves2004historical,baines2014upwellings,baines2017autumn,reeves2004historical}, with perhaps greater importance of winter foraging in temperate regions (Figure \ref{fig4}).
+Better fitting models in general were those predicting a greater latitudinal foraging range, and foraging in more northerly waters (Supplemental Figures S7 and S8).
+Best-fit model distributions are largely consistent with our current understanding of blue whale distributions in the northeast Atlantic \cite{reeves2004historical,baines2014upwellings,baines2017autumn,reeves2004historical}, with perhaps greater importance of winter foraging in temperate regions (Figure \ref{fig4}).
 
@@
 Beginning in the winter of 1889/1890, the observed $\delta^{13}$C values are relatively low, and remain constant for approximately six months, before increasing rapidly in the winter of 1890/1891. 
-Low $\delta^{13}$C values are found in northern waters, but these regions also show large temporal fluctuations in $\delta^{13}$Cplk values (Supplementary Figure S4), and the observed values cannot be simulated purely from movements within the known geographic range.
+Low $\delta^{13}$C values are found in northern waters, but these regions also show large temporal fluctuations in $\delta^{13}$Cplk values (Supplemental Figure S4), and the observed values cannot be simulated purely from movements within the known geographic range.
 We infer that the $\delta^{13}$C record in the last period of life reflects pregnancy and calf-rearing. 
@@
 After six months the whale began ingesting protein with high $\delta^{13}$C  values, indicating feeding in low latitude waters shortly followed by a final northward migration. 
-Blue whales have a 10-12 month gestation period, with calving occurring in sub-tropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
+Blue whales have a 10-12 month gestation period, with calving occurring in sub-tropical waters, and calves are weaned after 6-7 months \cite{handbook}. 
 We therefore interpret the final year of the isotopic record as indicating: (a) birth of a calf in the winter of 1889/1890, presumably in sub-tropical waters, (b) six months of weaning during which time the maternal whale did not feed, followed by (c) a short period of feeding in sub-tropical waters in the second half of 1890, potentially during a final northward migration, and (d) eventual stranding during the return to northern feeding grounds in early 1891.
@@
 Although based on a single individual, our results highlight some interesting patterns worthy of further study.
-Firstly, next to nothing is known about juvenile behaviour in blue whales - calves become independent of their mothers at around seven months, and become sexually mature around 8-10 years old\cite{handbook}.
+Firstly, next to nothing is known about juvenile behaviour in blue whales - calves become independent of their mothers at around seven months, and become sexually mature around 8-10 years old \cite{handbook}.
 Our analyses suggest that between these points, juvenile North Atlantic blue whales may spend extended periods in warm, highly productive waters associated with upwelling systems, before migrating to higher latitude feeding grounds.
@@
 Whaling was an intense pressure for blue whales during the period we are analysing. 
-Before whaling in the North Atlantic began in 1868\cite{reilly2008balaenoptera}, there were an estimated 10,000-15,000 blue whales in the region\cite{sigurjonsson1995life}. 
-In the early 20th century, fisheries moved outside the area because stocks were so depleted\cite{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed\cite{sigurjonsson1995life}.
+Before whaling in the North Atlantic began in 1868 \cite{reilly2008balaenoptera}, there were an estimated 10,000-15,000 blue whales in the region \cite{sigurjonsson1995life}. 
+In the early 20th century, fisheries moved outside the area because stocks were so depleted \cite{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed \cite{sigurjonsson1995life}.
 The NHM blue whale thus represents an individual from a species at the brink of extirpation.
 Despite dramatic changes in anthropogenic pressures since 1891, the NHM whale and present-day blue whales in the North Atlantic have similar spatial ecologies; there are still populations of blue whales in the regions indicated by our model.
-This may reflect behavioural responses to whaling, as blue whales were more widely distributed in the past\cite{reeves2004historical}, or the selective extinction of populations outside of these areas.
+This may reflect behavioural responses to whaling, as blue whales were more widely distributed in the past \cite{reeves2004historical}, or the selective extinction of populations outside of these areas.
 
@@
 Such techniques can also be applied to extinct species behaviours and used to understand how populations were affected by past pressures such as hunting, and present-day pressures such as global change.
-Our results also add to previous research\cite{lister2011natural,ryan2013stable} highlighting museums as rich archives of individual-level behavioural information.
+Our results also add to previous research \cite{lister2011natural,ryan2013stable} highlighting museums as rich archives of individual-level behavioural information.
 
-\section{Methods}
-%<3000 words
-% no figures no tables
+\section{Acknowledgments}
+This work was funded by the British Ecological Society (grant: 5771/6815). 
+We thank C.J.
+Somes for providing  $\delta^{15}$N POM data, Bastian Hambach and Megan Spencer at the University of Southampton SEAPORT isotope laboratory for assistance with stable isotope analyses, and Andrew Yool for allowing us to use and share NEMO-MEDUSA outputs.
 
-\subsection{Stable isotope extractions from baleen}
+\section{Author Contributions}
+CT, NC, AJ, RS, and KC collected baleen from the NHM collections. 
+KC and CT extracted and analysed samples.
+CT created the movement model.
+EC, NC, SM and RS helped parameterise the model.
+AJ and NC created the figures. 
+CT and NC wrote the paper.
+All authors edited drafts and approved the final version.
+
+\section{Declaration of Interests}
+The authors declare no competing interests.
+
+\bibliographystyle{currentbiol}
+\bibliography{blue-whale}
+
+\section{STAR Methods}
+%Specific headers
+
+\subsection{Contact for reagent and resource sharing}
+Further information and requests for resources and reagents should be directed to and will be fulfilled by the Lead Contact, Natalie Cooper (natalie.cooper@nhm.ac.uk).
+
+\subsection{Method details}
+\subsubsection{Stable isotope extractions from baleen}
 Baleen was collected from the Natural History Museum, London (specimen NHMUK.1892.3.1.1). 
@@
 97 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal (gingival) section that contains the most recent tissue. 
-Baleen grows at a constant rate, so the samples are equally spaced through time\cite{best1996stable}. 
+Baleen grows at a constant rate, so the samples are equally spaced through time \cite{best1996stable}. 
 Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the University of Southampton SEAPORT Stable Isotope Ratio Mass Spectrometry Laboratory (Southampton, UK), using a Vario Isotope select elemental analyser, coupled to an Isoprime 100 isotope mass spectrometer.
 Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
-C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples\cite{hobson1998stable}.
-All data are available from the NHM Data Portal\cite{data-set}.
+C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples \cite{hobson1998stable}.
+All data are available from the NHM Data Portal \cite{data-set}.
 
-\subsection{Time calibrating stable isotope profiles}
-Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates\cite{hobson1998stable,busquets2017estimating}.
+\subsubsection{Time calibrating stable isotope profiles}
+Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates \cite{hobson1998stable,busquets2017estimating}.
 Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
-We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Supplementary Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated for six recent blue whales from the Californian Pacific\cite{busquets2017estimating}.
+We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis \cite{cardona2017temporal} (Supplemental Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated for six recent blue whales from the Californian Pacific \cite{busquets2017estimating}.
 Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891. 
 Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase two (70cm to 18cm on the baleen plate), but no clear relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase one (the oldest 30cm of the baleen plate, i.e.
-100cm - 70cm; Supplementary Figure S2) or the final 500 days of the whale's life (youngest 18cm of the baleen plate; Supplementary Figure S2).
+100cm - 70cm; Supplemental Figure S2) or the final 500 days of the whale's life (youngest 18cm of the baleen plate; Supplemental Figure S2).
 
-\subsection{Baseline isotope comparisons}
-Isotope-enabled biogeochemical ocean models\cite{magozzi2017using,schmittner2016complementary} were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Supplementary Figure S3).
-Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by C.J.
-Somes (\textit{pers.comm}) based on a 5$^{\circ}$ resolution biogeochemical model (Supplementary Figure S3).
-$\delta^{13}$C POM values were simulated at 1$^{\circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model\cite{magozzi2017using,yool2013medusa}. 
+\subsubsection{Baseline isotope comparisons}
+Isotope-enabled biogeochemical ocean models \cite{magozzi2017using,schmittner2016complementary} were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Supplemental Figure S3).
+Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by C.J. Somes (\textit{pers.comm}) based on a 5$^{\circ}$ resolution biogeochemical model (Supplemental Figure S3). 
+$\delta^{13}$C POM values were simulated at 1$^{\circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model \cite{magozzi2017using,yool2013medusa}. 
 Simulated  $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c. 60$^{\circ}$N, and relatively negative in the central and southern North Atlantic. 
 Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
-In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Supplementary Figure S3). 
-The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature\cite{magozzi2017using,laws1995dependence}.
+In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Supplemental Figure S3). 
+The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \cite{magozzi2017using,laws1995dependence}.
 Thus temporal variations in $\delta^{13}$C POM values were superimposed on latitudinal gradients. 
@@
 
-\subsection{Agent-based whale movement model}
+\subsection{Quantification and statistical analysis}
+\subsubsection{Agent-based whale movement model}
 We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
-At each time point and location in the simulations, we extracted time averaged $\delta^{13}$C values from the spatio-temporal model\cite{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figures \ref{fig2}, \ref{fig3}, and Supplementary Figure S4).
+At each time point and location in the simulations, we extracted time averaged $\delta^{13}$C values from the spatio-temporal model \cite{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figures \ref{fig2}, \ref{fig3}, and Supplemental Figure S4).
 Movement was coded as a set of probabilistic rules. 
-The parameters for these were taken from the literature on blue whale behaviour (e.g.\cite{handbook}). 
+The parameters for these were taken from the literature on blue whale behaviour (e.g. \cite{handbook}). 
 All terms were expressed as probability distributions, yielding multiple potential movement tracks.
@@
 Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
-(ii) Sea surface temperature\cite{yool2013medusa} ($^{\circ}$C). 
+(ii) Sea surface temperature \cite{yool2013medusa} ($^{\circ}$C). 
 When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.
-(iii) Water depth\cite{bathy} (m; extracted via the marmap package in R\cite{marmap}). 
+(iii) Water depth \cite{bathy} (m; extracted via the marmap package in R \cite{marmap}). 
 Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
-(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities\cite{yool2013medusa}). 
+(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities \cite{yool2013medusa}). 
 This was included as a proxy for zooplankton food availability. 
@@
 
-We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid-Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplementary Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and sub-tropical latitudes.
+We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid-Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and sub-tropical latitudes.
 Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four years of simulations.
-For a full description of the model and its parameters see Supplementary Methods: Supplementary Figure S5 and Supplementary Table S1. 
+For a full description of the model and its parameters see Supplemental Methods: Supplemental Figure S5 and Supplemental Table S1. 
 We simulated whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.
-For each movement simulation we extracted phytoplankton $\delta^{13}$C values at each time point and location from the models described above\cite{magozzi2017using}. 
+For each movement simulation we extracted phytoplankton $\delta^{13}$C values at each time point and location from the models described above \cite{magozzi2017using}. 
 We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).
 
-\subsection{Data Availability}
+\subsection{Data and Software Availability}
 Data are available from the NHM Data Portal (https://doi.org/10.5519/0093278). 
 R code is available from GitHub (https://github.com/nhcooper123/blue-whale-bes).
-% References
-
-\bibliographystyle{nature}
-\bibliography{blue-whale}
-
-%\section{Supplementary Information}
-
-\section{Acknowledgments}
-% adjectives not allowed!
-This work was funded by the British Ecological Society (grant: 5771/6815). 
-We thank C.J.
-Somes for providing  $\delta^{15}$N POM data, Bastian Hambach and Megan Spencer at the University of Southampton SEAPORT isotope laboratory for assistance with stable isotope analyses, and Andrew Yool for allowing us to use and share NEMO-MEDUSA outputs.
-
-\section{Author contributions}
-% rough draft at present feel free to edit
-CT, NC, AJ, RS, and KC collected baleen from the NHM collections. 
-KC and CT extracted and analysed samples.
-CT created the movement model.
-EC, NC, SM and RS helped parameterise the model.
-AJ and NC created the figures. 
-CT and NC wrote the paper.
-All authors edited drafts and approved the final version.
-
-%\section{Author information}
-%Reprints and permissions information is available at www.nature.com/reprints.
-%The authors declare no competing financial interests.
-%Correspondence and requests for materials should be addressed to natalie.cooper@nhm.ac.uk.
 
@@
 
-%\section{Figures}
+%\section{Tables}
 

### 2018-02-14 manuscript/PNAS/blue-whale-ms.tex
Add files for PNAS submission

--- 2018-02-14 manuscript/CurrentBiology/blue-whale-ms.tex
@@
 \usepackage{ms}            % load the template
-%\usepackage[round]{natbib} % author-year citations
+\usepackage[round, numbers]{natbib} % author-year citations
 %\usepackage[superscript,biblabel]{cite} % for superscript citations
@@
   $^1$ Ocean and Earth Science, University of Southampton Waterfront Campus, Southampton, SO14 3ZH, UK.\\
-  $^2$ School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
+  $^2$ Department of Zoology, School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
   $^3$ Department of Life Sciences, Natural History Museum London, Cromwell Road, London, SW7 5BD, UK.\\ 
@@
   $^5$ Department of Geology and Geophysics, University of Utah, Salt Lake City, UT 84112-0102, USA.\\
-  $^*$Correspondence: natalie.cooper@nhm.ac.uk; @nhcooper123. 
+  $^*$Corresponding author: +44(0)207942 5083; natalie.cooper@nhm.ac.uk. 
 }}
@@
 
+\textbf{Classification:} \textsc{Biological sciences:} Ecology.
+
 \mstitlepage
@@
 \addtolength{\parskip}{.9em}
-% Summary 250 words max
+% 250 words max
 
@@
 
-\section{Summary}
+\section{Abstract}
 Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change. 
@@
 
+\section{Significance statement}
+XXX 120 max
+
 \newpage
-\section{Introduction}
 Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites \cite{runge2014conserving}.
@@
 
-\section{Author Contributions}
-CT, NC, AJ, RS, and KC collected baleen from the NHM collections. 
-KC and CT extracted and analysed samples.
-CT created the movement model.
-EC, NC, SM and RS helped parameterise the model.
-AJ and NC created the figures. 
-CT and NC wrote the paper.
-All authors edited drafts and approved the final version.
-
-\section{Declaration of Interests}
-The authors declare no competing interests.
-
-\bibliographystyle{currentbiol}
-\bibliography{blue-whale}
-
-\section{STAR Methods}
-%Specific headers
-
-\subsection{Contact for reagent and resource sharing}
-Further information and requests for resources and reagents should be directed to and will be fulfilled by the Lead Contact, Natalie Cooper (natalie.cooper@nhm.ac.uk).
-
-\subsection{Method details}
-\subsubsection{Stable isotope extractions from baleen}
+\section{Methods}
+\subsection{Stable isotope extractions from baleen}
 Baleen was collected from the Natural History Museum, London (specimen NHMUK.1892.3.1.1). 
@@
 
-\subsubsection{Time calibrating stable isotope profiles}
+\subsection{Time calibrating stable isotope profiles}
 Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates \cite{hobson1998stable,busquets2017estimating}.
@@
 
-\subsubsection{Baseline isotope comparisons}
+\subsection{Baseline isotope comparisons}
 Isotope-enabled biogeochemical ocean models \cite{magozzi2017using,schmittner2016complementary} were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Supplemental Figure S3).
@@
 
-\subsection{Quantification and statistical analysis}
-\subsubsection{Agent-based whale movement model}
+\subsection{Agent-based whale movement model}
 We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
@@
 
-\subsection{Data and Software Availability}
+\subsection{Data and code availability}
 Data are available from the NHM Data Portal (https://doi.org/10.5519/0093278). 
 R code is available from GitHub (https://github.com/nhcooper123/blue-whale-bes).
+
+\bibliographystyle{pnas-new}
+\bibliography{blue-whale-abbreviations}
 

### 2018-02-14 manuscript/PNAS/pnas-blue-whale-ms.tex
Add files for PNAS submission

--- 2018-02-14 manuscript/PNAS/blue-whale-ms.tex
@@
-% Preamble
-\documentclass[a4paper,12pt]{article}
+\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
+% Use the lineno option to display guide line numbers if required.
+% Note that the use of elements such as single-column equations
+% may affect the guide line number alignment.
 
-\usepackage[osf]{mathpazo} % palatino
-\usepackage{ms}            % load the template
-\usepackage[round, numbers]{natbib} % author-year citations
-%\usepackage[superscript,biblabel]{cite} % for superscript citations
-\usepackage{graphicx}
-\usepackage{parskip} 
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{textcomp} % for parts per mille symbol     
-\pagenumbering{arabic}    
-\linespread{1.66}
+\templatetype{pnasresearcharticle} % Choose template 
+% {pnasresearcharticle} = Template for a two-column research article
+% {pnasmathematics} = Template for a one-column mathematics article
+% {pnasinvited} = Template for a PNAS invited submission
 
-% remove brackets from bibliography list
-\makeatletter 
-\renewcommand\@biblabel[1]{#1} 
-\makeatother
+\title{Reconstructing the last known movements of one of Nature's giants}
 
-% -------------------------------------------------------------------------------------------------------
-% some custom commands
+% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
+\author[a]{Clive N. Trueman}
+\author[b]{Andrew L. Jackson} 
+\author[a]{Katharyn S. Chadwick}
+\author[c,d]{Ellen J. Coombs}
+\author[a,e]{Sarah Magozzi}
+\author[c]{Richard C. Sabin}
+\author[c,1]{Natalie Cooper}
 
-% Add support for highlighting
-\usepackage{color}
-\newcommand{\hilight}[1]{\colorbox{yellow}{#1}}
+\affil[a]{Ocean and Earth Science, University of Southampton Waterfront Campus, Southampton, SO14 3ZH, UK.}
+\affil[b]{Department of Zoology, School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.}
+\affil[c]{Department of Life Sciences, Natural History Museum London, Cromwell Road, London, SW7 5BD, UK.}
+\affil[d]{Department of Earth Sciences, University College London, Gower Street, London, WC1E 6BT, UK.}
+\affil[e]{Department of Geology and Geophysics, University of Utah, Salt Lake City, UT 84112-0102, USA.}
+\affil[1]{Corresponding author}
 
-% -------------------------------------------------------------------------------------------------------
+% Please give the surname of the lead author for the running footer
+\leadauthor{Cooper}
 
-% Title page information
-\title{Reconstructing the last known movements of one of Nature's giants}
-% 150 characters max
+% Please add here a significance statement to explain the relevance of your work
+\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality.
+The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership.
+The Significance Statement appears in the paper itself and is required for all research papers.}
 
-\author{
-  Clive N. Trueman$^{1}$, Andrew L. Jackson$^{2}$, Katharyn S. Chadwick$^{1}$,\\ 
-  Ellen J. Coombs$^{3,4}$, Sarah Magozzi$^{1,5}$, Richard C. Sabin$^{3}$ \\
-  and Natalie Cooper$^{3*}$
-}
-\date{}
-\affiliation{\noindent{\footnotesize
-  $^1$ Ocean and Earth Science, University of Southampton Waterfront Campus, Southampton, SO14 3ZH, UK.\\
-  $^2$ Department of Zoology, School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
-  $^3$ Department of Life Sciences, Natural History Museum London, Cromwell Road, London, SW7 5BD, UK.\\ 
-  $^4$ Department of Earth Sciences, University College London, Gower Street, London, WC1E 6BT, UK.\\
-  $^5$ Department of Geology and Geophysics, University of Utah, Salt Lake City, UT 84112-0102, USA.\\
-  $^*$Corresponding author: +44(0)207942 5083; natalie.cooper@nhm.ac.uk. 
-}}
+% Please include corresponding author, author contribution and author declaration information
+\authorcontributions{Please provide details of author contributions here.}
+\authordeclaration{We have no conflicts of interest.}
+\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: natalie.cooper@nhm.ac.uk.}
 
-\vfill
+% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
+\keywords{stable isotope analysis $|$ movement models $|$ museum collections $|$ blue whale $|$ baleen}
 
-%\runninghead{}
-\keywords{stable isotope analysis, movement models, museum collections, blue whale, baleen}
-
-% End of preamble
-
-\begin{document}
-\modulolinenumbers[1]   % Line numbering on every line
-
-\textbf{Classification:} \textsc{Biological sciences:} Ecology.
-
-\mstitlepage
-
-\parindent = 1.5em
-\addtolength{\parskip}{.9em}
-% 250 words max
-
-\raggedright
-
-\section{Abstract}
+\begin{abstract}
 Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change. 
@@
 Combining biochemical tracers with simulation modeling unlocks the information recorded in these irreplaceable archives.
+\end{abstract}
 
-\section{Significance statement}
-XXX 120 max
+\dates{This manuscript was compiled on \today}
+\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
 
-\newpage
+\begin{document}
+
+% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
+% You should only change this length when you've finalised the article contents.
+\verticaladjustment{-2pt}
+
+\maketitle
+\thispagestyle{firststyle}
+\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
+
+\dropcap{T}
+
 Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites \cite{runge2014conserving}.
@@
 
-\section{Results and Discussion}
+\section*{Results and Discussion}
 
@@
 
-\section{Acknowledgments}
-This work was funded by the British Ecological Society (grant: 5771/6815). 
-We thank C.J.
-Somes for providing  $\delta^{15}$N POM data, Bastian Hambach and Megan Spencer at the University of Southampton SEAPORT isotope laboratory for assistance with stable isotope analyses, and Andrew Yool for allowing us to use and share NEMO-MEDUSA outputs.
-
-\section{Methods}
-\subsection{Stable isotope extractions from baleen}
+% Methods section
+\matmethods{
+\subsection*{Stable isotope extractions from baleen}
 Baleen was collected from the Natural History Museum, London (specimen NHMUK.1892.3.1.1). 
@@
 
-\subsection{Time calibrating stable isotope profiles}
+\subsection*{Time calibrating stable isotope profiles}
 Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates \cite{hobson1998stable,busquets2017estimating}.
@@
 
-\subsection{Baseline isotope comparisons}
+\subsection*{Baseline isotope comparisons}
 Isotope-enabled biogeochemical ocean models \cite{magozzi2017using,schmittner2016complementary} were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Supplemental Figure S3).
@@
 
-\subsection{Agent-based whale movement model}
+\subsection*{Agent-based whale movement model}
 We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
@@
 
-\subsection{Data and code availability}
+\subsection*{Data and code availability}
 Data are available from the NHM Data Portal (https://doi.org/10.5519/0093278). 
 R code is available from GitHub (https://github.com/nhcooper123/blue-whale-bes). 
+}
 
-\bibliographystyle{pnas-new}
+\showmatmethods % Display the Materials and Methods section
+
+\acknow{This work was funded by the British Ecological Society (grant: 5771/6815). 
+We thank C.J.
+Somes for providing  $\delta^{15}$N POM data, Bastian Hambach and Megan Spencer at the University of Southampton SEAPORT isotope laboratory for assistance with stable isotope analyses, and Andrew Yool for allowing us to use and share NEMO-MEDUSA outputs.}
+
+\showacknow % Display the acknowledgments section
+
+% \pnasbreak splits and balances the columns before the references.
+% If you see unexpected formatting errors, try commenting out this line
+% as it can run into problems with floats and footnotes on the final page.
+% \pnasbreak
+
+% Bibliography
 \bibliography{blue-whale-abbreviations}
 
-%\section{Figure legends}
-
-%\section{Tables}
-
 \end{document}

### 2018-02-14 manuscript/CurrentBiology/blue-whale-ms.tex
Edits for PNAS formatting

@@
 \bibliographystyle{currentbiol}
-\bibliography{blue-whale}
+\bibliography{blue-whale-abbreviations}
 

### 2018-02-14 quarantine/Figure-best-simulation.R
Remove quarantined files

@@ -1,273 +1,0 @@
-rm(list = ls())
-
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Import and Process the empirical data
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-# Import Whale data
-whale_isos <- read.csv("data/KC.NHM.full.csv", header = TRUE, 
-                       stringsAsFactors = FALSE)
-
-KC7 <- filter(whale_isos, Whale == "KC7")
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# convert sample number to days using following parameters
-samp_resolution <- 1 # samples per centimetre
-
-growth_rate <- 13.5  # baleen growth rate in centimetres per year
-
-days <- KC7$Samp.No * samp_resolution * 365.25 / growth_rate
-
-days <- days - days[1]
-
-KC7$days <- days
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# convert to calendar date
-
-# reverse time
-rev_days <- days - days[length(days)]
-
-last_sample <- as.Date("1891-03-01")
-
-sample_date <- last_sample + rev_days
-
-KC7$rev_days <- rev_days
-KC7$date <- sample_date
-KC7$year <- year(sample_date)
-
-
-new_years <- with(KC7, seq(min(year), max(year), by = 1))
-
-new_years_chr <- paste(new_years, "01", "01", sep = "-")
-
-new_years_date <- as.Date(new_years_chr)
-
-new_years_day <- tail(KC7$days,1) - 
-  (as.numeric(last_sample - new_years_date ))
-
-new_years_samp <- round(new_years_day[2:length(new_years_day)] * growth_rate / 
-                          (samp_resolution * 365.25))
-
-new_years_all <- round(new_years_day * growth_rate / 
-                         (samp_resolution * 365.25))
-
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Import and process the simulation data
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# ----
-
-# which Rep will we plot
-do_rep = 43
-
-# Map plot of all simulated tracks for the full migratory model
-
-resTrack <- read.csv("data/record_migrate5.csv", header=TRUE)
-
-# remove NAs from the dataset
-# resTrack <- filter(resTrack, !is.na(d13C))
-
[…]

### 2018-02-14 quarantine/Figure2.R

@@ -1,346 +1,0 @@
-# Author: Clive Trueman
-# Adapted by: Andrew Jackson June 2017 & Natalie Cooper Oct 2017.
-# About: Script to generate figure 1
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Load libraries
-
-#library(raster)
-#library(gstat)
-#library(sp)
-##library(maps)
-#library(mapdata)
-#library(RColorBrewer)
-
-#library(tidyr)
-#library(dplyr)
-#library(ggplot2)
-#library(magrittr)
-#library(viridis)
-
-library(tidyverse)
-library(lubridate)
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Figure 2 - Map plot of all simulated tracks for the full migratory model
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-resTrack <- read_csv("data/Model.sims.csv")
-res <- resTrack
-
-# cut out last rep if it didn't work
-resTrack <- slice(res, -c(3213051:3213211))
-
-
-
-res <- filter(resTrack, Rep == 1)
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-x <- 
-  res %>%
-  group_by(Rep) %>%
-  summarise(max = max(count2)) %>%
-  #mutate(days = 1:max)
-  map(mutate, days = 1:max)
-
-remove <- x$Rep[x$max < 3019]
-
-resTrack <- slice(resTrack, -remove)
-
-
-
-# remove NAs from the dataset
-# resTrack <- filter(resTrack, !is.na(d13C))
-
-# create a new sequence of days on which to predict our
-# loess smoothed function and add it to the resTrack data.frame for 
-# each Rep
-days <- rep(seq(from = 1, to = 3019), length(resTrack$Rep))
-# resTrack$days 
-resTrack3 <- 
-  resTrack %>% 
-  group_by(Rep) %>% 
-  data.frame(days)
-
-
-# jitter the points
-resTrack$jLat <- jitter(resTrack$Lat, factor = 2)
-resTrack$jLon <- jitter(resTrack$Lon, factor = 1)
-
-# data required to add a vertical line indicating the start of migration
-migrate <- as.Date("1889-06-01")
-first_sim_day <- as.Date("1885-01-01")
-migrate_day <- as.numeric(migrate - first_sim_day)
-
[…]

### 2018-02-14 quarantine/Figure3.R

@@ -1,226 +1,0 @@
-# Author: Clive Trueman
-# Adapted by: Andrew Jackson June 2017 & Natalie Cooper Oct 2017.
-# About: Script to generate figure 1
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Load libraries
-
-#library(raster)
-#library(gstat)
-#library(sp)
-##library(maps)
-#library(mapdata)
-#library(RColorBrewer)
-
-#library(tidyr)
-#library(dplyr)
-#library(ggplot2)
-#library(magrittr)
-
-library(viridis)
-library(tidyverse)
-library(lubridate)
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Add raw whale isotope data
-# See figure 1 code for more detail
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# read in data and extract just the blue whale
-whale_isos <- read_csv("data/raw-whale-isotope-data.csv")
-KC7 <- filter(whale_isos, whale_ID == "KC7")
-growth_rate <- 13.5  # baleen growth rate in centimetres per year
-last_sample <- ymd("1891-03-01") # final date of last keratin sample
-
-KC7 <- 
-  KC7 %>%
-  mutate(days = sample_cm * 365.25 / growth_rate) %>%
-  mutate(days = days - days[1]) %>%
-  mutate(rev_days = days - days[length(days)]) %>%
-  mutate(sample_date = last_sample + rev_days) %>%
-  mutate(year = year(sample_date))
-
-# Create a variable to label the years
-year_labels <- with(KC7, seq(min(year), max(year), by = 1))
-year_breaks <- which(duplicated(KC7$year) == FALSE)
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# plot the isotope data
-point_size <- 4
-
-isop <- ggplot(KC7, aes(x = rev(sample_cm))) + 
-  theme_classic(base_size = 16) + 
-  xlab("Baleen sample (cm from youngest sample)") + 
-  scale_x_continuous(breaks = seq(0, 100, 20), 
-                     labels = seq(100, 0, -20),
-                     sec.axis = sec_axis(~., 
-                                         breaks = rev(year_breaks),
-                                         labels = rev(year_labels),
-                                         name = "Year"))
-
-isop <- isop + geom_line(aes(y = d13C)) + 
[…]

### 2018-02-14 quarantine/Whale model code NEE.R

@@ -1,951 +1,0 @@
-# Whale individual based movement model
-# Author: Clive Trueman (trueman@noc.soton.ac.uk)
-# This version Sept. 2017
-
-
-
-# run the Lat.Lon code to get the km conversion first.
-# Degree to km function
-
-# Calculate distance in kilometers between two points
-earth.dist <- function (long1, lat1, long2, lat2)
-{
-	
-rad <- pi/180
-a1 <- lat1 * rad
-a2 <- long1 * rad
-b1 <- lat2 * rad
-b2 <- long2 * rad
-dlon <- b2 - a2
-dlat <- b1 - a1
-a <- (sin(dlat/2))^2 + cos(a1) * cos(b1) * (sin(dlon/2))^2
-c <- 2 * atan2(sqrt(a), sqrt(1 - a))
-R <- 6378.145
-d <- R * c
-return(d)
-}
-
-# build array for distance vectors
-Lat.Lon<-array(data=0, dim=c(180, 360,2))
-
-	#longitude distance varies with lat
-	Lt=vector(length=179)
-	count=1
-	
-	for(Lat in 89:-90){
-		x<-earth.dist(-180,Lat, -179,Lat)
-		Lt[count]<-x
-		count=count+1
-		}
-
-Lat.Lon[1:180,1,2]<-0
-	Lat.Lon[1:180,2,2]<-Lt
-	for(Lon in 3:359){
-	Lat.Lon[1:180,Lon,2]<-Lat.Lon[,Lon-1,2]+Lat.Lon[,2,2]
-	}
-	
-		#latitude distance constant at 111km
-
-	Lt2=rep(0,179)
-	count=2
-		for(Lat in -89:89){
-
-		Lt2[count]<-Lt2[count-1]+111
-		count=count+1
-		}
-
-Lat.Lon[,,1]<-Lt2
-
-
-Lon2<-Lat.Lon[,1:359,2]/Lat.Lon[,359,2]*360
-
-
-#########################################################
-
-library(R.matlab)
-library(maps)
-library(mapdata)
-library(scales)
-library(TTR)
-require(maptools)
-require(raster) 
-
-
-
-# link to monthly isoscape layers 
[…]

### 2018-02-14 quarantine/best_fit_All.R

@@ -1,630 +1,0 @@
-library(plyr)
-library(maps)
-library(mapdata)
-library(R.utils)
-
-
-#read in the measured whale data
-setwd("/Users/trueman/Desktop/R scripts/Whale")
-data<-read.csv("KC.NHM.fullB.csv", header=TRUE)
-colz<-colorRampPalette(c("brown", "cadet blue", "pink"))(12)
-colz<-seq(1:6)
-fin<-data[data$Spp=="fin",]
-minke<-data[data$Spp=="minke",]
-sei<-data[data$Spp=="sei",]
-blue<-data[data$Spp=="blue",]
-
-
-##HERE fix day numbers reflecting monthly samples (for loess sampling - fixed to have same no.s)
-predict.days<-seq(from=30, to=(8*365)+100, by=31)
-blue$Day.sim<-rev(predict.days)
-plot(y=blue$d13C, x=blue$Day.sim, type="l")
-par(new=TRUE)
-plot(y=blue$d15N, x=blue$Day.sim, type="l", axes=FALSE, col="red")
-
-#read in combined model files (BIIIIG)
-setwd("/Users/trueman/Desktop/R scripts/migrate.files/Model.files")
-resTrack<-read.csv("model.sims.csv", header=TRUE)
-# cut out last rep if it didn't work
-
-resTrack<-resTrack[-c(3213051:3213211),]
-
-
-##HERE fix day numbers reflecting monthly samples (for loess sampling)
-
-lengthS=length(resTrack$d13C[resTrack$Rep==1])
-mo_no<-lengthS/30+1
-sampleD<-rnorm(mo_no,30,1)
-
-sampledays<-as.integer(cumsum(sampleD))
-
-
-
-
-#extract model values only for days of sample - as defined above (cound change to an average if wanted)
-test<-ddply(resTrack, "Rep", function(x) {
-	newSeries<-x[x$count2%in%sampledays,]
-})
-
-test2<-ddply(test, "Rep", function(x) {
-	newSeries<-x[x$count2<3021,]
-})
-
-
-#run loess through series and predict for same days as measured
-
-test3<-ddply(test2, "Rep", function(x){
-	lo<-predict(loess(x$d13C~x$count2, span=0.05))[1:97]
-	
-})
-test3<-t(test3)
[…]

### 2018-02-14 quarantine/best_fit_All_NEW.R

@@ -1,793 +1,0 @@
-library(plyr)
-library(maps)
-library(mapdata)
-library(R.utils)
-library(raster)
-library(RColorBrewer)
-
-#read in the measured whale data
-setwd("/Users/trueman/Desktop/R scripts/Whale")
-data<-read.csv("KC.NHM.fullB.csv", header=TRUE)
-colz<-colorRampPalette(c("brown", "cadet blue", "pink"))(12)
-colz<-seq(1:6)
-fin<-data[data$Spp=="fin",]
-minke<-data[data$Spp=="minke",]
-sei<-data[data$Spp=="sei",]
-blue<-data[data$Spp=="blue",]
-
-
-##HERE fix day numbers reflecting monthly samples (for loess sampling - fixed to have same no.s)
-predict.days<-seq(from=30, to=(8*365)+100, by=31)
-blue$Day.sim<-rev(predict.days)
-plot(y=blue$d13C, x=blue$Day.sim, type="l")
-par(new=TRUE)
-plot(y=blue$d15N, x=blue$Day.sim, type="l", axes=FALSE, col="red")
-
-
-
-##### files to account for trophic level time integrations
-rm(list=ls())
-#Read in environmental values
-setwd("/Users/trueman/Desktop/R scripts/migrate.files/Model.files/")
-
-#plkd13c values
-d13Cjan<-raster("d13C_Jan.gri")
-d13Cfeb<-raster("d13C_Feb.gri") 
-d13Cmar<-raster("d13C_Mar.gri")
-d13Capr<-raster("d13C_Apr.gri")
-d13Cmay<-raster("d13C_May.gri")
-d13Cjun<-raster("d13C_Jun.gri")
-d13Cjul<-raster("d13C_Jul.gri")
-d13Caug<-raster("d13C_Aug.gri")
-d13Csep<-raster("d13C_Sep.gri")
-d13Coct<-raster("d13C_Oct.gri")
-d13Cnov<-raster("d13C_Nov.gri")
-d13Cdec<-raster("d13C_Dec.gri")
-
-TL1<-stack(d13Cjan, d13Cfeb, d13Cmar, d13Capr, d13Cmay, d13Cjun, d13Cjul, d13Caug, d13Csep, d13Coct, d13Cnov, d13Cdec)
-
-
-
-#d13C of different trophic levels
-TL2 <- stack("Trolev2_d13C.grd")
-TL3 <- stack("Trolev3_d13C.grd")
[…]

### 2018-02-14 quarantine/code-for-figures-ggplot2.R

@@ -1,398 +1,0 @@
-# Author: Clive Trueman
-# Adapted by: Andrew Jackson 19-Jun-2017
-# About: script to generate figures for the blue whale paper
-
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-# setwd("/Users/trueman/Desktop/R scripts/migrate.files")
-
-
-library(raster)
-library(gstat)
-library(sp)
-library(maps)
-library(mapdata)
-
-library(tidyr)
-library(dplyr)
-library(ggplot2)
-library(magrittr)
-library(viridis)
-library(lubridate)
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-rm(list = ls())
-graphics.off()
-
-
-##plot model output annual average baseline images
-Atl_carbon   <- raster("data/Atl_Annual_d13C.grd")
-Atl_nitrogen <- raster("data/Atl_Annual_d15N.grd")
-
-image(Atl_carbon)
-image(Atl_nitrogen)
-
-# set the constant to add as a correction to the d13C data
-cnst <- 6
-
-
-##########read resident whale model files
-
-resNor    <- read.csv("data/Norway.csv", header = TRUE)
-resIre    <- read.csv("data/west.Ireland.csv", header = TRUE)
-resCan    <- read.csv("data/Canaries.csv", header = TRUE)
-resMidAtl <- read.csv("data/Mid_Atl.csv", header = TRUE)
-
-
-
-
-#plot loess fits to the 30 resident simulated whales for each region
-
-
-
-# setEPS()
-# postscript("resident.eps", width=10, height=5)
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-#norway resident
-
-# use piping to break up the data.frame by replicate, then 
-# predict the loess
-tmp <- resNor %>% group_by(Rep) %>% 
-  do(., data.frame(Z = predict(loess((.$d13C + cnst) ~ .$Day.No, span = 0.1))))
-
-# append this vector onto the original data.frame
-resNor$lo <- tmp$Z
-  
-# plot via ggplot
-nor_plot <- ggplot(resNor, aes(Day.No, lo, group = Rep)) + 
-  geom_line(col = viridis(3)[2], alpha = 0.25) + 
-  xlab("Time") + 
-  ylab(expression(paste(delta^{13}, "C (\u2030)"))) + 
-  theme_classic(base_size = 14)
-print(nor_plot)
-
[…]

### 2018-02-14 quarantine/fig.s4 code.R

@@ -1,119 +1,0 @@
-setwd("/Users/trueman/Desktop/R scripts/migrate.files")
-library(raster)
-
-
-##########read resident whale model files with the trophic attenuation d13C values
-setwd("/Users/trueman/Desktop/R scripts/migrate.files/Model.files/blue.res")
-
-resNor<-read.csv("Norway.Res.TL.csv", header=TRUE)
-resIre<-read.csv("Ireland.Res.TL.csv", header=TRUE)
-resCan<-read.csv("Canaries.Res.TL.csv", header=TRUE)
-resMidAtl<-read.csv("Atl.Res.TL.csv", header=TRUE)
-resCV<-read.csv("CV.TL.csv", header=TRUE)
-
-
-#remove samples with NAs
-resNor<-na.omit(resNor)
-resIre<-na.omit(resIre)
-resCan<-na.omit(resCan)
-resMidAtl<-na.omit(resMidAtl)
-resCV<-na.omit(resCV)
-
-
-
-#plot loess fits to the 30 resident simulated whales for each region
-
-
-
-setEPS()
-postscript("resident.eps", width=10, height=5)
-par(new=FALSE)
-par(mfrow=c(2,3))
-
-
-
-#norway resident
-
-#get series length
-inds<-length(unique(resNor$Rep))
-
-#run loess on day number
-lo<-loess(resNor$TL4[resNor$Rep==1]+6~resNor$Day.No[resNor$Rep==1], span=0.1)
-
-# plot blank
-plot(x=resNor$Day.No[resNor$Rep==1], y=resNor$TL4[resNor$Rep==1], ylim=c(-23,-12), main="Norwegian Sea" , type="p", cex=0, col="red", xlab="Day.No", ylab="d13C")
-#add loess line
-lines(resNor$Day.No[resNor$Rep==1],predict(lo), col="red", lwd=0.2)
-
-par(new=TRUE)
-# run loess and plot for each of the 30 simulations 
-
-for(w in 2:inds){
-	lo<-loess(resNor$TL4[resNor$Rep==w]+6~resNor$Day.No[resNor$Rep==w], span=0.1)
-lines(resNor$Day.No[resNor$Rep==w],predict(lo), col="red", lwd=0.2)
-
-par(new=TRUE)
-}
-par(new=FALSE)
-
-
[…]

### 2018-02-14 quarantine/figures-1-2-3.R

@@ -1,415 +1,0 @@
-# Author: Clive Trueman
-# Adapted by: Andrew Jackson 19-Jun-2017
-# About: script to generate figure 1
-
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-# setwd("/Users/trueman/Desktop/R scripts/migrate.files")
-
-
-library(raster)
-library(gstat)
-library(sp)
-library(maps)
-library(mapdata)
-library(RColorBrewer)
-
-library(tidyr)
-library(dplyr)
-library(ggplot2)
-library(magrittr)
-library(viridis)
-library(lubridate)
-
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-rm(list = ls())
-graphics.off()
-
-
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Panel A - raw whale isotope data with two y-axes
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-
-
-whale_isos <- read.csv("data/KC.NHM.full.csv", header = TRUE, 
-                       stringsAsFactors = FALSE)
-
-KC7 <- filter(whale_isos, Whale == "KC7")
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# convert sample number to days using following parameters
-samp_resolution <- 1 # samples per centimetre
-
-growth_rate <- 13.5  # baleen growth rate in centimetres per year
-
-days <- KC7$Samp.No * samp_resolution * 365.25 / growth_rate
-
-days <- days - days[1]
-
-KC7$days <- days
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# convert to calendar date
-
-# reverse time
-rev_days <- days - days[length(days)]
-
-last_sample <- as.Date("1891-03-01")
-
-sample_date <- last_sample + rev_days
-
-KC7$rev_days <- rev_days
-KC7$date <- sample_date
-KC7$year <- year(sample_date)
-
-
-new_years <- with(KC7, seq(min(year), max(year), by = 1))
-
-new_years_chr <- paste(new_years, "01", "01", sep = "-")
-
-new_years_date <- as.Date(new_years_chr)
-
-new_years_day <- tail(KC7$days,1) - 
-  (as.numeric(last_sample - new_years_date ))
-
-new_years_samp <- round(new_years_day[2:length(new_years_day)] * growth_rate / 
[…]

### 2018-02-15 manuscript/PNAS/CoverLetter.tex
Cover letter

\documentclass[11pt]{letter}
\usepackage[a4paper,left=2.5cm, right=2.5cm, top=1cm, bottom=1cm]{geometry}
\usepackage[osf]{mathpazo}
\usepackage{url}
\signature{Natalie Cooper on behalf of all coauthors}
\address{Natural History Museum, London \\ Cromwell Road \\ London, SW7 5BD \\ natalie.cooper@nhm.ac.uk}
\longindentation=0pt
\begin{document}

\begin{letter}{}
\opening{Dear Editors,}

Please find attached our letter, `Reconstructing the last known movements of one of Nature's giants', for consideration at PNAS.

We get few opportunities to study the behaviour of rare and enigmatic species in the wild, especially wide-ranging marine species. 
The blue whale is an excellent example. 
Despite famously known as the largest animal ever to have lived, we know very little about the behavioral ecology species as a whole, particularly in the North Atlantic and during juvenile life periods.

Museum collections offer an opportunity to gain information on the movements and life-history of such species. 
In 1891 a 25m long female blue whale stranded in Wexford, Ireland. 
From July 13th 2017, this whale has taken pride of place in the Natural History Museum's (NHM) Hintze Hall (we can provide a picture of this as a cover image if desired).
Natural stable isotope markers have been widely used to infer information about individual-level behavior from animal tissues, but the level of behavioral detail that can be inferred has been limited by a lack of contextual data. 
Here we describe an entirely new method for inferring individual level movement data from stable isotope records from incremental tissues, based on recent advances in computer simulations. 
As a high profile case study, we apply our method to the baleen of the NHM blue whale ``Hope'' to identify the most likely movements of this individual over the last seven years of her life. We identify three phases in the behaviour of the whale - a juvenile period spent in warm waters, short range seasonal migrations at northern latitudes, followed by isotopic signatures consistent with pregnancy, calf rearing in warmer waters, and then a return to her northern feeding grounds where she stranded. Our approach can be applied to any marine animal with incrementally grown tissues, and unlocks a wealth of ecological information stored in animal tissues. In the context of blue whales, we provide the first multi-annual record of the behavior of juvenile blue whales in the North Atlantic, and reveal crucial information about historical movements in blue whales at a point where the species was on the brink of extirpation in the North Atlantic.

Blue whales were the first species that humans legislated to save, and represent a big conservation success story, although North Atlantic stocks have not recovered to the same extent as other whales in the region. There is also evidence that blue whales are shifting their ranges due to warming waters off Iceland. Thus understanding historical movements of North Atlantic blue whales has never been more vital if we wish to preserve them for future generations.

We look forward to hearing from you,

\closing{Yours sincerely,}

\end{letter}
\end{document}

### 2018-02-15 manuscript/PNAS/blue-whale-ms.tex
Final PNAS submission edits

@@
 \section{Significance statement}
-XXX 120 max
+Blue whales, the largest animals to have ever lived, are rare and enigmatic. 
+We know little about their movements and life-history which hampers conservation efforts.
+Here we investigate historical blue whale movements using stable isotopes from the baleen of the Natural History Museum London's blue whale ``Hope'', a female whale that stranded in 1891 at the height of industrial whaling.
+We show that the whale spent at least two years in the sub-tropics as a juvenile, before beginning annual latitudinal migrations in the northeast Atlantic. 
+The whale then likely became pregnant and raised a calf in sub-tropical waters before stranding.
+Our results provide valuable insights into blue whale ecology, and also highlight museums as rich archives of individual-level behavioural information.
 

### 2018-05-02 manuscript/CurrentBiology/blue-whale-ms.tex
Edit abstract

@@
 \section{Summary}
-Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change. 
-Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions, but interpreting these records can be challenging.
-Here we combine stable isotope analyses of baleen with agent-based simulation models, to reconstruct seven years of movements of an iconic female blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
-We provide the longest continuous record of movement for an individual blue whale in the north Atlantic. 
-We show that the whale spent at least two years resident in sub-tropical waters as a juvenile, before commencing annual latitudinal migrations consistent with patterns of movements of blue whales in the northeast Atlantic inferred from observations, tagging and historic whaling data.
-The isotopic record is consistent with observations of opportunistic feeding throughout the year in blue whales, and suggests pregnancy and calf-rearing in the year before death.
-Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climate, ecological conditions and anthropogenic pressures, and reveal aspects of life history.
-Combining biochemical tracers with simulation modeling unlocks the information recorded in these irreplaceable archives.
+The spatial ecology of rare, migratory, oceanic mammals is challenging to study directly, leading to distinct knowledge gaps associated with multi-year migratory behavior at the individual scale.
+Animal tissues housed in museum collections provide biochemical records of behavior that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions, but interpreting these records is difficult.
+Many baleen whales face severe conservation threats, therefore knowledge of their movements and space use over multiple years are needed. 
+Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to infer seven years of movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
+We identify three distinct behavioral states that we interpret as residency in sub-tropical waters (potentially as a juvenile), three annual cycles of seasonal latitudinal migrations and pregnancy and calf-rearing in the year before death.
+Our approach combining simulation modeling and stable isotope data yields individual-level movement inferences that are consistent with the current population-based understanding of the movement ecology of blue whales in the northeast Atlantic ocean.
+Combining biochemical tracers with simulation modeling unlocks the information recorded in historic collections, providing a link between traditional population-based assessments of animal distributions and more recent methodologies drawing inferences from high-resolution observations of relatively few individuals.
+This approach can be adapted to infer movements in any animal where underlying stable isotope models and incrementally grown tissues are available.
 

### 2018-05-02 manuscript/CurrentBiology/blue-whale-ms.tex
Edits to intro

@@
 Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle \cite{robinson2009travelling}.
-Identifying threats to migratory species and developing effective conservation measures requires information on individual-level movements spanning multiple years, for both historical and present-day populations.
-Such data are scarce, especially for rare, wide-ranging and long-lived marine species \cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
+Identifying threats to migratory species, understanding species responses to change, and developing effective conservation measures all require information on individual-level movements spanning multiple years, ideally for both historical and present-day populations.
+With the development of electronic tagging technology, studies of animal populations have shifted from infrequent observations of many individuals with limited or no information about individual movement behavior to frequent observations of a few individuals, with detailed information about individual movement \cite{holdo2013inferring}.
+However, such data are scarce, especially for rare, wide-ranging and long-lived marine species \cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
 
 The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
-Despite hunting for blue whales being banned for over 50 years, blue whales are still listed as Endangered on the IUCN Red List \cite{reilly2008balaenoptera}.
-In the northeast Atlantic, blue whales appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations \cite{pike2009note}.  
-The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland \cite{pike2009note}.
-Following movements of individual blue whales is extremely challenging due to their scarcity and oceanic habit.
+Although hunting for blue whales has been banned for over 50 years, blue whales are still listed as Endangered on the IUCN Red List \cite{reilly2008balaenoptera}.
+Blue whales in the northeast Atlantic appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations \cite{pike2009note}.  
+The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, currently fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland \cite{pike2009note}.
+Their small population size and oceanic habit makes blue whale movements extremely hard to study.
+
+Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records. 
+While these methods are adding vital behavioural data to our understanding of whale movements \cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies \cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour \cite{walker2012review}.
+At present the longest period of continuous monitoring via satellite tagging of an individual blue whale in the north Atlantic, that we are aware of, is 177 days \cite{lesage2017foraging}, but most satellite deployments on blue whales report movement data for under three months \cite{heide2001new,silva2013north, bailey2009behavioural,lesage2017foraging}, and are unable to show individual movements between feeding and wintering areas.
+To our knowledge photographic identification has matched summer and winter locations of a single north-east Atlantic blue whale with sightings in Iceland and Mauritania (REF).
+
 Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
-A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggest that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde islands \cite{baines2014upwellings}.
-Northward migration in the spring may occur along mid-Atlantic corridors with peak sightings in the Azores around April-May.
+A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands \cite{baines2014upwellings}.
+Northward migration in the spring may occur along mid Atlantic corridors with peak sightings in the Azores around April-May \cite{silva2013north}. 
 Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas \cite{pike2009note}. 
-Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southward-migrating animals \cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
-While direct tagging methods are adding vital behavioural data to our understanding of whale movements \cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies \cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour \cite{walker2012review}.
-Conservation efforts are consequently hampered by the lack of detailed data on the distribution, behaviour and ecology of blue whales, particularly during the winter and for the juvenile phase of life.
-An alternative technique for investigating whale movements retrospectively is to use intrinsic biochemical information such as stable isotopes \cite{west2006stable,busquets2017estimating,hobson2008tracking}.
+Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals \cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
+Inferences concerning movements of blue whales in the northeast Atlantic are therefore largely drawn from disparate information sources and rarely detail movements of individual animals. 
+It is difficult to know whether spatial variations in observed distributions reflect movements undertaken by individual animals or the frequency that individuals may undertake movements. 
+Information on historical movement patterns of individual blue whales is particularly important for understanding the drivers of whale declines, and potential human-whale conflicts as populations recover.
+
+An alternative technique for investigating whale movements retrospectively is to use intrinsic biochemical information such as stable isotope compositions \cite{west2006stable,busquets2017estimating,hobson2008tracking}.
+Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioral information that are often under-utilised \cite{lister2011natural}.
 The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
@@
 `trophic geography' \cite{bird2018global}), but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values.
-Here we reconstruct movements of an historic whale by comparing isotopic compositions of samples from museum collections, rich archives of behavioural information \cite{lister2011natural}, to spatially referenced models of isotopic variation in diet across the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
-Mysticete (baleen) whales are characterized by the development of baleen plates, keratinous structures in the upper jaw used to filter food items from seawater. 
-Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert \cite{best1996stable}. 
-Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although as baleen is worn away at the tips over time, an individual's entire lifespan is rarely recorded.
+Here we use the isotopic compositions of whale tissues to reconstruct movement by comparing them to spatially referenced models of isotopic variation in diet across the species' range in the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
+
+Mysticete (baleen) whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater. 
+Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert \cite{best1996stable,hobson1998stable}. 
+Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although as baleen is worn away at the tips over time, so rarely records an individual's entire lifespan.
 Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales \cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}, but have yet to be applied to blue whales from the North Atlantic population.
-Here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale ``Hope'', a female whale estimated to be a young adult c.15 years at her death in 1891.
+Here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale ``Hope''. 
+Hope stranded in Wexford Ireland in March 1891 and is a young adult female, estimated to be at least 15 years at death, suggesting that her baleen will yield information about migratory behaviors in north Atlantic blue whales around the peak of industrial whaling.
 
@@
 Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton \cite{handbook}.
-Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on.
+Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on in the weeks prior to keratin growth.
 We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth, resulting in 97 discrete samples of baleen.
@@
 The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
-In behavioural phase one (from the start of the record in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Figure \ref{fig1}; Supplemental Figures S1 and S2).
+In behavioural phase one (from the start of the record to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Figure \ref{fig1}; Supplemental Figures S1 and S2).
 In behavioural phase two (summer 1886 to spring 1890) $\delta^{15}$N values are relatively high and $\delta^{13}$C values are relatively low with coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values.
-In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively high $\delta^{13}$C values in the second half of 1891.
+In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively high $\delta^{13}$C values in the second half of 1890.
 The final three months of the record show a progressive fall in $\delta^{13}$C values (Figure \ref{fig1}; Supplemental Figures S1 and S2).

### 2018-05-02 manuscript/CurrentBiology/blue-whale-ms.tex
Finish edits

@@
   $^5$ Department of Geology and Geophysics, University of Utah, Salt Lake City, UT 84112-0102, USA.\\
-  $^*$Correspondence: natalie.cooper@nhm.ac.uk; @nhcooper123. 
+  $^*$Correspondence and lead contact: natalie.cooper@nhm.ac.uk; @nhcooper123. 
 }}
@@
 \section{Summary}
-The spatial ecology of rare, migratory, oceanic mammals is challenging to study directly, leading to distinct knowledge gaps associated with multi-year migratory behavior at the individual scale.
-Animal tissues housed in museum collections provide biochemical records of behavior that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions, but interpreting these records is difficult.
+The spatial ecology of rare, migratory, oceanic mammals is challenging to study directly, leading to distinct knowledge gaps associated with multi-year migratory behaviour at the individual scale.
+Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions, but interpreting these records is difficult.
 Many baleen whales face severe conservation threats, therefore knowledge of their movements and space use over multiple years are needed. 
 Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to infer seven years of movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
-We identify three distinct behavioral states that we interpret as residency in sub-tropical waters (potentially as a juvenile), three annual cycles of seasonal latitudinal migrations and pregnancy and calf-rearing in the year before death.
+We identify three distinct behavioural states that we interpret as residency in sub-tropical waters (potentially as a juvenile), three annual cycles of seasonal latitudinal migrations and pregnancy and calf-rearing in the year before death.
 Our approach combining simulation modeling and stable isotope data yields individual-level movement inferences that are consistent with the current population-based understanding of the movement ecology of blue whales in the northeast Atlantic ocean.
@@
 This approach can be adapted to infer movements in any animal where underlying stable isotope models and incrementally grown tissues are available.
+
+\textbf{Keywords:} stable isotope analysis, movement models, museum collections, blue whale, baleen.
 
@@
 Identifying threats to migratory species, understanding species responses to change, and developing effective conservation measures all require information on individual-level movements spanning multiple years, ideally for both historical and present-day populations.
-With the development of electronic tagging technology, studies of animal populations have shifted from infrequent observations of many individuals with limited or no information about individual movement behavior to frequent observations of a few individuals, with detailed information about individual movement \cite{holdo2013inferring}.
+With the development of electronic tagging technology, studies of animal populations have shifted from infrequent observations of many individuals with limited or no information about individual movement behaviour to frequent observations of a few individuals, with detailed information about individual movement \cite{holdo2013inferring}.
 However, such data are scarce, especially for rare, wide-ranging and long-lived marine species \cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
@@
 An alternative technique for investigating whale movements retrospectively is to use intrinsic biochemical information such as stable isotope compositions \cite{west2006stable,busquets2017estimating,hobson2008tracking}.
-Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioral information that are often under-utilised \cite{lister2011natural}.
+Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioural information that are often under-utilised \cite{lister2011natural}.
 The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
@@
 Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert \cite{best1996stable,hobson1998stable}. 
-Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although as baleen is worn away at the tips over time, so rarely records an individual's entire lifespan.
+Baleen plates therefore offer a continuous isotopic record of behaviour reflecting multiple years of life of an individual whale, although as baleen is worn away at the tips over time, so rarely records an individual's entire lifespan.
 Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales \cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}, but have yet to be applied to blue whales from the North Atlantic population.
 Here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale ``Hope''. 
-Hope stranded in Wexford Ireland in March 1891 and is a young adult female, estimated to be at least 15 years at death, suggesting that her baleen will yield information about migratory behaviors in north Atlantic blue whales around the peak of industrial whaling.
+Hope stranded in Wexford Ireland in March 1891 and is a young adult female, estimated to be at least 15 years at death, suggesting that her baleen will yield information about migratory behaviours in north Atlantic blue whales around the peak of industrial whaling.
 
@@
 We develop a new approach to address this problem using temporally dynamic models of $\delta^{13}$C values in the global ocean \cite{magozzi2017using} (Supplemental Figure S3), coupled to an agent-based model of whale movements (see Supplemental Methods; Supplemental Table S1; Supplemental Figure S5).
-We use our coupled model to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of North Atlantic blue whales.
-We then compare simulated profiles (Figures \ref{fig2}, \ref{fig3} and Supplemental Figure S4) to the measured isotopic profile of the blue whale's baleen (Figure \ref{fig1}) to identify the most likely movement history.
-Blue whales are commonly assumed to feed mainly in summer in productive, high latitude regions and rely largely on stored lipid reserves to meet metabolic demands in lower latitudes in winter. 
-However, in winter blue whales in the northeast Atlantic are associated with upwellings \cite{baines2017autumn}, while northward migrations in spring coincide with the onset and development of phytoplankton blooms \cite{silva2013north,visser2011timing,busquets2017estimating}, strongly implying feeding throughout the year.
+We use our coupled model to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in different parts of the known geographic range of North Atlantic blue whales.
+We then compare simulated profiles (Figures \ref{fig2}, \ref{fig3} and Supplemental Figure S4) to the measured isotopic profile of the blue whale's baleen (Figure \ref{fig1}), using z-score transformations to account for uncertainty in baseline variation, to identify the most likely movement history.
+
+Many baleen whales, such as humpback whales (\textit{Megaptera novaeangliae}), rely largely on stored lipid reserves to meet metabolic demands in lower latitudes in winter \cite{irvine2017quantifying}.
+However, increasing evidence suggests that blue whales engage in opportunistic feeding throughout the year.
+Whale distributions during winter coincide with the onset and development of phytoplankton blooms and areas of high productivity associated with upwelling \cite{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating}, and satellite tagged individuals express fine scale movement behaviour associated with area restricted foraging in winter and summer \cite{lesage2017foraging,bailey2009behavioural}.
 We therefore assume opportunistic ingestion of protein throughout the year.
@@
 The relatively positive and seasonally-invariant $\delta^{13}$C values seen during behavioural phase one are only found in sub-tropical areas of the North Atlantic.  
-Our simulations show high variance during this period (Figure \ref{fig3}), reflecting a range of possible locations for the whale, although areas around the Mauritanian coast and Cape Verde islands, a known current and historic winter feeding area for blue whales  \cite{baines2014upwellings,reeves2004historical}, and potentially to the west of the Azores (Figure \ref{fig2}), most closely match the measured profile.
-The whale remained in relatively warm waters for at least two full years. 
-Hope was estimated to be 15 years old when she died (based on vertebral epiphyseal fusion; R.C. Sabin \textit{pers.comm}), so was approximately 3-5 years old during this period.
-During behavioural phase two, starting at an approximate age of 5-6 years, the observed low $\delta^{13}$C values imply foraging in colder, more northerly latitudes. 
+Our simulations identify a range of possible locations for the whale (Figure \ref{fig3}), reflecting a range of possible locations for the whale, although areas around the Mauritanian coast and Cape Verde islands, a known current and historic winter feeding area for blue whales \cite{baines2014upwellings,reeves2004historical}, and potentially to the west of the Azores (Figure \ref{fig2}), most closely match the measured profile.
+The whale remained in relatively warm waters for at least one full year. 
+Hope was estimated to be at least 15 years old when she died (based on vertebral epiphyseal fusion; R.C. Sabin \textit{pers.comm}), so probably older than 10 years old during this period.
+During behavioural phase two, starting at an approximate age of at least 11 years, the observed low $\delta^{13}$C values imply foraging in colder, more northerly latitudes. 
 The pronounced cyclical variations in $\delta^{13}$C values observed during behavioural phase two could reflect either the isotopic expression of the spring phytoplankton bloom in northern waters \cite{magozzi2017using}, latitudinal migrations, or a combination of both.
@@
   \includegraphics[width = \linewidth]{figures/Figure-4-monthly.png}
-  \caption{Simulated locations by month taken from the top 10\% best fitting migratory movement models for behavioral phase two (summer 1886 to spring 1890) only.}
+  \caption{Simulated locations by month taken from the top 10\% best fitting migratory movement models for behavioural phase two (summer 1886 to spring 1890) only.}
   \label{fig4}
@@
 The isotopic record of the last 500 days of the whale's life are difficult to simulate. 
-Beginning in the winter of 1889/1890, the observed $\delta^{13}$C values are relatively low, and remain constant for approximately six months, before increasing rapidly in the winter of 1890/1891. 
+Beginning in the winter of 1889/1890, the observed $\delta^{13}$C values are relatively low, and remain constant for four to six months, before increasing rapidly in the second half of of 1890. 
 Low $\delta^{13}$C values are found in northern waters, but these regions also show large temporal fluctuations in $\delta^{13}$Cplk values (Supplemental Figure S4), and the observed values cannot be simulated purely from movements within the known geographic range.
-We infer that the $\delta^{13}$C record in the last period of life reflects pregnancy and calf-rearing. 
-We suggest that the constant, depleted $\delta^{13}$C values reflect exclusive use of carbon reserves assimilated in northern latitudes (i.e. a period of time with no opportunistic foraging). 
-After six months the whale began ingesting protein with high $\delta^{13}$C  values, indicating feeding in low latitude waters shortly followed by a final northward migration. 
+We infer that the $\delta^{13}$C record in the last period of life is associated with pregnancy and calf-rearing. 
+We suggest that the constant, low $\delta^{13}$C values reflect exclusive use of carbon reserves assimilated in northern latitudes (i.e. a period of time with limited opportunistic foraging). 
+After around 4-6 months the whale began ingesting food with relatively high $\delta^{13}$C  values, indicating feeding in low latitude waters shortly followed by a final northward migration. 
 Blue whales have a 10-12 month gestation period, with calving occurring in sub-tropical waters, and calves are weaned after 6-7 months \cite{handbook}. 
-We therefore interpret the final year of the isotopic record as indicating: (a) birth of a calf in the winter of 1889/1890, presumably in sub-tropical waters, (b) six months of weaning during which time the maternal whale did not feed, followed by (c) a short period of feeding in sub-tropical waters in the second half of 1890, potentially during a final northward migration, and (d) eventual stranding during the return to northern feeding grounds in early 1891.
-
-Data on the movement patterns of blue whales, particularly of juveniles, are scarce. 
-Although based on a single individual, our results highlight some interesting patterns worthy of further study.
-Firstly, next to nothing is known about juvenile behaviour in blue whales - calves become independent of their mothers at around seven months, and become sexually mature around 8-10 years old \cite{handbook}.
-Our analyses suggest that between these points, juvenile North Atlantic blue whales may spend extended periods in warm, highly productive waters associated with upwelling systems, before migrating to higher latitude feeding grounds.
-Extended periods of juvenile residency in sub-tropical waters could have implications for blue whale conservation in the northeast Atlantic, as the upwelling regions around Mauritania and the Cape Verde islands are areas with emerging offshore oil and gas exploration, and are fishing, tourist and shipping hotspots, potentially increasing threats to juvenile whales.
+We infer a chronology reflecting three years of uninterrupted annual latitudinal migrations leading to pregnancy in the year 1889 and birth of a calf in the winter of 1889/1890, presumably in sub-tropical waters.
+Following birth, we infer a period of around 4-6 months of residency in sub-tropical waters where the whale was sustained largely from stored lipid reserves, which we interpret as reflecting weaning.
+We subsequently propose that the whale had a short period of feeding in sub-tropical waters in the second half of 1890, potentially during a final northward migration during the return to northern feeding grounds in early 1891.
+The inferred reproductive time-line is therefore consistent with the assumed reproductive ecology of blue whales globally \cite{handbook}. 
+Long-term, multi-annual data on the movement patterns and reproductive ecology of individual blue whales are scarce. 
+Although based on a single individual, our results confirm that seasonal differences in population-level distributions inferred from observations of multiple individuals over time and space are indeed expressed in a single individual, with a detailed record proving information on residency, seasonal migrations and reproductive behaviour.
 
@@
 In the early 20th century, fisheries moved outside the area because stocks were so depleted \cite{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed \cite{sigurjonsson1995life}.
-The NHM blue whale thus represents an individual from a species at the brink of extirpation.
-Despite dramatic changes in anthropogenic pressures since 1891, the NHM whale and present-day blue whales in the North Atlantic have similar spatial ecologies; there are still populations of blue whales in the regions indicated by our model.
-This may reflect behavioural responses to whaling, as blue whales were more widely distributed in the past \cite{reeves2004historical}, or the selective extinction of populations outside of these areas.
+The NHM blue whale thus represents an individual from a species at the brink of extirpation, potentially representing  more than 0.1\% of the entire population of blue whales in the northeast Atlantic at that time.
+Inferring spatio-temporal distributions of populations from high density observations of a few individuals (or, in this case one individual) will always be problematic, but the predictive power linking a tracked individual to the population distribution increases when the tracked animal has a strong preference for specific habitats and there is a patchy (clumped) distribution of those habitats \cite{holdo2013inferring}.
+In the case of blue whales, the primary driver influencing seasonal movements is assumed to be the seasonally variable distribution of food resources between high and low latitudes. 
+As blue whales have a relatively restricted diet that shows highly predictable spatio-temporal differences in abundance at least on ocean basin scales, we argue that the seasonal latitudinal migrations inferred during behavioural phase two are likely to be common movement traits within north Atlantic blue whales.
+This inference of course requires testing with additional isotopic records from individual whales and/or multi-year satellite tracking records.
 
-Our use of movement simulation modeling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviours over multiple years.
+Little is known about juvenile behaviour in blue whales in the Atlantic - calves become independent of their mothers at around seven months, and become sexually mature around 8-10 years old. 
+The earliest part of our record may represent residency in subtropical waters prior to the onset of seasonal migrations at sexual maturity, but determining the age of the whale at death is highly uncertain, and unfortunately the baleen record is not long enough to resolve this issue with certainty.
+Our movement models suggest that the upwelling regions around Mauritania and the Cape Verde Islands may have been important historic calving areas, consistent with modern observations of juvenile blue whales in the northeast Atlantic \cite{handbook}.
+These areas are now fishing, tourist and shipping hotspots with emerging offshore oil and gas exploration, potentially increasing threats to the recovery of the small population of blue whales in the north east Atlantic.
+
+Our movement simulation modeling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviours over multiple years.
 Such techniques can also be applied to extinct species behaviours and used to understand how populations were affected by past pressures such as hunting, and present-day pressures such as global change.
-Our results also add to previous research \cite{lister2011natural,ryan2013stable} highlighting museums as rich archives of individual-level behavioural information.
+Our results also add to previous research \cite{lister2011natural,ryan2013stable} highlighting museums as critical sources of individual-level spatial ecology information.
 
 \section{Acknowledgments}
-This work was funded by the British Ecological Society (grant: 5771/6815). 
+This work was funded by the British Ecological Society (grant: 5771/6815) to NC, CT and AJ. 
 We thank C.J.
@@
 \section{Author Contributions}
+CT, NC and AJ conceived the project.
 CT, NC, AJ, RS, and KC collected baleen from the NHM collections. 
@@
 97 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal (gingival) section that contains the most recent tissue. 
-Baleen grows at a constant rate, so the samples are equally spaced through time \cite{best1996stable}. 
+We assume that baleen grows at a relatively constant rate, so the samples are equally spaced through time \cite{best1996stable}. 
 Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the University of Southampton SEAPORT Stable Isotope Ratio Mass Spectrometry Laboratory (Southampton, UK), using a Vario Isotope select elemental analyser, coupled to an Isoprime 100 isotope mass spectrometer.
@@
 Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates \cite{hobson1998stable,busquets2017estimating}.
-Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
+Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values at 70 - 18cm distance from the gingival end of the plate (behavioural phase two). 
 We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis \cite{cardona2017temporal} (Supplemental Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated for six recent blue whales from the Californian Pacific \cite{busquets2017estimating}.
@@
 The scale and nature of temporal variation in $\delta^{13}$C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in $\delta^{13}$C POM values linked to strongly seasonal phytoplankton growth dynamics.
+The model simulations used are forced with decadal climatological data from 2000-2010 \cite{magozzi2017using}. 
+We do not draw inferences about location from absolute $\delta^{13}$C values and thus the reduction in oceanic $\delta^{13}$C POM values caused by release of fossil carbon into the atmosphere (Suess effect) does not influence inferences about location or movement.
+Ocean basin scale spatio-temporal differences in $\delta^{13}$C values, largely reflecting relative differences in sea surface temperatures between trophic, temperate and arctic waters, have remained essentially similar over millennial timescales (refs) therefore we are confident that the Magozzi \textit{et al} \cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.
+
 We therefore used monthly simulated $\delta^{13}$C POM values to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours.

### 2018-05-02 manuscript/CurrentBiology/CoverLetter.tex
Cover letter for Current Biology

--- 2018-02-15 manuscript/PNAS/CoverLetter.tex
@@
 
-Please find attached our letter, `Reconstructing the last known movements of one of Nature's giants', for consideration at PNAS.
+Please find attached our letter, `Reconstructing the last known movements of one of Nature's giants', for consideration at Current Biology.
 
@@
 The blue whale is an excellent example. 
-Despite famously known as the largest animal ever to have lived, we know very little about the behavioral ecology species as a whole, particularly in the North Atlantic and during juvenile life periods.
+Despite famously known as the largest animal ever to have lived, we know very little about the behavioral ecology species as a whole, particularly in the North Atlantic.
 
@@
 In 1891 a 25m long female blue whale stranded in Wexford, Ireland. 
-From July 13th 2017, this whale has taken pride of place in the Natural History Museum's (NHM) Hintze Hall (we can provide a picture of this as a cover image if desired).
+From July 13th 2017, this whale has taken pride of place in the Natural History Museum London's (NHM) Hintze Hall (we can provide a picture of this as a cover image if desired).
 Natural stable isotope markers have been widely used to infer information about individual-level behavior from animal tissues, but the level of behavioral detail that can be inferred has been limited by a lack of contextual data.

### 2018-05-04 manuscript/CurrentBiology/blue-whale-ms.tex
Add final refs for Current Biology submission

@@
 At present the longest period of continuous monitoring via satellite tagging of an individual blue whale in the north Atlantic, that we are aware of, is 177 days \cite{lesage2017foraging}, but most satellite deployments on blue whales report movement data for under three months \cite{heide2001new,silva2013north, bailey2009behavioural,lesage2017foraging}, and are unable to show individual movements between feeding and wintering areas.
-To our knowledge photographic identification has matched summer and winter locations of a single north-east Atlantic blue whale with sightings in Iceland and Mauritania (REF).
+To our knowledge photographic identification has matched summer and winter locations of a single north-east Atlantic blue whale with sightings in Iceland and Mauritania \cite{poster}.
 
@@
 We do not draw inferences about location from absolute $\delta^{13}$C values and thus the reduction in oceanic $\delta^{13}$C POM values caused by release of fossil carbon into the atmosphere (Suess effect) does not influence inferences about location or movement.
-Ocean basin scale spatio-temporal differences in $\delta^{13}$C values, largely reflecting relative differences in sea surface temperatures between trophic, temperate and arctic waters, have remained essentially similar over millennial timescales (refs) therefore we are confident that the Magozzi \textit{et al} \cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.
+Ocean basin scale spatio-temporal differences in $\delta^{13}$C values, largely reflecting relative differences in sea surface temperatures between trophic, temperate and arctic waters, have remained essentially similar over millennial timescales \cite{schmittner2016complementary} therefore we are confident that the Magozzi \textit{et al} \cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.
 

### 2018-05-29 manuscript/BiologyLetters/blue-whale-ms.tex
Bio Lett version initial attempt at pruning

--- 2018-05-04 manuscript/CurrentBiology/blue-whale-ms.tex
@@
   $^5$ Department of Geology and Geophysics, University of Utah, Salt Lake City, UT 84112-0102, USA.\\
-  $^*$Correspondence and lead contact: natalie.cooper@nhm.ac.uk; @nhcooper123. 
+  $^*$Corresponding author: natalie.cooper@nhm.ac.uk. 
 }}
@@
 \addtolength{\parskip}{.9em}
-% Summary 250 words max
 
@@
 
-\section{Summary}
-The spatial ecology of rare, migratory, oceanic mammals is challenging to study directly, leading to distinct knowledge gaps associated with multi-year migratory behaviour at the individual scale.
-Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions, but interpreting these records is difficult.
-Many baleen whales face severe conservation threats, therefore knowledge of their movements and space use over multiple years are needed. 
+\section{Abstract} % 200 words max (199 at present)
+%Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions, but interpreting these records is difficult.
+%Many baleen whales face severe conservation threats, therefore knowledge of their movements and space use over multiple years are needed. 
+The spatial ecology of rare, migratory, baleen whales is challenging to study directly, leading to gaps in our knowledge of their individual-level migratory behaviours.
 Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to infer seven years of movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
-We identify three distinct behavioural states that we interpret as residency in sub-tropical waters (potentially as a juvenile), three annual cycles of seasonal latitudinal migrations and pregnancy and calf-rearing in the year before death.
+We identify three distinct behavioural states that we interpret as residency in sub-tropical waters, three annual cycles of seasonal latitudinal migrations and pregnancy and calf-rearing in the year before death.
 Our approach combining simulation modeling and stable isotope data yields individual-level movement inferences that are consistent with the current population-based understanding of the movement ecology of blue whales in the northeast Atlantic ocean.
@@
 \section{Introduction}
-Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites \cite{runge2014conserving}.
-Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle \cite{robinson2009travelling}.
+%Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites \cite{runge2014conserving}.
+%Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle \cite{robinson2009travelling}.
 Identifying threats to migratory species, understanding species responses to change, and developing effective conservation measures all require information on individual-level movements spanning multiple years, ideally for both historical and present-day populations.
 With the development of electronic tagging technology, studies of animal populations have shifted from infrequent observations of many individuals with limited or no information about individual movement behaviour to frequent observations of a few individuals, with detailed information about individual movement \cite{holdo2013inferring}.
-However, such data are scarce, especially for rare, wide-ranging and long-lived marine species \cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
-
+However, such data are scarce, especially for rare, wide-ranging and long-lived marine species \cite{ryan2013stable,bailey2009behavioural}. 
 The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
 Although hunting for blue whales has been banned for over 50 years, blue whales are still listed as Endangered on the IUCN Red List \cite{reilly2008balaenoptera}.
-Blue whales in the northeast Atlantic appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations \cite{pike2009note}.  
+%Blue whales in the northeast Atlantic appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations \cite{pike2009note}.  
 The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, currently fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland \cite{pike2009note}.
@@
 Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records. 
-While these methods are adding vital behavioural data to our understanding of whale movements \cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies \cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour \cite{walker2012review}.
-At present the longest period of continuous monitoring via satellite tagging of an individual blue whale in the north Atlantic, that we are aware of, is 177 days \cite{lesage2017foraging}, but most satellite deployments on blue whales report movement data for under three months \cite{heide2001new,silva2013north, bailey2009behavioural,lesage2017foraging}, and are unable to show individual movements between feeding and wintering areas.
-To our knowledge photographic identification has matched summer and winter locations of a single north-east Atlantic blue whale with sightings in Iceland and Mauritania \cite{poster}.
+While these methods are adding vital behavioural data to our understanding of whale movements \cite{bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies \cite{bailey2009behavioural,mate2007evolution}, and may influence whale behaviour \cite{walker2012review}.
+%At present the longest period of continuous monitoring via satellite tagging of an individual blue whale in the north Atlantic, that we are aware of, is 177 days \cite{lesage2017foraging}, but most satellite deployments on blue whales report movement data for under three months \cite{heide2001new,silva2013north, bailey2009behavioural,lesage2017foraging}, and are unable to show individual movements between feeding and wintering areas.
+%To our knowledge photographic identification has matched summer and winter locations of a single north-east Atlantic blue whale with sightings in Iceland and Mauritania \cite{poster}.
 
-Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
-A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands \cite{baines2014upwellings}.
-Northward migration in the spring may occur along mid Atlantic corridors with peak sightings in the Azores around April-May \cite{silva2013north}. 
-Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas \cite{pike2009note}. 
-Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals \cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
-Inferences concerning movements of blue whales in the northeast Atlantic are therefore largely drawn from disparate information sources and rarely detail movements of individual animals. 
-It is difficult to know whether spatial variations in observed distributions reflect movements undertaken by individual animals or the frequency that individuals may undertake movements. 
-Information on historical movement patterns of individual blue whales is particularly important for understanding the drivers of whale declines, and potential human-whale conflicts as populations recover.
+%Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
+%A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands \cite{baines2014upwellings}.
+%Northward migration in the spring may occur along mid Atlantic corridors with peak sightings in the Azores around April-May \cite{silva2013north}. 
+%Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas \cite{pike2009note}. 
+%Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals \cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
+%Inferences concerning movements of blue whales in the northeast Atlantic are therefore largely drawn from disparate information sources and rarely detail movements of individual animals. 
+%It is difficult to know whether spatial variations in observed distributions reflect movements undertaken by individual animals or the frequency that individuals may undertake movements. 
+%Information on historical movement patterns of individual blue whales is particularly important for understanding the drivers of whale declines, and potential human-whale conflicts as populations recover.
 
 An alternative technique for investigating whale movements retrospectively is to use intrinsic biochemical information such as stable isotope compositions \cite{west2006stable,busquets2017estimating,hobson2008tracking}.
-Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioural information that are often under-utilised \cite{lister2011natural}.
+%Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioural information that are often under-utilised \cite{lister2011natural}.
 The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
-Stable isotope data therefore provide information on diet and location (i.e.
-`trophic geography' \cite{bird2018global}), but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values.
-Here we use the isotopic compositions of whale tissues to reconstruct movement by comparing them to spatially referenced models of isotopic variation in diet across the species' range in the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
+Stable isotope data therefore provide information on diet and location, but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values.
+Here we use the isotopic compositions of whale tissues to reconstruct movement by comparing them to spatially referenced models of isotopic variation in diet across the species' range in the northeast Atlantic \cite{hobson2008tracking,eisenmann2016isotopic}.
 
@@
 
-\section{Results and Discussion}
+\section{Materials and Methods}
 
-Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton \cite{handbook}.
-Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on in the weeks prior to keratin growth.
-We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth, resulting in 97 discrete samples of baleen.
-We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin.
-These isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e.
+\subsubsection{Baleen and oceanic stable isotope comparisons}
+%\subsubsection{Stable isotope extractions from baleen}
+Baleen was collected from the Natural History Museum, London (specimen NHMUK.1892.3.1.1).  
+97 samples were taken at 1cm intervals, starting at the proximal (gingival) section that contains the most recent tissue. 
+We assume that baleen grows at a relatively constant rate, so the samples are equally spaced through time \cite{best1996stable}. 
+Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the University of Southampton SEAPORT Stable Isotope Ratio Mass Spectrometry Laboratory (Southampton, UK), using a Vario Isotope select elemental analyser, coupled to an Isoprime 100 isotope mass spectrometer.
+Carbon and nitrogen isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e.
 relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope \cite{west2006stable}.
-Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; both isotopes provide information on the geographical locations of an individual, but nitrogen isotopes are more strongly fractionated during protein metabolism and thus contain additional trophic and physiological information \cite{west2006stable}.
+%Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; both isotopes provide information on the geographical locations of an individual, but nitrogen isotopes are more strongly fractionated during protein metabolism and thus contain additional trophic and physiological information \cite{west2006stable}.
+%Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
+%C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples \cite{hobson1998stable}.
+All data are available from the NHM Data Portal \cite{data-set}.
 
-Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplemental Methods; Supplemental Figures S1 and S2).
+%\subsubsection{Time calibrating stable isotope profiles}
+%Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates \cite{hobson1998stable,busquets2017estimating}.
+%Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values at 70 - 18cm distance from the gingival end of the plate (behavioural phase two). 
+We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis 
+%\cite{cardona2017temporal} 
+(Supplemental Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ % which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated for six recent blue whales from the Californian Pacific \cite{busquets2017estimating}.
+Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891. 
+%Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase two (70cm to 18cm on the baleen plate), but no clear relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase one (the oldest 30cm of the baleen plate, i.e.
+100cm - 70cm; Supplemental Figure S2) or the final 500 days of the whale's life (youngest 18cm of the baleen plate; Supplemental Figure S2).
+
+Relating the isotopic compositions of animal tissues to likely locations is complicated by a lack of knowledge of spatio-temporal variations in the isotopic composition of diet \cite{west2006stable}.
+We develop a new approach to address this problem using temporally dynamic models of $\delta^{13}$C values in the global ocean \cite{magozzi2017using} (Supplemental Figure S3), coupled to an agent-based model of whale movements (see Supplemental Methods; Supplemental Table S1; Supplemental Figure S5).
+We use our coupled model to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in different parts of the known geographic range of North Atlantic blue whales.
+We then compare simulated profiles (Figures \ref{fig2}, \ref{fig3} and Supplemental Figure S4) to the measured isotopic profile of the blue whale's baleen (Figure \ref{fig1}), using z-score transformations to account for uncertainty in baseline variation, to identify the most likely movement history.
+
+%Many baleen whales, such as humpback whales (\textit{Megaptera novaeangliae}), rely largely on stored lipid reserves to meet metabolic demands in lower latitudes in winter \cite{irvine2017quantifying}.
+%However, increasing evidence suggests that blue whales engage in opportunistic feeding throughout the year.
+%Whale distributions during winter coincide with the onset and development of phytoplankton blooms and areas of high productivity associated with upwelling \cite{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating}, and satellite tagged individuals express fine scale movement behaviour associated with area restricted foraging in winter and summer \cite{lesage2017foraging,bailey2009behavioural}.
+%We therefore assume opportunistic ingestion of protein throughout the year.
+
+%\subsubsection{Baseline isotope comparisons}
+Isotope-enabled biogeochemical ocean models \cite{magozzi2017using,schmittner2016complementary} were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Supplemental Figure S3).
+Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by C.J. Somes (\textit{pers.comm}) based on a 5$^{\circ}$ resolution biogeochemical model (Supplemental Figure S3). 
+$\delta^{13}$C POM values were simulated at 1$^{\circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model \cite{magozzi2017using,yool2013medusa}. 
+%Simulated $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c. 60$^{\circ}$N, and relatively negative in the central and southern North Atlantic. 
+%Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
+%In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Supplemental Figure S3). 
+%The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \cite{magozzi2017using,laws1995dependence}.
+%Thus temporal variations in $\delta^{13}$C POM values were superimposed on latitudinal gradients. 
+%The scale and nature of temporal variation in $\delta^{13}$C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in $\delta^{13}$C POM values linked to strongly seasonal phytoplankton growth dynamics.
+%The model simulations used are forced with decadal climatological data from 2000-2010 \cite{magozzi2017using}. 
+%We do not draw inferences about location from absolute $\delta^{13}$C values and thus the reduction in oceanic $\delta^{13}$C POM values caused by release of fossil carbon into the atmosphere (Suess effect) does not influence inferences about location or movement.
+Ocean basin scale spatio-temporal differences in $\delta^{13}$C values, largely reflecting relative differences in sea surface temperatures between trophic, temperate and arctic waters, have remained essentially similar over millennial timescales \cite{schmittner2016complementary} therefore we are confident that the Magozzi \textit{et al} \cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.
+
+We used monthly simulated $\delta^{13}$C POM values to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours.
+Whales feed on krill, not POM. 
+Assimilation of carbon into krill tissues will dampen the temporal variability seen in POM, effectively producing a temporal average over the timescale of isotopic turnover within krill. 
+We estimate turnover to be complete between four and six months and therefore we resampled the $\delta^{13}$C POM values in each one degree cell to reflect a six month moving average. 
+Carbon isotope values are also likely to be fractionated during transfer from plankton to krill as $^{12}$C  is preferentially lost through respiration. 
+The degree of such trophic fractionation is unclear, however and we do not draw interpretations based on absolute $\delta^{13}$C values, rather on the relative $\delta^{13}$C values across the length of the baleen plate.
+
+\subsubsection{Agent-based whale movement model}
+We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
+At each time point and location in the simulations, we extracted time averaged $\delta^{13}$C values from the spatio-temporal model \cite{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figures \ref{fig2}, \ref{fig3}, and Supplemental Figure S4).
+Movement was coded as a set of probabilistic rules. 
+The parameters for these were taken from the literature on blue whale behaviour (e.g. \cite{handbook}). 
+All terms were expressed as probability distributions, yielding multiple potential movement tracks.
+
+%In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and linear distance of movement are all influenced by the following. 
+%(i) Behavioural state (migrating north, migrating south, or foraging).
+%Northerly migrations were possible only in spring, and southerly migrations in autumn. 
+%Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
+%(ii) Sea surface temperature \cite{yool2013medusa} ($^{\circ}$C). 
+%When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.
+%(iii) Water depth \cite{bathy} (m; extracted via the marmap package in R \cite{marmap}). 
+%Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
+%(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities \cite{yool2013medusa}). 
+%This was included as a proxy for zooplankton food availability. 
+%Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state.
+
+%We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid-Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and sub-tropical latitudes.
+%Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four years of simulations.
+For a full description of the model and its parameters see Supplemental Methods: Supplemental Figure S5 and Supplemental Table S1. 
+We simulated whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.
+For each movement simulation we extracted phytoplankton $\delta^{13}$C values at each time point and location from the models described above \cite{magozzi2017using}. 
+We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).
+
+ We simulated 1200 individual movement patterns, then excluded simulations where the virtual whale stranded before reaching the 3019 days represented in the baleen.
+We then compared the remaining 1049 simulations and the resulting simulated baleen $\delta^{13}$C records, to the measured records with simple linear regressions.
+
+\section{Results}
+
+%Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton \cite{handbook}.
+%Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on in the weeks prior to keratin growth.
+%We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth, resulting in 97 discrete samples of baleen.
+%We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin.
+%These isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e.
+relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope \cite{west2006stable}.
+%Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; both isotopes provide information on the geographical locations of an individual, but nitrogen isotopes are more strongly fractionated during protein metabolism and thus contain additional trophic and physiological information \cite{west2006stable}.
+
+Throughout the baleen record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplemental Methods; Supplemental Figures S1 and S2).
 Therefore, given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$, we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over seven full years of the whale's life (early 1884 - spring 1891).
@@
 
-Relating the isotopic compositions of animal tissues to likely locations is complicated by a lack of knowledge of spatio-temporal variations in the isotopic composition of diet \cite{west2006stable,mcmahon2015millennial}.
-We develop a new approach to address this problem using temporally dynamic models of $\delta^{13}$C values in the global ocean \cite{magozzi2017using} (Supplemental Figure S3), coupled to an agent-based model of whale movements (see Supplemental Methods; Supplemental Table S1; Supplemental Figure S5).
-We use our coupled model to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in different parts of the known geographic range of North Atlantic blue whales.
-We then compare simulated profiles (Figures \ref{fig2}, \ref{fig3} and Supplemental Figure S4) to the measured isotopic profile of the blue whale's baleen (Figure \ref{fig1}), using z-score transformations to account for uncertainty in baseline variation, to identify the most likely movement history.
-
-Many baleen whales, such as humpback whales (\textit{Megaptera novaeangliae}), rely largely on stored lipid reserves to meet metabolic demands in lower latitudes in winter \cite{irvine2017quantifying}.
-However, increasing evidence suggests that blue whales engage in opportunistic feeding throughout the year.
-Whale distributions during winter coincide with the onset and development of phytoplankton blooms and areas of high productivity associated with upwelling \cite{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating}, and satellite tagged individuals express fine scale movement behaviour associated with area restricted foraging in winter and summer \cite{lesage2017foraging,bailey2009behavioural}.
-We therefore assume opportunistic ingestion of protein throughout the year.
-
 % figure 2
@@
 % figure 3
-\begin{figure}
- \centering
-  \includegraphics[width = \linewidth]{figures/Figure-3-blue-sims.png}
-  \caption{Correlations among simulated $\delta^{13}$C from the top 10\% best fitting migratory movement models (grey lines, right hand y-axis) and $\delta^{13}$C from baleen (black line, left hand y-axis; see Figure \ref{fig1}).
-  Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location, reflecting temporal integration of phytoplankton $\delta^{13}$C values within the food chain before ingestion by the whale as krill.
-  The end points of the simulations and empirical data have been aligned to coincide.
-}
-  \label{fig3}
-\end{figure}
+%\begin{figure}
+ %\centering
+  %\includegraphics[width = \linewidth]{figures/Figure-3-blue-sims.png}
+  %\caption{Correlations among simulated $\delta^{13}$C from the top 10\% best fitting migratory movement models (grey lines, right hand y-axis) and $\delta^{13}$C from baleen (black line, left hand y-axis; see Figure \ref{fig1}).
+  %Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location, reflecting temporal integration of phytoplankton $\delta^{13}$C values within the food chain before ingestion by the whale as krill.
+  %The end points of the simulations and empirical data have been aligned to coincide.
+%}
+%  \label{fig3}
+%\end{figure}
 
@@
 
-We simulated 1200 individual movement patterns, then excluded simulations where the virtual whale stranded before reaching the 3019 days represented in the baleen.
-We then compared the remaining 1049 simulations and the resulting simulated baleen $\delta^{13}$C records, to the measured records with simple linear regressions. 
 Simulated baleen $\delta^{13}$C profiles produce a good fit to measured profiles, the median $r^2$ value across 1049 simulated profiles was 0.49, and the maximum was 0.71 (Supplemental Figure S6).
@@
 Better fitting models in general were those predicting a greater latitudinal foraging range, and foraging in more northerly waters (Supplemental Figures S7 and S8).
-Best-fit model distributions are largely consistent with our current understanding of blue whale distributions in the northeast Atlantic \cite{reeves2004historical,baines2014upwellings,baines2017autumn,reeves2004historical}, with perhaps greater importance of winter foraging in temperate regions (Figure \ref{fig4}).
+%Best-fit model distributions are largely consistent with our current understanding of blue whale distributions in the northeast Atlantic \cite{reeves2004historical,baines2014upwellings,baines2017autumn,reeves2004historical}, with perhaps greater importance of winter foraging in temperate regions (Figure \ref{fig4}).
 
 % figure 4
-\begin{figure}
- \centering
-  \includegraphics[width = \linewidth]{figures/Figure-4-monthly.png}
-  \caption{Simulated locations by month taken from the top 10\% best fitting migratory movement models for behavioural phase two (summer 1886 to spring 1890) only.}
-  \label{fig4}
-\end{figure}
+%\begin{figure}
+ %\centering
+  %\includegraphics[width = \linewidth]{figures/Figure-4-monthly.png}
+  %\caption{Simulated locations by month taken from the top 10\% best fitting migratory movement models for behavioural phase two (summer 1886 to spring 1890) only.}
+  %\label{fig4}
+%\end{figure}
 
@@
 The inferred reproductive time-line is therefore consistent with the assumed reproductive ecology of blue whales globally \cite{handbook}.
+
+\section{Discussion}
 Long-term, multi-annual data on the movement patterns and reproductive ecology of individual blue whales are scarce. 
@@
 
+\section{Figure and Table captions}
+% 4 total, only 2 can be figures
+
 \section{Acknowledgments}
-This work was funded by the British Ecological Society (grant: 5771/6815) to NC, CT and AJ. 
-We thank C.J.
-Somes for providing  $\delta^{15}$N POM data, Bastian Hambach and Megan Spencer at the University of Southampton SEAPORT isotope laboratory for assistance with stable isotope analyses, and Andrew Yool for allowing us to use and share NEMO-MEDUSA outputs.
+%We thank C.J.
+Somes for $\delta^{15}$N POM data, Bastian Hambach and Megan Spencer at the University of Southampton SEAPORT isotope laboratory for assistance with stable isotope analyses, and Andrew Yool for allowing us to use and share NEMO-MEDUSA outputs.
+We thank C.J. Somes, Bastian Hambach, Megan Spencer and Andrew Yool for assistance with data and stable isotope analyses.
 
 \section{Author Contributions}
-CT, NC and AJ conceived the project.
-CT, NC, AJ, RS, and KC collected baleen from the NHM collections. 
+%CT, NC and AJ conceived the project.
+%CT, NC, AJ, RS, and KC collected baleen from the NHM collections. 
 KC and CT extracted and analysed samples.
@@
 EC, NC, SM and RS helped parameterise the model.
-AJ and NC created the figures. 
-CT and NC wrote the paper.
-All authors edited drafts and approved the final version.
+%AJ and NC created the figures. 
+%CT and NC wrote the paper.
+All authors edited drafts, are accountable for the content, and approved the final version of the manuscript.
 
-\section{Declaration of Interests}
-The authors declare no competing interests.
-
-\bibliographystyle{currentbiol}
-\bibliography{blue-whale-abbreviations}
-
-\section{STAR Methods}
-%Specific headers
-
-\subsection{Contact for reagent and resource sharing}
-Further information and requests for resources and reagents should be directed to and will be fulfilled by the Lead Contact, Natalie Cooper (natalie.cooper@nhm.ac.uk).
-
-\subsection{Method details}
-\subsubsection{Stable isotope extractions from baleen}
-Baleen was collected from the Natural History Museum, London (specimen NHMUK.1892.3.1.1). 
-The baleen plate was cleaned with ethanol to remove surface contaminants such as skin/gum or other lipids that can influence isotopic signals. 
-{\raise.17ex\hbox{$\scriptstyle\sim$}}1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
-97 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal (gingival) section that contains the most recent tissue. 
-We assume that baleen grows at a relatively constant rate, so the samples are equally spaced through time \cite{best1996stable}. 
-Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the University of Southampton SEAPORT Stable Isotope Ratio Mass Spectrometry Laboratory (Southampton, UK), using a Vario Isotope select elemental analyser, coupled to an Isoprime 100 isotope mass spectrometer.
-Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
-C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples \cite{hobson1998stable}.
-All data are available from the NHM Data Portal \cite{data-set}.
-
-\subsubsection{Time calibrating stable isotope profiles}
-Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates \cite{hobson1998stable,busquets2017estimating}.
-Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values at 70 - 18cm distance from the gingival end of the plate (behavioural phase two). 
-We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis \cite{cardona2017temporal} (Supplemental Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated for six recent blue whales from the Californian Pacific \cite{busquets2017estimating}.
-Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891. 
-Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase two (70cm to 18cm on the baleen plate), but no clear relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase one (the oldest 30cm of the baleen plate, i.e.
-100cm - 70cm; Supplemental Figure S2) or the final 500 days of the whale's life (youngest 18cm of the baleen plate; Supplemental Figure S2).
-
-\subsubsection{Baseline isotope comparisons}
-Isotope-enabled biogeochemical ocean models \cite{magozzi2017using,schmittner2016complementary} were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Supplemental Figure S3).
-Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by C.J. Somes (\textit{pers.comm}) based on a 5$^{\circ}$ resolution biogeochemical model (Supplemental Figure S3). 
-$\delta^{13}$C POM values were simulated at 1$^{\circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model \cite{magozzi2017using,yool2013medusa}. 
-Simulated  $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c. 60$^{\circ}$N, and relatively negative in the central and southern North Atlantic. 
-Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
-In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Supplemental Figure S3). 
-The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \cite{magozzi2017using,laws1995dependence}.
-Thus temporal variations in $\delta^{13}$C POM values were superimposed on latitudinal gradients. 
-The scale and nature of temporal variation in $\delta^{13}$C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in $\delta^{13}$C POM values linked to strongly seasonal phytoplankton growth dynamics.
-The model simulations used are forced with decadal climatological data from 2000-2010 \cite{magozzi2017using}. 
-We do not draw inferences about location from absolute $\delta^{13}$C values and thus the reduction in oceanic $\delta^{13}$C POM values caused by release of fossil carbon into the atmosphere (Suess effect) does not influence inferences about location or movement.
-Ocean basin scale spatio-temporal differences in $\delta^{13}$C values, largely reflecting relative differences in sea surface temperatures between trophic, temperate and arctic waters, have remained essentially similar over millennial timescales \cite{schmittner2016complementary} therefore we are confident that the Magozzi \textit{et al} \cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.
-
-We therefore used monthly simulated $\delta^{13}$C POM values to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours.
-Whales feed on krill, not POM. 
-Assimilation of carbon into krill tissues will dampen the temporal variability seen in POM, effectively producing a temporal average over the timescale of isotopic turnover within krill. 
-We estimate turnover to be complete between four and six months and therefore we resampled the $\delta^{13}$C POM values in each one degree cell to reflect a six month moving average. 
-Carbon isotope values are also likely to be fractionated during transfer from plankton to krill as $^{12}$C  is preferentially lost through respiration. 
-The degree of such trophic fractionation is unclear, however and we do not draw interpretations based on absolute $\delta^{13}$C values, rather on the relative $\delta^{13}$C  values across the length of the baleen plate.
-
-\subsection{Quantification and statistical analysis}
-\subsubsection{Agent-based whale movement model}
-We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
-At each time point and location in the simulations, we extracted time averaged $\delta^{13}$C values from the spatio-temporal model \cite{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figures \ref{fig2}, \ref{fig3}, and Supplemental Figure S4).
-Movement was coded as a set of probabilistic rules. 
-The parameters for these were taken from the literature on blue whale behaviour (e.g. \cite{handbook}). 
-All terms were expressed as probability distributions, yielding multiple potential movement tracks.
-
-In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and linear distance of movement are all influenced by the following. 
-(i) Behavioural state (migrating north, migrating south, or foraging).
-Northerly migrations were possible only in spring, and southerly migrations in autumn. 
-Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
-(ii) Sea surface temperature \cite{yool2013medusa} ($^{\circ}$C). 
-When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.
-(iii) Water depth \cite{bathy} (m; extracted via the marmap package in R \cite{marmap}). 
-Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
-(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities \cite{yool2013medusa}). 
-This was included as a proxy for zooplankton food availability. 
-Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state.
-
-We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid-Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and sub-tropical latitudes.
-Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four years of simulations.
-For a full description of the model and its parameters see Supplemental Methods: Supplemental Figure S5 and Supplemental Table S1. 
-We simulated whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.
-For each movement simulation we extracted phytoplankton $\delta^{13}$C values at each time point and location from the models described above \cite{magozzi2017using}. 
-We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).
-
-\subsection{Data and Software Availability}
+\section{Data accessibility}
 Data are available from the NHM Data Portal (https://doi.org/10.5519/0093278). 
@@
 
-%\section{Figure legends}
+\section{Funding}
+This work was funded by the British Ecological Society (grant: 5771/6815).
 
-%\section{Tables}
+\section{Competing interests}
+We have declare no competing interests.
+
+\section{Ethical statement}
+No ethical issues were applicable to this research.
+
+\bibliographystyle{vancouver}
+\bibliography{blue-whale-abbreviations}
+
+\section{mess}
+%1 Runge CA, Martin TG, Possingham HP, Willis SG, Fuller RA.
+%Conserving mobile species. Front Ecol Environ. 2014;12(7):395–402.
+%2 Robinson RA, Crick HQ, Learmonth JA, Maclean I, Thomas CD,
+%Bairlein F, et al. Travelling through a warming world: climate change
+%and migratory species. Endanger Spec Res. 2009;7(2):87–99.
+3 Holdo RM, Roach RR. Inferring animal population distributions from
+individual tracking data: theoretical insights and potential pitfalls. J Anim Ecol. 2013;82(1):175–181.
+4 Ryan C, McHugh B, Trueman CN, Sabin R, Deaville R, Harrod C, et al.
+Stable isotope analysis of baleen reveals resource partitioning among
+sympatric rorquals and population structure in fin whales. Mar Ecol
+Prog Ser. 2013;479:251–261.
+6 Bailey H, Mate BR, Palacios DM, Irvine L, Bograd SJ, Costa DP, et al. Behavioural estimation of blue whale movements in the Northeast
+Pacific from state-space model analysis of satellite tracks. Endanger Spec Res. 2009;10(93-106):1–1.
+7 Reilly S, Bannister J, Best P, Brown M, Brownell Jr R, Butterworth D, et al. Balaenoptera musculus. In: IUCN Red List of Threatened Species.
+Gland, Switzerland: IUCN;; 2008. .
+8 Pike DG, Vikingsson GA, Gunnlaugsson T, Øien N. A note on the distribution and abundance of blue whales (Balaenoptera musculus) in
+the Central and Northeast North Atlantic. NAMMCO Scientific
+Publications. 2009;7:19–29.
+%9 Borger L. Stuck in motion? Reconnecting questions and tools in
+%movement ecology. J Anim Ecol. 2016;85(1):5–10.
+%10 McDonald MA, Mesnick SL, Hildebrand JA. Biogeographic
+%characterization of blue whale song worldwide: using song to identify
+%populations. J Cet Res Man. 2006;8(1):55–65.
+11 Mate B, Mesecar R, Lagerquist B.
+The evolution of satellite-monitoredradio tags for large whales: One laboratory’s experience.
+Deep Sea Research Part II: Topical Studies in Oceanography.
+2007;54(3):224–247.
+12 Silva MA, Prieto R, Jonsen I, Baumgartner MF, Santos RS. North
+Atlantic blue and fin whales suspend their spring migration to forage
+in middle latitudes: building up energy reserves for the journey? PLoS ONE. 2013;8(10):e76507.
+13 Best PB, Mate B, Lagerquist B. Tag retention, wound healing, and
+subsequent reproductive history of southern right whales following
+satellite-tagging. Mar Mamm Sci. 2015;31(2):520–539.
+14 Walker KA, Trites AW, Haulena M, Weary DM. A review of the effects
+of different marking and tagging techniques on marine mammals.
+Wildlife Res. 2012;39(1):15–30.
+15 Lesage V, Gavrilchuk K, Andrews RD, Sears R. Foraging areas,
+migratory movements and winter destinations of blue whales from the
+western North Atlantic. Endanger Spec Res. 2017;34:27–43.
+%16 Heide-Jørgensen MP, Kleivane L, ØIen N, Laidre KL, Jensen MV. A
+%new technique for deploying satellite transmitters on baleen whales:
+%tracking a blue whale (Balaenoptera musculus) in the North Atlantic.
+%Marine Mamm Sci. 2001;17(4):949–954.
+%17 Sears R, Burton CLK, V´ıkingsson GA. Review of blue whale
+%430 (Balaenoptera musculus) photo-identification distribution data in the North Atlantic including the first long-range match between Iceland
+%and Mauritania. Available from:
+%https://www.rorqual.com/downloads/review-of-blue-whalebalaenoptera-
+%musculus-photoidentification-distribution-data-in-the435
+%north-atlantic-including-the-first-long-.pdf;
+%2005.
+18 Baines ME, Reichelt M. Upwellings, canyons and whales: An
+important winter habitat for balaenopterid whales off Mauritania,
+northwest Africa. J Cet Res Man. 2014;14:57–67.
+19 Reeves RR, Smith TD, Josephson EA, Clapham PJ, Woolmer G.
+Historical observations of humpback and blue whales in the North
+Atlantic Ocean: Clues to migratory routes and possibly additional
+feeding grounds. Marine Mamm Sci. 2004;20(4):774–786.
+20 Baines M, Reichelt M, Griffin D. An autumn aggregation of fin
+(Balaenoptera physalus) and blue whales (B. musculus) in the Porcupine
+Seabight, southwest of Ireland. Deep Sea Research Part II: Topical
+Studies in Oceanography. 2017;.
+21 Charif RA, Clark CW. Acoustic monitoring of large whales in deep
+waters north and west of the British Isles: 1996-2005. Cornell
+Laboratory of Ornithology Bioacoustics Research Program Technical
+Reports. 2009;p. 08–07.
+22 Visser F, Hartman KL, Pierce GJ, Valavanis VD, Huisman J. Timing of
+migratory baleen whales at the Azores in relation to the North Atlantic
+spring bloom. Marine Ecol Prog Ser. 2011;440:267–279.
+23 West JB, Bowen GJ, Cerling TE, Ehleringer JR. Stable isotopes as one of
+Nature’s ecological recorders. Trends Ecol Evol. 2006;21(7):408–414.
+24 Busquets-Vass G, Newsome SD, Calambokidis J, Serra-Valente G,
+Jacobsen JK, Aguiniga-Garcıa S, et al. Estimating blue whale skin
+isotopic incorporation rates and baleen growth rates: Implications for assessing diet and movement patterns in mysticetes. PLoS ONE.
+2017;12(5):e0177880.
+25 Hobson KA, Wassenaar LI, editors. Tracking Animal Migration with
+Stable Isotopes. Academic Press;; 2008.
+26 Lister AM, Group CCR, et al. Natural history collections as sources of long-term datasets. Trends Ecol Evol. 2011;26(4):153–154.
+29 Eisenmann P, Fry B, Holyoake C, Coughran D, Nicol S, Nash SB.
+Isotopic evidence of a wide spectrum of feeding strategies in southern
+hemisphere humpback whale baleen records. PLoS ONE. 2016;11(5):e0156698.
+30 Best P, Schell D. Stable isotopes in southern right whale (Eubalaena australis) baleen as indicators of seasonal movements, feeding and growth. Mar Biol. 1996;124(4):483–494.
+31 Hobson KA, Schell DM. Stable carbon and nitrogen isotope patterns in baleen from eastern Arctic bowhead whales (Balaena mysticetus). Can J Fish Aq Sci. 1998;55(12):2601–2607.
+32 Trueman CN, Jackson AL, Chadwick KS, Coombs EJ, Magozzi S, Sabin
+RC, et al. In: Dataset: Baleen stable isotope data. Natural History
+Museum Data Portal (data.nhm.ac.uk). https://doi.org/10.5519/0093278; 2018.
+35 Magozzi S, Yool A, Vander Zanden H, Wunder M, Trueman C. Using
+ocean models to predict spatial and temporal variation in marine
+carbon isotopes. Ecosphere. 2017;8(5):e01763.
+37 Schmittner A, Somes CJ. Complementary constraints from carbon
+(13C) and nitrogen (15N) isotopes on the glacial ocean’s soft-tissue
+biological pump. Paleoceanography. 2016;31(6):669–693.
+38 Yool A, Popova E, Anderson T. MEDUSA-2.0: an intermediate
+complexity biogeochemical model of the marine carbon cycle for
+climate change and ocean acidification studies. Geosci Model Devel.
+2013;6(5):1767–1811.
+39 Wilson DE, Mittermeier RA, editors. Handbook of Mammals of the
+World. 4: Sea Mammals. Barcelona, Spain; Lynx Edicions; 2014.
+40 Sigurjonsson J. On the life history and autecology of North Atlanticrorquals. Devel in Mar Biol. 1995;4:425–441.
 

### 2018-10-12 manuscript/NatureComms/blue-whale-ms-oct18.tex
Fix abstract

--- 2018-02-01 manuscript/revision/blue-whale-ms-v2.tex
@@
 \section{Abstract}
-Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}. 
-Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}, but interpreting these records can be challenging.
-Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct seven years of movements of an iconic female blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
-We provide the longest continuous record of movement for an individual blue whale in the north Atlantic. 
-We show that the whale spent at least two years resident in sub-tropical waters as a juvenile, before commencing annual latitudinal migrations consistent with patterns of movements of blue whales in the northeast Atlantic inferred from observations, tagging and historic whaling data.
-The isotopic record is consistent with observations of opportunistic feeding throughout the year in blue whales, and suggests pregnancy and calf-rearing in the year before death.
-Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climate, ecological conditions and anthropogenic pressures, and reveal aspects of life history.
-Combining biochemical tracers with simulation modeling unlocks the information recorded in these irreplaceable archives.
+The spatial ecology of rare, migratory oceanic mammals such as the blue whale (\textit{Balaenoptera musculus}) is difficult to study directly, leading to distinct knowledge gaps associated with migratory behavior at the individual scale.
+Recent advances in satellite telemetry have improved our understanding of individual-level movement, but tags rarely provide data for more than a few months, and tagging cannot provide retrospective information.
+Incrementally-grown tissues provide alternative, biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales, but inferring detailed spatio-temporal information from biochemical tracers is challenging.
+Here we present a new approach combining stable isotope analyses of incrementally-grown tissues with simulation models to infer individual-level animal movements.  
+We use our new method to reconstruct multiple years of movement histories of individual modern and historic blue whales, including an iconic blue whale that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
+We recover inferences of individual space use and between-individual variation in movements that are consistent with satellite-tag based data for modern blue whales in the northeast Pacific Ocean. 
+For the historic whale, inferred movements over seven years of life of a single individual are also consistent with our limited understanding of the movement ecology of blue whales in the northeast Atlantic Ocean.
+Combining biochemical tracers in incrementally-grown tissues with simulation modeling unlocks the information recorded in historic collections, and provides inferences that are complimentary to and comparable with tag-derived movement data.
 

### 2018-10-15 manuscript/NatureComms/blue-new.tex
Nature Comms version with new pacific analyses

% Preamble
\documentclass[a4paper,12pt]{article}

\usepackage[osf]{mathpazo} % palatino
\usepackage{ms}            % load the template
\usepackage[round]{natbib} % author-year citations
\usepackage[superscript,biblabel]{cite} % for superscript citations
\usepackage{graphicx}
\usepackage{parskip} 
\usepackage{caption}
\usepackage{subcaption}
\usepackage{textcomp} % for parts per mille symbol     
\pagenumbering{arabic}    
\linespread{1.66}

% Title page information
\title{Reconstructing the last known movements of one of Nature's giants}
% 90 characters max

\author{
  Clive N. Trueman$^{1}$, Andrew L. Jackson$^{2}$, Katharyn S. Chadwick$^{1}$,\\ 
  Ellen J. Coombs$^{3,4}$, Sarah Magozzi$^{1,5}$, Richard C. Sabin$^{3}$ \\
  and Natalie Cooper$^{3*}$
}
\date{}
\affiliation{\noindent{\footnotesize
  $^1$ Ocean and Earth Science, University of Southampton Waterfront Campus, Southampton, SO14 3ZH, UK.\\
  $^2$ School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
  $^3$ Department of Life Sciences, Natural History Museum London, Cromwell Road, London, SW7 5BD, UK.\\ 
  $^4$ Department of Earth Sciences, University College London, Gower Street, London, WC1E 6BT, UK.\\
  $^5$ Department of Geology and Geophysics, University of Utah, Salt Lake City, UT 84112-0102, USA.\\
}}

\vfill

%\runninghead{}
%\keywords{}
%}

% End of preamble

\begin{document}
\modulolinenumbers[1]   % Line numbering on every line

\mstitlepage

\parindent = 1.5em
\addtolength{\parskip}{.9em}

\raggedright

\section{Abstract} % Check number of words allowed...

The spatial ecology of rare, migratory oceanic mammals such as the blue
whale (\textit{Balaenoptera musculus}) is difficult to study directly, leading to distinct knowledge gaps associated with migratory behavior at the individual scale. 
Recent advances in satellite telemetry have improved our understanding of individual-level movement, but tags rarely provide data for more than a few months, and tagging cannot provide retrospective information. 
Incrementally-grown tissues provide alternative, biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales, but inferring detailed spatio-temporal information from biochemical tracers is challenging. 
Here we present a new approach combining stable isotope analyses of incrementally-grown tissues with simulation models to infer individual-level animal movements. 
We use our new method to reconstruct multiple years of movement histories of individual modern and historic blue whales, including an iconic blue whale that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.

We recover inferences of individual space use and between-individual variation in movements that are consistent with satellite-tag based data for modern blue whales in the northeast Pacific Ocean. 
For the historic whale, inferred movements over seven years of life of a single individual are also consistent with our limited understanding of the movement ecology of blue whales in the northeast Atlantic Ocean.
Combining biochemical tracers in incrementally-grown tissues with simulation modeling unlocks the information recorded in historic collections, and provides inferences that are complimentary to and comparable with tag-derived movement data.

\section{Main text}\label{background}

Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}. 
Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle \cite{robinson2009travelling}. 
Identifying threats to migratory species, understanding species responses to change and developing effective conservation measures all require information on the movements of individual animals over multiple years, ideally for both historic and present-day populations. 
With the development of electronic tagging technology, studies of the distributions of animals have largely shifted from infrequent observations of many individuals with limited or no information about individual movement behavior, to frequent observations of a few individuals, with detailed information about individual movement\cite{holdo2013inferring}. 
Despite the tremendous advances made using direct telemetry devices, tags are expensive, with a relatively high failure rate \cite{bailey2009behavioural,best2015tag,mate2007evolution}. 
Data on individual-level, multi-annual movements remain scarce, especially for rare, wide-ranging and long-lived marine species such as baleen whales (Mysticeti)\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.

An alternative technique for investigating animal movements retrospectively is to use intrinsic biochemical information such as stable isotope compositions \cite{west2006stable,busquets2017estimating,hobson2008tracking}. 
Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioral information that are often under-utilised\cite{lister2011natural}. 
The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
A large literature describes the use of stable isotope markers to link the composition of a tissue to a geographic source at a single point in time\cite{hobson2008tracking}. 
Relating the isotopic compositions of marine animal tissues to the likely location of tissue growth is complicated by a lack of knowledge of spatial variations in the isotopic composition of diet (the isotopic baseline)\cite{west2006stable,mcmahon2015millennial}. 
This uncertainty is compounded when multiple samples are taken through time in the same individual, as temporal variation in the isotopic baseline must also be considered explicitly. 
Consequently, relatively little attention has been given to the potential to infer individual movement histories by reconstructing time series of multiple discrete origins from sequentially sampled incrementally-grown tissues.

Here we introduce a new approach to infer location from the isotopic compositions of sequentially-sampled incrementally-grown tissues by drawing on newly developed models predicting spatio-temporal variation in phytoplankton stable carbon isotope composition at global and monthly resolution\cite{magozzi2017using} (Fig. S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig. S5).
We use our coupled models to simulate time series of carbon stable isotope compositions expected under different movement behaviours. 
We then compare simulated isotopic profiles with the measured profiles to identify potential movement histories most consistent with the measured isotopic profile.

We apply our method to blue whales (\textit{Balaenoptera musculus}), the
largest animal to have ever lived. 
Mysticete whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater. 
Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable,hobson1998stable}. 
Baleen plates therefore offer a continuous isotopic record of behavior typically reflecting multiple years of life of an individual whale. 
Baleen is worn away at the tips over time, so a baleen plate reflects the most recent years of life, and rarely records an individual's entire lifespan. 
Among the mysticete whales, blue whales are a particularly attractive target for model-based isotope movement work as they have a consistent, low trophic level diet and feed continuously through the year, likely driven by high energetic costs of maintaining extreme muscle mass %(Goldbogen et al 2015).

Similar to other large balaenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes %(Hucke-Gaete et al 2018). 
As blue whales feed throughout the year to accommodate energetic costs, migration routes and feeding areas for blue whales are thought to be shaped by the year-round location of highly productive regions %(Branch et al 2007). 
Recent studies tracking movements of individual blue whales primarily in the northeast Pacific have demonstrated high levels of among-individual variation in movement histories including suspending migration for opportunistic feeding and skipped migration (i.e. year round residency in either summer or winter
areas) \cite{busquets2017estimating}. 
The northeast Pacific population of blue whales is perhaps the best studied in terms of individual-level movements with a recent history of deployment of over 177 satellite tags \cite{irvine2017quantifying, mate2007evolution}.
In the northeast Pacific, satellite tags show individuals occupying most of the exclusive economic zone (EEZ) and into the Gulf of Alaska \cite{irvine2017quantifying} with greatest overlapping areas in the southern half of the EEZ, along the continental slope and shelf region between the California Channel Islands and Cape Mendocino. 
Winter feeding is less well characterized, but winter feeding grounds are known
in the Gulf of California and in the eastern tropical Pacific, specifically the Costa Rica Dome where production is maintained by seasonal upwelling (REFS). % Refs???
However, most satellite deployments on blue whales report movement data for periods of time under six months \cite{heide2001new, silva2013north, bailey2009behavioural, lesage2017foraging, irvine2017quantifying}, and are unable to show individual movements between summer and winter feeding or breeding areas. 
The longest period of continuous monitoring via satellite tagging of an individual blue whale that we are aware of in the northeast Pacific is 504 days \cite{irvine2017quantifying} and in the North Atlantic is 177 days \cite{lesage2017foraging}.

The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now around 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland \cite{pike2009note}. 
The small population size and oceanic habit makes blue whale movements extremely hard to study. 
A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic also
track regions of high production throughout the year, at least some
whales wintering in the upwelling systems between Mauritania and the
Cape Verde Islands \cite{baines2014upwellings}. 
Northward migration in the spring may occur along mid Atlantic corridors with peak sightings in the Azores around April-May \cite{silva2013north}. 
Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas \cite{pike2009note}. 
Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals \cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
To our knowledge one photographic identification has matched summer and winter locations of a single north-east Atlantic blue whale with sightings in Iceland and Mauritania \cite{poster}.

Inferences concerning movements of blue whales in general, and in the northeast Atlantic in particular, are therefore largely drawn from disparate information sources and rarely detail movements of individual animals over timescales of a year or more. 
It is particularly difficult to establish the degree of connectivity between summer and winter feeding areas, and the level of temporal consistency in movement behaviour within individuals. 
Information on historical movement patterns of individual blue whales is important for understanding the drivers of whale declines, and potential for human-whale conflicts as populations recover.

Here we infer individual level high-resolution movement records for seven blue whales for which baleen stable isotope data were available.
Isotope data for six whales stranding on the west coast of the USA or Mexico were reported in Busquets-Vass \textit{et al}.\cite{busquets2017estimating}, and we report new isotopic data from a whale stranding off the coast of Wexford Ireland in March 1891, and currently on display in the Natural History Museum, London (NHM).
The NHM whale ``Hope'' is a female estimated to be a young adult at least 15 years at death. 
Baleen from the NHM whale therefore yields information about migratory behaviors in North Atlantic blue whales around the peak of industrial whaling.

\section{Results and Discussion}

\subsection{Matching model to Pacific data.}

Estimates of baleen growth rates for the six modern whales from the northeast Pacific were inferred from cyclical fluctuations in nitrogen isotope ratios and reported in Busquets-Vass \textit{et al}\cite{busquets2017estimating}. 
We used these growth rates to reconstruct a potential timeline for baleen isotope records.

% Pacific whales figure
\begin{figure}
 \centering
  \includegraphics[width = \linewidth]{figures/figure-Pacific-draft.eps}
  \caption{Correlations among simulated $\delta^{13}$C from the best fitting migratory movement models (grey lines, right hand y-axis) and $\delta^{13}$C from baleen (black line, left hand y-axis) for six Pacific blue whales taken from \cite{busquets2017estimating}. 
  Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location, reflecting temporal integration of phytoplankton $\delta^{13}$C values within the food chain before ingestion by the whale as krill. 
  The end points of the simulations and empirical data have been aligned to coincide.
}
  \label{pacific}
\end{figure}

In four of the six cases, measured baleen carbon isotope profiles could be replicated from model simulations with relatively high precision (Fig. \ref{pacific}
; best fit r\textsuperscript{2} values ranging from 0.43 to 0.74).
Whales D and F showed similar carbon isotope records and were best fit by the same movement simulation. 
We extracted monthly locations associated with the best fitting simulation models for each whale (Fig. \ref{pacific}).
Best fitting simulation models for whales D and F imply a relatively southerly distribution with few locations north of Baja, California and relatively limited seasonal latitudinal migration. 
The measured carbon isotope record for whale A was best fitted by simulations with a southerly distribution, but with an additional strong seasonal latitudinal migration, and with more northerly locations between April and October. 
Whale C was best fit by models with relatively northerly locations and strong seasonal latitudinal migration. 
We were unable to find convincing fits to baleen profiles for whales B and E suggesting that longer simulations with a greater variation in movements were needed. 
The inferred movement records are consistent with the known spatial distributions and relatively high level of individual variation in spatial distributions among blue whales in the Pacific northwest \cite{irvine2017quantifying, mate2007evolution}.

\subsection{Atlantic whale baleen measured isotopic record}

For Hope, the historic North Atlantic whale, \(\delta^{15}\)N values record regular cyclical fluctuations throughout the length of the baleen plate, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods). 
Therefore, given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm\(y^{- 1}\), we reconstructed a timeline for \(\delta^{13}\)C and \(\delta^{15}\)N fluctuations in the baleen over seven full years of the whale's life (early 1884 - spring 1891). 
The \(\delta^{13}\)C and \(\delta^{15}\)N profiles show two distinct phases in the whale's behaviour. 
In behavioural phase one (from the start of the record to spring 1886), we find relatively stable, elevated \(\delta^{13}\)C values, and relatively low \(\delta^{15}\)N values (Fig. \ref{fig1}; Figs. S1 and S2). 
In behavioural phase two (summer 1886 to spring 1890) \(\delta^{15}\)N values are relatively elevated and \(\delta^{13}\)C values are relatively depleted with coincident cyclical fluctuations in both \(\delta^{13}\)C and \(\delta^{15}\)N values. 
In the last year of life the cyclical pattern is disrupted, with constant low \(\delta^{13}\)C values for approximately six months in the first half of 1890, before a rapid switch to relatively positive \(\delta^{13}\)C values in the second half of 1890. 
The final three months of the record show a progressive fall in \(\delta^{13}\)C values (Fig. \ref{fig1}; Figs. S1 and S2).

% figure 1
\begin{figure}
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
  \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis). Samples were taken longitudinally through the baleen plate (n = 97 samples from a single baleen plate for both isotopes). There is strong annual periodicity and cross-correlation (Supplemental Figures S1 and S2) in both isotopes. The approximate relationship to years assuming a growth rate of 13.5cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
  \label{fig1}
\end{figure}

% figure 2
\begin{figure}
 \centering
 \includegraphics[width = \linewidth]{figures/Figure-2-points.png}
  \caption{Simulated locations of the whale taken from the top 10\% best fitting migratory movement models. 
  Colours reflect the behavioural phase. 
  Phase one is early 1884 to spring 1886, phase two is summer 1886 to spring 1890, and phase three is spring 1890 to spring 1891.}
  \label{fig2}
\end{figure}

The relatively positive and seasonally-invariant \(\delta^{13}\)C values seen during behavioural phase one are only found in subtropical areas of the North Atlantic. 
Our simulations identify a range of possible locations for the whale (Fig. \ref{fig3}), although areas around the Mauritanian coast and Cape Verde Islands, a known current and historic winter feeding area for blue whales \cite{baines2014upwellings,reeves2004historical}, and potentially to the west of the Azores (Figure \ref{fig2}), most closely match the measured profile. 
The whale remained in these warm waters for at least one full year. 
Hope was estimated to be at least 15 years old when she died (based on vertebral epiphyseal fusion; R.C. Sabin \textit{pers.comm}), so was probably older than 10 years old, and therefore sexually mature, during this period. 
During behavioral phase two, the observed low \(\delta^{13}\)C values imply foraging in colder, more northerly latitudes. 
The pronounced cyclical variations in \(\delta^{13}\)C values observed during behavioral phase two could reflect either the isotopic expression of the spring phytoplankton bloom in northern waters\cite{magozzi2017using}, latitudinal migrations, or a combination of both.

% figure 3
\begin{figure}
 \centering
  \includegraphics[width = \linewidth]{figures/Figure-3-blue-sims.png}
  \caption{Correlations among simulated $\delta^{13}$C from the top 10\% best fitting migratory movement models (grey lines, right hand y-axis) and $\delta^{13}$C from baleen (black line, left hand y-axis; see Figure \ref{fig1}). 
  Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location, reflecting temporal integration of phytoplankton $\delta^{13}$C values within the food chain before ingestion by the whale as krill. 
  The end points of the simulations and empirical data have been aligned to coincide.
}
  \label{fig3}
\end{figure}

%We simulated 1200 individual movement patterns, then excluded simulations where the virtual whale stranded before reaching the 3019 days represented in the baleen.
%We then compared the remaining 1049 simulations and the resulting simulated baleen $\delta^{13}$C records, to the measured records with simple linear regressions. 
% CHECK THIS!!!!
We simulated 1000 individual movement patterns and compared the resulting simulated baleen \(\delta^{13}\)C records to the measured records with simple linear regressions. 
%the median $r^2$ value across 1049 simulated profiles was 0.49, and the maximum was 0.71 (Supplemental Fig. S6).
Simulated baleen \(\delta^{13}\)C profiles produce a good fit to measured profiles, the median \(r^{2}\) value across 1000 simulated profiles was 0.49, and the maximum was 0.74 (Fig. S6, distribution of \(r^{2}\) values). 
The top 10\% best fitting simulated profiles are shown in Fig. \ref{fig2}. 
Best fitting models predict residency in the Cape Verde region in behavioral phase one. 
Behavioral phase two is best simulated by seasonal migrations between summer foraging in northern areas (e.g. Norwegian Sea/Barents Sea/Iceland region), and winter foraging in a broad region between the UK and more southerly, subtropical waters. 
Better-fitting models in general were those predicting a greater latitudinal foraging range, and foraging in more northerly waters (Fig. S7 and S8). 
Best-fit model distributions are largely consistent with current understanding of blue whale distributions in the northeast Atlantic \cite{reeves2004historical,baines2014upwellings,baines2017autumn,reeves2004historical} with perhaps greater importance of winter foraging in temperate regions (Fig. \ref{fig4}).

% figure 4
\begin{figure}
 \centering
  \includegraphics[width = \linewidth]{figures/Figure-4-monthly.png}
  \caption{Simulated locations by month taken from the top 10\% best fitting migratory movement models for behavioural phase two (summer 1886 to spring 1890) only.}
  \label{fig4}
\end{figure}

The last 500 days of Hope's life are difficult to simulate.
Beginning in the winter of 1889/1890, the observed \(\delta^{13}\)C values are relatively low, and remain constant for c.4-6 months, before increasing rapidly in the second half of 1890. 
Low \(\delta^{13}\)C values are found in northern waters, but these areas also show large temporal fluctuations in \(\delta^{13}\)C\textsubscript{plk} values (Fig. S4), and the observed values cannot be simulated purely from movements within the known geographic range. 
We suggest that the \(\delta^{13}\)C record in the last period of life may be associated with pregnancy and calf rearing. 
Constant, low \(\delta^{13}\)C values may reflect later metabolic release of carbon reserves assimilated from northern latitudes (i.e. a period of time with limited opportunistic foraging). 
After c.4-6 months the whale began ingesting food with relatively high \(\delta^{13}\)C values, indicating feeding in low latitude waters, shortly followed by a final northward migration. Blue whales have a 10-12 month gestation period, with calving occurring in subtropical waters, and calves are weaned after 6-7 months \cite{handbook}.
We therefore infer a movement chronology reflecting three years of uninterrupted annual latitudinal migrations leading to pregnancy in the
year 1889 and birth in the winter of 1889/1890. 
Following birth, we infer a period of c.4-6 months of residency in sub-tropical waters where the whale was sustained largely from stored lipid reserves, which we interpret as reflecting weaning. 
Subsequently we propose that the whale had a short period of feeding in sub-tropical waters in the second half of 1890 potentially during a final northward migration and eventual stranding during the return to northern feeding grounds in early 1891.

Whaling was an intense pressure for blue whales during the period we are
analysing. 
Before whaling in the North Atlantic began in 1868 \cite{reilly2008balaenoptera}, there were and estimated 10,000-15,000 blue whales in the region \cite{sigurjonsson1995life}. 
In the early 20th century, fisheries moved outside the area because stocks were so depleted \cite{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed \cite{sigurjonsson1995life}. 
The NHM blue whale was thus an individual from a species at the brink of local extinction, potentially representing more than 0.1\% of the entire population of blue whales in the northeast Atlantic.
Inferring spatio-temporal distributions of populations from high density observations of a few individuals (or, in this case one individual) will always be problematic, but the predictive power linking a tracked individual to the population distribution increases when the tracked animal has a strong preference for specific habitats and there is a patchy (clumped) distribution of those habitats \cite{holdo2013inferring}. 
In the case of blue whales, the primary driver influencing seasonal movements is assumed to be the seasonally variable distribution of food resources between high and low latitudes. 
As blue whales have a relatively restricted diet that shows highly predictable spatio-temporal differences in abundance at least on ocean basin scales, we argue that the seasonal latitudinal migrations inferred during behavioural phase two are likely to be common movement traits within North Atlantic blue whales. 
This inference of course requires testing with additional isotopic records from individual whales, observational sightings data and multi-year satellite tracking records.

Our movement models suggest that the upwelling regions around Mauritania and the Cape Verde Islands may have been important historic calving areas, consistent with modern observations of juvenile blue whales in the northeast Atlantic \cite{handbook}.
These areas are now fishing, tourist and shipping hotspots with emerging offshore oil and gas exploration, potentially increasing threats to the recovery of the small population of blue whales in the northeast Atlantic. 
Although hunting for blue whales has been banned for over 50 years, the species is still listed as Endangered on the IUCN Red List, and expected recoveries following the cessation of active hunting may be impeded by other human activities such as modifications to the food web or more direct impacts of noise and shipping.
Developing an understanding of the nature and plasticity of individual level movements in populations of blue (and other mysticete) whales would aid management decision making in the conservation of this endangered species \cite{irvine2017quantifying}.

Long-term, multi-annual data on the movement patterns and reproductive ecology of individual blue whales are scarce. 
Although based on a few individuals, our results confirm that sequential sampling of stable isotope compositions in whale baleen, combined with simulation modeling can yield plausible inferences of individual whale movements that are consistent with population-level satellite tag data and/or inferences drawn from seasonal differences in distributions inferred from observations of multiple individuals over time and space. 
Our movement simulation modeling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviors over multiple years. 
By unlocking information contained in incrementally-grown tissues we hope that a more detailed picture of individual movement behaviour in modern and historic mysticete whales (and other animals) can be developed.

\section{Methods}\label{methods}

\subsection{Stable isotope extractions from baleen}\label{stable-isotope-extractions-from-baleen}

Baleen was collected from the Natural History Museum, London (specimen NHMUK.1892.3.1.1). 
The baleen plate was cleaned with ethanol to remove surface contaminants such as skin/gum or other lipids that can influence isotopic signals. 
{\raise.17ex\hbox{$\scriptstyle\sim$}}1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
97 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal (gingival) section that contains the most recent tissue. 
Baleen grows at a constant rate, so the samples are equally spaced through time\cite{best1996stable}. 
Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the University of Southampton SEAPORT Stable Isotope Ratio Mass Spectrometry Laboratory (Southampton, UK), using a Vario Isotope select elemental analyser, coupled to an Isoprime 100 isotope mass spectrometer. 
Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples\cite{hobson1998stable}. 
All data are available from the NHM Data Portal\cite{data-set}.

\subsection{Time calibrating stable isotope
profiles}\label{time-calibrating-stable-isotope-profiles}

Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates\cite{hobson1998stable,busquets2017estimating}. 
Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
We calculated isotopic periodicity within the NHM baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Supplementary Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated by Busquets-Vass \textit{et al.}\cite{busquets2017estimating}.  
Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891. 
%Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase two (70cm to 18cm on the baleen plate), but no clear relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase one (the oldest 30cm of the baleen plate, i.e. 100cm - 70cm; Supplementary Figure S2) or the final 500 days of the whale's life (youngest 18cm of the baleen plate; Supplementary Figure S2).

\subsection{Baseline isotope
comparisons}\label{baseline-isotope-comparisons}

Isotope-enabled biogeochemical ocean models \cite{magozzi2017using,schmittner2016complementary} were used to characterize
the isotopic composition of phytoplankton expected in different
potential foraging grounds (Fig. S3). 
Annual average \(\delta^{15}\)N POM (particulate organic matter) values were provided by C.J. Somes (\textit{pers.comm}) based on a 5\({}^{\circ}\) resolution biogeochemical model (Fig. S3). \(\delta^{13}\)C POM values were simulated at 1\({}^{\circ}\) and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model \cite{magozzi2017using,yool2013medusa}. 
Simulated \(\delta^{15}\)N POM values are relatively positive in the northeast Atlantic north of c. 60\({}^{\circ}\)N, and relatively negative in the central and southern North Atlantic. 
Annual average \(\delta^{13}\)C POM values largely vary with latitude, with more negative values in more northerly regions. 
In the central North Atlantic, \(\delta^{13}\)C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S3). 
The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \cite{magozzi2017using,laws1995dependence}. 
Thus temporal variations in \(\delta^{13}\)C POM values are superimposed on latitudinal gradients. 
The scale and nature of temporal variation in \(\delta^{13}\)C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in \(\delta^{13}\)C POM values linked to strongly seasonal phytoplankton growth dynamics.
The model simulations used are forced with decadal climatological data from 2000-2010\cite{magozzi2017using}. 
We do not draw inferences about location from absolute \(\delta^{13}\)C values and thus the reduction in oceanic \(\delta^{13}\)C POM values caused by release of fossil carbon into the atmosphere (Suess effect) does not influence inferences about location or movement. 
Ocean basin scale spatio-temporal differences in \(\delta^{13}\)C values largely reflect relative differences in sea surface temperatures between trophic, temperate and arctic waters, therefore we are confident that the Magozzi \textit{et al}.\cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.

We used \(\delta^{13}\)C POM values modeled at monthly resolution to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours. 
The stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on in the weeks prior to keratin growth. 
Assimilation of carbon into krill tissues will dampen the temporal variability seen in POM, effectively producing a temporal average over the timescale of isotopic turnover within krill. 
We estimate turnover to be complete between two and four months and therefore we resampled the \(\delta^{13}\)C POM values in each one degree cell to reflect an average of isotopic compositions in phytoplankton in the two months prior to the sampling date. 
Average values were weighted according to the proportional plankton biomass estimated for each month. 
Carbon isotope values are also likely to be fractionated during transfer from plankton to krill as \textsuperscript{12}C is preferentially lost through respiration. 
The degree of such trophic fractionation is unclear, however and we do not draw interpretations based on absolute \(\delta^{13}\)C values, rather on the relative \(\delta^{13}\)C values across the length of the baleen plate. 
We assume opportunistic ingestion of protein throughout the year \cite{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating, lesage2017foraging, bailey2009behavioural}.
% Branch et al 2007; Hucke-Gaete et al 2017

\subsection{Agent-based whale movement
model}\label{agent-based-whale-movement-model}

We simulated whale movements with the likelihood, direction, and extent of movement influenced by behavioral state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
Movement was coded as a set of probabilistic rules, informed by the literature on blue whale behavior (e.g. \cite{handbook}).
All terms were expressed as probability distributions, yielding multiple potential movement tracks.

In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and linear distance of movement are all influenced by the following. 
(i) Behavioural state (migrating north, migrating south, or foraging).
Northerly migrations were possible only in spring, and southerly migrations in autumn. 
Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
(ii) Sea surface temperature \cite{yool2013medusa} ($^{\circ}$C). 
When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.  
(iii) Water depth (m; from the General Bathymetric Chart of the Oceans (GEBCO) global bathymetry dataset\cite{bathy}, extracted via the marmap package in R \cite{marmap}). 
Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities \cite{yool2013medusa}). 
This was included as a proxy for zooplankton food availability. 
Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state.

For Pacific whales where satellite data offers a rich view of individual movements, we coded our movement models to reflect either residency in southern waters (between Baja California and the Costa Rica Dome) or seasonal latitudinal movements within the California current system. 
We simulated 250 movement tracks of four years duration. 
We allowed start and end dates to vary by aligning measured profiles to sections of simulated profiles of identical lengths with sequentially changing start dates and saved best fitting models.

For the historic North Atlantic whale, we simulated the isotopicexpression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four
years of simulations. 
For a full description of the model and its parameters see Supplementary Methods: Fig. S5 and Table S1. 
We simulated seven years of whale movements 1000 times and two years of residency 30 times for each residency hotspot (Fig. S4).
%We simulated whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.
% CHECK THISSSSSSS!!!!

At each time point and location in the simulations, we extracted time averaged \(\delta^{13}\)C values from the spatio-temporal model\cite{magozzi2017using}resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs X, \ref{fig2}, and S4). % CHECK FIGS
We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).

%\subsection{Data Code and Materials}\label{data-code-and-materials}
\subsection{Data Availability}
Data are available from the NHM Data Portal (https://doi.org/10.5519/0093278). 
R code is available from GitHub (https://github.com/nhcooper123/blue-whale-bes)(Zenodo DOI will be added before publication).

% References

%\bibliographystyle{nature}
\bibliographystyle{authordate1}
\bibliography{blue-whale-abbreviations}

\section{Acknowledgments}\label{acknowledgments}
This work was funded by the British Ecological Society (grant: 5771/6815). 
We thank C.J. Somes for providing  $\delta^{15}$N POM data, Bastian Hambach and Megan Spencer at the University of Southampton SEAPORT isotope laboratory for assistance with stable isotope analyses, and Andrew Yool for allowing us to use and share NEMO-MEDUSA outputs.

\section{Author contributions}\label{author-contributions}
CT, NC, AJ, RS, and KC collected baleen from the NHM collections. 
KC and CT extracted and analysed samples.
CT created the movement model.
EC, NC, SM and RS helped parameterise the model.
AJ and NC created the figures. 
CT and NC wrote the paper.
All authors edited drafts and approved the final version.

\section{Competing interests}
The authors declare no competing interests.

\end{document}

### 2018-10-15 manuscript/NatureComms/blue-whale-ms-oct18.tex
Nature Comms version with new pacific analyses

@@
 Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle\cite{robinson2009travelling}.
-Identifying threats to migratory species and developing effective conservation measures requires information on individual-level movements spanning multiple years, for both historical and present-day populations.
-Such data are scarce, especially for rare, wide-ranging and long-lived marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
+Identifying threats to migratory species, understanding species responses to change and developing effective conservation measures all require information on the movements of individual animals over multiple years, ideally for both historic and present-day populations.
+With the development of electronic tagging technology, studies of the distributions of animals have largely shifted from infrequent observations of many individuals with limited or no information about individual movement behavior, to frequent observations of a few individuals, with detailed information about individual movement (Holdo et al., 2013).
+Despite the tremendous advances made using direct telemetry devices, tags are expensive, with a relatively high failure rate. 
+Data on individual level, multi-annual movements remain scarce, especially for rare, wide-ranging and long-lived marine species such as baleen whales (Mysticeti)\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
+
+An alternative technique for investigating animal movements retrospectively is to use intrinsic biochemical information such as stable isotope compositions. 
+Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioral information that are often under-utilised.
+The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
+A large literature describes the use of stable isotope markers to link the composition of a tissue to a geographic source at a single point in time (Hobson and Wassenaar 2008). 
+Relating the isotopic compositions of marine animal tissues to the likely location of tissue growth is complicated by a lack of knowledge of spatial variations in the isotopic composition of diet (the isotopic baseline).
+This uncertainty is compounded when multiple samples are taken through time in the same individual, as temporal variation in the isotopic baseline must also be considered explicitly. 
+Consequently, relatively little attention has been given to the potential to infer individual movement histories by reconstructing time series of multiple discrete origins from sequentially sampled incrementally-grown tissues.
+
+Here we introduce a new approach to infer location from the isotopic compositions of sequentially-sampled incrementally-grown tissues by drawing on newly developed models predicting spatio-temporal variation in phytoplankton stable carbon isotope composition at global and monthly resolution (Magozzi et al., 2017; Fig.
+S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig.
+S5).
+We use our coupled models to simulate time series of carbon stable isotope compositions expected under different movement behaviours.
+We then compare simulated and measured with the measured isotopic profiles to identify potential movement histories most consistent with the measured isotopic profile.
+
+We apply our method to blue whales (\textit{Balaenoptera musculus}), the largest animal to have ever lived. 
+Mysticete whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater. 
+Baleen is ideal for stable isotope studies because keratin grows continuously through an individual’s life, and once laid down it is metabolically inert (Schell, Schell Hobson ref). 
+Baleen plates therefore offer a continuous isotopic record of behavior typically reflecting multiple years of life of an individual whale. 
+B, although baleen is worn away at the tips over time, so a baleen plate reflects the most recent years of life, and rarely records an individual’s entire lifespan. 
+Among the mysticete whales, blue whales are a particularly attractive target for model-based isotope movement work as they have a consistent, low trophic level diet and feed continuously through the year, perhaps driven by high energetic costs of maintaining extreme muscle mass (Goldbogen et al 2015).
+Similar to other large balenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes
+Mysticete whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater.
+Baleen is ideal for stable isotope studies because keratin grows continuously through an individual’s life, and once laid down it is metabolically inert (Schell, Schell Hobson ref).
+Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although baleen is worn away at the tips over time, so rarely records an individual’s entire lifespan.
+Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales (e.g.
+Schell, Schell and Hobson), but have yet to be applied to blue whales from the North Atlantic population.
+we are not aware of any examples where baleen stable isotope data have been used to directly inform spatial management.
+Relating marine animal tissue isotopic compositions to likely locations is complicated by a lack of knowledge of spatio-temporal variations in the isotopic composition of diet.
+Here we introduce a new approach to infer location from tissue isotope data drawing on newly developed models predicting spatio-temporal variation in phytoplankton stable carbon isotope composition at global and monthly resolution (Fig.
+S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig.
+S5).
+We use our coupled model to simulate time series of carbon stable isotope compositions expected under different movement behaviours.
+We are then able to compare simulated and measured with the measured isotopic profiles to identify the most likely movement history.
+
+We apply this method to blue whales (Balaenoptera musculus), the largest animal to have ever lived.
+Blue whales are an attractive target for isotope-based movement work as they have a consistent, low trophic level diet and feed continuously through the year.
+Isotope data can be further complicated by seasonal or individual variations in diet compositions.
+Here we apply this method to baleen from the Although hunting for blue whale has been banned for over 50 years, blue whales are still listed as Endangered in the IUCN Red List, and expected recoveries following the cessation of active hunting may be impeded by other human activities such as modifications to the food web or more direct impacts of noise and shipping.
+Developing an understanding of the nature and plasticity of individual level movements in populations of blue (and other baleen) whales would aid management decision making in the conservation of this endangered species (Irvine et al 2017).
+ (Hucke-Gaete et al 2018).
+As blue whales feed throughout the year to accommodate energetic costs, migration routes and feeding areas for blue whales are thought to be shaped by the year-round location of highly productive regions (Branch et al 2007).
+Recent studies tracking movements of individual blue whales primarily in the northeast Pacific have demonstrated high levels of among-individual variation in movement histories including suspending migration for opportunistic feeding and skipped migration (i.e.
+year round residency in either summer or wint
+er areas).
+most satellite deployments on blue whales report movement data for periods of time under 6 months (Heide-Jørgensen et al., 2001; Silva et al 2013; Bailey et al., 2009; Lesage et al 2017; Irvine et al 2017), and are unable to show individual movements between summer and winter feeding or breeding areas.
+T.
+
+The NorthEast Pacific population of blue whales is perhaps the best studied in terms of individual level movements with a recent history of deployment of over 177 satellite tags (Mate et al 2007; Irvine et al 2017).
+ Similar to other large balenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes.
+Blue whales, however are thought to feed throughout the year, perhaps driven by high energetic costs of maintaining extreme muscle mass (Goldbogen et al 2015).
+Migration routes and feeding areas are therefore thought to be shaped by the year-round location of highly productive regions (Branch et al 2007).
+The northeast Pacific population of blue whales is perhaps the best studied in terms of individual level movements with a recent history of deployment of over 177 satellite tags (Mate et al 2007; Irvine et al 2017).
+In the nNortheast Pacific, satellite tags show individuals occupying most of the exclusive economic zone (EEZ) and into the Gulf of Alaska (Irvine et al 2017 +) with greatest overlapping areas in the southern half of the EEZ, along the continental slope and shelf region between the California Channel Islands and Cape Mendocino.
+Whales were present in the EEZ throughout the year, but were mainly present between July and November.
+Whales tended to occupy more northerly portions of the range later in the year before departing the EEZ presumably for winter feeding grounds in late autumn.
+Winter feeding is less well characterized, but winter feeding grounds are known in the Gulf of California and in the eastern tropical Pacific, specifically the Costa Rica Dome  where production is maintained by seasonal upwelling (REFS).
+However, most satellite deployments on blue whales report movement data for periods of time under 6 months (Heide-Jørgensen et al., 2001; Silva et al 2013; Bailey et al., 2009; Lesage et al 2017; Irvine et al 2017), and are unable to show individual movements between summer and winter feeding or breeding areas.
+The longest period of continuous monitoring via satellite tagging of an individual blue whale that we are aware of in the northeast Pacific is 504 days (Irvine et al 2017) and in the north Atlantic is 177 days (Lesage et al 2017).
+
+The blue whale (Balaenoptera musculus), the largest animal to have ever lived, is one such migratory marine species.
+Although hunting for blue whale has been banned for over 50 years, blue whales are still listed as Endangered in the IUCN Red List, and expected recoveries following the cessation of active hunting may be impeded by other human activities such as modifications to the food web or more direct impacts of noise and shipping.
+Developing an understanding of the nature and plasticity of individual level movements in populations of blue (and other baleen) whales would aid management decision making in the conservation of this endangered species (Irvine et al 2017).
+The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now around 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland (Pike, Víkingsson, Gunnlaugsson, & Øien, 2009).
+The small population size and oceanic habit makes blue whale movements extremely hard to study.
+
+The NorthEast Pacific population of blue whales is perhaps the best studied in terms of individual level movements with a recent history of deployment of over 177 satellite tags (Mate et al 2007; Irvine et al 2017).
+Similar to other large balenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes.
+Blue whales, however are thought to feed throughout the year, perhaps driven by high energetic costs of maintaining extreme muscle mass (Goldbogen et al 2015).
+Migration routes and feeding areas are therefore thought to be shaped by the year-round location of highly productive regions (Branch et al 2007).
+In the Northeast Pacific satellite tags show individuals occupying most of the exclusive economic zone (EEZ) and into the Gulf of Alaska (Irvine et al 2017 +) with greatest overlapping areas in the southern half of the EEZ, along the continental slope and shelf region between the California Channel Islands and Cape Mendocino.
+Whales were present in the EEZ throughout the year, but were mainly present between July and November.
+Whales tended to occupy more northerly portions of the range later in the year before departing the EEZ presumably for winter feeding grounds in late autumn.
+Winter feeding is less well characterized, but winter feeding grounds are known in the Gulf of California and in the eastern tropical Pacific, specifically the Costa Rica Dome  where production is maintained by seasonal upwelling (REFS).
+
+A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic also track regions of high production throughout the year, at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands (M.E Baines & Reichelt, 2014).
+Northward migration in the spring may occur along mid Atlantic corridors with peak sightings in the Azores around April-May (Silva et al 2014).
+Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas (Pike, Víkingsson, Gunnlaugsson, & Øien, 2009).
+Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals (M.
+E Baines, Reichelt, & Griffin, 2017; Charif & Clark, 2009; Reeves, Smith, Josephson, Clapham, & Woolmer, 2004; Visser, Hartman, Pierce, Valavanis, & Huisman, 2011).
+To our knowledge one photographic identification has matched summer and winter locations of a single north-east Atlantic blue whale with sightings in Iceland and Mauritania (REF).
+
+Inferences concerning movements of blue whales in general and in the northeast Atlantic in particular are therefore largely drawn from disparate information sources and rarely detail movements of individual animals over timescales of a year or more.
+It is particularly difficult to establish the degree of connectivity between summer and winter feeding areas, and the level of temporal consistency in movement behaviour within individuals.
+Information on historical movement patterns of individual blue whales is important for understanding the drivers of whale declines, and potential for human-whale conflicts as populations recover.
+
+Here we infer individual level high- resolution movement records for seven blue whales for whom baleen stable isotope data were available.
+Isotope data forro 6 whales stranding on the west coast of the USA or Mexico were reported in Busquetts- Vass et al 2016, and we report new isotopic data from a whale stranding off the coast of Wexford Ireland in March 1891, and currently on display in the Natural History Museum London’s (NHM) blue whale ‘Hope’.
+Hope The NHM whale (Cat No XXXX, unofficially christened ‘Hope”) stranded in Wexford Ireland in March 1891 and is a female estimated to be a young adult at least 15 years at  death, suggesting that her.
+Bbaleen from the NHM whale will therefore yields information about migratory behaviors in north Atlantic blue whales around the peak of industrial whaling.
 

### 2018-10-15 manuscript/NatureComms/blue-whale-supplementary-oct18.tex
Initial commit of new Oct 2018 version
Nature Comms version with new pacific analyses

--- 2018-02-14 manuscript/NatureEE/blue-whale-supplementary-v2.tex
@@
 % Preamble
-\documentclass[a4paper,12pt]{article}
+\documentclass[a4paper,10pt]{article}
 
-\usepackage[osf]{mathpazo} % palatino
+%\usepackage[osf]{mathpazo} % palatino
 %\usepackage{ms1}            % load the template
@@
 \pagenumbering{arabic}
-\linespread{1.66}
+
+% Weird Current Biology rules about times and single line spacing
+\linespread{1}
+\usepackage{times}
 
@@
 % Title page information
-\title{Supplementary Information from:\\
+\title{Supplemental Information from:\\
 \textit{Reconstructing the last known movements of one of Nature's giants}}
@@
 \begin{document}
-%\title{Supplementary Information - Cooper et al. 2017}
 %\modulolinenumbers[1]   % Line numbering on every line
@@
 
-\section*{Supplementary Methods}
+\section*{Supplemental Methods}
 
@@
 
-\section*{Supplementary Results}
+\section*{Supplemental Results: NHM blue whale}
 
@@
 
+\newpage
+
+\section*{Supplemental Results: Pacific blue whales from Busquets-Vass et al. 2017}
+
+\begin{landscape}
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[width=\linewidth]{figures/top25.moveWhaleA_TL2.eps}
+      \caption{Simulated locations by month taken from the top 25\% best fitting migratory movement model simulations for Pacific blue whale A.}
+      \label{figwhaleA}
+  \end{figure}
+\end{landscape}
+
+\begin{landscape}
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[width=\linewidth]{figures/top25.moveWhaleB_TL2.eps}
+      \caption{Simulated locations by month taken from the top 25\% best fitting migratory movement model simulations for Pacific blue whale B.}
+      \label{figwhaleB}
+  \end{figure}
+\end{landscape}
+
+\begin{landscape}
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[width=\linewidth]{figures/top25.moveWhaleC_TL2.eps}
+      \caption{Simulated locations by month taken from the top 25\% best fitting migratory movement model simulations for Pacific blue whale C.}
+      \label{figwhaleC}
+  \end{figure}
+\end{landscape}
+
+\begin{landscape}
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[width=\linewidth]{figures/top25.moveWhaleD_TL2.eps}
+      \caption{Simulated locations by month taken from the top 25\% best fitting migratory movement model simulations for Pacific blue whale D.}
+      \label{figwhaleD}
+  \end{figure}
+\end{landscape}
+
+\begin{landscape}
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[width=\linewidth]{figures/top10.moveWhaleE_TL2.eps}
+      \caption{Simulated locations by month taken from the top 10\% best fitting migratory movement model simulations for Pacific blue whale E.}
+      \label{figwhaleE}
+  \end{figure}
+\end{landscape}
+
+\begin{landscape}
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[width=\linewidth]{figures/top25.moveWhaleF_TL2.eps}
+      \caption{Simulated locations by month taken from the top 25\% best fitting migratory movement model simulations for Pacific blue whale F.}
+      \label{figwhaleF}
+  \end{figure}
+\end{landscape}
+
 % References
@@
 \bibliographystyle{nature}
-\bibliography{blue-whale}
+\bibliography{blue-whale-abbreviations}
 

### 2018-10-16 manuscript/NatureComms/blue-new.tex
Abstract edits

@@
 
-\section{Abstract} % Check number of words allowed...
+\section{Abstract} % 150 words only
 
-The spatial ecology of rare, migratory oceanic mammals such as the blue
-whale (\textit{Balaenoptera musculus}) is difficult to study directly, leading to distinct knowledge gaps associated with migratory behavior at the individual scale. 
-Recent advances in satellite telemetry have improved our understanding of individual-level movement, but tags rarely provide data for more than a few months, and tagging cannot provide retrospective information.
-Incrementally-grown tissues provide alternative, biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales, but inferring detailed spatio-temporal information from biochemical tracers is challenging.
+The spatial ecology of rare, migratory oceanic mammals such as blue
+whales (\textit{Balaenoptera musculus}) is difficult to study directly. 
+Recent advances in satellite telemetry have improved our understanding of individual-level migratory behavior, but tags generally last only a few months, and cannot provide retrospective information. 
+Incrementally-grown tissues provide an alternative means of reconstructing individual-level movements over long timescales, but inferring spatio-temporal information from biochemical tracers is challenging.
 Here we present a new approach combining stable isotope analyses of incrementally-grown tissues with simulation models to infer individual-level animal movements. 
-We use our new method to reconstruct multiple years of movement histories of individual modern and historic blue whales, including an iconic blue whale that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London.
-
-We recover inferences of individual space use and between-individual variation in movements that are consistent with satellite-tag based data for modern blue whales in the northeast Pacific Ocean. 
-For the historic whale, inferred movements over seven years of life of a single individual are also consistent with our limited understanding of the movement ecology of blue whales in the northeast Atlantic Ocean.
-Combining biochemical tracers in incrementally-grown tissues with simulation modeling unlocks the information recorded in historic collections, and provides inferences that are complimentary to and comparable with tag-derived movement data.
+We use our new method to reconstruct multiple years of movement histories of individual modern and historic blue whales, including an iconic blue whale which stranded in 1891 and is now on display at the Natural History Museum, London.
+Our method unlocks the information recorded in historic collections, and provides inferences that are complimentary to and comparable with tag-derived movement data.
 

### 2018-11-02 code/01-extract-models.R
Updated to 2018 analyses

@@ -2,14 +2,11 @@
 # About: script to extract top 10% and bottom 10% of models from 
 # simulations and extract latitude and longitude info from them
 # Tidied by Natalie Cooper Nov 2017
-# Edits Oct 2018 to use different plankton isotopes
-
+# Changes to fitting Nov 2018
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Load libraries
 library(plyr)
 # note if using group_by in dplyr later you'll need to detach plyr
-#library(maps)
-#library(mapdata)
 library(tidyverse)
 library(raster)
 
@@ -29,8 +26,8 @@ TL2 <- stack("data/TL2_raster.grd")
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Read in model simulations output
 # Takes a bit of time
-resTrack <- read_csv("data/model.sims.full.csv")
-
+#resTrack <- read_csv("data/model.sims.full.csv")
+resTrack <- read_csv("data/model.sims.csv")
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Extract model values only for days of sample - 
 # as defined above (could change to an average if wanted)
@@ -48,29 +45,14 @@ for(i in 1:nrow(test2)){
                                   y = test2[i, c("Lon", "Lat")])
 }
 
-test3 <- NA
-# Run loess through series and predict for same days as measured
-# for the six month sliding window isoscape
-test3 <- ddply(test2, "Rep", function(x){
-  lo <- predict(loess(x$TL2 ~ x$count2, span = 0.3))[1:97]
-})
-
-# Transpose
-test3 <- t(test3)
-
-# Add column names
-colnames(test3) <- 1:length(unique(resTrack$Rep)) - 1
-test3 <- as.data.frame(test3[-1, ])
-
-# Link to the measured data to allow regression comparison
-test3$Day <- test2$count2[test2$Rep == 1][1:97]
-test3$Blue <- rev(blue$d13C)
+test3 <- test2
 
 # Run regressions for each model simulation and save r square values
 r2 <- vector()
 
 for(run in 1:length(unique(resTrack$Rep))){
-	r2[run] <- summary(lm(test3[, run] ~ test3$Blue))$adj.r.squared
+	sub <- test3[test3$Rep == unique(resTrack$Rep)[run], ]
[…]

### 2018-11-02 code/Figure-3-code.R
Update code based on new smoothed values

@@ -15,7 +15,7 @@ blue <- filter(whale_isos, whale_ID == "KC7")
 top100 <- read.csv("data/top10smooth.csv")
 
 # Fix day numbers to match in both
-blue$Day.sim <- rev(unique(top100$Day))
+blue$Day.sim <- rev(unique(top100$count2))
 
 # Create three plots, one for blue whale and one for simulations, and 
 # a base plot with axes etc.
@@ -34,7 +34,7 @@ base_plot <-
   theme(panel.background = element_blank())
 
 top_plot <- 
-  ggplot(top100, aes(x = Day, y = d13C_smooth+2, group = Rep)) +
+  ggplot(top100, aes(x = count2, y = d13C+2, group = Rep)) +
   geom_line(alpha = 0.2, col = "grey") +
   theme_classic(base_size = 16) + 
   xlab("") + 

### 2018-11-02 manuscript/NatureComms/blue-new.tex
Updates based on CT comments

@@
 \usepackage{ms}            % load the template
-\usepackage[round]{natbib} % author-year citations
+%\usepackage[round]{natbib} % author-year citations
 \usepackage[superscript,biblabel]{cite} % for superscript citations
@@
 
-We apply our method to blue whales (\textit{Balaenoptera musculus}), the
-largest animal to have ever lived. 
+We apply our method to blue whales (\textit{Balaenoptera musculus}), the largest animal to have ever lived. 
 Mysticete whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater. 
@@
 Baleen is worn away at the tips over time, so a baleen plate reflects the most recent years of life, and rarely records an individual's entire lifespan. 
-Among the mysticete whales, blue whales are a particularly attractive target for model-based isotope movement work as they have a consistent, low trophic level diet and feed continuously through the year, likely driven by high energetic costs of maintaining extreme muscle mass %(Goldbogen et al 2015).
+Among the mysticete whales, blue whales are a particularly attractive target for model-based isotope movement work as they have a consistent, low trophic level diet and feed continuously through the year, likely driven by high energetic costs of maintaining extreme muscle mass\cite{goldbogen2015}.
 
-Similar to other large balaenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes %(Hucke-Gaete et al 2018). 
-As blue whales feed throughout the year to accommodate energetic costs, migration routes and feeding areas for blue whales are thought to be shaped by the year-round location of highly productive regions %(Branch et al 2007).
+Similar to other large balaenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes\cite{huckgaete2018}. 
+As blue whales feed throughout the year to accommodate energetic costs, migration routes and feeding areas for blue whales are thought to be shaped by the year-round location of highly productive regions\cite{branch2007}.
 Recent studies tracking movements of individual blue whales primarily in the northeast Pacific have demonstrated high levels of among-individual variation in movement histories including suspending migration for opportunistic feeding and skipped migration (i.e.
@@
 Winter feeding is less well characterized, but winter feeding grounds are known
-in the Gulf of California and in the eastern tropical Pacific, specifically the Costa Rica Dome where production is maintained by seasonal upwelling (REFS). % Refs???
+in the Gulf of California and in the eastern tropical Pacific, specifically the Costa Rica Dome where production is maintained by seasonal upwelling.
 However, most satellite deployments on blue whales report movement data for periods of time under six months \cite{heide2001new, silva2013north, bailey2009behavioural, lesage2017foraging, irvine2017quantifying}, and are unable to show individual movements between summer and winter feeding or breeding areas.
@@
 The small population size and oceanic habit makes blue whale movements extremely hard to study. 
-A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic also
-track regions of high production throughout the year, at least some
-whales wintering in the upwelling systems between Mauritania and the
-Cape Verde Islands \cite{baines2014upwellings}. 
+A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic also track regions of high production throughout the year, at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands \cite{baines2014upwellings}.
 Northward migration in the spring may occur along mid Atlantic corridors with peak sightings in the Azores around April-May \cite{silva2013north}. 
@@
 
-%We simulated 1200 individual movement patterns, then excluded simulations where the virtual whale stranded before reaching the 3019 days represented in the baleen.
-%We then compared the remaining 1049 simulations and the resulting simulated baleen $\delta^{13}$C records, to the measured records with simple linear regressions. 
-% CHECK THIS!!!!
-We simulated 1000 individual movement patterns and compared the resulting simulated baleen \(\delta^{13}\)C records to the measured records with simple linear regressions. 
-%the median $r^2$ value across 1049 simulated profiles was 0.49, and the maximum was 0.71 (Supplemental Fig. S6).
-Simulated baleen \(\delta^{13}\)C profiles produce a good fit to measured profiles, the median \(r^{2}\) value across 1000 simulated profiles was 0.49, and the maximum was 0.74 (Fig.
-S6, distribution of \(r^{2}\) values).
+We simulated 1200 individual movement patterns, then excluded simulations where the virtual whale stranded before reaching the 3019 days represented in the baleen.
+We then compared the remaining 1049 simulations and the resulting simulated baleen $\delta^{13}$C records, to the measured records with simple linear regressions. 
+Simulated baleen \(\delta^{13}\)C profiles produce a good fit to measured profiles, the median \(r^{2}\) value across 1049 simulated profiles was 0.31, and the maximum was 0.65 (Fig. S6). 
 The top 10\% best fitting simulated profiles are shown in Fig. \ref{fig2}. 
@@
 Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891.
-%Cross-correlation analysis demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase two (70cm to 18cm on the baleen plate), but no clear relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase one (the oldest 30cm of the baleen plate, i.e.
-100cm - 70cm; Supplementary Figure S2) or the final 500 days of the whale's life (youngest 18cm of the baleen plate; Supplementary Figure S2).
 
@@
 The degree of such trophic fractionation is unclear, however and we do not draw interpretations based on absolute \(\delta^{13}\)C values, rather on the relative \(\delta^{13}\)C values across the length of the baleen plate.
-We assume opportunistic ingestion of protein throughout the year \cite{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating, lesage2017foraging, bailey2009behavioural}.
-% Branch et al 2007; Hucke-Gaete et al 2017
+We assume opportunistic ingestion of protein throughout the year \cite{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating, lesage2017foraging, bailey2009behavioural, branch2007, huckgaete2018}.
 
@@
 We simulated seven years of whale movements 1000 times and two years of residency 30 times for each residency hotspot (Fig. S4).
-%We simulated whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.
-% CHECK THISSSSSSS!!!!
+We simulated whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.
 
@@
 
-%\bibliographystyle{nature}
-\bibliographystyle{authordate1}
+\bibliographystyle{nature}
+%\bibliographystyle{authordate1}
 \bibliography{blue-whale-abbreviations}

### 2018-11-02 code/Figure-1-code.R
Edits to names to reflect additional figure

@@ -1,71 +1,0 @@
-# Author: Clive Trueman
-# Adapted by: Andrew Jackson June 2017 & Natalie Cooper Oct 2017.
-# About: Script to generate figure 1
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Load libraries
-library(tidyverse)
-library(lubridate)
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Figure 1 - raw whale isotope data with two y-axes
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# read in data and extract just the blue whale
-whale_isos <- read_csv("data/raw-whale-isotope-data.csv")
-
-KC7 <- filter(whale_isos, whale_ID == "KC7")
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# convert sample number to days based on growth rate of baleen
-growth_rate <- 13.5  # baleen growth rate in centimetres per year
-last_sample <- ymd("1891-03-01") # final date of last keratin sample
-
-KC7 <- 
-  KC7 %>%
-  mutate(days = sample_cm * 365.25 / growth_rate) %>%
-  mutate(days = days - days[1]) %>%
-  mutate(rev_days = days - days[length(days)]) %>%
-  mutate(sample_date = last_sample + rev_days) %>%
-  mutate(year = year(sample_date))
-
-# Create a variable to label the years
-year_labels <- with(KC7, seq(min(year), max(year), by = 1))
-year_breaks <- which(duplicated(KC7$year) == FALSE)
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# plot the isotope data
-point_size <- 4
-
-isop <- ggplot(KC7, aes(x = rev(sample_cm))) + 
-  theme_classic(base_size = 16) + 
-  xlab("Baleen sample (cm from youngest sample)") + 
-  scale_x_continuous(breaks = seq(0, 100, 20), 
-                     labels = seq(100, 0, -20),
-                     sec.axis = sec_axis(~., 
-                                         breaks = rev(year_breaks),
-                                         labels = rev(year_labels),
-                                         name = "Year"))
-
-isop <- isop + geom_line(aes(y = d13C)) + 
-  geom_point(aes(y = d13C), size = point_size, shape = 21, fill = "black") +
[…]

### 2018-11-02 code/Figure-2-code.R

@@ -1,46 +1,71 @@
 # Author: Clive Trueman
-# About: script to plot track for top 10% of models
-# Tidied by Natalie Cooper Nov 2017
+# Adapted by: Andrew Jackson June 2017 & Natalie Cooper Oct 2017.
+# About: Script to generate figure 1
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Load libraries
-library(maps)
-library(mapdata)
-library(spatstat)
-library(viridis)
-
-# Read in data
-top10 <- read.csv("data/top10percent.csv")
-
-# Pick colours
-mycols <- c(viridis_pal()(10)[1], viridis_pal()(10)[5], viridis_pal()(10)[9])
-
-# Jitter points of lat and long to get smooth density plot
-jLat <- jitter(top10$Lat, factor = 2)
-jLon <- jitter(top10$Lon, factor = 1)
+library(tidyverse)
+library(lubridate)
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Points plot with phases coloured
-png("manuscript/revision/figures/Figure-2-points.png", width = 800, height = 600)
+# Figure 1 - raw whale isotope data with two y-axes
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# read in data and extract just the blue whale
+whale_isos <- read_csv("data/raw-whale-isotope-data.csv")
 
-# Create empty backgrounds
-plot(x = NA, y = NA, xlim = c(-80, 50), ylim = c(0, 80), 
-     xlab = "", ylab = "", axes = FALSE)
-rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], 
-     col = "grey90")
+KC7 <- filter(whale_isos, whale_ID == "KC7")
 
-# Add points coloured by phase
-points(jLon, jLat, pch = 16, col = mycols[top10$phase], cex = 0.2)
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# convert sample number to days based on growth rate of baleen
+growth_rate <- 13.5  # baleen growth rate in centimetres per year
+last_sample <- ymd("1891-03-01") # final date of last keratin sample
 
-# Add world map
-map('world', col = "black", fill = TRUE, add = TRUE, lwd = 0.25)
-# If you get this error, 
-# Error in as_mapper(.f, ...) : argument ".f" is missing, with no default
[…]

### 2018-11-02 code/Figure-3-code.R

@@ -1,76 +1,46 @@
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Code for Figure 3
-# Adapted from code by Clive Trueman by Natalie Cooper Nov 2017
+# Author: Clive Trueman
+# About: script to plot track for top 10% of models
+# Tidied by Natalie Cooper Nov 2017
+
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Load libraries
-library(tidyverse)
-library(gtable)
-library(grid)
+library(maps)
+library(mapdata)
+library(spatstat)
+library(viridis)
 
-# Read in the measured whale data and select blue whale
-whale_isos <- read_csv("data/raw-whale-isotope-data.csv")
-blue <- filter(whale_isos, whale_ID == "KC7")
+# Read in data
+top10 <- read.csv("data/top10percent.csv")
 
-# Read in top 100 models
-top100 <- read.csv("data/top10smooth.csv")
+# Pick colours
+mycols <- c(viridis_pal()(10)[1], viridis_pal()(10)[5], viridis_pal()(10)[9])
 
-# Fix day numbers to match in both
-blue$Day.sim <- rev(unique(top100$count2))
+# Jitter points of lat and long to get smooth density plot
+jLat <- jitter(top10$Lat, factor = 2)
+jLon <- jitter(top10$Lon, factor = 1)
 
-# Create three plots, one for blue whale and one for simulations, and 
-# a base plot with axes etc.
-base_plot <- 
-  ggplot(blue, aes(x = Day.sim, y = d13C)) +
-  #geom_line(size = 1.5) +
-  theme_classic(base_size = 16) + 
-  xlab("Days") + 
-  ylab(expression(paste(delta^{13}, "C (\u2030)"))) +
-  scale_y_continuous(limits = c(-20, -16.5),
-                     sec.axis = sec_axis(~.+4, 
-                                breaks = c(-16, -15, -14, -13),
-                                labels = c(-30, -25, -20, -15),
-                                name = expression(paste("simulated ", 
-                                                        delta^{13}, "C (\u2030)")))) +
-  theme(panel.background = element_blank())
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Points plot with phases coloured
+png("manuscript/revision/figures/Figure-2-points.png", width = 800, height = 600)
 
-top_plot <- 
[…]

### 2018-11-02 code/Figure-4-code.R

@@ -1,57 +1,76 @@
-# Author: Clive Trueman
-# About: script to plot track for top 10% of models by month
-# for behavioural phase 2.
-# Tidied by Natalie Cooper Jan 2018
-
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Code for Figure 3
+# Adapted from code by Clive Trueman by Natalie Cooper Nov 2017
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Load libraries
-library(maps)
-library(mapdata)
-library(spatstat)
-library(viridis)
+library(tidyverse)
+library(gtable)
+library(grid)
 
-# Read in data
-mid.top10 <- read.csv("data/mid.top10percent.csv")
+# Read in the measured whale data and select blue whale
+whale_isos <- read_csv("data/raw-whale-isotope-data.csv")
+blue <- filter(whale_isos, whale_ID == "KC7")
 
-# Pick colours to match phases in Figure 2
-mycols <- c(viridis_pal()(10)[1], viridis_pal()(10)[5], viridis_pal()(10)[9])
+# Read in top 100 models
+top100 <- read.csv("data/top10smooth.csv")
 
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Points plot with phases coloured
-png("manuscript/revision/figures/Figure-4-monthly.png", width = 800, height = 600)
+# Fix day numbers to match in both
+blue$Day.sim <- rev(unique(top100$count2))
 
-# Make 12 plots
-par(mfrow = c(3, 4))
-par(mar = c(1, 1, 2, 1))
+# Create three plots, one for blue whale and one for simulations, and 
+# a base plot with axes etc.
+base_plot <- 
+  ggplot(blue, aes(x = Day.sim, y = d13C)) +
+  #geom_line(size = 1.5) +
+  theme_classic(base_size = 16) + 
+  xlab("Days") + 
+  ylab(expression(paste(delta^{13}, "C (\u2030)"))) +
+  scale_y_continuous(limits = c(-20, -16.5),
+                     sec.axis = sec_axis(~.+4, 
+                                breaks = c(-16, -15, -14, -13),
+                                labels = c(-30, -25, -20, -15),
+                                name = expression(paste("simulated ", 
+                                                        delta^{13}, "C (\u2030)")))) +
+  theme(panel.background = element_blank())
 
-# Loop through months  
[…]

### 2018-11-02 code/Figure-5-code.R

@@ -1,0 +1,57 @@
+# Author: Clive Trueman
+# About: script to plot track for top 10% of models by month
+# for behavioural phase 2.
+# Tidied by Natalie Cooper Jan 2018
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Load libraries
+library(maps)
+library(mapdata)
+library(spatstat)
+library(viridis)
+
+# Read in data
+mid.top10 <- read.csv("data/mid.top10percent.csv")
+
+# Pick colours to match phases in Figure 2
+mycols <- c(viridis_pal()(10)[1], viridis_pal()(10)[5], viridis_pal()(10)[9])
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Points plot with phases coloured
+png("manuscript/revision/figures/Figure-4-monthly.png", width = 800, height = 600)
+
+# Make 12 plots
+par(mfrow = c(3, 4))
+par(mar = c(1, 1, 2, 1))
+
+# Loop through months  
+for (Mn in 1:12) {
+  
+  # Make background of plots
+  plot(x = NA, y = NA, xlim = c(-80, 50), ylim = c(0, 80), 
+       xlab = "", ylab = "", axes = FALSE, main = month.abb[Mn], cex.main = 2)
+  rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], 
+       col = "grey90")
+  
+  # Select month
+  Month <- mid.top10[mid.top10$Month == Mn, ]
+  
+  # Jitter points of lat and long to get smooth density plot
+  jLat <- jitter(Month$Lat, factor = 2)
+  jLon <- jitter(Month$Lon, factor = 1)
+  
+  # Add points coloured by phase
+  points(jLon, jLat, pch = 16, col = mycols[2], cex = 0.2)
+
+  # Add world map
+  map('world', col = "black", fill = TRUE, add = TRUE, lwd = 0.25)
+  # If you get this error, 
+  # Error in as_mapper(.f, ...) : argument ".f" is missing, with no default
+  # You need to detach the package purrr (detach(package:purrr)) and reload
+  # the maps library (library(map))
+  
+  par(new=FALSE)
+}
+
+dev.off()
+

### 2018-11-02 code/01-extract-models.R
Remove un-needed line

@@ -26,7 +26,6 @@ TL2 <- stack("data/TL2_raster.grd")
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Read in model simulations output
 # Takes a bit of time
-#resTrack <- read_csv("data/model.sims.full.csv")
 resTrack <- read_csv("data/model.sims.csv")
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Extract model values only for days of sample - 

### 2018-11-02 manuscript/PRSB/blue-new-supplementary.tex
Add PRSB version

--- 2018-02-14 manuscript/NatureEE/blue-whale-supplementary-v2.tex
@@
 % Preamble
-\documentclass[a4paper,12pt]{article}
-
+\documentclass[a4paper,10pt]{article}
+\usepackage[margin = 2cm]{geometry}
 \usepackage[osf]{mathpazo} % palatino
@@
 \usepackage{pdflscape} % landscape   
+\usepackage[multidot]{grffile} % deal with weird image issue where it can't cope if names have "." in them
+
 \pagenumbering{arabic}
-\linespread{1.66}
 
@@
 
+\newpage
+\section*{Supplemental Results: Pacific blue whales from Busquets-Vass et al. 2017}
+
+\begin{landscape}
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[angle = 90, origin = c, scale = 0.65]{figures/top25.moveWhaleA_TL2.pdf}
+      \caption{Locations corresponding to the top 25\% best-fitting migratory movement model simulations for Pacific blue whale A. Data from Busquets-Vass et al. 2017.}
+      \label{figwhaleA}
+  \end{figure}
+\end{landscape}
+
+\begin{landscape}
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[width=\linewidth, scale = 0.55]{figures/top25.moveWhaleB_TL2.pdf}
+      \caption{Locations corresponding to the top 25\% best-fitting migratory movement model simulations for Pacific blue whale B. Data from Busquets-Vass et al. 2017.}
+      \label{figwhaleB}
+  \end{figure}
+\end{landscape}
+
+\begin{landscape}
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[angle = 90, origin = c, scale = 0.65]{figures/top25.moveWhaleC_TL2.pdf}
+      \caption{Locations corresponding to the top 25\% best-fitting migratory movement model simulations for Pacific blue whale C. Data from Busquets-Vass et al. 2017.}
+      \label{figwhaleC}
+  \end{figure}
+\end{landscape}
+
+\begin{landscape}
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[angle = 90, origin = c, scale = 0.65]{figures/top25.moveWhaleD_TL2.pdf}
+      \caption{Locations corresponding to the top 25\% best-fitting migratory movement model simulations for Pacific blue whale D. Data from Busquets-Vass et al. 2017.}
+      \label{figwhaleD}
+  \end{figure}
+\end{landscape}
+
+\begin{landscape}
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[angle = 90, origin = c, scale = 0.65]{figures/top10.moveWhaleE_TL2.pdf}
+      \caption{Locations corresponding to the top 25\% best-fitting migratory movement model simulations for Pacific blue whale E. Data from Busquets-Vass et al. 2017.}
+      \label{figwhaleE}
+  \end{figure}
+\end{landscape}
+
+\begin{landscape}
+    \begin{figure}[!htbp]
+    \centering
+      \includegraphics[angle = 90, origin = c, scale = 0.65]{figures/top25.moveWhaleF_TL2.pdf}
+      \caption{Locations corresponding to the top 25\% best-fitting migratory movement model simulations for Pacific blue whale F. Data from Busquets-Vass et al. 2017.}
+      \label{figwhaleF}
+  \end{figure}
+\end{landscape}
+
 % References

### 2018-11-02 manuscript/PRSB/blue-new.tex
Add PRSB version

--- 2018-11-02 manuscript/NatureComms/blue-new.tex
@@
 %\usepackage[round]{natbib} % author-year citations
-\usepackage[superscript,biblabel]{cite} % for superscript citations
 \usepackage{graphicx}
@@
 
-\section{Abstract} % 150 words only
+\section{Abstract} % 200 words only
 
@@
 
-\section{Main text}\label{background}
+\textbf{Keywords: carbon stable isotopes, movement models}
+
+\section{Background}\label{background}
 
@@
 Baleen from the NHM whale therefore yields information about migratory behaviors in North Atlantic blue whales around the peak of industrial whaling.
+
+\section{Methods}\label{methods}
+
+\subsection{Stable isotope extractions from baleen}\label{stable-isotope-extractions-from-baleen}
+
+Baleen was collected from the Natural History Museum, London (specimen NHMUK.1892.3.1.1). 
+The baleen plate was cleaned with ethanol to remove surface contaminants such as skin/gum or other lipids that can influence isotopic signals. 
+{\raise.17ex\hbox{$\scriptstyle\sim$}}1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
+97 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal (gingival) section that contains the most recent tissue. 
+Baleen grows at a constant rate, so the samples are equally spaced through time\cite{best1996stable}. 
+Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the University of Southampton SEAPORT Stable Isotope Ratio Mass Spectrometry Laboratory (Southampton, UK), using a Vario Isotope select elemental analyser, coupled to an Isoprime 100 isotope mass spectrometer.
+Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
+C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples\cite{hobson1998stable}.
+All data are available from the NHM Data Portal\cite{data-set}.
+
+\subsection{Time calibrating stable isotope
+profiles}\label{time-calibrating-stable-isotope-profiles}
+
+Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates\cite{hobson1998stable,busquets2017estimating}.
+Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
+We calculated isotopic periodicity within the NHM baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Supplementary Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated by Busquets-Vass \textit{et al.}\cite{busquets2017estimating}.
+Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891.
+
+\subsection{Baseline isotope
+comparisons}\label{baseline-isotope-comparisons}
+
+Isotope-enabled biogeochemical ocean models \cite{magozzi2017using,schmittner2016complementary} were used to characterize
+the isotopic composition of phytoplankton expected in different
+potential foraging grounds (Fig. S3). 
+Annual average \(\delta^{15}\)N POM (particulate organic matter) values were provided by C.J.
+Somes (\textit{pers.comm}) based on a 5\({}^{\circ}\) resolution biogeochemical model (Fig.
+S3).
+\(\delta^{13}\)C POM values were simulated at 1\({}^{\circ}\) and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model \cite{magozzi2017using,yool2013medusa}.
+Simulated \(\delta^{15}\)N POM values are relatively positive in the northeast Atlantic north of c. 60\({}^{\circ}\)N, and relatively negative in the central and southern North Atlantic. 
+Annual average \(\delta^{13}\)C POM values largely vary with latitude, with more negative values in more northerly regions. 
+In the central North Atlantic, \(\delta^{13}\)C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S3). 
+The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \cite{magozzi2017using,laws1995dependence}.
+Thus temporal variations in \(\delta^{13}\)C POM values are superimposed on latitudinal gradients. 
+The scale and nature of temporal variation in \(\delta^{13}\)C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in \(\delta^{13}\)C POM values linked to strongly seasonal phytoplankton growth dynamics.
+The model simulations used are forced with decadal climatological data from 2000-2010\cite{magozzi2017using}. 
+We do not draw inferences about location from absolute \(\delta^{13}\)C values and thus the reduction in oceanic \(\delta^{13}\)C POM values caused by release of fossil carbon into the atmosphere (Suess effect) does not influence inferences about location or movement.
+Ocean basin scale spatio-temporal differences in \(\delta^{13}\)C values largely reflect relative differences in sea surface temperatures between trophic, temperate and arctic waters, therefore we are confident that the Magozzi \textit{et al}.\cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.
+
+We used \(\delta^{13}\)C POM values modeled at monthly resolution to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours. 
+The stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on in the weeks prior to keratin growth. 
+Assimilation of carbon into krill tissues will dampen the temporal variability seen in POM, effectively producing a temporal average over the timescale of isotopic turnover within krill. 
+We estimate turnover to be complete between two and four months and therefore we resampled the \(\delta^{13}\)C POM values in each one degree cell to reflect an average of isotopic compositions in phytoplankton in the two months prior to the sampling date.
+Average values were weighted according to the proportional plankton biomass estimated for each month. 
+Carbon isotope values are also likely to be fractionated during transfer from plankton to krill as \textsuperscript{12}C is preferentially lost through respiration. 
+The degree of such trophic fractionation is unclear, however and we do not draw interpretations based on absolute \(\delta^{13}\)C values, rather on the relative \(\delta^{13}\)C values across the length of the baleen plate.
+We assume opportunistic ingestion of protein throughout the year \cite{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating, lesage2017foraging, bailey2009behavioural, branch2007, huckgaete2018}.
+
+\subsection{Agent-based whale movement
+model}\label{agent-based-whale-movement-model}
+
+We simulated whale movements with the likelihood, direction, and extent of movement influenced by behavioral state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
+Movement was coded as a set of probabilistic rules, informed by the literature on blue whale behavior (e.g. \cite{handbook}).
+All terms were expressed as probability distributions, yielding multiple potential movement tracks.
+
+In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and linear distance of movement are all influenced by the following. 
+(i) Behavioural state (migrating north, migrating south, or foraging).
+Northerly migrations were possible only in spring, and southerly migrations in autumn. 
+Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
+(ii) Sea surface temperature \cite{yool2013medusa} ($^{\circ}$C). 
+When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.
+(iii) Water depth (m; from the General Bathymetric Chart of the Oceans (GEBCO) global bathymetry dataset\cite{bathy}, extracted via the marmap package in R \cite{marmap}). 
+Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
+(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities \cite{yool2013medusa}). 
+This was included as a proxy for zooplankton food availability. 
+Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state.
+
+For Pacific whales where satellite data offers a rich view of individual movements, we coded our movement models to reflect either residency in southern waters (between Baja California and the Costa Rica Dome) or seasonal latitudinal movements within the California current system.
+We simulated 250 movement tracks of four years duration. 
+We allowed start and end dates to vary by aligning measured profiles to sections of simulated profiles of identical lengths with sequentially changing start dates and saved best fitting models.
+
+For the historic North Atlantic whale, we simulated the isotopicexpression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
+Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four
+years of simulations. 
+For a full description of the model and its parameters see Supplementary Methods: Fig. S5 and Table S1. 
+We simulated seven years of whale movements 1000 times and two years of residency 30 times for each residency hotspot (Fig. S4).
+We simulated whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.
+
+At each time point and location in the simulations, we extracted time averaged \(\delta^{13}\)C values from the spatio-temporal model\cite{magozzi2017using}resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs X, \ref{fig2}, and S4).
+% CHECK FIGS
+We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).
 
@@
 
-\section{Methods}\label{methods}
-
-\subsection{Stable isotope extractions from baleen}\label{stable-isotope-extractions-from-baleen}
-
-Baleen was collected from the Natural History Museum, London (specimen NHMUK.1892.3.1.1). 
-The baleen plate was cleaned with ethanol to remove surface contaminants such as skin/gum or other lipids that can influence isotopic signals. 
-{\raise.17ex\hbox{$\scriptstyle\sim$}}1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
-97 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal (gingival) section that contains the most recent tissue. 
-Baleen grows at a constant rate, so the samples are equally spaced through time\cite{best1996stable}. 
-Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the University of Southampton SEAPORT Stable Isotope Ratio Mass Spectrometry Laboratory (Southampton, UK), using a Vario Isotope select elemental analyser, coupled to an Isoprime 100 isotope mass spectrometer.
-Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
-C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples\cite{hobson1998stable}.
-All data are available from the NHM Data Portal\cite{data-set}.
-
-\subsection{Time calibrating stable isotope
-profiles}\label{time-calibrating-stable-isotope-profiles}
-
-Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates\cite{hobson1998stable,busquets2017estimating}.
-Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
-We calculated isotopic periodicity within the NHM baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Supplementary Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated by Busquets-Vass \textit{et al.}\cite{busquets2017estimating}.
-Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891.
-
-\subsection{Baseline isotope
-comparisons}\label{baseline-isotope-comparisons}
-
-Isotope-enabled biogeochemical ocean models \cite{magozzi2017using,schmittner2016complementary} were used to characterize
-the isotopic composition of phytoplankton expected in different
-potential foraging grounds (Fig. S3). 
-Annual average \(\delta^{15}\)N POM (particulate organic matter) values were provided by C.J.
-Somes (\textit{pers.comm}) based on a 5\({}^{\circ}\) resolution biogeochemical model (Fig.
-S3).
-\(\delta^{13}\)C POM values were simulated at 1\({}^{\circ}\) and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model \cite{magozzi2017using,yool2013medusa}.
-Simulated \(\delta^{15}\)N POM values are relatively positive in the northeast Atlantic north of c. 60\({}^{\circ}\)N, and relatively negative in the central and southern North Atlantic. 
-Annual average \(\delta^{13}\)C POM values largely vary with latitude, with more negative values in more northerly regions. 
-In the central North Atlantic, \(\delta^{13}\)C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S3). 
-The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \cite{magozzi2017using,laws1995dependence}.
-Thus temporal variations in \(\delta^{13}\)C POM values are superimposed on latitudinal gradients. 
-The scale and nature of temporal variation in \(\delta^{13}\)C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in \(\delta^{13}\)C POM values linked to strongly seasonal phytoplankton growth dynamics.
-The model simulations used are forced with decadal climatological data from 2000-2010\cite{magozzi2017using}. 
-We do not draw inferences about location from absolute \(\delta^{13}\)C values and thus the reduction in oceanic \(\delta^{13}\)C POM values caused by release of fossil carbon into the atmosphere (Suess effect) does not influence inferences about location or movement.
-Ocean basin scale spatio-temporal differences in \(\delta^{13}\)C values largely reflect relative differences in sea surface temperatures between trophic, temperate and arctic waters, therefore we are confident that the Magozzi \textit{et al}.\cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.
-
-We used \(\delta^{13}\)C POM values modeled at monthly resolution to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours. 
-The stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on in the weeks prior to keratin growth. 
-Assimilation of carbon into krill tissues will dampen the temporal variability seen in POM, effectively producing a temporal average over the timescale of isotopic turnover within krill. 
-We estimate turnover to be complete between two and four months and therefore we resampled the \(\delta^{13}\)C POM values in each one degree cell to reflect an average of isotopic compositions in phytoplankton in the two months prior to the sampling date.
-Average values were weighted according to the proportional plankton biomass estimated for each month. 
-Carbon isotope values are also likely to be fractionated during transfer from plankton to krill as \textsuperscript{12}C is preferentially lost through respiration. 
-The degree of such trophic fractionation is unclear, however and we do not draw interpretations based on absolute \(\delta^{13}\)C values, rather on the relative \(\delta^{13}\)C values across the length of the baleen plate.
-We assume opportunistic ingestion of protein throughout the year \cite{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating, lesage2017foraging, bailey2009behavioural, branch2007, huckgaete2018}.
-
-\subsection{Agent-based whale movement
-model}\label{agent-based-whale-movement-model}
-
-We simulated whale movements with the likelihood, direction, and extent of movement influenced by behavioral state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
-Movement was coded as a set of probabilistic rules, informed by the literature on blue whale behavior (e.g. \cite{handbook}).
-All terms were expressed as probability distributions, yielding multiple potential movement tracks.
-
-In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and linear distance of movement are all influenced by the following. 
-(i) Behavioural state (migrating north, migrating south, or foraging).
-Northerly migrations were possible only in spring, and southerly migrations in autumn. 
-Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
-(ii) Sea surface temperature \cite{yool2013medusa} ($^{\circ}$C). 
-When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.
-(iii) Water depth (m; from the General Bathymetric Chart of the Oceans (GEBCO) global bathymetry dataset\cite{bathy}, extracted via the marmap package in R \cite{marmap}). 
-Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
-(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities \cite{yool2013medusa}). 
-This was included as a proxy for zooplankton food availability. 
-Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state.
-
-For Pacific whales where satellite data offers a rich view of individual movements, we coded our movement models to reflect either residency in southern waters (between Baja California and the Costa Rica Dome) or seasonal latitudinal movements within the California current system.
-We simulated 250 movement tracks of four years duration. 
-We allowed start and end dates to vary by aligning measured profiles to sections of simulated profiles of identical lengths with sequentially changing start dates and saved best fitting models.
-
-For the historic North Atlantic whale, we simulated the isotopicexpression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
-Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four
-years of simulations. 
-For a full description of the model and its parameters see Supplementary Methods: Fig. S5 and Table S1. 
-We simulated seven years of whale movements 1000 times and two years of residency 30 times for each residency hotspot (Fig. S4).
-We simulated whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.
-
-At each time point and location in the simulations, we extracted time averaged \(\delta^{13}\)C values from the spatio-temporal model\cite{magozzi2017using}resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs X, \ref{fig2}, and S4).
-% CHECK FIGS
-We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).
-
-%\subsection{Data Code and Materials}\label{data-code-and-materials}
-\subsection{Data Availability}
-Data are available from the NHM Data Portal (https://doi.org/10.5519/0093278). 
-R code is available from GitHub (https://github.com/nhcooper123/blue-whale-bes)(Zenodo DOI will be added before publication).
-
 % References
@@
 
+\section{Competing interests}
+We have no competing interests.
+
 \section{Author contributions}\label{author-contributions}
@@
 
-\section{Competing interests}
-The authors declare no competing interests.
+\subsection{Data Code and Materials}\label{data-code-and-materials}
+Data are available from the NHM Data Portal (https://doi.org/10.5519/0093278). 
+R code is available from GitHub (https://github.com/nhcooper123/blue-whale-bes)(Zenodo DOI will be added before publication).
 

### 2018-11-07 code/pacific-figure-andrew-temp/plot-figure-x.Rmd
added a temp folder with my code and data for the pacific figure
if we are happy with this, i can move the data to the data folder and
move this up a folder and adjust the paths accordingly. For now ive
just dumped it in here as a block. One question mark is over the
workflow, which doesnt show where the files test2A.csv etc… come from,
which i think is from some of Clive’s original code. I can check this
too.

---
title: "Plot-figure-X"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
---
title: "generates Figure X"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

## Setup

Load required libraries.

```{r setup}

library(ggplot2)
library(dplyr)

```

## Read in the data

This data comes is derived from processing the Pacific whale movement models. These are the best fit simulations in terms of $R^2$ values obtained from linear regression of the normalised model predictions on the normalised empirical data.

```{r import-empirical-data}

# data in local csv file
pacific_data <- read.csv("pacific-blue-whale-isotope-data.csv", 
                         header = TRUE, stringsAsFactors = FALSE)

# convert ID to factor
# pacific_data$whale_ID <- factor(pacific_data$whale_ID)

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
# set the day no based on estimated growth rate given in Busquetts-Vass
GRx<-15.5
GRA<-13.5
GRB<-14.8
GRC<-17.5

# bundle into an ordered vector
growth_rates <- c(GRA, GRB, GRC, GRx, GRx, GRx)

# create an empty vector in the empirical data.frame to record day number (time)
pacific_data$day_no <- 0
pacific_data$day_no_f <- 0 # flipped time vecrot

# loop over each individual and calculate their spefic time vector 
# based on their growth rates
for (i in 1:length(growth_rates)){

  # find indices of focal whale individual
  idx <- pacific_data$whale_ID == LETTERS[i]

  # conver their sampled baleen cm to time via growth rate
  pacific_data$day_no[idx] <- as.integer(pacific_data$sample_cm[idx] * 
                                           growth_rates[i]) + 1

  # flip time into day_no_f vector
  pacific_data$day_no_f[idx] <- -1 * (pacific_data$day_no[idx] - 
                                        max(pacific_data$day_no[idx]))

}

# 
# # baleen$day.no<-baleen$cm*GRx
# 
# colz<-colorRampPalette(c("brown", "cadet blue", "pink"))(12)
# colz<-seq(1:6)
# A<-baleen[baleen$Baleen_code=="A",]
# B<-baleen[baleen$Baleen_code =="B",]
# C<-baleen[baleen$Baleen_code =="C",]
# D<-baleen[baleen$Baleen_code =="D",]
# E<-baleen[baleen$Baleen_code =="E",]
# Fw<-baleen[baleen$Baleen_code =="F",]
# 
# A$day.no<-as.integer(A$cm*GRA)+1
# B$day.no<-as.integer(B$cm*GRB)+1
# C$day.no<-as.integer(C$cm*GRC)+1
# D$day.no<-as.integer(D$cm*GRx)+1
# E$day.no<-as.integer(E$cm*GRx)+1
# Fw$day.no<-as.integer(Fw$cm*GRx)+1
# 
# # fix so that day numbers start at the oldest end (i.e. time moves forwards as in simulated baleen)
# A$day.no.F<-((A$day.no)-max(A$day.no, na.rm=TRUE))*-1
# B$day.no.F<-((B$day.no)-max(B$day.no, na.rm=TRUE))*-1
# C$day.no.F<-((C$day.no)-max(C$day.no, na.rm=TRUE))*-1
# D$day.no.F<-((D$day.no)-max(D$day.no, na.rm=TRUE))*-1
# E$day.no.F<-((E$day.no)-max(E$day.no, na.rm=TRUE))*-1
# Fw$day.no.F<-((Fw$day.no)-max(Fw$day.no, na.rm=TRUE))*-1

```

## Quick plot of empirical data

```{r quick-plot}

p1 <- ggplot(pacific_data, aes(x = day_no_f, y = d13C)) + 
  facet_wrap(~whale_ID) + 
  geom_line(col = "red") + 
  geom_point(col = "red", size = 1)

print(p1)

```

## Import simulated data

```{r}
# Read in the corresponding best fit simulated data

best.start.All<-c(7, 1, 6, 7, 1, 5)
end.All<-c(41, 67, 67, 55, 71, 58)
best.sim.All<-c(197, 14, 121, 48, 41, 48)

# they are in csv files named test2A, test2B etc...
read_in_these <- paste0("test2", LETTERS[1:6], ".csv")

# read in and bind into a single data.frame
sim_data <- list()

# loop over csv files, import and subset for the "best sim"
for (i in 1:length(read_in_these)){

  tmp <- read.csv(read_in_these[i], 
                                 header = TRUE, 
                                 stringsAsFactors = FALSE)

  # add the appropriate whale_ID letter
  tmp$whale_ID <- LETTERS[i]

  # subset by the index for the best sim in each case
  # this is hard-coded from above
  tmp2 <- subset(tmp, tmp$Rep == best.sim.All[i])

  # further select a subset window of this time series
  tmp3 <- tmp2[best.start.All[i]:(end.All[i]+(best.start.All[i]-1)),]

  # add the time vectors from the empirical data that should match
  tmp3$day_no <- pacific_data$day_no[pacific_data$whale_ID == LETTERS[i]]
  tmp3$day_no_f <- pacific_data$day_no_f[pacific_data$whale_ID == LETTERS[i]]

  print(unique(tmp3$whale_ID))

  # write this subset of data into the list sim_data and for ad hoc reasons
  # remove the NAs from the end
  sim_data[[i]] <- tmp3[!is.na(tmp3$whale_ID),]

  # remove tmp for next iteration of loop
  rm(tmp, tmp2, tmp3)
}

# cant use lapply easily owing to the required subsetting
# sim_data <- dplyr::bind_rows(lapply(read_in_these, 
#                                     read.csv,
#                                     header = T,
#                                     stringsAsFactors = FALSE))

# convert list to single data.frame
sim_data_df <- bind_rows(sim_data)

# coerce whale_ID to factor
sim_data_df$whale_ID <- factor(sim_data_df$whale_ID)

# remove no longer required objects
rm(sim_data)

```

## Quick plot simulated data

```{r quick-simulated}

p2 <- ggplot(sim_data_df, aes(x = day_no, y = TL2)) + 
  facet_wrap(~whale_ID) + 
  geom_line()

print(p2)

```

## plot both together

```{r plot-both-together}

# merge them 
merged_whales <- pacific_data %>% 
  left_join(sim_data_df, by = c("whale_ID", "day_no", "day_no_f"))

merged_whales <- merged_whales %>% group_by(whale_ID) %>% 
  mutate(
  z.d13C.obs = scale(d13C.x),
  z.d13C.sim = scale(TL2)
)

p3 <-  ggplot(merged_whales, aes(x = day_no_f, y = scale(d13C.x))) + 
  facet_wrap(~whale_ID) + 
  geom_line(col = "red") + 
  geom_point(col = "red", size = 1) + 
  geom_line(aes(x = day_no, y = scale(TL2)), col = "black") + 
  xlab("Day number") + 
  ylab("z-scored d13C")

print(p3)

# plot the data z-scored by group

p4 <-  ggplot(merged_whales, aes(x = day_no_f, y = z.d13C.obs)) + 
  facet_wrap(~whale_ID) + 
  geom_line(col = "red") + 
  geom_point(col = "red", size = 1) + 
  geom_line(aes(x = day_no, y = z.d13C.sim), col = "black") + 
  xlab("Day number") + 
  ylab("z-scored d13C")

print(p4)

ggsave("pacific-fits.pdf", plot = p4)
```

### 2018-11-08 code/pacific-figure-andrew-temp/plot-figure-x.Rmd
minor updates on colours

@@
   facet_wrap(~whale_ID) + 
-  geom_line(col = "red") + 
-  geom_point(col = "red", size = 1) + 
-  geom_line(aes(x = day_no, y = z.d13C.sim), col = "black") + 
+  geom_line(col = "black") + 
+  geom_point(col = "black", size = 1) + 
+  geom_line(aes(x = day_no, y = z.d13C.sim), col = "grey60") + 
   xlab("Day number") + 
-  ylab("z-scored d13C")
+  ylab("z-scored d13C") + 
+  theme_classic() + 
+  theme(strip.background = element_blank())
 
@@
 
-ggsave("pacific-fits.pdf", plot = p4)
+ggsave("pacific-fits.eps", plot = p4)
 ```

### 2019-01-17 oldversions/blue-new.tex
Move old manuscript versions

--- 2018-11-02 manuscript/NatureComms/blue-new.tex
@@
  \centering
-  \includegraphics[width = \linewidth]{figures/figure-Pacific-draft.eps}
-  \caption{Correlations among simulated $\delta^{13}$C from the best fitting migratory movement models (grey lines, right hand y-axis) and $\delta^{13}$C from baleen (black line, left hand y-axis) for six Pacific blue whales taken from \cite{busquets2017estimating}.
+  \includegraphics[width = \linewidth]{figures/Figure-pacific-fits-zscore.pdf}
+  \caption{Correlations among simulated $\delta^{13}$C from the best fitting migratory movement models (black lines) and $\delta^{13}$C from baleen (red lines and dots) for six Pacific blue whales taken from \cite{busquets2017estimating}.
+  Each time series of $\delta^{13}$C have been independently z-scored to normalise both the mean and standard deviations to facilitate visualisation of the correlation.
   Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location, reflecting temporal integration of phytoplankton $\delta^{13}$C values within the food chain before ingestion by the whale as krill.
@@
 Samples were taken longitudinally through the baleen plate (n = 97 samples from a single baleen plate for both isotopes).
-There is strong annual periodicity and cross-correlation (Supplemental Figures S1 and S2) in both isotopes.
+There is strong annual periodicity and cross-correlation (Supplemental Figs S1 and S2) in both isotopes.
 The approximate relationship to years assuming a growth rate of 13.5cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
@@
 Our simulations identify a range of possible locations for the whale (Fig.
-\ref{fig3}), although areas around the Mauritanian coast and Cape Verde Islands, a known current and historic winter feeding area for blue whales \cite{baines2014upwellings,reeves2004historical}, and potentially to the west of the Azores (Figure \ref{fig2}), most closely match the measured profile.
+\ref{fig3}), although areas around the Mauritanian coast and Cape Verde Islands, a known current and historic winter feeding area for blue whales \cite{baines2014upwellings,reeves2004historical}, and potentially to the west of the Azores (Fig.
+\ref{fig2}), most closely match the measured profile.
 The whale remained in these warm waters for at least one full year. 
@@
   \includegraphics[width = \linewidth]{figures/Figure-3-blue-sims.png}
-  \caption{Correlations among simulated $\delta^{13}$C from the top 10\% best fitting migratory movement models (grey lines, right hand y-axis) and $\delta^{13}$C from baleen (black line, left hand y-axis; see Figure \ref{fig1}).
+  \caption{Correlations among simulated $\delta^{13}$C from the top 10\% best fitting migratory movement models (grey lines, right hand y-axis) and $\delta^{13}$C from baleen (black line, left hand y-axis; see Fig.
+\ref{fig1}).
   Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location, reflecting temporal integration of phytoplankton $\delta^{13}$C values within the food chain before ingestion by the whale as krill.
@@
 Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
-We calculated isotopic periodicity within the NHM baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Supplementary Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated by Busquets-Vass \textit{et al.}\cite{busquets2017estimating}.
+We calculated isotopic periodicity within the NHM baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Supplemental Fig.
+S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated by Busquets-Vass \textit{et al.}\cite{busquets2017estimating}.
 Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891.
@@
 
-For the historic North Atlantic whale, we simulated the isotopicexpression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
+For the historic North Atlantic whale, we simulated the isotopicexpression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Fig.
+S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
 Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four
@@
 
-At each time point and location in the simulations, we extracted time averaged \(\delta^{13}\)C values from the spatio-temporal model\cite{magozzi2017using}resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs X, \ref{fig2}, and S4).
-% CHECK FIGS
-We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).
+At each time point and location in the simulations, we extracted time averaged \(\delta^{13}\)C values from the spatio-temporal model\cite{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs \ref{fig2}, and S4).
+We then compared the simulated stable isotope profiles (Fig. \ref{fig3}) to the profile of the blue whale (Fig. \ref{fig1}).
 

### 2019-01-17 code/pacific-figure-andrew-temp/plot-figure-x.Rmd

--- 2018-11-07 code/pacific-figure-andrew-temp/plot-figure-x.Rmd
@@
 # plot the data z-scored by group
-
 p4 <-  ggplot(merged_whales, aes(x = day_no_f, y = z.d13C.obs)) + 
   facet_wrap(~whale_ID) + 
-  geom_line(col = "red") + 
-  geom_point(col = "red", size = 1) + 
-  geom_line(aes(x = day_no, y = z.d13C.sim), col = "black") + 
+  geom_line(col = "black") + 
+  geom_point(col = "black", size = 1) + 
+  geom_line(aes(x = day_no, y = z.d13C.sim), col = "grey30") + 
   xlab("Day number") + 
-  ylab("z-scored d13C")
+  ylab(expression(paste("z-scored ", delta^{13}, "C"))) + 
+  theme_classic() + 
+  theme(strip.background = element_rect(colour=NA, fill=NA))
 
@@
 
-ggsave("pacific-fits.pdf", plot = p4)
+ggsave("pacific-fits-zscore.eps", plot = p4)
+
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Also try to plot the raw data but shifted and with 
+# a second y-axis as per Figure-2-code.R
+# AJ - THIS IS MORE DIFFICULT THAN A QUICK BIT OF CODE CAN FIX 
+# AS THE SHIFT NEEDS TO BE APPLIED SEPARATELY TO EACH WHALE AND HENCE
+# FACET. IM GOING TO COMMENT OUT AND STICK WITH THE Z-SCORED OPTION 
+# JUST ABOVE.
+
+# add the d15N data but need to scale it to d13C
+# v_shift <- mean(merged_whales$d13C.x) - mean(merged_whales$TL2)
+# 
+# p5 <-  ggplot(merged_whales, aes(x = day_no_f, y = d13C.x)) + 
+#   facet_wrap(~whale_ID) + 
+#   geom_line(col = "red") + 
+#   geom_point(col = "red", size = 1) + 
+#   geom_line(aes(x = day_no, y = TL2), col = "black") + 
+#   xlab("Day number") + 
+#   ylab(expression(paste(delta^{13}, "C (\u2030) Observed"))) + 
+#   scale_y_continuous(
+#     sec.axis = sec_axis(~.+v_shift, 
+#                         name = expression(paste(delta^{13}, 
+#                                                 "C (\u2030) Simulated"))))
+# 
+# print(p5)
+# 
+# ggsave("pacific-fits-shifted.eps", plot = p4)
+
 ```

### 2019-01-17 manuscript/PeerJ/blue-new.tex
Minor format changes

--- 2018-11-02 manuscript/PRSB/blue-new.tex
@@
   $^5$ Department of Geology and Geophysics, University of Utah, Salt Lake City, UT 84112-0102, USA.\\
+  $*$Email address: natalie.cooper@nhm.ac.uk
 }}
@@
 
-\section{Abstract} % 200 words only
+\section{Abstract}
 
@@
 
-\section{Background}\label{background}
+\section{Introduction}\label{background}
 
@@
 
-\section{Methods}\label{methods}
+\section{Materials and Methods}\label{methods}
 
@@
 C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples\cite{hobson1998stable}.
-All data are available from the NHM Data Portal\cite{data-set}.
+All data are available from the NHM Data Portal\cite{data-set} (https://doi.org/10.5519/0093278).
 
@@
 At each time point and location in the simulations, we extracted time averaged \(\delta^{13}\)C values from the spatio-temporal model\cite{magozzi2017using}resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs X, \ref{fig2}, and S4).
-% CHECK FIGS
 We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).
@@
 
-% References
-
-\bibliographystyle{nature}
-%\bibliographystyle{authordate1}
-\bibliography{blue-whale-abbreviations}
-
 \section{Acknowledgments}\label{acknowledgments}
@@
 
-\section{Competing interests}
-We have no competing interests.
-
-\section{Author contributions}\label{author-contributions}
-CT, NC, AJ, RS, and KC collected baleen from the NHM collections. 
-KC and CT extracted and analysed samples.
-CT created the movement model.
-EC, NC, SM and RS helped parameterise the model.
-AJ and NC created the figures. 
-CT and NC wrote the paper.
-All authors edited drafts and approved the final version.
+% References
+\bibliographystyle{refs}
+%\bibliographystyle{authordate1}
+\bibliography{blue-whale-abbreviations}
 

### 2019-01-17 manuscript/PeerJ/blue-new.tex
Add GitHub ref

@@
 
+R code for all analyses is available from GitHub (https://github.com/nhcooper123/blue-whale-bes; \cite{github})
+
 \section{Results and Discussion}
@@
 Data are available from the NHM Data Portal (https://doi.org/10.5519/0093278). 
-R code is available from GitHub (https://github.com/nhcooper123/blue-whale-bes)(Zenodo DOI will be added before publication).
+R code is available from GitHub (https://github.com/nhcooper123/blue-whale-bes)(Zenodo DOI 10.5281/zenodo.2542777).
 

### 2019-01-17 manuscript/PeerJ/blue-new-supplementary.tex
Change ref file
Update to author year citations

--- 2018-11-02 manuscript/PRSB/blue-new-supplementary.tex
@@
 \usepackage[osf]{mathpazo} % palatino
-%\usepackage{ms1}            % load the template
-%\usepackage[round]{natbib} % author-year citations
-\usepackage[superscript,biblabel]{cite} % for superscript citations
+\usepackage[round]{natbib} % author-year citations
 \usepackage{graphicx}
@@
 Across the baleen sample, $\delta^{15}$N values show clear cycles with periods of relatively high $\delta^{15}$N values interspersed by relatively short periods where $\delta^{15}$N values are c.1\text{\textperthousand} lower.
-Fourier analysis (periodogram function in the R package TSA\cite{Chan:2012aa}) revealed a strong periodic repetition with a 13.3cm periodicity. 
+Fourier analysis (periodogram function in the R package TSA \citep{Chan:2012aa}) revealed a strong periodic repetition with a 13.3cm periodicity. 
 The oldest (most distal) 35cm of the baleen plate records relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.
@@
 \subsection*{BASELINE ISOTOPE COMPARISONS}
-Isotope-enabled biogeochemical ocean models\cite{magozzi2017using,schmittner2016complementary} were used to characterise the isotopic composition of phytoplankton expected in different potential foraging grounds.
+Isotope-enabled biogeochemical ocean models \citep{magozzi2017using,schmittner2016complementary} were used to characterise the isotopic composition of phytoplankton expected in different potential foraging grounds.
 Annual average $\delta^{15}$N POM values were provided by C.J. Somes (\textit{pers.comm}) based on a 5$^{\circ}$ resolution biogeochemical model. 
-$\delta^{13}$C POM values were simulated using an isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using}.
+$\delta^{13}$C POM values were simulated using an isotopic extension to the NEMO-MEDUSA model \citep{magozzi2017using}.
 
@@
     \caption{Spatial model (isoscape) of model-simulated biomass, using weighted annual average $\delta^{13}$C (left) and $\delta^{15}$N (right) values from particulate organic matter (POM).
-$\delta^{13}$C values are from Magozzi et al.
-2017\cite{magozzi2017using} and $\delta^{15}$N from Schmittner \& Somes 2016\cite{schmittner2016complementary} and C.J.
+$\delta^{13}$C values are from \cite{magozzi2017using} and $\delta^{15}$N from \cite{schmittner2016complementary} and C.J.
 Somes (\textit{pers.
@@
 
-The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature\cite{laws1995dependence}.
+The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \citep{laws1995dependence}.
 Thus temporal variations in $\delta^{13}$C POM values are superimposed on latitudinal gradients. 
@@
 The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Figure 3). 
-Blue whales have a gestation period of 10-12 months, and calves suckle for approximately 6-7 months \cite{handbook}. 
+Blue whales have a gestation period of 10-12 months, and calves suckle for approximately 6-7 months \citep{handbook}. 
 The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer and autumn 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the assumed reproductive ecology of blue whales.
@@
 \subsection*{AGENT-BASED WHALE MOVEMENT MODEL}
-We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent-based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model\cite{magozzi2017using,yool2013medusa}, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA\cite{yool2013medusa}, and bathymetry from\cite{bathy} and extracted using the marmap R pakckage\cite{marmap} (Figure \ref{figs5}).
+We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent-based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model \citep{magozzi2017using,yool2013medusa}, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA \citep{yool2013medusa}, and bathymetry from \cite{bathy} and extracted using the marmap R pakckage \citep{marmap} (Figure \ref{figs5}).
 
@@
 \newpage
-\bibliographystyle{nature}
+\bibliographystyle{refs}
 \bibliography{blue-whale-abbreviations}

### 2019-01-17 manuscript/PeerJ/blue-new.tex
Update to author year citations

@@
 \usepackage{ms}            % load the template
-%\usepackage[round]{natbib} % author-year citations
+\usepackage[round]{natbib} % author-year citations
 \usepackage{graphicx}
@@
 
-Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}. 
-Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle \cite{robinson2009travelling}.
+Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites \citep{runge2014conserving}. 
+Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle \citep{robinson2009travelling}.
 Identifying threats to migratory species, understanding species responses to change and developing effective conservation measures all require information on the movements of individual animals over multiple years, ideally for both historic and present-day populations.
-With the development of electronic tagging technology, studies of the distributions of animals have largely shifted from infrequent observations of many individuals with limited or no information about individual movement behavior, to frequent observations of a few individuals, with detailed information about individual movement\cite{holdo2013inferring}.
-Despite the tremendous advances made using direct telemetry devices, tags are expensive, with a relatively high failure rate \cite{bailey2009behavioural,best2015tag,mate2007evolution}. 
-Data on individual-level, multi-annual movements remain scarce, especially for rare, wide-ranging and long-lived marine species such as baleen whales (Mysticeti)\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.
+With the development of electronic tagging technology, studies of the distributions of animals have largely shifted from infrequent observations of many individuals with limited or no information about individual movement behavior, to frequent observations of a few individuals, with detailed information about individual movement \citep{holdo2013inferring}.
+Despite the tremendous advances made using direct telemetry devices, tags are expensive, with a relatively high failure rate \citep{bailey2009behavioural,best2015tag,mate2007evolution}. 
+Data on individual-level, multi-annual movements remain scarce, especially for rare, wide-ranging and long-lived marine species such as baleen whales (Mysticeti) \citep{ryan2013stable,hall2005stable,bailey2009behavioural}.
 
-An alternative technique for investigating animal movements retrospectively is to use intrinsic biochemical information such as stable isotope compositions \cite{west2006stable,busquets2017estimating,hobson2008tracking}.
-Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioral information that are often under-utilised\cite{lister2011natural}.
+An alternative technique for investigating animal movements retrospectively is to use intrinsic biochemical information such as stable isotope compositions \citep{west2006stable,busquets2017estimating,hobson2008tracking}.
+Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioral information that are often under-utilised \citep{lister2011natural}.
 The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
-A large literature describes the use of stable isotope markers to link the composition of a tissue to a geographic source at a single point in time\cite{hobson2008tracking}. 
-Relating the isotopic compositions of marine animal tissues to the likely location of tissue growth is complicated by a lack of knowledge of spatial variations in the isotopic composition of diet (the isotopic baseline)\cite{west2006stable,mcmahon2015millennial}.
+A large literature describes the use of stable isotope markers to link the composition of a tissue to a geographic source at a single point in time \citep{hobson2008tracking}. 
+Relating the isotopic compositions of marine animal tissues to the likely location of tissue growth is complicated by a lack of knowledge of spatial variations in the isotopic composition of diet (the isotopic baseline) \citep{west2006stable,mcmahon2015millennial}.
 This uncertainty is compounded when multiple samples are taken through time in the same individual, as temporal variation in the isotopic baseline must also be considered explicitly. 
@@
 
-Here we introduce a new approach to infer location from the isotopic compositions of sequentially-sampled incrementally-grown tissues by drawing on newly developed models predicting spatio-temporal variation in phytoplankton stable carbon isotope composition at global and monthly resolution\cite{magozzi2017using} (Fig.
+Here we introduce a new approach to infer location from the isotopic compositions of sequentially-sampled incrementally-grown tissues by drawing on newly developed models predicting spatio-temporal variation in phytoplankton stable carbon isotope composition at global and monthly resolution \citep{magozzi2017using} (Fig.
 S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig.
@@
 Mysticete whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater. 
-Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable,hobson1998stable}. 
+Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert \citep{best1996stable,hobson1998stable}. 
 Baleen plates therefore offer a continuous isotopic record of behavior typically reflecting multiple years of life of an individual whale. 
 Baleen is worn away at the tips over time, so a baleen plate reflects the most recent years of life, and rarely records an individual's entire lifespan. 
-Among the mysticete whales, blue whales are a particularly attractive target for model-based isotope movement work as they have a consistent, low trophic level diet and feed continuously through the year, likely driven by high energetic costs of maintaining extreme muscle mass\cite{goldbogen2015}.
+Among the mysticete whales, blue whales are a particularly attractive target for model-based isotope movement work as they have a consistent, low trophic level diet and feed continuously through the year, likely driven by high energetic costs of maintaining extreme muscle mass \citep{goldbogen2015}.
 
-Similar to other large balaenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes\cite{huckgaete2018}. 
-As blue whales feed throughout the year to accommodate energetic costs, migration routes and feeding areas for blue whales are thought to be shaped by the year-round location of highly productive regions\cite{branch2007}.
+Similar to other large balaenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes \citep{huckgaete2018}. 
+As blue whales feed throughout the year to accommodate energetic costs, migration routes and feeding areas for blue whales are thought to be shaped by the year-round location of highly productive regions \citep{branch2007}.
 Recent studies tracking movements of individual blue whales primarily in the northeast Pacific have demonstrated high levels of among-individual variation in movement histories including suspending migration for opportunistic feeding and skipped migration (i.e.
 year round residency in either summer or winter
-areas) \cite{busquets2017estimating}. 
-The northeast Pacific population of blue whales is perhaps the best studied in terms of individual-level movements with a recent history of deployment of over 177 satellite tags \cite{irvine2017quantifying, mate2007evolution}.
-In the northeast Pacific, satellite tags show individuals occupying most of the exclusive economic zone (EEZ) and into the Gulf of Alaska \cite{irvine2017quantifying} with greatest overlapping areas in the southern half of the EEZ, along the continental slope and shelf region between the California Channel Islands and Cape Mendocino.
+areas) \citep{busquets2017estimating}. 
+The northeast Pacific population of blue whales is perhaps the best studied in terms of individual-level movements with a recent history of deployment of over 177 satellite tags \citep{irvine2017quantifying, mate2007evolution}.
+In the northeast Pacific, satellite tags show individuals occupying most of the exclusive economic zone (EEZ) and into the Gulf of Alaska \citep{irvine2017quantifying} with greatest overlapping areas in the southern half of the EEZ, along the continental slope and shelf region between the California Channel Islands and Cape Mendocino.
 Winter feeding is less well characterized, but winter feeding grounds are known
 in the Gulf of California and in the eastern tropical Pacific, specifically the Costa Rica Dome where production is maintained by seasonal upwelling.
-However, most satellite deployments on blue whales report movement data for periods of time under six months \cite{heide2001new, silva2013north, bailey2009behavioural, lesage2017foraging, irvine2017quantifying}, and are unable to show individual movements between summer and winter feeding or breeding areas.
-The longest period of continuous monitoring via satellite tagging of an individual blue whale that we are aware of in the northeast Pacific is 504 days \cite{irvine2017quantifying} and in the North Atlantic is 177 days \cite{lesage2017foraging}.
+However, most satellite deployments on blue whales report movement data for periods of time under six months \citep{heide2001new, silva2013north, bailey2009behavioural, lesage2017foraging, irvine2017quantifying}, and are unable to show individual movements between summer and winter feeding or breeding areas.
+The longest period of continuous monitoring via satellite tagging of an individual blue whale that we are aware of in the northeast Pacific is 504 days \citep{irvine2017quantifying} and in the North Atlantic is 177 days \citep{lesage2017foraging}.
 
-The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now around 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland \cite{pike2009note}.
+The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now around 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland \citep{pike2009note}.
 The small population size and oceanic habit makes blue whale movements extremely hard to study. 
-A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic also track regions of high production throughout the year, at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands \cite{baines2014upwellings}.
-Northward migration in the spring may occur along mid Atlantic corridors with peak sightings in the Azores around April-May \cite{silva2013north}. 
-Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas \cite{pike2009note}. 
-Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals \cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
-To our knowledge one photographic identification has matched summer and winter locations of a single north-east Atlantic blue whale with sightings in Iceland and Mauritania \cite{poster}.
+A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic also track regions of high production throughout the year, at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands \citep{baines2014upwellings}.
+Northward migration in the spring may occur along mid Atlantic corridors with peak sightings in the Azores around April-May \citep{silva2013north}. 
+Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas \citep{pike2009note}. 
+Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals \citep{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
+To our knowledge one photographic identification has matched summer and winter locations of a single north-east Atlantic blue whale with sightings in Iceland and Mauritania \citep{poster}.
 
@@
 Here we infer individual level high-resolution movement records for seven blue whales for which baleen stable isotope data were available.
-Isotope data for six whales stranding on the west coast of the USA or Mexico were reported in Busquets-Vass \textit{et al}.\cite{busquets2017estimating}, and we report new isotopic data from a whale stranding off the coast of Wexford Ireland in March 1891, and currently on display in the Natural History Museum, London (NHM).
+Isotope data for six whales stranding on the west coast of the USA or Mexico were reported in \cite{busquets2017estimating}, and we report new isotopic data from a whale stranding off the coast of Wexford Ireland in March 1891, and currently on display in the Natural History Museum, London (NHM).
 The NHM whale ``Hope'' is a female estimated to be a young adult at least 15 years at death. 
@@
 97 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal (gingival) section that contains the most recent tissue. 
-Baleen grows at a constant rate, so the samples are equally spaced through time\cite{best1996stable}. 
+Baleen grows at a constant rate, so the samples are equally spaced through time \citep{best1996stable}. 
 Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the University of Southampton SEAPORT Stable Isotope Ratio Mass Spectrometry Laboratory (Southampton, UK), using a Vario Isotope select elemental analyser, coupled to an Isoprime 100 isotope mass spectrometer.
 Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
-C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples\cite{hobson1998stable}.
-All data are available from the NHM Data Portal\cite{data-set} (https://doi.org/10.5519/0093278).
+C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples \citep{hobson1998stable}.
+All data are available from the NHM Data Portal \citep{data-set} (https://doi.org/10.5519/0093278).
 
@@
 
-Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates\cite{hobson1998stable,busquets2017estimating}.
+Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates \citep{hobson1998stable,busquets2017estimating}.
 Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
-We calculated isotopic periodicity within the NHM baleen sample using Fourier Transform analysis\cite{cardona2017temporal} (Supplementary Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated by Busquets-Vass \textit{et al.}\cite{busquets2017estimating}.
+We calculated isotopic periodicity within the NHM baleen sample using Fourier Transform analysis \citep{cardona2017temporal} (Supplementary Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated by \cite{busquets2017estimating}.
 Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891.
@@
 
-Isotope-enabled biogeochemical ocean models \cite{magozzi2017using,schmittner2016complementary} were used to characterize
+Isotope-enabled biogeochemical ocean models \citep{magozzi2017using,schmittner2016complementary} were used to characterize
 the isotopic composition of phytoplankton expected in different
@@
 S3).
-\(\delta^{13}\)C POM values were simulated at 1\({}^{\circ}\) and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model \cite{magozzi2017using,yool2013medusa}.
+\(\delta^{13}\)C POM values were simulated at 1\({}^{\circ}\) and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model \citep{magozzi2017using,yool2013medusa}.
 Simulated \(\delta^{15}\)N POM values are relatively positive in the northeast Atlantic north of c. 60\({}^{\circ}\)N, and relatively negative in the central and southern North Atlantic. 
@@
 In the central North Atlantic, \(\delta^{13}\)C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S3). 
-The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \cite{magozzi2017using,laws1995dependence}.
+The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \citep{magozzi2017using,laws1995dependence}.
 Thus temporal variations in \(\delta^{13}\)C POM values are superimposed on latitudinal gradients. 
 The scale and nature of temporal variation in \(\delta^{13}\)C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in \(\delta^{13}\)C POM values linked to strongly seasonal phytoplankton growth dynamics.
-The model simulations used are forced with decadal climatological data from 2000-2010\cite{magozzi2017using}. 
+The model simulations used are forced with decadal climatological data from 2000-2010 \citep{magozzi2017using}. 
 We do not draw inferences about location from absolute \(\delta^{13}\)C values and thus the reduction in oceanic \(\delta^{13}\)C POM values caused by release of fossil carbon into the atmosphere (Suess effect) does not influence inferences about location or movement.
-Ocean basin scale spatio-temporal differences in \(\delta^{13}\)C values largely reflect relative differences in sea surface temperatures between trophic, temperate and arctic waters, therefore we are confident that the Magozzi \textit{et al}.\cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.
+Ocean basin scale spatio-temporal differences in \(\delta^{13}\)C values largely reflect relative differences in sea surface temperatures between trophic, temperate and arctic waters, therefore we are confident that the \cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.
 
@@
 The degree of such trophic fractionation is unclear, however and we do not draw interpretations based on absolute \(\delta^{13}\)C values, rather on the relative \(\delta^{13}\)C values across the length of the baleen plate.
-We assume opportunistic ingestion of protein throughout the year \cite{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating, lesage2017foraging, bailey2009behavioural, branch2007, huckgaete2018}.
+We assume opportunistic ingestion of protein throughout the year \citep{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating, lesage2017foraging, bailey2009behavioural, branch2007, huckgaete2018}.
 
@@
 We simulated whale movements with the likelihood, direction, and extent of movement influenced by behavioral state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability).
-Movement was coded as a set of probabilistic rules, informed by the literature on blue whale behavior (e.g. \cite{handbook}).
+Movement was coded as a set of probabilistic rules, informed by the literature on blue whale behavior (e.g. \citep{handbook}).
 All terms were expressed as probability distributions, yielding multiple potential movement tracks.
@@
 Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
-(ii) Sea surface temperature \cite{yool2013medusa} ($^{\circ}$C). 
+(ii) Sea surface temperature \citep{yool2013medusa} ($^{\circ}$C). 
 When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.
-(iii) Water depth (m; from the General Bathymetric Chart of the Oceans (GEBCO) global bathymetry dataset\cite{bathy}, extracted via the marmap package in R \cite{marmap}). 
+(iii) Water depth (m; from the General Bathymetric Chart of the Oceans (GEBCO) global bathymetry dataset\citep{bathy}, extracted via the marmap package in R \citep{marmap}). 
 Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
-(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities \cite{yool2013medusa}). 
+(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities \citep{yool2013medusa}). 
 This was included as a proxy for zooplankton food availability. 
@@
 
-For the historic North Atlantic whale, we simulated the isotopicexpression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
+For the historic North Atlantic whale, we simulated the isotopicexpression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \citep{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
 Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four
@@
 
-At each time point and location in the simulations, we extracted time averaged \(\delta^{13}\)C values from the spatio-temporal model\cite{magozzi2017using}resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs X, \ref{fig2}, and S4).
+At each time point and location in the simulations, we extracted time averaged \(\delta^{13}\)C values from the spatio-temporal model \citep{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs X, \ref{fig2}, and S4).
 We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).
 
-R code for all analyses is available from GitHub (https://github.com/nhcooper123/blue-whale-bes; \cite{github})
+R code for all analyses is available from GitHub (https://github.com/nhcooper123/blue-whale-bes; \citep{github})
 
@@
 
-Estimates of baleen growth rates for the six modern whales from the northeast Pacific were inferred from cyclical fluctuations in nitrogen isotope ratios and reported in Busquets-Vass \textit{et al}\cite{busquets2017estimating}.
+Estimates of baleen growth rates for the six modern whales from the northeast Pacific were inferred from cyclical fluctuations in nitrogen isotope ratios and reported in \cite{busquets2017estimating}.
 We used these growth rates to reconstruct a potential timeline for baleen isotope records.
@@
 We were unable to find convincing fits to baleen profiles for whales B and E suggesting that longer simulations with a greater variation in movements were needed. 
-The inferred movement records are consistent with the known spatial distributions and relatively high level of individual variation in spatial distributions among blue whales in the Pacific northwest \cite{irvine2017quantifying, mate2007evolution}.
+The inferred movement records are consistent with the known spatial distributions and relatively high level of individual variation in spatial distributions among blue whales in the Pacific northwest \citep{irvine2017quantifying, mate2007evolution}.
 
@@
 Our simulations identify a range of possible locations for the whale (Fig.
-\ref{fig3}), although areas around the Mauritanian coast and Cape Verde Islands, a known current and historic winter feeding area for blue whales \cite{baines2014upwellings,reeves2004historical}, and potentially to the west of the Azores (Figure \ref{fig2}), most closely match the measured profile.
+\ref{fig3}), although areas around the Mauritanian coast and Cape Verde Islands, a known current and historic winter feeding area for blue whales \citep{baines2014upwellings,reeves2004historical}, and potentially to the west of the Azores (Figure \ref{fig2}), most closely match the measured profile.
 The whale remained in these warm waters for at least one full year. 
@@
 During behavioral phase two, the observed low \(\delta^{13}\)C values imply foraging in colder, more northerly latitudes. 
-The pronounced cyclical variations in \(\delta^{13}\)C values observed during behavioral phase two could reflect either the isotopic expression of the spring phytoplankton bloom in northern waters\cite{magozzi2017using}, latitudinal migrations, or a combination of both.
+The pronounced cyclical variations in \(\delta^{13}\)C values observed during behavioral phase two could reflect either the isotopic expression of the spring phytoplankton bloom in northern waters \citep{magozzi2017using}, latitudinal migrations, or a combination of both.
 
@@
 Better-fitting models in general were those predicting a greater latitudinal foraging range, and foraging in more northerly waters (Fig. S7 and S8). 
-Best-fit model distributions are largely consistent with current understanding of blue whale distributions in the northeast Atlantic \cite{reeves2004historical,baines2014upwellings,baines2017autumn,reeves2004historical} with perhaps greater importance of winter foraging in temperate regions (Fig.
+Best-fit model distributions are largely consistent with current understanding of blue whale distributions in the northeast Atlantic \citep{reeves2004historical,baines2014upwellings,baines2017autumn,reeves2004historical} with perhaps greater importance of winter foraging in temperate regions (Fig.
 \ref{fig4}).
@@
 After c.4-6 months the whale began ingesting food with relatively high \(\delta^{13}\)C values, indicating feeding in low latitude waters, shortly followed by a final northward migration.
-Blue whales have a 10-12 month gestation period, with calving occurring in subtropical waters, and calves are weaned after 6-7 months \cite{handbook}.
+Blue whales have a 10-12 month gestation period, with calving occurring in subtropical waters, and calves are weaned after 6-7 months \citep{handbook}.
 We therefore infer a movement chronology reflecting three years of uninterrupted annual latitudinal migrations leading to pregnancy in the
@@
 analysing. 
-Before whaling in the North Atlantic began in 1868 \cite{reilly2008balaenoptera}, there were and estimated 10,000-15,000 blue whales in the region \cite{sigurjonsson1995life}. 
-In the early 20th century, fisheries moved outside the area because stocks were so depleted \cite{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed \cite{sigurjonsson1995life}.
+Before whaling in the North Atlantic began in 1868 \citep{reilly2008balaenoptera}, there were and estimated 10,000-15,000 blue whales in the region \citep{sigurjonsson1995life}. 
+In the early 20th century, fisheries moved outside the area because stocks were so depleted \citep{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed \citep{sigurjonsson1995life}.
 The NHM blue whale was thus an individual from a species at the brink of local extinction, potentially representing more than 0.1\% of the entire population of blue whales in the northeast Atlantic.
-Inferring spatio-temporal distributions of populations from high density observations of a few individuals (or, in this case one individual) will always be problematic, but the predictive power linking a tracked individual to the population distribution increases when the tracked animal has a strong preference for specific habitats and there is a patchy (clumped) distribution of those habitats \cite{holdo2013inferring}.
+Inferring spatio-temporal distributions of populations from high density observations of a few individuals (or, in this case one individual) will always be problematic, but the predictive power linking a tracked individual to the population distribution increases when the tracked animal has a strong preference for specific habitats and there is a patchy (clumped) distribution of those habitats \citep{holdo2013inferring}.
 In the case of blue whales, the primary driver influencing seasonal movements is assumed to be the seasonally variable distribution of food resources between high and low latitudes. 
@@
 
-Our movement models suggest that the upwelling regions around Mauritania and the Cape Verde Islands may have been important historic calving areas, consistent with modern observations of juvenile blue whales in the northeast Atlantic \cite{handbook}.
+Our movement models suggest that the upwelling regions around Mauritania and the Cape Verde Islands may have been important historic calving areas, consistent with modern observations of juvenile blue whales in the northeast Atlantic \citep{handbook}.
 These areas are now fishing, tourist and shipping hotspots with emerging offshore oil and gas exploration, potentially increasing threats to the recovery of the small population of blue whales in the northeast Atlantic.
 Although hunting for blue whales has been banned for over 50 years, the species is still listed as Endangered on the IUCN Red List, and expected recoveries following the cessation of active hunting may be impeded by other human activities such as modifications to the food web or more direct impacts of noise and shipping.
-Developing an understanding of the nature and plasticity of individual level movements in populations of blue (and other mysticete) whales would aid management decision making in the conservation of this endangered species \cite{irvine2017quantifying}.
+Developing an understanding of the nature and plasticity of individual level movements in populations of blue (and other mysticete) whales would aid management decision making in the conservation of this endangered species \citep{irvine2017quantifying}.
 

### 2019-02-14 manuscript/PeerJ/blue-new.tex
Edits from CT

@@
 Incrementally-grown tissues provide an alternative means of reconstructing individual-level movements over long timescales, but inferring spatio-temporal information from biochemical tracers is challenging.
-Here we present a new approach combining stable isotope analyses of incrementally-grown tissues with simulation models to infer individual-level animal movements. 
-We use our new method to reconstruct multiple years of movement histories of individual modern and historic blue whales, including an iconic blue whale which stranded in 1891 and is now on display at the Natural History Museum, London.
-Our method unlocks the information recorded in historic collections, and provides inferences that are complimentary to and comparable with tag-derived movement data.
+Here we outline a new approach combining stable isotope analyses of incrementally-grown tissues with simulation models to test hypotheses relating to individual-level animal movements. 
+We illustrate our new method by identifying the most likely multi-year movement history of a historic blue whale that stranded in 1891 and is now on display at the Natural History Museum, London. 
+Our method unlocks movement information coded in biochemical compositions of incremental tissues including those housed in historic collections, and provides inferences that are complimentary to and comparable with tag-derived movement data.
 
@@
 This uncertainty is compounded when multiple samples are taken through time in the same individual, as temporal variation in the isotopic baseline must also be considered explicitly. 
-Consequently, relatively little attention has been given to the potential to infer individual movement histories by reconstructing time series of multiple discrete origins from sequentially sampled incrementally-grown tissues.
+Consequently, relatively little attention has been given to the potential to infer individual movement histories by reconstructing time series of multiple discrete origins from sequentially sampled incrementally-grown tissues
+% [trueman st john glew, Tokyo eel, darnaude, brennan?].
 
-Here we introduce a new approach to infer location from the isotopic compositions of sequentially-sampled incrementally-grown tissues by drawing on newly developed models predicting spatio-temporal variation in phytoplankton stable carbon isotope composition at global and monthly resolution \citep{magozzi2017using} (Fig.
-S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig.
-S5).
-We use our coupled models to simulate time series of carbon stable isotope compositions expected under different movement behaviours. 
-We then compare simulated isotopic profiles with the measured profiles to identify potential movement histories most consistent with the measured isotopic profile.
+Recently, mechanistic models have been developed providing relatively accurate predictions of isotopic variability at high temporal and spatial resolution. 
+These models can be coupled to Lagrangian or agent-based models of animal movement to predict isotopic compositions or trajectories expected to be recorded in animal tissues under differing movement scenarios.
+Here we illustrate the potential of this conceptual approach by assessing the relative likelihood of differing, plausible hypotheses of movement behavior for a single individual blue whale based on carbon isotope compositions sampled across a baleen plate.
+We draw on newly developed models predicting spatio-temporal variation in phytoplankton stable carbon isotope composition at global and monthly resolution \citep{magozzi2017using} (Figure S3), coupled to an agent-based model of whale movements (see Table S1; Figure S5).
+We specifically ask whether the isotopic profile observed in baleen can be simulated using agent based movement models informed by current understanding of animal movements and isotopic variability.
 
-We apply our method to blue whales (\textit{Balaenoptera musculus}), the largest animal to have ever lived. 
+We apply our method a blue whales (\textit{Balaenoptera musculus}), the largest animal to have ever lived. 
 Mysticete whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater. 
@@
 areas) \citep{busquets2017estimating}. 
-The northeast Pacific population of blue whales is perhaps the best studied in terms of individual-level movements with a recent history of deployment of over 177 satellite tags \citep{irvine2017quantifying, mate2007evolution}.
-In the northeast Pacific, satellite tags show individuals occupying most of the exclusive economic zone (EEZ) and into the Gulf of Alaska \citep{irvine2017quantifying} with greatest overlapping areas in the southern half of the EEZ, along the continental slope and shelf region between the California Channel Islands and Cape Mendocino.
-Winter feeding is less well characterized, but winter feeding grounds are known
-in the Gulf of California and in the eastern tropical Pacific, specifically the Costa Rica Dome where production is maintained by seasonal upwelling.
 However, most satellite deployments on blue whales report movement data for periods of time under six months \citep{heide2001new, silva2013north, bailey2009behavioural, lesage2017foraging, irvine2017quantifying}, and are unable to show individual movements between summer and winter feeding or breeding areas.
@@
 
-Here we infer individual level high-resolution movement records for seven blue whales for which baleen stable isotope data were available.
-Isotope data for six whales stranding on the west coast of the USA or Mexico were reported in \cite{busquets2017estimating}, and we report new isotopic data from a whale stranding off the coast of Wexford Ireland in March 1891, and currently on display in the Natural History Museum, London (NHM).
+Here we aim to show that stable carbon isotope tracers can be used to test hypotheses about individual-level movement behavior, and can be applied to recent or historic archived samples. 
+We apply our method to infer the most likely movement history for a blue whale stranding off the coast of Wexford Ireland in March 1891, and currently on display in the Natural History Museum, London (NHM).
 The NHM whale ``Hope'' is a female estimated to be a young adult at least 15 years at death. 
-Baleen from the NHM whale therefore yields information about migratory behaviors in North Atlantic blue whales around the peak of industrial whaling.
+Baleen from the NHM whale therefore yields information about about a single blue whale living in the North Atlantic around the peak of industrial whaling.
 
@@
 
-\subsection{Stable isotope extractions from baleen}\label{stable-isotope-extractions-from-baleen}
+\subsection{Stable isotope extractions from baleen}
+\label{stable-isotope-extractions-from-baleen}
 
@@
 
-\subsection{Time calibrating stable isotope
-profiles}\label{time-calibrating-stable-isotope-profiles}
+\subsection{Time calibrating stable isotope profiles}
+\label{time-calibrating-stable-isotope-profiles}
 
-Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates \citep{hobson1998stable,busquets2017estimating}.
+Seasonal variations in isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates \citep{hobson1998stable,busquets2017estimating}.
 Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
-We calculated isotopic periodicity within the NHM baleen sample using Fourier Transform analysis \citep{cardona2017temporal} (Supplementary Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated by \cite{busquets2017estimating}.
+We calculated isotopic periodicity within the NHM baleen sample using Fourier Transform analysis \citep{cardona2017temporal} (Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated by \cite{busquets2017estimating}.
 Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891.
 
-\subsection{Baseline isotope
-comparisons}\label{baseline-isotope-comparisons}
+\subsection{Baseline isotope comparisons}
+\label{baseline-isotope-comparisons}
 
-Isotope-enabled biogeochemical ocean models \citep{magozzi2017using,schmittner2016complementary} were used to characterize
-the isotopic composition of phytoplankton expected in different
-potential foraging grounds (Fig. S3). 
-Annual average \(\delta^{15}\)N POM (particulate organic matter) values were provided by C.J.
-Somes (\textit{pers.comm}) based on a 5\({}^{\circ}\) resolution biogeochemical model (Fig.
-S3).
+Isotope-enabled biogeochemical ocean models \citep{magozzi2017using,schmittner2016complementary} were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Figure S3).
+Annual average \(\delta^{15}\)N POM (particulate organic matter) values were provided by C.J. Somes (\textit{pers.comm}) based on a 5\({}^{\circ}\) resolution biogeochemical model (Figure S3). 
 \(\delta^{13}\)C POM values were simulated at 1\({}^{\circ}\) and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model \citep{magozzi2017using,yool2013medusa}. 
@@
 Annual average \(\delta^{13}\)C POM values largely vary with latitude, with more negative values in more northerly regions. 
-In the central North Atlantic, \(\delta^{13}\)C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S3). 
+In the central North Atlantic, \(\delta^{13}\)C POM values are relatively positive in the west, reflecting warm gulf stream waters (Figure S3). 
 The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \citep{magozzi2017using,laws1995dependence}.
@@
 The model simulations used are forced with decadal climatological data from 2000-2010 \citep{magozzi2017using}.
-We do not draw inferences about location from absolute \(\delta^{13}\)C values and thus the reduction in oceanic \(\delta^{13}\)C POM values caused by release of fossil carbon into the atmosphere (Suess effect) does not influence inferences about location or movement.
-Ocean basin scale spatio-temporal differences in \(\delta^{13}\)C values largely reflect relative differences in sea surface temperatures between trophic, temperate and arctic waters, therefore we are confident that the \cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.
+
+The isotopic composition of phytoplankton is influenced by the release of fossil carbon with low \(\delta^{13}\)C values into the atmosphere (Suess effect) and the effect of increased concentrations of dissolved CO2 on isotopic fractionation during phytoplankton growth
+%(Young et al 2013)
+, potentially complicating the application of a model trained on recent data to historic samples. 
+Ocean basin scale spatio-temporal differences in \(\delta^{13}\)C values largely reflect relative differences in sea surface temperatures between trophic, temperate and arctic waters and over seasons, and we assume that any regional variation in isotopic effects due to CO\textsubscript{2}-influenced changes in phytoplankton composition and growth rate are minor compared to latitudinal and seasonal variations in \(\delta^{13}\)C POM values.
+We do not draw inferences about location from absolute \(\delta^{13}\)C values and thus the reduction in oceanic \(\delta^{13}\)C POM values does not directly influence inferences about location or movement.
+We are therefore confident that the \cite{magozzi2017using} model can be applied to interpret historic baleen isotope data at least at ocean-basin scale spatial resolution.
 
@@
 Carbon isotope values are also likely to be fractionated during transfer from plankton to krill as \textsuperscript{12}C is preferentially lost through respiration. 
-The degree of such trophic fractionation is unclear, however and we do not draw interpretations based on absolute \(\delta^{13}\)C values, rather on the relative \(\delta^{13}\)C values across the length of the baleen plate.
-We assume opportunistic ingestion of protein throughout the year \citep{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating, lesage2017foraging, bailey2009behavioural, branch2007, huckgaete2018}.
+The degree of such trophic fractionation is unclear, however and as we do not draw interpretations based on absolute \(\delta^{13}\)C values, rather on the relative \(\delta^{13}\)C values across the length of the baleen plate, we do not need to quantify this trophic enrichment effect.
+We assume that carbon available for synthesis of keratin is drawn from carbon released by respiration of diet captured opportunistically throughout the year \citep{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating, lesage2017foraging, bailey2009behavioural, branch2007, huckgaete2018}.
 
-\subsection{Agent-based whale movement
-model}\label{agent-based-whale-movement-model}
+\subsection{Agent-based whale movement model}
+\label{agent-based-whale-movement-model}
+
+We simulated the likely isotopic expression associated with different movement behaviours by building an agent-based movement model within the R coding environment, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA \citep{yool2013medusa}, and bathymetry from the General Bathymetric Chart of the Oceans (GEBCO) \citep{bathy}, extracted via the marmap package in R \citep{marmap} (Figure S5).
+% check figure S5
 
@@
 
-In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and linear distance of movement are all influenced by the following. 
-(i) Behavioural state (migrating north, migrating south, or foraging).
-Northerly migrations were possible only in spring, and southerly migrations in autumn. 
+In the models, the likelihood of moving, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and linear distance of movement are all influenced by the following. 
+(i) Behavioural state (migrating north, migrating south, or foraging, fixed according to month as defined by the operator).
+Directly, northerly migrations were coded to occur in spring, and southerly migrations in autumn.
 Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
@@
 Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state. 
+At each daily step, the probability of movement, movement direction and distance traveled are sampled from probability distributions to allow individual variation.
 
-For Pacific whales where satellite data offers a rich view of individual movements, we coded our movement models to reflect either residency in southern waters (between Baja California and the Costa Rica Dome) or seasonal latitudinal movements within the California current system.
-We simulated 250 movement tracks of four years duration. 
-We allowed start and end dates to vary by aligning measured profiles to sections of simulated profiles of identical lengths with sequentially changing start dates and saved best fitting models.
+Initial boundary conditions are defined with a maximum temperature of 25$^{\circ}$C and minimum temperature of 3$^{\circ}$C. 
+The likelihood of movement (i.e.
+whether to move or not from the current location) is sampled from a binomial distribution with the probability of movement influenced by behavioural state and external conditions.
+The maximum daily movement distance permitted in each behavioural state is defined as a random sample of a Gaussian distribution with specified mean and standard deviation (see Table S1).
 
-For the historic North Atlantic whale, we simulated the isotopicexpression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \citep{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
-Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four
-years of simulations. 
-For a full description of the model and its parameters see Supplementary Methods: Fig. S5 and Table S1. 
-We simulated seven years of whale movements 1000 times and two years of residency 30 times for each residency hotspot (Fig. S4).
-We simulated whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.
+The agent-based mode provides daily location data points which are then used to extract time averaged \(\delta^{13}\)C values from the isotopic extension to the NEMO-MEDUSA model and assembled to build simulated isotopic profiles across baleen plates.
+We created experimental movement models by manipulating the relative duration of the foraging and migratory behavioural states to produce simulated records of isotopic composition of carbon in baleen expected under differing movement trajectories.
+We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \citep{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
 
-At each time point and location in the simulations, we extracted time averaged \(\delta^{13}\)C values from the spatio-temporal model \citep{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs X, \ref{fig2}, and S4).
-We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).
+We simulated two years of residency 30 times for each residency hotspot (Figure S4). 
+We simulated seven years of whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations. 
+We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile measured in the blue whale baleen (Figure \ref{fig1}) with simple linear regressions.
 
+For a full description of the model and its parameters see Figure S5 and Table S1. 
 R code for all analyses is available from GitHub (https://github.com/nhcooper123/blue-whale-bes; \citep{github})
 
-\section{Results and Discussion}
+\section{Results}
 
-\subsection{Matching model to Pacific data.}
+\subsection{Baleen stable isotope profiles}
+The baleen plate yielded 97 discrete samples of baleen.
+\(\delta^{15}\)N values measured in the baleen plate display regular cyclical fluctuations throughout the length of the plate.
+Fourier analysis (periodogram function in the R package TSA \citep{Chan:2012aa}) revealed a strong periodic repetition with a 13.3cm periodicity, with a mean spacing of 13.5cm, assumed to represent annual periodicity.
+Therefore, given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm y$^{-1}$, we reconstructed a timeline for \(\delta^{13}\)C and \(\delta^{15}\)N fluctuations in the baleen over seven full years of the whale’s life (early 1884 - spring 1891).
 
-Estimates of baleen growth rates for the six modern whales from the northeast Pacific were inferred from cyclical fluctuations in nitrogen isotope ratios and reported in \cite{busquets2017estimating}.
-We used these growth rates to reconstruct a potential timeline for baleen isotope records.
+\(\delta^{13}\)C values are relatively high and constant in the oldest (most distal) 35cm of the baleen plate, associated with a weakening of the periodic fluctuation seen in \(\delta^{15}\)N values, and an overall reduction in \(\delta^{15}\)N values.
+This is followed by a clear change towards repeated fluctuations in \(\delta^{13}\)C values in the middle 50cm of the baleen plate, with a similar periodicity of 13.5cm (Figure \ref{fig1}). 
+A second, distinct change in the pattern of \(\delta^{13}\)C values along the baleen plate occurs at around 84cm from the distal end, with a transition to uniform, relatively low \(\delta^{13}\)C values followed by an abrupt transition to positive \(\delta^{13}\)C values and a decline in \(\delta^{13}\)C values in the most recent (most gingival) 3cm of the record.
 
-% Pacific whales figure
-\begin{figure}
- \centering
-  \includegraphics[width = \linewidth]{figures/figure-Pacific-draft.eps}
-  \caption{Correlations among simulated $\delta^{13}$C from the best fitting migratory movement models (grey lines, right hand y-axis) and $\delta^{13}$C from baleen (black line, left hand y-axis) for six Pacific blue whales taken from \cite{busquets2017estimating}.
-  Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location, reflecting temporal integration of phytoplankton $\delta^{13}$C values within the food chain before ingestion by the whale as krill.
-  The end points of the simulations and empirical data have been aligned to coincide.
-}
-  \label{pacific}
-\end{figure}
-
-In four of the six cases, measured baleen carbon isotope profiles could be replicated from model simulations with relatively high precision (Fig. \ref{pacific}
-; best fit r\textsuperscript{2} values ranging from 0.43 to 0.74).
-Whales D and F showed similar carbon isotope records and were best fit by the same movement simulation. 
-We extracted monthly locations associated with the best fitting simulation models for each whale (Fig. \ref{pacific}).
-Best fitting simulation models for whales D and F imply a relatively southerly distribution with few locations north of Baja, California and relatively limited seasonal latitudinal migration. 
-The measured carbon isotope record for whale A was best fitted by simulations with a southerly distribution, but with an additional strong seasonal latitudinal migration, and with more northerly locations between April and October.
-Whale C was best fit by models with relatively northerly locations and strong seasonal latitudinal migration. 
-We were unable to find convincing fits to baleen profiles for whales B and E suggesting that longer simulations with a greater variation in movements were needed. 
-The inferred movement records are consistent with the known spatial distributions and relatively high level of individual variation in spatial distributions among blue whales in the Pacific northwest \citep{irvine2017quantifying, mate2007evolution}.
-
-\subsection{Atlantic whale baleen measured isotopic record}
-
-For Hope, the historic North Atlantic whale, \(\delta^{15}\)N values record regular cyclical fluctuations throughout the length of the baleen plate, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods).
-Therefore, given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm\(y^{- 1}\), we reconstructed a timeline for \(\delta^{13}\)C and \(\delta^{15}\)N fluctuations in the baleen over seven full years of the whale's life (early 1884 - spring 1891).
-The \(\delta^{13}\)C and \(\delta^{15}\)N profiles show two distinct phases in the whale's behaviour. 
-In behavioural phase one (from the start of the record to spring 1886), we find relatively stable, elevated \(\delta^{13}\)C values, and relatively low \(\delta^{15}\)N values (Fig.
-\ref{fig1}; Figs.
-S1 and S2).
-In behavioural phase two (summer 1886 to spring 1890) \(\delta^{15}\)N values are relatively elevated and \(\delta^{13}\)C values are relatively depleted with coincident cyclical fluctuations in both \(\delta^{13}\)C and \(\delta^{15}\)N values.
-In the last year of life the cyclical pattern is disrupted, with constant low \(\delta^{13}\)C values for approximately six months in the first half of 1890, before a rapid switch to relatively positive \(\delta^{13}\)C values in the second half of 1890.
-The final three months of the record show a progressive fall in \(\delta^{13}\)C values (Fig. \ref{fig1}; Figs. S1 and S2).
+The \(\delta^{13}\)C and \(\delta^{15}\)N profiles therefore divide the isotopic record into two distinct phases that we assume reflect changes in the whale’s behaviour. 
+In behavioural phase one (from the start of the record to spring 1886), we find relatively stable, elevated \(\delta^{13}\)C values, and relatively low \(\delta^{15}\)N values (Figure \ref{fig1}; Figures S1 and S2).
+In behavioural phase two (summer 1886 to spring 1890) \(\delta^{15}\)N values are relatively high and \(\delta^{13}\)C values are relatively low with coincident cyclical fluctuations in both \(\delta^{13}\)C and \(\delta^{15}\)N values.
+In the last year of life the cyclical pattern is disrupted, with constant low \(\delta^{13}\)C values for approximately six months in the first half of 1890, before a rapid switch to relatively high \(\delta^{13}\)C values in the second half of 1890.
+The final three months of the record show a progressive fall in \(\delta^{13}\)C values (Figure \ref{fig1}; Figures S1 and S2). 
+Cross-correlation analysis demonstrates a strong negative covariance between \(\delta^{13}\)C and \(\delta^{15}\)N values within behavioural phase two (Figure S2), but no relationship between \(\delta^{13}\)C and \(\delta^{15}\)N values exists during behavioural phase one.
 
@@
 Samples were taken longitudinally through the baleen plate (n = 97 samples from a single baleen plate for both isotopes).
-There is strong annual periodicity and cross-correlation (Supplemental Figures S1 and S2) in both isotopes.
+There is strong annual periodicity and cross-correlation (Figures S1 and S2) in both isotopes.
 The approximate relationship to years assuming a growth rate of 13.5cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
@@
 \end{figure}
+
+\subsection{Agent-based whale movement model}
 
@@
 
-The relatively positive and seasonally-invariant \(\delta^{13}\)C values seen during behavioural phase one are only found in subtropical areas of the North Atlantic. 
-Our simulations identify a range of possible locations for the whale (Fig.
-\ref{fig3}), although areas around the Mauritanian coast and Cape Verde Islands, a known current and historic winter feeding area for blue whales \citep{baines2014upwellings,reeves2004historical}, and potentially to the west of the Azores (Figure \ref{fig2}), most closely match the measured profile.
-The whale remained in these warm waters for at least one full year. 
-Hope was estimated to be at least 15 years old when she died (based on vertebral epiphyseal fusion; R.C.
-Sabin \textit{pers.comm}), so was probably older than 10 years old, and therefore sexually mature, during this period.
-During behavioral phase two, the observed low \(\delta^{13}\)C values imply foraging in colder, more northerly latitudes. 
-The pronounced cyclical variations in \(\delta^{13}\)C values observed during behavioral phase two could reflect either the isotopic expression of the spring phytoplankton bloom in northern waters \citep{magozzi2017using}, latitudinal migrations, or a combination of both.
+We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions that would be encountered by whales foraging within broad geographic regions (west UK/Irish shelf, Norwegian/Barents Sea, Canaries/west Azores, Mid-Atlantic ridge, Cape Verde/Mauritanian upwelling area; Figure S4).
+Strong seasonal dynamics in $\delta^{13}$C values are evident in all northerly regions, characterized by a rapid increase to annual maximum $\delta^{13}$C values associated with the onset of the spring phytoplankton bloom, followed by a more gradual decline in $\delta^{13}$C values towards minima in winter conditions (Figure S4).
+In temperate latitudes around the British Isles, seasonal dynamic cycles are present, but dampened (Figure S4), whereas in sub-tropical regions exemplified by the Canaries, Azores and particularly Mauritanian upwelling areas, $\delta^{13}$C values are relatively high and constant throughout the year (Figure S4).
+
+Adding seasonal north-south migrations within mid-high latitude regions to foraging models yielded simulated profiles with regular isotopic fluctuations of relatively high amplitude reflecting $\delta^{13}$C minima preceding the June bloom (Figure S4, Figure \ref{fig1}).
+
+\subsection{Comparing the observed baleen isotopic profile to the model}
+In the measured profile, behavioural phase one is characterised by relatively high and constant $\delta^{13}$C values.
+$\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity is absent (Figure S1).  
+The relatively high and seasonally-invariant $\delta^{13}$C values seen during behavioural phase one are only found in subtropical areas of the North Atlantic. 
+Our simulations identify a range of possible locations for the whale (Figure \ref{fig2}), although areas around the Mauritanian coast and Cape Verde Islands, a known current and historic winter feeding area for blue whales \citep{baines2014upwellings,reeves2004historical}, and potentially to the west of the Azores (Figure \ref{fig2}), most closely match the measured profile.
+Temporal dynamics in $\delta^{13}$C values observed in the measured baleen during phase two cannot be reproduced in the simulated resident whales (Figure \ref{fig3}, Figure S4).
+During behavioral phase two, the observed low $\delta^{13}$C values imply foraging in colder, more northerly latitudes. 
+The pronounced cyclical variations in $\delta^{13}$C values observed during behavioral phase two could reflect either the isotopic expression of the spring phytoplankton bloom in northern waters \citep{magozzi2017using}, latitudinal migrations, or a combination of both.
+Accordingly we modeled whale movements allowing three years of residency in warm waters of the Azores or Cape Verde/Mauritania region followed by a further 3.5 years of seasonal north-south migration in the northeast northern Atlantic.
+We simulated 1200 individual movement patterns, then excluded simulations where the virtual whale stranded before reaching the 3019 days represented in the baleen.
+We then compared the remaining 1049 simulations and the resulting simulated baleen $\delta^{13}$C records, to the measured records with simple linear regressions. 
+Simulated baleen $\delta^{13}$C profiles produce a good fit to measured profiles, the median r\textsuperscript{2} value across 1049 simulated profiles was 0.31, and the maximum was 0.65 (Figure S6). 
+The top 10\% best fitting simulated profiles are shown in Figure \ref{fig2}. 
+Best fitting models predict residency in the Cape Verde region in behavioral phase one.
+Behavioral phase two is best simulated by seasonal migrations between summer foraging in northern areas (e.g.
+Norwegian Sea/Barents Sea/Iceland region), and winter foraging in a broad region between the UK and more southerly, subtropical waters.
+Better-fitting models in general were those predicting a greater latitudinal foraging range, and foraging in more northerly waters (Figures S7 and S8).
+Best-fit model distributions are largely consistent with current understanding of blue whale distributions in the northeast Atlantic \citep{reeves2004historical,baines2014upwellings,baines2017autumn,reeves2004historical} with perhaps greater importance of winter foraging in temperate regions (Figure \ref{fig4}).
 
@@
 
-We simulated 1200 individual movement patterns, then excluded simulations where the virtual whale stranded before reaching the 3019 days represented in the baleen.
-We then compared the remaining 1049 simulations and the resulting simulated baleen $\delta^{13}$C records, to the measured records with simple linear regressions. 
-Simulated baleen \(\delta^{13}\)C profiles produce a good fit to measured profiles, the median \(r^{2}\) value across 1049 simulated profiles was 0.31, and the maximum was 0.65 (Fig. S6). 
-The top 10\% best fitting simulated profiles are shown in Fig. \ref{fig2}. 
-Best fitting models predict residency in the Cape Verde region in behavioral phase one. 
-Behavioral phase two is best simulated by seasonal migrations between summer foraging in northern areas (e.g.
-Norwegian Sea/Barents Sea/Iceland region), and winter foraging in a broad region between the UK and more southerly, subtropical waters.
-Better-fitting models in general were those predicting a greater latitudinal foraging range, and foraging in more northerly waters (Fig. S7 and S8). 
-Best-fit model distributions are largely consistent with current understanding of blue whale distributions in the northeast Atlantic \citep{reeves2004historical,baines2014upwellings,baines2017autumn,reeves2004historical} with perhaps greater importance of winter foraging in temperate regions (Fig.
-\ref{fig4}).
-
 % figure 4
@@
 
+\subsection{Discussion}
+Long-term, multi-annual data on the movement patterns and reproductive ecology of individual migratory marine animals are scarce, and stable isotope data provide a promising source of additional indirect information regarding spatio-temporal behavior, but interpreting such profiles is difficult, particularly in marine environments, because of unknown temporal and spatial variation in isotopic baselines.
+
+Here we show how simulation models can be used to infer the isotopic expression expected from particular spatio-temporal foraging behaviours, and thus can be used used to interpret measured isotope data.
+We tested whether the isotopic variations observed across 6-7 years of growth of a baleen plate in a single blue whale could be simulated by agent based models coding for year round residency in known hotspots.
+We found that the isotopic record measured during behavioural phase one could be simulated through year round residency in relatively warm, subtropical waters with limited seasonal phytoplankton blooms.
+The measured profile is therefore consistent with residency in warm sub-tropical waters for at least one full year.
+Hope was estimated to be at least 15 years old when she died (based on vertebral epiphyseal fusion; R.C.
+Sabin pers.comm), so was probably older than 10 years old, and therefore sexually mature, during this period.
+
+The measured isotopic variations seen in behavioural phase two, characterized by repeated high amplitude cycles in $\delta^{13}$C values were best matched by movement models coded for annual latitudinal migrations with best fitting models predicting greater latitudinal foraging ranges.
+We therefore infer that Hope conducted at least three annual latitudinal migration cycles.
 The last 500 days of Hope's life are difficult to simulate. 
-Beginning in the winter of 1889/1890, the observed \(\delta^{13}\)C values are relatively low, and remain constant for c.4-6 months, before increasing rapidly in the second half of 1890. 
-Low \(\delta^{13}\)C values are found in northern waters, but these areas also show large temporal fluctuations in \(\delta^{13}\)C\textsubscript{plk} values (Fig.
-S4), and the observed values cannot be simulated purely from movements within the known geographic range.
-We suggest that the \(\delta^{13}\)C record in the last period of life may be associated with pregnancy and calf rearing. 
-Constant, low \(\delta^{13}\)C values may reflect later metabolic release of carbon reserves assimilated from northern latitudes (i.e. a period of time with limited opportunistic foraging). 
-After c.4-6 months the whale began ingesting food with relatively high \(\delta^{13}\)C values, indicating feeding in low latitude waters, shortly followed by a final northward migration.
+Beginning in the winter of 1889/1890, the observed $\delta^{13}$C values are relatively low, and remain constant for c.4-6 months, before increasing rapidly in the second half of 1890.
+Low $\delta^{13}$C values are found in northern waters, but these areas also show large temporal fluctuations in $\delta^{13}$Cplk values (Figure S4), and the observed values cannot be simulated purely from movements within the known geographic range.
+We tentatively suggest that the $\delta^{13}$C record in the last period of life may be associated with pregnancy and calf rearing. 
+Constant, low $\delta^{13}$C values may reflect later metabolic release of carbon reserves assimilated from northern latitudes (i.e. a period of time with limited opportunistic foraging). 
+After c.4-6 months the whale began ingesting food with relatively high $\delta^{13}$C values, indicating feeding in low latitude waters, shortly followed by a final northward migration. 
 Blue whales have a 10-12 month gestation period, with calving occurring in subtropical waters, and calves are weaned after 6-7 months \citep{handbook}.
-We therefore infer a movement chronology reflecting three years of uninterrupted annual latitudinal migrations leading to pregnancy in the
-year 1889 and birth in the winter of 1889/1890. 
+We therefore infer a movement chronology reflecting three years of uninterrupted annual latitudinal migrations leading to pregnancy in the year 1889 and birth in the winter of 1889/1890. 
 Following birth, we infer a period of c.4-6 months of residency in sub-tropical waters where the whale was sustained largely from stored lipid reserves, which we interpret as reflecting weaning. 
@@
 
-Whaling was an intense pressure for blue whales during the period we are
-analysing. 
+Whaling was an intense pressure for blue whales during the period we are analysing. 
 Before whaling in the North Atlantic began in 1868 \citep{reilly2008balaenoptera}, there were and estimated 10,000-15,000 blue whales in the region \citep{sigurjonsson1995life}. 
@@
 The NHM blue whale was thus an individual from a species at the brink of local extinction, potentially representing more than 0.1\% of the entire population of blue whales in the northeast Atlantic. 
-Inferring spatio-temporal distributions of populations from high density observations of a few individuals (or, in this case one individual) will always be problematic, but the predictive power linking a tracked individual to the population distribution increases when the tracked animal has a strong preference for specific habitats and there is a patchy (clumped) distribution of those habitats \citep{holdo2013inferring}.
+Inferring spatio-temporal distributions of populations from high density observations of a few individuals (or, in this case one individual) will always be problematic, but the predictive power associated with linking a tracked individual to the population distribution increases when the tracked animal has a strong preference for specific habitats and there is a patchy (clumped) distribution of those habitats \citep{holdo2013inferring}.
 In the case of blue whales, the primary driver influencing seasonal movements is assumed to be the seasonally variable distribution of food resources between high and low latitudes. 
@@
 This inference of course requires testing with additional isotopic records from individual whales, observational sightings data and multi-year satellite tracking records. 
-
-Our movement models suggest that the upwelling regions around Mauritania and the Cape Verde Islands may have been important historic calving areas, consistent with modern observations of juvenile blue whales in the northeast Atlantic \citep{handbook}.
-These areas are now fishing, tourist and shipping hotspots with emerging offshore oil and gas exploration, potentially increasing threats to the recovery of the small population of blue whales in the northeast Atlantic.
-Although hunting for blue whales has been banned for over 50 years, the species is still listed as Endangered on the IUCN Red List, and expected recoveries following the cessation of active hunting may be impeded by other human activities such as modifications to the food web or more direct impacts of noise and shipping.
 Developing an understanding of the nature and plasticity of individual level movements in populations of blue (and other mysticete) whales would aid management decision making in the conservation of this endangered species \citep{irvine2017quantifying}.
 
-Long-term, multi-annual data on the movement patterns and reproductive ecology of individual blue whales are scarce. 
-Although based on a few individuals, our results confirm that sequential sampling of stable isotope compositions in whale baleen, combined with simulation modeling can yield plausible inferences of individual whale movements that are consistent with population-level satellite tag data and/or inferences drawn from seasonal differences in distributions inferred from observations of multiple individuals over time and space.
+\subsection{Conclusion}
+Temporal variations in stable isotope compositions of incrementally grown tissues offer a potentially valuable record of movement in migratory organisms, but interpreting such profiles is extremely difficult, particularly in marine environments.
+Here we show how a simulation framework can be used to help interpret measured data through in silico experimentation.
+By varying parameters of agent-based movement models coupled to models predicting temporal and spatial variation in plankton carbon isotope data we identify combinations of movement behaviours producing simulated baleen isotope records that are most consistent with measured data.
+Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four years of simulations. 
+Our results confirm that sequential sampling of stable isotope compositions in whale baleen, combined with simulation modeling can yield plausible inferences of individual whale movements that are consistent with assumed movement behaviours.
 Our movement simulation modeling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviors over multiple years.

### 2019-02-14 manuscript/PeerJ/blue-new-supplementary.tex
Remove text as this is now in main doc

@@
 
-\section*{Supplemental Methods}
-
-\subsection*{CHARACTERISATION OF THE MEASURED BALEEN}
-The baleen plate yielded 97 discrete samples of baleen. 
-Across the baleen sample, $\delta^{15}$N values show clear cycles with periods of relatively high $\delta^{15}$N values interspersed by relatively short periods where $\delta^{15}$N values are c.1\text{\textperthousand} lower.
-Fourier analysis (periodogram function in the R package TSA \citep{Chan:2012aa}) revealed a strong periodic repetition with a 13.3cm periodicity. 
-The oldest (most distal) 35cm of the baleen plate records relatively positive and constant $\delta^{13}$C values, associated with a weakening of the periodic fluctuation seen in $\delta^{15}$N values, and an overall reduction in $\delta^{15}$N values.
-$\delta^{13}$C values also show repeated fluctuations in the middle 50cm of the baleen plate (behavioural phase two), with a similar periodicity of 13.5cm (Figure \ref{figs1}). 
-A distinct change in the isotopic pattern along the baleen plate occurs at around 84cm from the distal end in behavioural phase two, with a transition to uniform, relatively low $\delta^{13}$C values followed by an abrupt transition to positive $\delta^{13}$C values and a decline in $\delta^{13}$C values in the most recent (most gingival) 3cm of the record.
-
-% figure s1
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width = \linewidth]{figures/Figure-S1-periodograms.png}
-  \caption{Fourier analysis of $\delta^{15}$N (A, B) and $\delta^{13}$C (C, D) values of the full (A, C) and behavioural phase two only (B, D) components of the baleen plate.
-Inset values are the highest frequency spectral densities (in cm), excluding the values at the edge of the baleen (i.e.
-80cm), and at the sharp increase of $\delta^{13}$C seen at 60cm.
-Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent in behavioural phase two.}
-  \label{figs1}
-\end{figure}
-
-Cross-correlation analysis demonstrates a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N  values within behavioural phase two (Figure \ref{figs2}), but no relationship between $\delta^{13}$C and $\delta^{15}$N values exists during behavioural phase one.
-
-% figure s2
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width = \linewidth]{figures/Figure-S2-cross-cor.png}
-  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for the three behavioural phases, showing limited covariation in behavioural phases one and three, but strong negative covariance in behavioural phase two with a periodicity of around 13cm.
-  }
-  \label{figs2}
-\end{figure}
-
-\subsection*{BASELINE ISOTOPE COMPARISONS}
-Isotope-enabled biogeochemical ocean models \citep{magozzi2017using,schmittner2016complementary} were used to characterise the isotopic composition of phytoplankton expected in different potential foraging grounds.
-Annual average $\delta^{15}$N POM values were provided by C.J. Somes (\textit{pers.comm}) based on a 5$^{\circ}$ resolution biogeochemical model. 
-$\delta^{13}$C POM values were simulated using an isotopic extension to the NEMO-MEDUSA model \citep{magozzi2017using}.
-
-Simulated $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c.60$^{\circ}$N, and relatively negative in the central and southern North Atlantic, associated with high levels of nitrogen fixation (Figure \ref{figs3}).
-Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. 
-In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Figure \ref{figs3}). 
-The temporally coincident periodic, inverse fluctuations in $\delta^{13}$C and $\delta^{15}$N values in the second phase of the baleen isotopic record are therefore consistent with migration between waters north of 60$^{\circ}$N and more southerly waters in the North Atlantic, with northerly occupancy in summer months and southerly occupancy in winter months.
-Incorporation of carbon and nitrogen with isotopic ratios consistent with southerly latitudes during winter months indicates continued feeding during migration and winter residency.
-
-\begin{figure}[!htbp]
-    \centering
-    \begin{subfigure}[t]{0.45\textwidth}
-        \centering
-        \includegraphics[width=\linewidth]{figures/Figure-S3-plankton-d13C-map.png} 
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[t]{0.45\textwidth}
-        \centering
-        \includegraphics[width=\linewidth]{figures/Figure-S3-plankton-d15N-map.png} 
-    \end{subfigure}
-    \caption{Spatial model (isoscape) of model-simulated biomass, using weighted annual average $\delta^{13}$C (left) and $\delta^{15}$N (right) values from particulate organic matter (POM).
-$\delta^{13}$C values are from \cite{magozzi2017using} and $\delta^{15}$N from \cite{schmittner2016complementary} and C.J.
-Somes (\textit{pers.
-comm.}).}
-    \label{figs3}
-\end{figure}
-
-The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \citep{laws1995dependence}.
-Thus temporal variations in $\delta^{13}$C POM values are superimposed on latitudinal gradients. 
-The scale and nature of temporal variation in $\delta^{13}$C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in $\delta^{13}$C POM values linked to strongly seasonal phytoplankton growth dynamics.
-We therefore used monthly simulated $\delta^{13}$C POM values to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours.
-
-\subsection*{ISOTOPE MODEL SIMULATIONS}
-We initially simulated temporal variations in baseline (phytoplankton) isotopic compositions that would be encountered by whales foraging within broad geographic regions (west UK/Irish shelf, Norwegian/Barents Sea, Canaries/west Azores, Mid-Atlantic ridge, Cape Verde/Mauritanian upwelling area; Figure \ref{figs4}).
-Strong seasonal dynamics in $\delta^{13}$C values are evident in all northerly regions, characterized by a rapid increase to annual maximum $\delta^{13}$C values associated with the onset of the spring phytoplankton bloom, followed by a more gradual decline in $\delta^{13}$C values towards minima in winter conditions (Figure \ref{figs4}).
-In temperate latitudes around the British Isles, seasonal dynamic cycles are present, but dampened (Figure \ref{figs4}), whereas in sub-tropical regions exemplified by the Canaries, Azores and particularly Mauritanian upwelling areas, $\delta^{13}$C values are relatively positive and constant throughout the year (Figure \ref{figs4}).
-
-  \begin{figure}[!htbp]
-    \centering
-      \includegraphics[width=\linewidth]{figures/Figure-S4-regions-d13C.png}
-      \caption{$\delta^{13}$C values expected in baleen plates for whales simulated with resident foraging models (i.e.
-restricting movements to broad regions) for west UK/Irish shelf (West Ireland), Norwegian/Barents Sea (Norwegian Sea), Canaries/west Azores (Canaries), Mid-Atlantic ridge (Mid-Atlantic), and the Cape Verde/Mauritanian upwelling area (Cape Verde/Mauritania), over a simulated two year period with daily resolution and 30 replicate simulations for each region.
-      Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location.}
-      \label{figs4}
-  \end{figure}
-
-\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase one}
-Behavioural phase one is characterised by approximately 24 months of relatively positive and constant $\delta^{13}$C values. 
-$\delta^{15}$N values in this phase are relatively low and the isotopic cyclicity is absent (Figure \ref{figs1}). 
-Positive and temporally constant $\delta^{13}$C values are characteristic of sub-tropical regions particularly upwelling areas around Mauritania, Cape Verde and the Azores region.
-
-\subsubsection*{Matching isotope model simulations with whale isotopes in behavioural phase two}
-Temporal dynamics in $\delta^{13}$C values simulated in northerly latitudes cannot be reproduced from simulated temporal trends in $\delta^{13}$C values in krill within static areas (Figure \ref{figs4}, Figure 3).
-Adding seasonal north-south migrations within mid-high latitude regions to foraging models, closely matches the measured profile, including $\delta^{13}$C minima preceding the June bloom (Figure \ref{figs4}, Figure 1).
-
-Accordingly we modeled whale movements allowing three years of residency in warm waters of the Azores or Cape Verde/Mauritania region followed by a further 3.5 years of seasonal north-south migration in the northeast northern Atlantic, followed by residency in sub-tropical waters in 1890 and rapid migration returning to more northerly waters in the spring of 1891.
-The resulting simulated profiles closely match the measured baleen plate, capturing all major temporal fluctuations (Figure 3). 
-Blue whales have a gestation period of 10-12 months, and calves suckle for approximately 6-7 months \citep{handbook}. 
-The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer and autumn 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the assumed reproductive ecology of blue whales.
-
-\subsection*{AGENT-BASED WHALE MOVEMENT MODEL}
-We simulated the likely isotopic expression associated with feeding and movement behaviours by building an agent-based model sampling $\delta^{13}$C POM values predicted by the isotopic extension to the NEMO-MEDUSA model \citep{magozzi2017using,yool2013medusa}, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA \citep{yool2013medusa}, and bathymetry from \cite{bathy} and extracted using the marmap R pakckage \citep{marmap} (Figure \ref{figs5}).
-
-The agent-based model developed employs a simple set of movement rules dictated by behavioural state which is itself fixed according to calendar month by the operator. 
-At each daily step, the probability of movement, movement direction and distance traveled are sampled from probability distributions to allow individual variation.
-
-Initial boundary conditions are defined with a maximum temperature of 25$^{\circ}$C and minimum temperature of 3$^{\circ}$C. 
-The likelihood of movement (i.e.
-whether to move or not from the current location) is sampled from a binomial distribution with the probability of movement influenced by behavioural state and external conditions.
-The maximum daily movement distance permitted in each behavioural state is defined as a random sample of a Gaussian distribution with specified mean and standard deviation (see Table \ref{tables1}).
-
-The behavioural states employed in the model simulations are foraging, migrating north and migrating south. 
-When foraging, the probability of movement and likely direction of movement is primarily controlled by the biomass of phytoplankton in the current and adjacent 1$^{\circ}$ cells. 
-When migrating, a fixed higher probability of movement is given to southerly or northerly movement directions, and the probability of movement is influenced by the ambient sea surface temperature and phytoplankton biomass.
-
-Parameters defined for the agent-based model best matching the measured profile are summarised in Table \ref{tables1}.
-
 \newpage
 
-\begin{landscape}
-
-  \vspace{-1cm}
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[trim = {3cm 0 0 0}, clip]{figures/flow-diagram-model.pdf}
-  %\vspace{-1.2cm}
-  \linespread{0.8}
-  \caption{
-  Schematic structure of the agent-based model used to generate hypothetical movement tracks. 
-  At the start of each simulation, max and min sea surface temperature (SST) and food availability thresholds are fixed for all individuals.
-Behavioural states are defined \textit{a priori} based on day of the year.
-Maximum movement distances under each behavioural state are defined for each individual based on a random sample from a Gaussian distribution.
-The model runs with a daily time step.
-Current location and day of simulation is noted, and the behavioural state defined.
-Environmental conditions (SST, food availability and depth) are sensed in the current and eight adjacent 1$^{\circ}$ grid cells.
-The probability of movement is sampled from a binomial distribution defined according to behavioural state and environmental conditions in the current location.
-The direction of movement is sampled based on discrete probabilities associated to each of the eight adjacent 1$^{\circ}$ cells, defined according to behavioural state and environmental conditions.
-Finally the distance moved is sampled from a Poisson distribution conditional on the maximum distance defined for each individual.
-The new location is then assigned from a linear vector originating at the current location with magnitude defined as the distance moved and direction as 45$^{\circ}$ increments reflecting the relative orientation of each of the eight potential adjacent cells.}
-  \label{figs5}
-\end{figure}
-
-\newpage
-
+% table s1
 \centering
@@
   See Figure \ref{figs5} and text for more details. 
-  %$^*$During nursing minimum sea surface temperature is set to 18$^{\circ}$C. 
   $^{*}$During behavioural phase one, mean maximum daily distance traveled is set to 20km.}
@@
 \end{table}
+
+\newpage
+% figure s1
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width = \linewidth]{figures/Figure-S1-periodograms.png}
+  \caption{Fourier analysis of $\delta^{15}$N (A, B) and $\delta^{13}$C (C, D) values of the full (A, C) and behavioural phase two only (B, D) components of the baleen plate.
+Inset values are the highest frequency spectral densities (in cm), excluding the values at the edge of the baleen (i.e.
+80cm), and at the sharp increase of $\delta^{13}$C seen at 60cm.
+Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent in behavioural phase two.}
+  \label{figs1}
+\end{figure}
+
+% figure s2
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width = \linewidth]{figures/Figure-S2-cross-cor.png}
+  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for the three behavioural phases, showing limited covariation in behavioural phases one and three, but strong negative covariance in behavioural phase two with a periodicity of around 13cm.
+  }
+  \label{figs2}
+\end{figure}
+
+% figure s3
+\begin{figure}[!htbp]
+    \centering
+    \begin{subfigure}[t]{0.45\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{figures/Figure-S3-plankton-d13C-map.png} 
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[t]{0.45\textwidth}
+        \centering
+        \includegraphics[width=\linewidth]{figures/Figure-S3-plankton-d15N-map.png} 
+    \end{subfigure}
+    \caption{Spatial model (isoscape) of model-simulated biomass, using weighted annual average $\delta^{13}$C (left) and $\delta^{15}$N (right) values from particulate organic matter (POM).
+$\delta^{13}$C values are from \cite{magozzi2017using} and $\delta^{15}$N from \cite{schmittner2016complementary} and C.J.
+Somes (\textit{pers.
+comm.}).}
+    \label{figs3}
+\end{figure}
+
+% figure s4 
+  \begin{figure}[!htbp]
+    \centering
+      \includegraphics[width=\linewidth]{figures/Figure-S4-regions-d13C.png}
+      \caption{$\delta^{13}$C values expected in baleen plates for whales simulated with resident foraging models (i.e.
+restricting movements to broad regions) for west UK/Irish shelf (West Ireland), Norwegian/Barents Sea (Norwegian Sea), Canaries/west Azores (Canaries), Mid-Atlantic ridge (Mid-Atlantic), and the Cape Verde/Mauritanian upwelling area (Cape Verde/Mauritania), over a simulated two year period with daily resolution and 30 replicate simulations for each region.
+      Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location.}
+      \label{figs4}
+  \end{figure}
+
+% figure s5   
+\newpage
+
+\begin{landscape}
+
+  \vspace{-1cm}
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[trim = {3cm 0 0 0}, clip]{figures/flow-diagram-model.pdf}
+  %\vspace{-1.2cm}
+  \linespread{0.8}
+  \caption{
+  Schematic structure of the agent-based model used to generate hypothetical movement tracks. 
+  At the start of each simulation, max and min sea surface temperature (SST) and food availability thresholds are fixed for all individuals.
+Behavioural states are defined \textit{a priori} based on day of the year.
+Maximum movement distances under each behavioural state are defined for each individual based on a random sample from a Gaussian distribution.
+The model runs with a daily time step.
+Current location and day of simulation is noted, and the behavioural state defined.
+Environmental conditions (SST, food availability and depth) are sensed in the current and eight adjacent 1$^{\circ}$ grid cells.
+The probability of movement is sampled from a binomial distribution defined according to behavioural state and environmental conditions in the current location.
+The direction of movement is sampled based on discrete probabilities associated to each of the eight adjacent 1$^{\circ}$ cells, defined according to behavioural state and environmental conditions.
+Finally the distance moved is sampled from a Poisson distribution conditional on the maximum distance defined for each individual.
+The new location is then assigned from a linear vector originating at the current location with magnitude defined as the distance moved and direction as 45$^{\circ}$ increments reflecting the relative orientation of each of the eight potential adjacent cells.}
+  \label{figs5}
+\end{figure}
 
@@
 
-\section*{Supplemental Results: NHM blue whale}
-
+% figure s6
   \begin{figure}[!htbp]
@@
 
+% figure s7
     \begin{figure}[!htbp]
@@
 
+% figure s8
 \begin{landscape}
@@
 
-\newpage
-\section*{Supplemental Results: Pacific blue whales from Busquets-Vass et al. 2017}
-
-\begin{landscape}
-    \begin{figure}[!htbp]
-    \centering
-      \includegraphics[angle = 90, origin = c, scale = 0.65]{figures/top25.moveWhaleA_TL2.pdf}
-      \caption{Locations corresponding to the top 25\% best-fitting migratory movement model simulations for Pacific blue whale A. Data from Busquets-Vass et al. 2017.}
-      \label{figwhaleA}
-  \end{figure}
-\end{landscape}
-
-\begin{landscape}
-    \begin{figure}[!htbp]
-    \centering
-      \includegraphics[width=\linewidth, scale = 0.55]{figures/top25.moveWhaleB_TL2.pdf}
-      \caption{Locations corresponding to the top 25\% best-fitting migratory movement model simulations for Pacific blue whale B. Data from Busquets-Vass et al. 2017.}
-      \label{figwhaleB}
-  \end{figure}
-\end{landscape}
-
-\begin{landscape}
-    \begin{figure}[!htbp]
-    \centering
-      \includegraphics[angle = 90, origin = c, scale = 0.65]{figures/top25.moveWhaleC_TL2.pdf}
-      \caption{Locations corresponding to the top 25\% best-fitting migratory movement model simulations for Pacific blue whale C. Data from Busquets-Vass et al. 2017.}
-      \label{figwhaleC}
-  \end{figure}
-\end{landscape}
-
-\begin{landscape}
-    \begin{figure}[!htbp]
-    \centering
-      \includegraphics[angle = 90, origin = c, scale = 0.65]{figures/top25.moveWhaleD_TL2.pdf}
-      \caption{Locations corresponding to the top 25\% best-fitting migratory movement model simulations for Pacific blue whale D. Data from Busquets-Vass et al. 2017.}
-      \label{figwhaleD}
-  \end{figure}
-\end{landscape}
-
-\begin{landscape}
-    \begin{figure}[!htbp]
-    \centering
-      \includegraphics[angle = 90, origin = c, scale = 0.65]{figures/top10.moveWhaleE_TL2.pdf}
-      \caption{Locations corresponding to the top 25\% best-fitting migratory movement model simulations for Pacific blue whale E. Data from Busquets-Vass et al. 2017.}
-      \label{figwhaleE}
-  \end{figure}
-\end{landscape}
-
-\begin{landscape}
-    \begin{figure}[!htbp]
-    \centering
-      \includegraphics[angle = 90, origin = c, scale = 0.65]{figures/top25.moveWhaleF_TL2.pdf}
-      \caption{Locations corresponding to the top 25\% best-fitting migratory movement model simulations for Pacific blue whale F. Data from Busquets-Vass et al. 2017.}
-      \label{figwhaleF}
-  \end{figure}
-\end{landscape}
-
 % References
-\newpage
 \bibliographystyle{refs}

### 2019-02-15 code/Figure-2-code.R
Update paths for PeerJ submission

@@ -67,5 +67,5 @@ isop <- isop + geom_vline(xintercept = rev(year_breaks),
 
 print(isop)
 
-ggsave("manuscript/revision/figures/Figure-1-raw-dC-dN-data.png", isop, 
+ggsave("manuscript/PeerJ/figures/Figure-1-raw-dC-dN-data.png", isop, 
        device = png(width = 600, height = 400))

### 2019-02-15 code/Figure-3-code.R

@@ -21,7 +21,7 @@ jLon <- jitter(top10$Lon, factor = 1)
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Points plot with phases coloured
-png("manuscript/revision/figures/Figure-2-points.png", width = 800, height = 600)
+png("manuscript/PeerJ/figures/Figure-2-points.png", width = 900, height = 900, units = "px")
 
 # Create empty backgrounds
 plot(x = NA, y = NA, xlim = c(-80, 50), ylim = c(0, 80), 

### 2019-02-15 code/Figure-4-code.R

@@ -72,5 +72,5 @@ gall <- gtable_add_grob(g1, list(g2$grobs[[which(g2$layout$name == "panel")]],
                         name = 1:2)
 plot(gall)
 
-ggsave("manuscript/figures/Figure-3-blue-sims.png", gall, 
+ggsave("manuscript/PeerJ/figures/Figure-3-blue-sims.png", gall, 
        device = png(width = 600, height = 400))

### 2019-02-15 code/Figure-5-code.R

@@ -18,7 +18,7 @@ mycols <- c(viridis_pal()(10)[1], viridis_pal()(10)[5], viridis_pal()(10)[9])
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Points plot with phases coloured
-png("manuscript/revision/figures/Figure-4-monthly.png", width = 800, height = 600)
+png("manuscript/PeerJ/figures/Figure-4-monthly.png", width = 900, height = 900)
 
 # Make 12 plots
 par(mfrow = c(3, 4))

### 2019-02-15 code/Figure-S3-code.R

@@ -39,7 +39,7 @@ p <-
           axis.ticks = element_blank(),
           axis.title = element_blank())
 
-ggsave("manuscript/revision/figures/Figure-S3-plankton-d13C-map.png", p, 
+ggsave("manuscript/PeerJ/figures/Figure-S3-plankton-d13C-map.png", p, 
        device = png(width = 600, height = 400))
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
@@ -67,5 +67,5 @@ p2 <-
           axis.ticks = element_blank(),
           axis.title = element_blank())
 
-ggsave("manuscript/revision/figures/Figure-S3-plankton-d15N-map.png", p2, 
+ggsave("manuscript/PeerJ/figures/Figure-S3-plankton-d15N-map.png", p2, 
        device = png(width = 600, height = 400))

### 2019-02-15 code/Figure-S6-code.R

@@ -21,5 +21,5 @@ p1 <-
   expand_limits(x = c(0, 1),  y = c(0, 50))
 
 
-ggsave("manuscript/figures/Figure-S6-r2.png", p1, 
+ggsave("manuscript/PeerJ/figures/Figure-S6-r2.png", p1, 
        device = png(width = 600, height = 400))

### 2019-02-15 code/Figure-2-code.R
Change names to correct figure numbers

@@ -1,71 +1,46 @@
 # Author: Clive Trueman
-# Adapted by: Andrew Jackson June 2017 & Natalie Cooper Oct 2017.
-# About: Script to generate figure 1
+# About: script to plot track for top 10% of models
+# Tidied by Natalie Cooper Nov 2017
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Load libraries
-library(tidyverse)
-library(lubridate)
+library(maps)
+library(mapdata)
+library(spatstat)
+library(viridis)
+
+# Read in data
+top10 <- read.csv("data/top10percent.csv")
+
+# Pick colours
+mycols <- c(viridis_pal()(10)[1], viridis_pal()(10)[5], viridis_pal()(10)[9])
+
+# Jitter points of lat and long to get smooth density plot
+jLat <- jitter(top10$Lat, factor = 2)
+jLon <- jitter(top10$Lon, factor = 1)
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Figure 1 - raw whale isotope data with two y-axes
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# read in data and extract just the blue whale
-whale_isos <- read_csv("data/raw-whale-isotope-data.csv")
+# Points plot with phases coloured
+png("manuscript/PeerJ/figures/Figure-2-points.png", width = 900, height = 900, units = "px")
 
-KC7 <- filter(whale_isos, whale_ID == "KC7")
+# Create empty backgrounds
+plot(x = NA, y = NA, xlim = c(-80, 50), ylim = c(0, 80), 
+     xlab = "", ylab = "", axes = FALSE)
+rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], 
+     col = "grey90")
 
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# convert sample number to days based on growth rate of baleen
-growth_rate <- 13.5  # baleen growth rate in centimetres per year
-last_sample <- ymd("1891-03-01") # final date of last keratin sample
+# Add points coloured by phase
+points(jLon, jLat, pch = 16, col = mycols[top10$phase], cex = 0.2)
 
-KC7 <- 
-  KC7 %>%
-  mutate(days = sample_cm * 365.25 / growth_rate) %>%
-  mutate(days = days - days[1]) %>%
[…]

### 2019-02-15 code/Figure-3-code.R

@@ -1,46 +1,76 @@
-# Author: Clive Trueman
-# About: script to plot track for top 10% of models
-# Tidied by Natalie Cooper Nov 2017
-
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Code for Figure 3
+# Adapted from code by Clive Trueman by Natalie Cooper Nov 2017
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Load libraries
-library(maps)
-library(mapdata)
-library(spatstat)
-library(viridis)
+library(tidyverse)
+library(gtable)
+library(grid)
 
-# Read in data
-top10 <- read.csv("data/top10percent.csv")
+# Read in the measured whale data and select blue whale
+whale_isos <- read_csv("data/raw-whale-isotope-data.csv")
+blue <- filter(whale_isos, whale_ID == "KC7")
 
-# Pick colours
-mycols <- c(viridis_pal()(10)[1], viridis_pal()(10)[5], viridis_pal()(10)[9])
+# Read in top 100 models
+top100 <- read.csv("data/top10smooth.csv")
 
-# Jitter points of lat and long to get smooth density plot
-jLat <- jitter(top10$Lat, factor = 2)
-jLon <- jitter(top10$Lon, factor = 1)
+# Fix day numbers to match in both
+blue$Day.sim <- rev(unique(top100$count2))
 
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Points plot with phases coloured
-png("manuscript/PeerJ/figures/Figure-2-points.png", width = 900, height = 900, units = "px")
+# Create three plots, one for blue whale and one for simulations, and 
+# a base plot with axes etc.
+base_plot <- 
+  ggplot(blue, aes(x = Day.sim, y = d13C)) +
+  #geom_line(size = 1.5) +
+  theme_classic(base_size = 16) + 
+  xlab("Days") + 
+  ylab(expression(paste(delta^{13}, "C (\u2030)"))) +
+  scale_y_continuous(limits = c(-20, -16.5),
+                     sec.axis = sec_axis(~.+4, 
+                                breaks = c(-16, -15, -14, -13),
+                                labels = c(-30, -25, -20, -15),
+                                name = expression(paste("simulated ", 
+                                                        delta^{13}, "C (\u2030)")))) +
+  theme(panel.background = element_blank())
 
-# Create empty backgrounds
[…]

### 2019-02-15 code/Figure-4-code.R

@@ -1,76 +1,57 @@
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Code for Figure 3
-# Adapted from code by Clive Trueman by Natalie Cooper Nov 2017
+# Author: Clive Trueman
+# About: script to plot track for top 10% of models by month
+# for behavioural phase 2.
+# Tidied by Natalie Cooper Jan 2018
+
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Load libraries
-library(tidyverse)
-library(gtable)
-library(grid)
+library(maps)
+library(mapdata)
+library(spatstat)
+library(viridis)
 
-# Read in the measured whale data and select blue whale
-whale_isos <- read_csv("data/raw-whale-isotope-data.csv")
-blue <- filter(whale_isos, whale_ID == "KC7")
+# Read in data
+mid.top10 <- read.csv("data/mid.top10percent.csv")
 
-# Read in top 100 models
-top100 <- read.csv("data/top10smooth.csv")
+# Pick colours to match phases in Figure 2
+mycols <- c(viridis_pal()(10)[1], viridis_pal()(10)[5], viridis_pal()(10)[9])
 
-# Fix day numbers to match in both
-blue$Day.sim <- rev(unique(top100$count2))
+# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+# Points plot with phases coloured
+png("manuscript/PeerJ/figures/Figure-4-monthly.png", width = 900, height = 900)
 
-# Create three plots, one for blue whale and one for simulations, and 
-# a base plot with axes etc.
-base_plot <- 
-  ggplot(blue, aes(x = Day.sim, y = d13C)) +
-  #geom_line(size = 1.5) +
-  theme_classic(base_size = 16) + 
-  xlab("Days") + 
-  ylab(expression(paste(delta^{13}, "C (\u2030)"))) +
-  scale_y_continuous(limits = c(-20, -16.5),
-                     sec.axis = sec_axis(~.+4, 
-                                breaks = c(-16, -15, -14, -13),
-                                labels = c(-30, -25, -20, -15),
-                                name = expression(paste("simulated ", 
-                                                        delta^{13}, "C (\u2030)")))) +
-  theme(panel.background = element_blank())
+# Make 12 plots
+par(mfrow = c(3, 4))
+par(mar = c(1, 1, 2, 1))
 
-top_plot <- 
[…]

### 2019-02-15 code/Figure-5-code.R

@@ -1,57 +1,0 @@
-# Author: Clive Trueman
-# About: script to plot track for top 10% of models by month
-# for behavioural phase 2.
-# Tidied by Natalie Cooper Jan 2018
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Load libraries
-library(maps)
-library(mapdata)
-library(spatstat)
-library(viridis)
-
-# Read in data
-mid.top10 <- read.csv("data/mid.top10percent.csv")
-
-# Pick colours to match phases in Figure 2
-mycols <- c(viridis_pal()(10)[1], viridis_pal()(10)[5], viridis_pal()(10)[9])
-
-# = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-# Points plot with phases coloured
-png("manuscript/PeerJ/figures/Figure-4-monthly.png", width = 900, height = 900)
-
-# Make 12 plots
-par(mfrow = c(3, 4))
-par(mar = c(1, 1, 2, 1))
-
-# Loop through months  
-for (Mn in 1:12) {
-  
-  # Make background of plots
-  plot(x = NA, y = NA, xlim = c(-80, 50), ylim = c(0, 80), 
-       xlab = "", ylab = "", axes = FALSE, main = month.abb[Mn], cex.main = 2)
-  rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], 
-       col = "grey90")
-  
-  # Select month
-  Month <- mid.top10[mid.top10$Month == Mn, ]
-  
-  # Jitter points of lat and long to get smooth density plot
-  jLat <- jitter(Month$Lat, factor = 2)
-  jLon <- jitter(Month$Lon, factor = 1)
-  
-  # Add points coloured by phase
-  points(jLon, jLat, pch = 16, col = mycols[2], cex = 0.2)
-
-  # Add world map
-  map('world', col = "black", fill = TRUE, add = TRUE, lwd = 0.25)
-  # If you get this error, 
-  # Error in as_mapper(.f, ...) : argument ".f" is missing, with no default
-  # You need to detach the package purrr (detach(package:purrr)) and reload
-  # the maps library (library(map))
-  
-  par(new=FALSE)
-}
-
-dev.off()
-

### 2019-07-11 manuscript/PeerJ/blue-new.tex

@@
 This uncertainty is compounded when multiple samples are taken through time in the same individual, as temporal variation in the isotopic baseline must also be considered explicitly. 
-Consequently, relatively little attention has been given to the potential to infer individual movement histories by reconstructing time series of multiple discrete origins from sequentially sampled incrementally-grown tissues
-% [trueman st john glew, Tokyo eel, darnaude, brennan?].
+Consequently, relatively little attention has been given to the potential to infer individual movement histories by reconstructing time series of multiple discrete origins from sequentially sampled incrementally-grown tissues \citep{trueman2019, sakomoto2018, darnoude2014}.
 
 Recently, mechanistic models have been developed providing relatively accurate predictions of isotopic variability at high temporal and spatial resolution. 
-These models can be coupled to Lagrangian or agent-based models of animal movement to predict isotopic compositions or trajectories expected to be recorded in animal tissues under differing movement scenarios.
+These models can be coupled to Lagrangian or agent-based models of animal movement to predict isotopic compositions or trajectories expected to be recorded in animal tissues under differing movement scenarios \citep{carpenter2019}.
 Here we illustrate the potential of this conceptual approach by assessing the relative likelihood of differing, plausible hypotheses of movement behavior for a single individual blue whale based on carbon isotope compositions sampled across a baleen plate.
@@
 
-The isotopic composition of phytoplankton is influenced by the release of fossil carbon with low \(\delta^{13}\)C values into the atmosphere (Suess effect) and the effect of increased concentrations of dissolved CO2 on isotopic fractionation during phytoplankton growth
-%(Young et al 2013)
-, potentially complicating the application of a model trained on recent data to historic samples. 
+The isotopic composition of phytoplankton is influenced by the release of fossil carbon with low \(\delta^{13}\)C values into the atmosphere (Suess effect) and the effect of increased concentrations of dissolved CO2 on isotopic fractionation during phytoplankton growth \citep{young2013}, potentially complicating the application of a model trained on recent data to historic samples.
 Ocean basin scale spatio-temporal differences in \(\delta^{13}\)C values largely reflect relative differences in sea surface temperatures between trophic, temperate and arctic waters and over seasons, and we assume that any regional variation in isotopic effects due to CO\textsubscript{2}-influenced changes in phytoplankton composition and growth rate are minor compared to latitudinal and seasonal variations in \(\delta^{13}\)C POM values.
@@
 
-We simulated the likely isotopic expression associated with different movement behaviours by building an agent-based movement model within the R coding environment, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA \citep{yool2013medusa}, and bathymetry from the General Bathymetric Chart of the Oceans (GEBCO) \citep{bathy}, extracted via the marmap package in R \citep{marmap} (Figure S5).
-% check figure S5
+We simulated the likely isotopic expression associated with different movement behaviours by building an agent-based movement model within the R coding environment, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA \citep{yool2013medusa}, and bathymetry from the General Bathymetric Chart of the Oceans (GEBCO) \citep{bathy}, extracted via the marmap package in R \citep{marmap} (Figure S5; Table S1).
 

### 2019-07-16 code/Figure-S1-S2-code.R
Reomve C and add phase 3

@@ -51,7 +51,7 @@ periodogramText <- function(p, k){
 png(filename = "manuscript/PeerJ/figures/Figure-S1-periodograms.png", 
     width = 1000, height = 600)
 
-par(mfrow = c(2, 2))
+par(mfrow = c(1, 3))
 
 p1 <- periodogram(KC7$d15N, main = "A - Full d15N")
 periodogramText(p1, k = 2)
@@ -59,11 +59,8 @@ periodogramText(p1, k = 2)
 p2 <- periodogram(KC7$d15N[18:70], main = "B - Phase 2 d15N")
 periodogramText(p2, k = 1)
 
-p3 <- periodogram(KC7$d13C, main = "C - Full d13C")
-periodogramText(p3, k = 2)
-
-p4 <- periodogram(KC7$d13C[18:70], main = "D - Phase 2 d13C")
-periodogramText(p4, k = 2)
+p3 <- periodogram(KC7$d15N[70:n], main = "B - Phase 3 d15N")
+periodogramText(p3, k = 1)
 
 dev.off()
 

### 2019-07-16 manuscript/PeerJ/blue-new-supplementary.tex
Edit to order of figures
Edit figs S1 and S2

@@
 \newpage
+\begin{landscape}
 % figure s1
@@
     \centering
-  \includegraphics[width = \linewidth]{figures/Figure-S1-periodograms.png}
-  \caption{Fourier analysis of $\delta^{15}$N (A, B) and $\delta^{13}$C (C, D) values of the full (A, C) and behavioural phase two only (B, D) components of the baleen plate.
-Inset values are the highest frequency spectral densities (in cm), excluding the values at the edge of the baleen (i.e.
-80cm), and at the sharp increase of $\delta^{13}$C seen at 60cm.
-Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but is weaker or absent in behavioural phase two.}
+      \includegraphics[width=\linewidth]{figures/Figure-S3-plankton-d13C-map.png} 
+    \caption{Spatial model (isoscape) of model-simulated biomass, using weighted annual average $\delta^{13}$C values from \cite{magozzi2017using}.} 
     \label{figs1}
 \end{figure}
+\end{landscape}
+\newpage
 
@@
   \centering
-  \includegraphics[width = \linewidth]{figures/Figure-S2-cross-cor.png}
-  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for the three behavioural phases, showing limited covariation in behavioural phases one and three, but strong negative covariance in behavioural phase two with a periodicity of around 13cm.
-  }
+  \includegraphics[width = \linewidth]{figures/Figure-S1-periodograms.png}
+  \caption{Fourier analysis of $\delta^{15}$N values of the full (A) and behavioural phase two only (B) and behavioural phase three only (C) components of the baleen plate.
+Inset values are the highest frequency spectral densities (in cm).
+Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered in behavioural phases one and two, but not in behavioural phase three.}
   \label{figs2}
@@
 % figure s3   
-\begin{figure}[!htbp]
-    \centering
-    \begin{subfigure}[t]{0.45\textwidth}
-        \centering
-        \includegraphics[width=\linewidth]{figures/Figure-S3-plankton-d13C-map.png} 
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[t]{0.45\textwidth}
-        \centering
-        \includegraphics[width=\linewidth]{figures/Figure-S3-plankton-d15N-map.png} 
-    \end{subfigure}
-    \caption{Spatial model (isoscape) of model-simulated biomass, using weighted annual average $\delta^{13}$C (left) and $\delta^{15}$N (right) values from particulate organic matter (POM).
-$\delta^{13}$C values are from \cite{magozzi2017using} and $\delta^{15}$N from \cite{schmittner2016complementary} and C.J.
-Somes (\textit{pers.
-comm.}).}
-    \label{figs3}
-\end{figure}
-
-% figure s4 
-  \begin{figure}[!htbp]
-    \centering
-      \includegraphics[width=\linewidth]{figures/Figure-S4-regions-d13C.png}
-      \caption{$\delta^{13}$C values expected in baleen plates for whales simulated with resident foraging models (i.e.
-restricting movements to broad regions) for west UK/Irish shelf (West Ireland), Norwegian/Barents Sea (Norwegian Sea), Canaries/west Azores (Canaries), Mid-Atlantic ridge (Mid-Atlantic), and the Cape Verde/Mauritanian upwelling area (Cape Verde/Mauritania), over a simulated two year period with daily resolution and 30 replicate simulations for each region.
-      Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location.}
-      \label{figs4}
-  \end{figure}
-
-% figure s5   
 \newpage
@@
 The new location is then assigned from a linear vector originating at the current location with magnitude defined as the distance moved and direction as 45$^{\circ}$ increments reflecting the relative orientation of each of the eight potential adjacent cells.}
-  \label{figs5}
+  \label{figs3}
 \end{figure}
@@
 \end{landscape}
+
+% figure s4 
+  \begin{figure}[!htbp]
+    \centering
+      \includegraphics[width=\linewidth]{figures/Figure-S4-regions-d13C.png}
+      \caption{$\delta^{13}$C values expected in baleen plates for whales simulated with resident foraging models (i.e.
+restricting movements to broad regions) for west UK/Irish shelf (West Ireland), Norwegian/Barents Sea (Norwegian Sea), Canaries/west Azores (Canaries), Mid-Atlantic ridge (Mid-Atlantic), and the Cape Verde/Mauritanian upwelling area (Cape Verde/Mauritania), over a simulated two year period with daily resolution and 30 replicate simulations for each region.
+      Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location.}
+      \label{figs4}
+  \end{figure}
+
+\newpage
+
+% figure s5
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width = \linewidth]{figures/Figure-S2-cross-cor.png}
+  \caption{Cross correlation between $\delta^{13}$C and $\delta^{15}$N values for the three behavioural phases, showing limited covariation in behavioural phases one and three, but strong negative covariance in behavioural phase two with a periodicity of around 13cm.
+  }
+  \label{figs5}
+\end{figure}
 
@@
 
+  \newpage
+
 % figure s7
@@
   \end{figure}
+
+\newpage
 

### 2019-07-16 manuscript/PeerJ/blue-new-supplementary.tex
Edit figs S1 and S2

@@
   \includegraphics[width = \linewidth]{figures/Figure-S1-periodograms.png}
-  \caption{Fourier analysis of $\delta^{15}$N values of the full (A) and behavioural phase two only (B) and behavioural phase three only (C) components of the baleen plate.
+  \caption{Fourier analysis of $\delta^{15}$N values of the full (A) and behavioural phase one only (B) and behavioural phase two only (C) components of the baleen plate.
 Inset values are the highest frequency spectral densities (in cm).
-Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered in behavioural phases one and two, but not in behavioural phase three.}
+Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered, but weaker in behavioural  phase one.
+Note that we do not separately graph behavioural phase three here because it does not show any periodicity.}
   \label{figs2}

### 2019-07-16 code/Figure-S1-S2-code.R
Changes to S1

@@ -56,11 +56,11 @@ par(mfrow = c(1, 3))
 p1 <- periodogram(KC7$d15N, main = "A - Full d15N")
 periodogramText(p1, k = 2)
 
-p2 <- periodogram(KC7$d15N[18:70], main = "B - Phase 2 d15N")
-periodogramText(p2, k = 1)
-
-p3 <- periodogram(KC7$d15N[70:n], main = "B - Phase 3 d15N")
+p3 <- periodogram(KC7$d15N[1:17], main = "B - Phase 1 d15N")
 periodogramText(p3, k = 1)
+
+p2 <- periodogram(KC7$d15N[18:70], main = "C - Phase 2 d15N")
+periodogramText(p2, k = 1)
 
 dev.off()
 

### 2019-07-17 code/Figure-1-code.R
Add phases to Figure

@@ -6,6 +6,7 @@
 # Load libraries
 library(tidyverse)
 library(lubridate)
+library(ggrepel)
 
 # = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
 # Figure 1 - raw whale isotope data with two y-axes
@@ -64,6 +65,19 @@ isop <- isop +
 
 isop <- isop + geom_vline(xintercept = rev(year_breaks),
                           color = "grey", lty = 2)
+
+# Add lines for phases
+isop <-
+isop +
+  annotate("rect", xmin = 0, xmax = 30, ymin = -16.8, ymax = -16.6,
+         alpha = 0.2) +
+  annotate("text", x = 15, y = -16.7, label = "phase 1") +
+  annotate("rect", xmin = 30, xmax = 83, ymin = -16.7, ymax = -16.5,
+           alpha = 0.2) +
+  annotate("text", x = 54, y = -16.6, label = "phase 2")  +
+annotate("rect", xmin = 83, xmax = 100, ymin = -16.8, ymax = -16.6,
+         alpha = 0.2) +
+  annotate("text", x = 91.5, y = -16.7, label = "phase 3")  
 
 print(isop)
 

### 2019-07-17 code/Figure-S1-S2-code.R
Update supp figures

@@ -51,15 +51,12 @@ periodogramText <- function(p, k){
 png(filename = "manuscript/PeerJ/figures/Figure-S1-periodograms.png", 
     width = 1000, height = 600)
 
-par(mfrow = c(1, 3))
+par(mfrow = c(1, 2))
 
-p1 <- periodogram(KC7$d15N, main = "A - Full d15N")
+p1 <- periodogram(KC7$d15N, main = "Full profile")
 periodogramText(p1, k = 2)
 
-p3 <- periodogram(KC7$d15N[1:17], main = "B - Phase 1 d15N")
-periodogramText(p3, k = 1)
-
-p2 <- periodogram(KC7$d15N[18:70], main = "C - Phase 2 d15N")
+p2 <- periodogram(KC7$d15N[18:70], main = "Phase 2 only")
 periodogramText(p2, k = 1)
 
 dev.off()

### 2019-07-18 manuscript/PeerJ/blue-new-supplementary.tex
Fix supplementary

--- 2019-07-16 manuscript/PeerJ/blue-new-supplementary.tex
@@
   \includegraphics[width = \linewidth]{figures/Figure-S1-periodograms.png}
-  \caption{Fourier analysis of $\delta^{15}$N values of the full (A) and behavioural phase two only (B) and behavioural phase three only (C) components of the baleen plate.
+  \caption{Fourier analysis of $\delta^{15}$N values of the full (A) and behavioural phase two only (B) components of the baleen plate.
 Inset values are the highest frequency spectral densities (in cm).
-Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered in behavioural phases one and two, but not in behavioural phase three.}
+Periodic fluctuation with a distance of 13.5cm along the baleen plate is consistently recovered in behavioural phase two where the whale engaged in latitudinal migrations.
+Note that we do not separately graph behavioural phases one or three here because they do not show clear periodicity.}
   \label{figs2}
@@
   \centering
-  \includegraphics[trim = {3cm 0 0 0}, clip]{figures/flow-diagram-model.pdf}
+  \includegraphics[trim = {3cm 0 0 0}, clip]{figures/flowdiagrammodel.pdf}
   %\vspace{-1.2cm}
